---
title: "Where Abstract Preference Outperforms Corpus Frequency"
subtitle: "Items where RelFreq fails in 125M models — traced across model sizes"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 6, dpi = 150,
                      out.width = "100%")
```

# Overview

We use the **125M model** as a discovery set within each corpus: identify
binomials where, at the final training checkpoint, the corpus relative frequency
(**RelFreq**) predicts the *wrong* preference direction. We then trace those
exact same binomials in the 350M and 1.3B models to ask whether the pattern
replicates, and whether abstract preference (**AbsPref / GenPref**) consistently
explains the deviation.

**Comparison structure**:
- Item selection is based on the 125M model's final checkpoint, separately per
  corpus (BabyLM and C4).
- The same items are then looked at in 350M and 1.3B within the same corpus.
- BabyLM and C4 items are kept separate throughout.

Three analyses:

1. **Disagreement items (§1)**: binomials where RelFreq is wrong in the 125M
   model — how often does AbsPref predict the right direction there, and do
   350M/1.3B also deviate from RelFreq for the same items?
2. **Near-flat RelFreq items (§2)**: binomials where the 125M corpus frequency
   signal is near-chance — how do AbsPref and OverallFreq jointly determine
   preferences in all three model sizes for those items?
3. **Case studies (§3)**: full training trajectories for the 5–10 strongest
   disagreement binomials from 125M, shown for all model sizes.

------------------------------------------------------------------------

```{r libraries}
library(tidyverse)
library(duckdb)
library(future)
library(furrr)
library(ggrepel)
library(scales)
library(kableExtra)
```

```{r config}
SCRIPT_DIR <- {
  knitr_input <- tryCatch(knitr::current_input(), error = function(e) NULL)
  if (!is.null(knitr_input) && nchar(knitr_input) > 0) {
    dirname(normalizePath(knitr_input, mustWork = FALSE))
  } else if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
    dirname(normalizePath(rstudioapi::getActiveDocumentContext()$path, mustWork = FALSE))
  } else {
    normalizePath(".", mustWork = FALSE)
  }
}

DATA_DIR          <- normalizePath(file.path(SCRIPT_DIR, "../Data/processed"), mustWork = FALSE)
TRAIN_CSV         <- file.path(DATA_DIR, "training_attested.csv")
EXPOSURES_CSV     <- normalizePath(file.path(DATA_DIR, "checkpoint_results_with_exposures.csv"),
                                   mustWork = FALSE)
EXPOSURES_CSV_FWD <- gsub("\\\\", "/", EXPOSURES_CSV)

N_TRAJ_CHECKPOINTS <- 20
SIZE_LEVELS        <- c("125M", "350M", "1.3B")
SIZE_COLOURS       <- c("125M" = "#74add1", "350M" = "#007030", "1.3B" = "#154733")
SIZE_SHAPES        <- c("125M" = 21, "350M" = 22, "1.3B" = 24)

# |log(alpha/beta)| below this = "near-flat RelFreq"
FLAT_THRESHOLD <- 0.3   # roughly P(alpha) in [0.43, 0.57]
# Number of case-study binomials
N_CASE_BINOMS  <- 8

# Parallelism
N_CORES <- parallel::detectCores(logical = FALSE)
message(sprintf("Detected %d physical cores", N_CORES))
plan(multicore, workers = N_CORES)

# Parquet cache path (same location as the CSV, faster for repeated reads)
EXPOSURES_PARQUET     <- sub("\\.csv$", ".parquet", EXPOSURES_CSV)
EXPOSURES_PARQUET_FWD <- gsub("\\\\", "/", EXPOSURES_PARQUET)

stopifnot(file.exists(TRAIN_CSV), file.exists(EXPOSURES_CSV))
```

```{r helpers}
parse_corpus <- function(x) case_when(
  grepl("babylm", x, TRUE) ~ "BabyLM",
  grepl("opt.c4", x, TRUE) ~ "C4",
  TRUE ~ NA_character_
)
parse_size <- function(x) case_when(
  grepl("125m",    x, TRUE) ~ "125M",
  grepl("350m",    x, TRUE) ~ "350M",
  grepl("1[._]3b", x, TRUE) ~ "1.3B",
  TRUE ~ NA_character_
)
sample_steps <- function(model_df, n = N_TRAJ_CHECKPOINTS) {
  steps <- sort(unique(model_df$step))
  idx   <- unique(round(seq(1, length(steps), length.out = min(n, length(steps)))))
  steps[idx]
}

# ── One-time Parquet conversion ───────────────────────────────────────────────
# Converts the 458 MB CSV to Parquet on first run. Subsequent runs skip this.
# DuckDB reads Parquet ~5-10x faster than CSV (columnar, no parsing overhead).
ensure_parquet <- function() {
  if (file.exists(EXPOSURES_PARQUET)) {
    message(sprintf("Parquet cache found: %s", EXPOSURES_PARQUET))
    return(invisible(NULL))
  }
  message("Converting exposures CSV to Parquet (one-time)...")
  t0  <- proc.time()
  con <- dbConnect(duckdb::duckdb())
  on.exit(dbDisconnect(con, shutdown = TRUE))
  dbExecute(con, sprintf("SET threads = %d", N_CORES))
  dbExecute(con, sprintf(
    "COPY (SELECT * FROM read_csv_auto('%s')) TO '%s' (FORMAT PARQUET, COMPRESSION ZSTD)",
    EXPOSURES_CSV_FWD, EXPOSURES_PARQUET_FWD
  ))
  elapsed <- (proc.time() - t0)[["elapsed"]]
  message(sprintf("Parquet written in %.1f s", elapsed))
}

# ── Batch DuckDB query ────────────────────────────────────────────────────────
# Fetches alpha_seen / beta_seen / tokens for all rows in meta_df (a data frame
# with columns `model` and `step`) in a single file scan.
# binom_filter: if supplied, only those binomials are returned (saves memory for
#               the trajectory load where we only need a handful of binomials).
batch_dyn_freq <- function(meta_df, binom_filter = NULL) {
  stopifnot(all(c("model", "step") %in% names(meta_df)))

  con <- dbConnect(duckdb::duckdb())
  on.exit(dbDisconnect(con, shutdown = TRUE))
  dbExecute(con, sprintf("SET threads = %d", N_CORES))

  src <- if (file.exists(EXPOSURES_PARQUET))
    sprintf("read_parquet('%s')", EXPOSURES_PARQUET_FWD)
  else
    sprintf("read_csv_auto('%s')", EXPOSURES_CSV_FWD)
  dbExecute(con, sprintf("CREATE VIEW exp AS SELECT * FROM %s", src))

  # Load the model × step pairs as an in-memory table for an efficient JOIN
  dbWriteTable(con, "pairs", meta_df |> distinct(model, step), overwrite = TRUE)

  binom_clause <- if (!is.null(binom_filter)) {
    binom_list <- paste0("'", gsub("'", "''", binom_filter), "'", collapse = ", ")
    sprintf("AND e.binom IN (%s)", binom_list)
  } else ""

  freq_raw <- dbGetQuery(con, sprintf("
    SELECT e.model, e.step, e.binom,
           ANY_VALUE(e.alpha_seen) AS alpha_seen,
           ANY_VALUE(e.beta_seen)  AS beta_seen,
           ANY_VALUE(e.tokens)     AS tokens
    FROM exp e
    INNER JOIN pairs p ON e.model = p.model AND e.step = p.step
    %s
    GROUP BY e.model, e.step, e.binom
  ", binom_clause))

  as_tibble(freq_raw) |>
    mutate(
      log_freq_ratio = if_else(alpha_seen > 0 & beta_seen > 0,
                               log(alpha_seen / beta_seen), NA_real_),
      freq_prob_c    = plogis(log_freq_ratio) - 0.5,
      log_total      = if_else(alpha_seen + beta_seen > 0,
                               log(alpha_seen + beta_seen), NA_real_)
    ) |>
    select(-alpha_seen, -beta_seen)
}

# ── Join freq data onto training data and compute derived columns ─────────────
attach_freq <- function(train_subset, freq_df) {
  train_subset |>
    left_join(freq_df, by = c("model", "step", "binom")) |>
    mutate(signed_gap = preference - log_freq_ratio) |>
    group_by(model, step) |>
    mutate(log_total_c = as.numeric(scale(log_total))) |>
    ungroup()
}
```

------------------------------------------------------------------------

# Data Loading

```{r parquet-cache}
ensure_parquet()
```

```{r load-train}
message("Loading training data...")
train <- read_csv(TRAIN_CSV, show_col_types = FALSE)

MODELS_LIST <- sort(unique(train$model))
MODELS_LIST <- MODELS_LIST[!is.na(parse_corpus(MODELS_LIST))]
message(sprintf("%d models: %s", length(MODELS_LIST), paste(MODELS_LIST, collapse = ", ")))
```

```{r ck-index}
ck_meta <- map_dfr(MODELS_LIST, function(m) {
  steps <- sample_steps(filter(train, model == m))
  tibble(model = m, ck_idx = seq_along(steps), step = steps)
})

final_ck_meta <- ck_meta |>
  group_by(model) |>
  filter(ck_idx == max(ck_idx)) |>
  ungroup()
```

```{r load-final}
# Single DuckDB scan for all models' final checkpoints — one file read total.
message("Fetching final-checkpoint exposures (single DuckDB scan)...")
t0 <- proc.time()
freq_final <- batch_dyn_freq(final_ck_meta)
message(sprintf("  Done in %.1f s (%d rows)", (proc.time() - t0)[["elapsed"]], nrow(freq_final)))

dat_final <- train |>
  inner_join(final_ck_meta |> select(model, step, ck_idx), by = c("model", "step")) |>
  attach_freq(freq_final) |>
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS)
  ) |>
  filter(!is.na(log_freq_ratio), !is.na(GenPref_centered), !is.na(preference)) |>
  mutate(
    pref_dir    = sign(preference),
    freq_dir    = sign(log_freq_ratio),
    abspref_dir = sign(GenPref_centered),
    relfreq_ok  = (pref_dir == freq_dir),
    abspref_ok  = (pref_dir == abspref_dir)
  )

message(sprintf("Final dataset: %d rows, %d models, %d unique binomials",
  nrow(dat_final), n_distinct(dat_final$model), n_distinct(dat_final$binom)))
```

## Identify items from 125M model

```{r items-from-125m}
# 125M final-checkpoint data — used as the discovery set for item selection
dat_125m <- dat_final |>
  filter(model_size == "125M", pref_dir != 0, freq_dir != 0)

# Per-corpus, per-binomial summary from 125M
items_125m <- dat_125m |>
  group_by(corpus, binom) |>
  summarise(
    log_freq_ratio   = first(log_freq_ratio),   # corpus-level (constant per binom×corpus)
    GenPref_centered = first(GenPref_centered),  # constant per binom
    preference_125m  = first(preference),
    signed_gap_125m  = first(signed_gap),
    relfreq_ok_125m  = first(relfreq_ok),
    abspref_ok_125m  = first(abspref_ok),
    .groups          = "drop"
  ) |>
  mutate(
    freq_dir    = sign(log_freq_ratio),
    abspref_dir = sign(GenPref_centered),
    pref_dir_125m = sign(preference_125m)
  )

cat(sprintf("125M items at final checkpoint:\n"))
cat(sprintf("  BabyLM: %d binomials\n", sum(items_125m$corpus == "BabyLM")))
cat(sprintf("  C4:     %d binomials\n", sum(items_125m$corpus == "C4")))
cat(sprintf("\nRelFreq predicts wrong direction in 125M:\n"))
cat(sprintf("  BabyLM: %d (%.1f%%)\n",
  sum(!items_125m$relfreq_ok_125m & items_125m$corpus == "BabyLM"),
  mean(!items_125m$relfreq_ok_125m[items_125m$corpus == "BabyLM"]) * 100))
cat(sprintf("  C4:     %d (%.1f%%)\n",
  sum(!items_125m$relfreq_ok_125m & items_125m$corpus == "C4"),
  mean(!items_125m$relfreq_ok_125m[items_125m$corpus == "C4"]) * 100))
```

------------------------------------------------------------------------

# Section 1: Items Where 125M RelFreq Is Wrong — Traced Across Model Sizes

For each corpus, we take all binomials where RelFreq predicts the wrong
direction in the 125M model. We then ask:

- Does AbsPref predict the right direction for those items in 125M?
- Do 350M and 1.3B also show a preference deviation for those same items
  (i.e., does the disagreement with RelFreq replicate at larger scales)?

```{r disagree-selection}
disagree_125m <- items_125m |>
  filter(!relfreq_ok_125m) |>
  arrange(corpus, desc(abs(signed_gap_125m)))

cat(sprintf("Disagreement items (RelFreq wrong in 125M): %d BabyLM, %d C4\n",
  sum(disagree_125m$corpus == "BabyLM"),
  sum(disagree_125m$corpus == "C4")))
cat(sprintf("AbsPref correct in 125M: BabyLM %.0f%%, C4 %.0f%%\n",
  mean(disagree_125m$abspref_ok_125m[disagree_125m$corpus == "BabyLM"]) * 100,
  mean(disagree_125m$abspref_ok_125m[disagree_125m$corpus == "C4"]) * 100))
```

## Preference across model sizes for 125M-disagreement items

Each binomial row shows the model's preference (log-odds) for 125M, 350M,
and 1.3B. The dashed vertical line marks where RelFreq falls (same for all
sizes since same training corpus). We expect preferences to align with
AbsPref (direction indicated by colour) rather than with RelFreq.

```{r plot-disagree-trace, fig.height=10}
# Get all-model-size data for these items
# Only pull columns from disagree_125m that aren't already in dat_final
# (GenPref_centered and log_freq_ratio already exist there from the training join)
dat_disagree_trace <- dat_final |>
  semi_join(disagree_125m, by = c("corpus", "binom")) |>
  left_join(disagree_125m |> select(corpus, binom, signed_gap_125m, abspref_ok_125m),
            by = c("corpus", "binom")) |>
  mutate(
    binom        = fct_reorder(binom, abs(signed_gap_125m)),
    abspref_label = if_else(GenPref_centered > 0, "\u03b1 preferred", "\u03b2 preferred")
  )

for (corp in c("BabyLM", "C4")) {
  d <- filter(dat_disagree_trace, corpus == corp)
  if (nrow(d) == 0) next

  # RelFreq reference lines (corpus-level, constant per binom)
  d_freq <- d |> distinct(binom, log_freq_ratio, signed_gap_125m)

  p <- ggplot(d, aes(y = binom, x = preference, colour = model_size, shape = model_size)) +
    geom_vline(xintercept = 0, linetype = "solid", colour = "grey60", linewidth = 0.4) +
    geom_segment(data = d_freq,
                 aes(y = binom, yend = binom, x = 0, xend = log_freq_ratio),
                 colour = "#cc8800", linewidth = 0.7, linetype = "dashed",
                 inherit.aes = FALSE) +
    geom_point(size = 3.5, stroke = 0.8) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_shape_manual(values = SIZE_SHAPES, name = "Model size") +
    scale_x_continuous("Model preference (log-odds; + = prefer alpha form)") +
    labs(
      title    = sprintf("%s — 125M disagreement items across model sizes", corp),
      subtitle = "Dashed gold segments = corpus RelFreq direction; dots = model preference per size"
    ) +
    theme_bw(base_size = 11) +
    theme(
      axis.title.y    = element_blank(),
      strip.text      = element_text(face = "bold"),
      legend.position = "bottom"
    )
  print(p)
  cat("\n\n")
}
```

## Does AbsPref predict the right direction across model sizes?

Two complementary views of the same question. The **heatmap** shows every
disagreement item × model-size cell: green = AbsPref predicts the correct
preference direction; red = AbsPref wrong. The **bar chart** summarises the
percentage of items where AbsPref is correct for each model size, so we can
see whether the effect strengthens or weakens with scale.

```{r plot-abspref-heatmap, fig.height=10}
dat_abspref_grid <- dat_final |>
  semi_join(disagree_125m, by = c("corpus", "binom")) |>
  left_join(disagree_125m |> select(corpus, binom, signed_gap_125m),
            by = c("corpus", "binom")) |>
  mutate(
    binom          = fct_reorder(binom, abs(signed_gap_125m)),
    abspref_result = if_else(abspref_ok, "AbsPref correct", "AbsPref wrong")
  )

for (corp in c("BabyLM", "C4")) {
  d <- filter(dat_abspref_grid, corpus == corp)
  if (nrow(d) == 0) next

  p <- ggplot(d, aes(x = model_size, y = binom, fill = abspref_result)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(
      label = sprintf("%.2f", preference),
      colour = abspref_result
    ), size = 3.0, fontface = "bold") +
    scale_fill_manual(
      values = c("AbsPref correct" = "#a8ddb5", "AbsPref wrong" = "#fc8d59"),
      name   = NULL
    ) +
    scale_colour_manual(
      values = c("AbsPref correct" = "#084081", "AbsPref wrong" = "#7f0000"),
      guide  = "none"
    ) +
    scale_x_discrete(position = "top") +
    labs(
      title    = sprintf("%s — AbsPref direction agreement by model size", corp),
      subtitle = "Cell value = model preference (log-odds). Items ordered by |signed gap| in 125M (largest at top).",
      x        = "Model size", y = NULL
    ) +
    theme_bw(base_size = 11) +
    theme(
      legend.position  = "bottom",
      axis.text.y      = element_text(size = 9),
      panel.grid       = element_blank(),
      plot.title       = element_text(face = "bold")
    )

  print(p)
  cat("\n\n")
}
```

```{r plot-abspref-pct-bar, fig.height=4}
abspref_pct <- dat_abspref_grid |>
  group_by(corpus, model_size) |>
  summarise(
    n          = n(),
    pct_correct = mean(abspref_ok, na.rm = TRUE) * 100,
    se          = sqrt(pct_correct / 100 * (1 - pct_correct / 100) / n) * 100,
    .groups    = "drop"
  )

ggplot(abspref_pct, aes(x = model_size, y = pct_correct, fill = corpus)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  geom_errorbar(aes(ymin = pct_correct - 1.96 * se, ymax = pct_correct + 1.96 * se),
                position = position_dodge(0.7), width = 0.25) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey40") +
  annotate("text", x = 0.5, y = 52, label = "chance (50%)", hjust = 0,
           colour = "grey40", size = 3.5) +
  geom_text(aes(label = sprintf("%.0f%%\n(n=%d)", pct_correct, n),
                y = pct_correct + 1.96 * se + 3),
            position = position_dodge(0.7), size = 3.0, vjust = 0) +
  scale_fill_manual(values = c(BabyLM = "#154733", C4 = "#F0B323"), name = "Corpus") +
  scale_y_continuous("AbsPref predicts correct direction (%)",
                     limits = c(0, 110), breaks = seq(0, 100, 20)) +
  labs(
    title    = "% of 125M-disagreement items where AbsPref is correct — by model size",
    subtitle = "Items selected from 125M only; same items traced in 350M and 1.3B",
    x        = "Model size"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "right")
```

## Full replication table

```{r disagree-replicate-table}
replicate_summary <- dat_final |>
  semi_join(disagree_125m, by = c("corpus", "binom")) |>
  group_by(corpus, binom) |>
  summarise(
    n_sizes_abspref_ok = sum(abspref_ok),
    n_sizes_relfreq_ok = sum(relfreq_ok),
    n_sizes            = n_distinct(model_size),
    pref_125m          = preference[model_size == "125M"][1],
    pref_350m          = preference[model_size == "350M"][1],
    pref_13b           = preference[model_size == "1.3B"][1],
    .groups            = "drop"
  ) |>
  left_join(items_125m |> select(corpus, binom, GenPref_centered,
                                  log_freq_ratio, signed_gap_125m),
            by = c("corpus", "binom")) |>
  arrange(corpus, desc(abs(signed_gap_125m)))

replicate_summary |>
  mutate(across(c(pref_125m, pref_350m, pref_13b, log_freq_ratio, signed_gap_125m,
                   GenPref_centered), \(x) round(x, 2))) |>
  select(corpus, binom, log_freq_ratio, GenPref_centered,
         pref_125m, pref_350m, pref_13b,
         n_sizes_abspref_ok, n_sizes_relfreq_ok, n_sizes) |>
  rename(
    Corpus = corpus, Binomial = binom,
    `RelFreq\nlog(α/β)` = log_freq_ratio,
    `GenPref\n(centered)` = GenPref_centered,
    `Pref\n125M` = pref_125m, `Pref\n350M` = pref_350m, `Pref\n1.3B` = pref_13b,
    `N sizes:\nAbsPref OK` = n_sizes_abspref_ok,
    `N sizes:\nRelFreq OK` = n_sizes_relfreq_ok,
    `N sizes\ntotal` = n_sizes
  ) |>
  kbl(caption = paste(
    "Items where 125M disagrees with RelFreq, traced across model sizes.",
    "Positive preference = prefers alpha form.")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 11)
```

------------------------------------------------------------------------

# Section 2: Near-Flat RelFreq Items (Identified in 125M)

Items where the corpus RelFreq is near-chance in the 125M model
(`|log(α/β)| <` `r FLAT_THRESHOLD`). These are the same items in 350M and
1.3B (same training corpus). When the frequency signal is flat, how do
AbsPref and OverallFreq jointly determine preferences — and does this
replicate across model sizes?

```{r flat-selection}
flat_125m <- items_125m |>
  filter(abs(log_freq_ratio) < FLAT_THRESHOLD)

cat(sprintf("Near-flat items in 125M: BabyLM = %d, C4 = %d\n",
  sum(flat_125m$corpus == "BabyLM"), sum(flat_125m$corpus == "C4")))

dat_flat <- dat_final |>
  semi_join(flat_125m, by = c("corpus", "binom")) |>
  filter(!is.na(GenPref_centered), !is.na(preference), !is.na(log_total))
```

## Scatter: preference ~ AbsPref, overlaid across model sizes

Same items, same corpus — the three model sizes are overlaid so you can see
whether the AbsPref–preference relationship is consistent.

```{r plot-flat-scatter, fig.height=6}
ggplot(dat_flat, aes(x = GenPref_centered, y = preference,
                     colour = model_size, fill = model_size)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_smooth(aes(group = model_size), method = "lm", se = TRUE,
              linewidth = 0.9, alpha = 0.12) +
  geom_point(shape = 21, size = 2.0, alpha = 0.55, colour = "grey30", stroke = 0.3) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~ corpus) +
  labs(
    title    = "Near-flat RelFreq items: preference ~ AbsPref across model sizes",
    subtitle = sprintf("Items selected where |log(\u03b1/\u03b2)| < %.1f in 125M; same items shown for all sizes",
                       FLAT_THRESHOLD),
    x = "GenPref (centered)", y = "Model preference (log-odds of alpha)"
  ) +
  theme_bw(base_size = 12) +
  theme(strip.text = element_text(face = "bold"), legend.position = "bottom")
```

## AbsPref × OverallFreq — faceted by model size

```{r plot-flat-interaction, fig.height=8}
dat_flat_grp <- dat_flat |>
  group_by(corpus) |>
  mutate(
    exposure_grp = if_else(
      log_total > median(log_total, na.rm = TRUE), "High overall freq", "Low overall freq"
    )
  ) |>
  ungroup()

ggplot(dat_flat_grp, aes(x = GenPref_centered, y = preference,
                          colour = exposure_grp, linetype = exposure_grp)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, alpha = 0.12) +
  geom_point(aes(shape = model_size), size = 1.8, alpha = 0.5) +
  scale_colour_manual(values = c("High overall freq" = "#154733", "Low overall freq" = "#74add1"),
                      name = "Overall freq") +
  scale_linetype_manual(values = c("High overall freq" = "solid", "Low overall freq" = "dashed"),
                        name = "Overall freq") +
  scale_shape_manual(values = SIZE_SHAPES, name = "Model size") +
  facet_grid(corpus ~ model_size) +
  labs(
    title    = "AbsPref \u00d7 OverallFreq for near-flat RelFreq items",
    subtitle = "Solid = above-median overall freq; dashed = below. Same items in each model-size panel.",
    x = "GenPref (centered)", y = "Model preference (log-odds)"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "bottom")
```

## OLS coefficients per corpus × model size

```{r flat-ols-table}
flat_ols <- dat_flat |>
  group_by(corpus, model_size) |>
  group_modify(function(d, k) {
    fit <- lm(preference ~ GenPref_centered * log_total_c, data = d)
    cf  <- coef(summary(fit))
    as_tibble(cf, rownames = "term") |>
      rename(estimate = Estimate, se = `Std. Error`, t = `t value`, p = `Pr(>|t|)`) |>
      filter(term != "(Intercept)") |>
      mutate(
        term = recode(term,
          "GenPref_centered"             = "AbsPref",
          "log_total_c"                  = "OverallFreq (log, z)",
          "GenPref_centered:log_total_c" = "AbsPref \u00d7 OverallFreq"
        ),
        R2 = round(summary(fit)$r.squared, 3),
        n  = nrow(d)
      )
  }) |>
  ungroup() |>
  mutate(
    estimate = round(estimate, 3),
    se       = round(se, 3),
    p_label  = case_when(p < .001 ~ "< .001", p < .01 ~ "< .01",
                         p < .05 ~ "< .05",   TRUE ~ sprintf("%.3f", p))
  )

flat_ols |>
  select(corpus, model_size, term, estimate, se, p_label, R2, n) |>
  rename(Corpus = corpus, Size = model_size, Term = term,
         Estimate = estimate, SE = se, p = p_label) |>
  kbl(caption = paste(
    "OLS: preference ~ AbsPref * OverallFreq, near-flat RelFreq items only.",
    "Same items used across 125M/350M/1.3B within each corpus.")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

------------------------------------------------------------------------

# Section 3: Case Studies — Full Training Trajectories

Take the top **`r N_CASE_BINOMS`** 125M-disagreement items per corpus (largest
`|signed_gap|` in 125M, AbsPref predicts correct direction) and plot their
full training trajectories for all three model sizes.

For each binomial:
- **Coloured solid lines** = model preference (log-odds) across training, one per model size
- **Dashed gold line** = corpus RelFreq (same for all sizes; cumulative count ratio)
- **Horizontal purple line** = AbsPref (constant, not training-dependent)

```{r select-case-binoms}
case_binoms <- disagree_125m |>
  filter(abspref_ok_125m) |>
  group_by(corpus) |>
  slice_max(abs(signed_gap_125m), n = N_CASE_BINOMS) |>
  ungroup()

cat(sprintf("Case study binomials: %d BabyLM, %d C4\n",
  sum(case_binoms$corpus == "BabyLM"), sum(case_binoms$corpus == "C4")))
print(case_binoms |> select(corpus, binom, log_freq_ratio, GenPref_centered, signed_gap_125m))
```

```{r load-case-traj}
case_binom_vec <- unique(case_binoms$binom)
message(sprintf("Loading trajectories for %d binomials (single DuckDB scan)...", length(case_binom_vec)))

# One scan of the Parquet file for all 20 checkpoints × all models, filtered to
# case binomials — replaces the previous loop of 120 separate DuckDB connections.
t0 <- proc.time()
freq_traj <- batch_dyn_freq(ck_meta, binom_filter = case_binom_vec)
message(sprintf("  DuckDB fetch done in %.1f s (%d rows)", (proc.time() - t0)[["elapsed"]], nrow(freq_traj)))

dat_traj <- train |>
  filter(binom %in% case_binom_vec) |>
  inner_join(ck_meta, by = c("model", "step")) |>
  attach_freq(freq_traj) |>
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS)
  ) |>
  filter(!is.na(tokens), !is.na(log_freq_ratio))

message(sprintf("Trajectory dataset: %d rows", nrow(dat_traj)))
```

```{r plot-case-traj, fig.height=5, results='asis'}
for (corp in c("BabyLM", "C4")) {
  these_binoms <- case_binoms |> filter(corpus == corp) |> pull(binom)

  for (binom_i in these_binoms) {
    d <- filter(dat_traj, corpus == corp, binom == binom_i, !is.na(log_freq_ratio))
    if (nrow(d) == 0) {
      cat(sprintf("\n*No data for %s / %s.*\n\n", corp, binom_i))
      next
    }

    abspref_val <- d$GenPref_centered[!is.na(d$GenPref_centered)][1]

    p <- ggplot(d, aes(x = tokens, y = preference,
                        colour = model_size, fill = model_size, shape = model_size)) +
      # Zero line
      geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.35) +
      # AbsPref reference
      geom_hline(yintercept = abspref_val, colour = "#9e3d8e",
                 linetype = "dotdash", linewidth = 0.8, alpha = 0.85) +
      annotate("text", x = -Inf, y = abspref_val, hjust = -0.05, vjust = -0.45,
               label = sprintf("AbsPref = %.2f", abspref_val), colour = "#9e3d8e", size = 3.2) +
      # Corpus RelFreq trajectory (same for all sizes — use one model's data)
      geom_line(data = d |> filter(model_size == first(levels(d$model_size))) |>
                  distinct(tokens, log_freq_ratio),
                aes(x = tokens, y = log_freq_ratio), colour = "#cc8800",
                linetype = "dashed", linewidth = 0.9, inherit.aes = FALSE) +
      annotate("text", x = Inf, y = filter(d, model_size == first(levels(d$model_size))) |>
                 arrange(desc(tokens)) |> slice(1) |> pull(log_freq_ratio),
               hjust = 1.05, vjust = -0.4, label = "RelFreq", colour = "#cc8800", size = 3.2) +
      # Model preference per size
      geom_line(linewidth = 0.85) +
      geom_point(size = 2.0, stroke = 0.5, colour = "grey25") +
      scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
      scale_fill_manual(values = SIZE_COLOURS, name = "Model size") +
      scale_shape_manual(values = SIZE_SHAPES, name = "Model size") +
      scale_x_continuous("Training tokens",
                         labels = label_number(scale_cut = cut_short_scale())) +
      scale_y_continuous("Log-odds preference (+ = prefer alpha form)") +
      labs(
        title    = sprintf('%s \u2014 "%s"', corp, binom_i),
        subtitle = sprintf(
          "RelFreq log(\u03b1/\u03b2) = %.2f (predicts %s); AbsPref = %.2f (predicts %s)",
          mean(d$log_freq_ratio, na.rm = TRUE),
          if_else(mean(d$log_freq_ratio, na.rm = TRUE) > 0, "\u03b1", "\u03b2"),
          abspref_val,
          if_else(abspref_val > 0, "\u03b1", "\u03b2"))
      ) +
      theme_bw(base_size = 12) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

    print(p)
    cat("\n\n")
  }
}
```

------------------------------------------------------------------------

# Summary

```{r summary-table}
tibble(
  Analysis = c(
    "BabyLM: RelFreq wrong in 125M",
    "  \u2514 AbsPref correct in 125M",
    "  \u2514 Replicates in 350M/1.3B (AbsPref ok in all 3 sizes)",
    "C4: RelFreq wrong in 125M",
    "  \u2514 AbsPref correct in 125M",
    "  \u2514 Replicates in 350M/1.3B (AbsPref ok in all 3 sizes)",
    "BabyLM: near-flat RelFreq items",
    "C4: near-flat RelFreq items"
  ),
  N = c(
    sum(disagree_125m$corpus == "BabyLM"),
    sum(disagree_125m$abspref_ok_125m & disagree_125m$corpus == "BabyLM"),
    {
      r <- dat_final |>
        semi_join(filter(disagree_125m, corpus == "BabyLM"), by = c("corpus","binom")) |>
        group_by(binom) |>
        summarise(all_ok = all(abspref_ok))
      sum(r$all_ok)
    },
    sum(disagree_125m$corpus == "C4"),
    sum(disagree_125m$abspref_ok_125m & disagree_125m$corpus == "C4"),
    {
      r <- dat_final |>
        semi_join(filter(disagree_125m, corpus == "C4"), by = c("corpus","binom")) |>
        group_by(binom) |>
        summarise(all_ok = all(abspref_ok))
      sum(r$all_ok)
    },
    n_distinct(dat_flat$binom[dat_flat$corpus == "BabyLM"]),
    n_distinct(dat_flat$binom[dat_flat$corpus == "C4"])
  )
) |>
  kbl() |>
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```
