---
title: "Where Abstract Preference Outperforms Corpus Frequency"
subtitle: "Items where RelFreq fails in 125M models — traced across model sizes"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 6, dpi = 150,
                      out.width = "100%")
```

# Overview

We use the **125M model** as a discovery set within each corpus: identify
binomials where, at the final training checkpoint, the corpus relative frequency
(**RelFreq**) predicts the *wrong* preference direction. We then trace those
exact same binomials in the 350M and 1.3B models to ask whether the pattern
replicates, and whether abstract preference (**AbsPref / GenPref**) consistently
explains the deviation.

**Comparison structure**:
- Item selection is based on the 125M model's final checkpoint, separately per
  corpus (BabyLM and C4).
- The same items are then looked at in 350M and 1.3B within the same corpus.
- BabyLM and C4 items are kept separate throughout.

Three analyses:

1. **Disagreement items (§1)**: binomials where RelFreq is wrong in the 125M
   model — how often does AbsPref predict the right direction there, and do
   350M/1.3B also deviate from RelFreq for the same items?
2. **Near-flat RelFreq items (§2)**: binomials where the 125M corpus frequency
   signal is near-chance — how do AbsPref and OverallFreq jointly determine
   preferences in all three model sizes for those items?
3. **Case studies (§3)**: full training trajectories for the 5–10 strongest
   disagreement binomials from 125M, shown for all model sizes.

------------------------------------------------------------------------

```{r libraries}
library(tidyverse)
library(duckdb)
library(future)
library(furrr)
library(patchwork)
library(ggrepel)
library(scales)
library(kableExtra)
```

```{r config}
SCRIPT_DIR <- {
  knitr_input <- tryCatch(knitr::current_input(), error = function(e) NULL)
  if (!is.null(knitr_input) && nchar(knitr_input) > 0) {
    dirname(normalizePath(knitr_input, mustWork = FALSE))
  } else if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
    dirname(normalizePath(rstudioapi::getActiveDocumentContext()$path, mustWork = FALSE))
  } else {
    normalizePath(".", mustWork = FALSE)
  }
}

DATA_DIR          <- normalizePath(file.path(SCRIPT_DIR, "../Data/processed"), mustWork = FALSE)
TRAIN_CSV         <- file.path(DATA_DIR, "training_attested.csv")
EXPOSURES_CSV     <- normalizePath(file.path(DATA_DIR, "checkpoint_results_with_exposures.csv"),
                                   mustWork = FALSE)
EXPOSURES_CSV_FWD <- gsub("\\\\", "/", EXPOSURES_CSV)

N_TRAJ_CHECKPOINTS <- 20
SIZE_LEVELS        <- c("125M", "350M", "1.3B")
SIZE_COLOURS       <- c("125M" = "#74add1", "350M" = "#007030", "1.3B" = "#154733")
SIZE_SHAPES        <- c("125M" = 21, "350M" = 22, "1.3B" = 24)

# |log(alpha/beta)| below this = "near-flat RelFreq"
FLAT_THRESHOLD <- 0.3   # roughly P(alpha) in [0.43, 0.57]
# Number of case-study binomials
N_CASE_BINOMS  <- 8

# Parallelism
N_CORES <- parallel::detectCores(logical = FALSE)
message(sprintf("Detected %d physical cores", N_CORES))
plan(multicore, workers = N_CORES)

# Parquet cache path (same location as the CSV, faster for repeated reads)
EXPOSURES_PARQUET     <- sub("\\.csv$", ".parquet", EXPOSURES_CSV)
EXPOSURES_PARQUET_FWD <- gsub("\\\\", "/", EXPOSURES_PARQUET)

stopifnot(file.exists(TRAIN_CSV), file.exists(EXPOSURES_CSV))
```

```{r helpers}
parse_corpus <- function(x) case_when(
  grepl("babylm", x, TRUE) ~ "BabyLM",
  grepl("opt.c4", x, TRUE) ~ "C4",
  TRUE ~ NA_character_
)
parse_size <- function(x) case_when(
  grepl("125m",    x, TRUE) ~ "125M",
  grepl("350m",    x, TRUE) ~ "350M",
  grepl("1[._]3b", x, TRUE) ~ "1.3B",
  TRUE ~ NA_character_
)
sample_steps <- function(model_df, n = N_TRAJ_CHECKPOINTS) {
  steps <- sort(unique(model_df$step))
  idx   <- unique(round(seq(1, length(steps), length.out = min(n, length(steps)))))
  steps[idx]
}

# ── One-time Parquet conversion ───────────────────────────────────────────────
# Converts the 458 MB CSV to Parquet on first run. Subsequent runs skip this.
# DuckDB reads Parquet ~5-10x faster than CSV (columnar, no parsing overhead).
ensure_parquet <- function() {
  if (file.exists(EXPOSURES_PARQUET)) {
    message(sprintf("Parquet cache found: %s", EXPOSURES_PARQUET))
    return(invisible(NULL))
  }
  message("Converting exposures CSV to Parquet (one-time)...")
  t0  <- proc.time()
  con <- dbConnect(duckdb::duckdb())
  on.exit(dbDisconnect(con, shutdown = TRUE))
  dbExecute(con, sprintf("SET threads = %d", N_CORES))
  dbExecute(con, sprintf(
    "COPY (SELECT * FROM read_csv_auto('%s')) TO '%s' (FORMAT PARQUET, COMPRESSION ZSTD)",
    EXPOSURES_CSV_FWD, EXPOSURES_PARQUET_FWD
  ))
  elapsed <- (proc.time() - t0)[["elapsed"]]
  message(sprintf("Parquet written in %.1f s", elapsed))
}

# ── Batch DuckDB query ────────────────────────────────────────────────────────
# Fetches alpha_seen / beta_seen / tokens for all rows in meta_df (a data frame
# with columns `model` and `step`) in a single file scan.
# binom_filter: if supplied, only those binomials are returned (saves memory for
#               the trajectory load where we only need a handful of binomials).
batch_dyn_freq <- function(meta_df, binom_filter = NULL) {
  stopifnot(all(c("model", "step") %in% names(meta_df)))

  con <- dbConnect(duckdb::duckdb())
  on.exit(dbDisconnect(con, shutdown = TRUE))
  dbExecute(con, sprintf("SET threads = %d", N_CORES))

  src <- if (file.exists(EXPOSURES_PARQUET))
    sprintf("read_parquet('%s')", EXPOSURES_PARQUET_FWD)
  else
    sprintf("read_csv_auto('%s')", EXPOSURES_CSV_FWD)
  dbExecute(con, sprintf("CREATE VIEW exp AS SELECT * FROM %s", src))

  # Load the model × step pairs as an in-memory table for an efficient JOIN
  dbWriteTable(con, "pairs", meta_df |> distinct(model, step), overwrite = TRUE)

  binom_clause <- if (!is.null(binom_filter)) {
    binom_list <- paste0("'", gsub("'", "''", binom_filter), "'", collapse = ", ")
    sprintf("AND e.binom IN (%s)", binom_list)
  } else ""

  freq_raw <- dbGetQuery(con, sprintf("
    SELECT e.model, e.step, e.binom,
           ANY_VALUE(e.alpha_seen) AS alpha_seen,
           ANY_VALUE(e.beta_seen)  AS beta_seen,
           ANY_VALUE(e.tokens)     AS tokens
    FROM exp e
    INNER JOIN pairs p ON e.model = p.model AND e.step = p.step
    %s
    GROUP BY e.model, e.step, e.binom
  ", binom_clause))

  as_tibble(freq_raw) |>
    mutate(
      log_freq_ratio = if_else(alpha_seen > 0 & beta_seen > 0,
                               log(alpha_seen / beta_seen), NA_real_),
      freq_prob_c    = plogis(log_freq_ratio) - 0.5,
      log_total      = if_else(alpha_seen + beta_seen > 0,
                               log(alpha_seen + beta_seen), NA_real_)
    )
  # alpha_seen and beta_seen are kept for downstream count annotations
}

# ── Join freq data onto training data and compute derived columns ─────────────
# training_attested.csv also has a `tokens` column, which conflicts with the
# `tokens` returned by batch_dyn_freq. We keep the freq_df version (checkpoint-
# specific token count from the exposures file) and drop the training one.
attach_freq <- function(train_subset, freq_df) {
  joined <- train_subset |>
    left_join(freq_df, by = c("model", "step", "binom"))

  # Resolve tokens conflict: freq_df tokens is .y; drop training tokens (.x)
  if ("tokens.y" %in% names(joined)) {
    joined <- joined |>
      select(-any_of("tokens.x")) |>
      rename(tokens = tokens.y)
  }

  joined |>
    mutate(signed_gap = preference - log_freq_ratio) |>
    group_by(model, step) |>
    mutate(log_total_c = as.numeric(scale(log_total))) |>
    ungroup()
}
```

------------------------------------------------------------------------

# Data Loading

```{r parquet-cache}
ensure_parquet()
```

```{r load-train}
message("Loading training data...")
train <- read_csv(TRAIN_CSV, show_col_types = FALSE)

MODELS_LIST <- sort(unique(train$model))
MODELS_LIST <- MODELS_LIST[!is.na(parse_corpus(MODELS_LIST))]
message(sprintf("%d models: %s", length(MODELS_LIST), paste(MODELS_LIST, collapse = ", ")))
```

```{r ck-index}
# 20-checkpoint sampled index — used for final-checkpoint item selection
ck_meta <- map_dfr(MODELS_LIST, function(m) {
  steps <- sample_steps(filter(train, model == m))
  tibble(model = m, ck_idx = seq_along(steps), step = steps)
})

final_ck_meta <- ck_meta |>
  group_by(model) |>
  filter(ck_idx == max(ck_idx)) |>
  ungroup()

# Full step index — every checkpoint in the training data, used for trajectories
ck_meta_full <- train |>
  filter(model %in% MODELS_LIST) |>
  distinct(model, step) |>
  arrange(model, step)

# Export ck_meta so compute_checkpoint_val_perplexity.py evaluates the same checkpoints
CK_META_CSV <- normalizePath(
  file.path(dirname(DATA_DIR), "training_quality", "ck_meta.csv"),
  mustWork = FALSE
)
write_csv(ck_meta, CK_META_CSV)
message(sprintf("ck_meta written to %s", CK_META_CSV))
```

```{r load-final}
# Single DuckDB scan for all models' final checkpoints — one file read total.
message("Fetching final-checkpoint exposures (single DuckDB scan)...")
t0 <- proc.time()
freq_final <- batch_dyn_freq(final_ck_meta)
message(sprintf("  Done in %.1f s (%d rows)", (proc.time() - t0)[["elapsed"]], nrow(freq_final)))

dat_final <- train |>
  inner_join(final_ck_meta |> select(model, step, ck_idx), by = c("model", "step")) |>
  attach_freq(freq_final) |>
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS)
  ) |>
  filter(!is.na(log_freq_ratio), !is.na(GenPref_centered), !is.na(preference)) |>
  mutate(
    pref_prob   = plogis(preference) - 0.5,          # P(alpha) - 0.5; same scale as GenPref_centered & freq_prob_c
    pref_dir    = sign(preference),
    freq_dir    = sign(log_freq_ratio),
    abspref_dir = sign(GenPref_centered),
    relfreq_ok  = (pref_dir == freq_dir),
    abspref_ok  = (pref_dir == abspref_dir)
  )

message(sprintf("Final dataset: %d rows, %d models, %d unique binomials",
  nrow(dat_final), n_distinct(dat_final$model), n_distinct(dat_final$binom)))

# Sanity check: exactly one step per model (final checkpoint only)
steps_per_model <- dat_final |> group_by(model) |> summarise(n_steps = n_distinct(step))
stopifnot("dat_final has more than one step per model" = all(steps_per_model$n_steps == 1))
```

## Identify items from 125M model

```{r items-from-125m}
# 125M final-checkpoint data — used as the discovery set for item selection
dat_125m <- dat_final |>
  filter(model_size == "125M", pref_dir != 0, freq_dir != 0)

# Per-corpus, per-binomial summary from 125M
items_125m <- dat_125m |>
  group_by(corpus, binom) |>
  summarise(
    log_freq_ratio   = first(log_freq_ratio),   # corpus-level (constant per binom×corpus)
    freq_prob_c      = first(freq_prob_c),       # P(alpha) - 0.5 for RelFreq
    GenPref_centered = first(GenPref_centered),  # constant per binom, already P(alpha) - 0.5
    log_total        = first(log_total),         # log(alpha_seen + beta_seen) at 125M final ckpt
    preference_125m  = first(preference),
    pref_prob_125m   = first(pref_prob),
    signed_gap_125m  = first(signed_gap),
    relfreq_ok_125m  = first(relfreq_ok),
    abspref_ok_125m  = first(abspref_ok),
    .groups          = "drop"
  ) |>
  mutate(
    freq_dir      = sign(log_freq_ratio),
    abspref_dir   = sign(GenPref_centered),
    pref_dir_125m = sign(preference_125m)
  )

cat(sprintf("125M items at final checkpoint:\n"))
cat(sprintf("  BabyLM: %d binomials\n", sum(items_125m$corpus == "BabyLM")))
cat(sprintf("  C4:     %d binomials\n", sum(items_125m$corpus == "C4")))
cat(sprintf("\nRelFreq predicts wrong direction in 125M:\n"))
cat(sprintf("  BabyLM: %d (%.1f%%)\n",
  sum(!items_125m$relfreq_ok_125m & items_125m$corpus == "BabyLM"),
  mean(!items_125m$relfreq_ok_125m[items_125m$corpus == "BabyLM"]) * 100))
cat(sprintf("  C4:     %d (%.1f%%)\n",
  sum(!items_125m$relfreq_ok_125m & items_125m$corpus == "C4"),
  mean(!items_125m$relfreq_ok_125m[items_125m$corpus == "C4"]) * 100))
```

------------------------------------------------------------------------

# Summary Statistics: When AbsPref and RelFreq Conflict

Across **all models and sizes** at the final checkpoint, we identify items
where AbsPref and RelFreq point in opposite directions and ask: when the model
follows AbsPref vs. when it follows RelFreq, what are the typical values of
AbsPref, RelFreq, and overall frequency?

```{r conflict-summary-stats}
# All final-checkpoint rows where AbsPref and RelFreq are in opposite directions.
# Values are *oriented* by the AbsPref direction (multiply by sign(GenPref_centered))
# so that AbsPref is always positive and RelFreq always negative, making means
# interpretable across items that differ in which ordering AbsPref favours.
dat_conflict <- dat_final |>
  filter(sign(GenPref_centered) != sign(freq_prob_c),
         GenPref_centered != 0, freq_prob_c != 0) |>
  mutate(
    model_follows = case_when(
      abspref_ok  & !relfreq_ok ~ "AbsPref",
      relfreq_ok  & !abspref_ok ~ "RelFreq",
      TRUE                      ~ "Neither / tie"
    ),
    orient        = sign(GenPref_centered),   # +1 or -1
    abspref_or    = GenPref_centered * orient, # always > 0
    relfreq_or    = freq_prob_c      * orient, # always < 0 (opposite direction)
    pref_or       = pref_prob        * orient  # > 0 if model follows AbsPref
  ) |>
  filter(model_follows != "Neither / tie")

conflict_summary <- dat_conflict |>
  group_by(corpus, model_size, model_follows) |>
  summarise(
    n              = n(),
    mean_abspref   = mean(abspref_or, na.rm = TRUE),  # oriented: always positive
    mean_relfreq   = mean(relfreq_or, na.rm = TRUE),  # oriented: always negative
    mean_pref      = mean(pref_or,    na.rm = TRUE),  # oriented: + if follows AbsPref
    mean_log_total = mean(log_total,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(c(mean_abspref, mean_relfreq, mean_pref, mean_log_total), \(x) round(x, 3)))

conflict_summary |>
  rename(
    Corpus                = corpus,
    Size                  = model_size,
    `Model follows`       = model_follows,
    N                     = n,
    `AbsPref\n(oriented)` = mean_abspref,
    `RelFreq\n(oriented)` = mean_relfreq,
    `Model pref\n(oriented)` = mean_pref,
    `ln(total)\nmean`     = mean_log_total
  ) |>
  kbl(caption = paste(
    "When AbsPref and RelFreq conflict: feature values by which signal the model follows.",
    "Values are oriented so AbsPref is always positive (multiply by sign(AbsPref));",
    "RelFreq is therefore always negative; model pref > 0 means it follows AbsPref.",
    "ln(total) = log(\u03b1-seen + \u03b2-seen).")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 11) |>
  collapse_rows(columns = 1:2, valign = "middle")
```

```{r conflict-summary-plot, fig.height=6}
conflict_summary_long <- dat_conflict |>
  group_by(corpus, model_follows) |>
  summarise(
    AbsPref          = mean(abspref_or, na.rm = TRUE),
    RelFreq          = mean(relfreq_or, na.rm = TRUE),
    `Model pref`     = mean(pref_or,    na.rm = TRUE),
    `ln(total freq)` = mean(log_total,  na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(c(AbsPref, RelFreq, `Model pref`, `ln(total freq)`),
               names_to = "metric", values_to = "value")

ggplot(conflict_summary_long,
       aes(x = model_follows, y = value, fill = model_follows)) +
  geom_col(width = 0.6) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey40") +
  scale_fill_manual(values = c("AbsPref" = "#9e3d8e", "RelFreq" = "#cc8800"),
                    name = "Model follows") +
  facet_grid(metric ~ corpus, scales = "free_y") +
  labs(
    title    = "Feature values when model follows AbsPref vs. RelFreq",
    subtitle = "Items where AbsPref and RelFreq conflict; values oriented so AbsPref > 0 (all models, final checkpoint)",
    x = NULL, y = "Mean value (AbsPref-oriented)"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "none",
        axis.text.x = element_text(face = "bold"))
```

## AbsPref vs. RelFreq following: counts and signal strengths

```{r diverge-summary}
dat_conflict |>
  group_by(corpus, model_size, model_follows) |>
  summarise(
    n            = n(),
    mean_abspref = mean(abs(GenPref_centered), na.rm = TRUE),
    mean_relfreq = mean(abs(freq_prob_c),      na.rm = TRUE),
    mean_gap     = mean(abs(GenPref_centered - freq_prob_c), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(across(c(mean_abspref, mean_relfreq, mean_gap), \(x) round(x, 3))) |>
  rename(
    Corpus          = corpus,
    Size            = model_size,
    `Model follows` = model_follows,
    N               = n,
    `|AbsPref|`     = mean_abspref,
    `|RelFreq|`     = mean_relfreq,
    `Gap`           = mean_gap
  ) |>
  kbl(caption = paste(
    "All conflict items (AbsPref and RelFreq in opposite directions).",
    "|AbsPref| = mean(|GenPref_centered|); |RelFreq| = mean(|freq_prob_c|);",
    "Gap = mean(|AbsPref - RelFreq|). All values on P(\u03b1)\u22120.5 scale.")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 11) |>
  collapse_rows(columns = 1:2, valign = "middle")
```

------------------------------------------------------------------------

# Section 1: Items Where AbsPref and RelFreq Conflict — Model Follows AbsPref

For each corpus, we select binomials where:
1. **AbsPref and RelFreq point in opposite directions** (`sign(GenPref_centered) ≠ sign(freq_prob_c)`)
2. **The 125M model's preference aligns with AbsPref** (not RelFreq)

These are the clearest cases where abstract preference overrides corpus frequency.
We then trace those same items in 350M and 1.3B to ask whether the pattern scales.

```{r disagree-selection}
disagree_125m <- items_125m |>
  filter(
    sign(GenPref_centered) != sign(freq_prob_c),  # AbsPref and RelFreq in opposite directions
    abspref_ok_125m                               # 125M model follows AbsPref (not RelFreq)
  ) |>
  arrange(corpus, desc(abs(signed_gap_125m)))

cat(sprintf("Items where AbsPref and RelFreq conflict AND 125M follows AbsPref:\n"))
cat(sprintf("  BabyLM: %d\n", sum(disagree_125m$corpus == "BabyLM")))
cat(sprintf("  C4:     %d\n", sum(disagree_125m$corpus == "C4")))
```

## Preference across model sizes for 125M-disagreement items

Each binomial row shows the model's preference (P(α) − 0.5 scale) for 125M, 350M,
and 1.3B. The dashed gold segment marks where RelFreq falls and the dashed purple
segment marks AbsPref (both same for all model sizes since same training corpus).
We expect preferences to align with AbsPref rather than RelFreq.

```{r plot-disagree-trace, fig.height=10}
# Get all-model-size data for these items
# Only pull columns from disagree_125m that aren't already in dat_final
# (GenPref_centered and log_freq_ratio already exist there from the training join)
# Top 15 per corpus by |signed_gap| in 125M
top15_disagree <- disagree_125m |>
  group_by(corpus) |>
  slice_max(abs(signed_gap_125m), n = 15) |>
  ungroup()

# dat_final is final-checkpoint only (verified by stopifnot in load-final chunk)
# Join log_total from items_125m (125M final-ckpt corpus total) for ordering/annotation
dat_disagree_trace <- dat_final |>
  semi_join(top15_disagree, by = c("corpus", "binom")) |>
  left_join(disagree_125m |> select(corpus, binom, signed_gap_125m, abspref_ok_125m),
            by = c("corpus", "binom")) |>
  left_join(items_125m |> select(corpus, binom, log_total_ref = log_total),
            by = c("corpus", "binom")) |>
  mutate(
    binom         = fct_reorder(binom, log_total_ref),   # high freq at top (fct_reorder ascending → top = high)
    abspref_label = if_else(GenPref_centered > 0, "\u03b1 preferred", "\u03b2 preferred")
  )

for (corp in c("BabyLM", "C4")) {
  d <- filter(dat_disagree_trace, corpus == corp)
  if (nrow(d) == 0) next

  # Reference lines: RelFreq and AbsPref (both on P(alpha)-0.5 scale)
  d_freq    <- d |> distinct(binom, freq_prob_c, signed_gap_125m)
  d_abspref <- d |> distinct(binom, GenPref_centered, signed_gap_125m)
  d_ann     <- d |> distinct(binom, log_total_ref) |>
    mutate(total_label = format(round(exp(log_total_ref)), big.mark = ",", scientific = FALSE))

  p <- ggplot(d, aes(y = binom, x = pref_prob, colour = model_size, shape = model_size)) +
    geom_vline(xintercept = 0, linetype = "solid", colour = "grey60", linewidth = 0.4) +
    geom_segment(data = d_freq,
                 aes(y = binom, yend = binom, x = 0, xend = freq_prob_c),
                 colour = "#cc8800", linewidth = 0.9, linetype = "dashed",
                 inherit.aes = FALSE) +
    geom_segment(data = d_abspref,
                 aes(y = binom, yend = binom, x = 0, xend = GenPref_centered),
                 colour = "#9e3d8e", linewidth = 0.9, linetype = "dashed",
                 inherit.aes = FALSE) +
    geom_point(size = 3.5, stroke = 0.8) +
    geom_text(data = d_ann,
              aes(y = binom, x = 0.53, label = total_label),
              inherit.aes = FALSE, hjust = 0, size = 2.6, colour = "grey40") +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_shape_manual(values = SIZE_SHAPES, name = "Model size") +
    scale_x_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form)") +
    coord_cartesian(xlim = c(-0.5, 0.5), clip = "off") +
    labs(
      title    = sprintf("%s \u2014 125M disagreement items across model sizes (ordered by overall freq)", corp),
      subtitle = "Gold dashed = RelFreq; purple dashed = AbsPref; dots = model pref per size; right labels = total corpus count"
    ) +
    theme_bw(base_size = 11) +
    theme(
      axis.title.y    = element_blank(),
      strip.text      = element_text(face = "bold"),
      legend.position = "bottom",
      plot.margin     = margin(5, 55, 5, 5, "pt")
    )
  print(p)
  cat("\n\n")
}
```

## Does AbsPref predict the right direction across model sizes?

Two complementary views of the same question. The **heatmap** shows every
disagreement item × model-size cell: green = AbsPref predicts the correct
preference direction; red = AbsPref wrong. The **bar chart** summarises the
percentage of items where AbsPref is correct for each model size, so we can
see whether the effect strengthens or weakens with scale.

```{r plot-abspref-heatmap, fig.height=8}
# dat_final is final-checkpoint only (verified by stopifnot in load-final chunk)
# Top 15 per corpus, matching the other Section 1 plots
dat_abspref_grid <- dat_final |>
  semi_join(top15_disagree, by = c("corpus", "binom")) |>
  left_join(disagree_125m |> select(corpus, binom, signed_gap_125m),
            by = c("corpus", "binom")) |>
  left_join(items_125m |> select(corpus, binom, log_total_ref = log_total),
            by = c("corpus", "binom")) |>
  mutate(abspref_result = if_else(abspref_ok, "AbsPref correct", "AbsPref wrong"))
# dat_final already carries GenPref_centered (P(alpha)-0.5) and freq_prob_c (same scale) from attach_freq

for (corp in c("BabyLM", "C4")) {
  d <- filter(dat_abspref_grid, corpus == corp)
  if (nrow(d) == 0) next

  # Build reference columns (AbsPref and RelFreq) as extra tiles — both on P(alpha)-0.5 scale
  d_refs <- d |>
    distinct(binom, signed_gap_125m, log_total_ref, GenPref_centered, freq_prob_c) |>
    pivot_longer(c(GenPref_centered, freq_prob_c),
                 names_to = "col", values_to = "val") |>
    mutate(
      col_label = recode(col, "GenPref_centered" = "AbsPref", "freq_prob_c" = "RelFreq"),
      col_label = factor(col_label, levels = c("AbsPref", "RelFreq"))
    )

  # Model preference tiles
  d_model <- d |>
    mutate(model_size = factor(as.character(model_size), levels = SIZE_LEVELS))

  # Order binomials high-to-low by overall frequency; label includes count
  binom_counts <- d |>
    distinct(binom, log_total_ref) |>
    arrange(log_total_ref)   # ascending so top of y-axis = highest freq
  binom_order  <- binom_counts$binom
  binom_labels <- setNames(
    sprintf("%s  (n=%s)", binom_counts$binom,
            format(round(exp(binom_counts$log_total_ref)), big.mark = ",", scientific = FALSE)),
    binom_counts$binom
  )

  d_refs$binom  <- factor(d_refs$binom,  levels = binom_order)
  d_model$binom <- factor(d_model$binom, levels = binom_order)

  p_refs <- ggplot(d_refs, aes(x = col_label, y = binom, fill = val)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(label = sprintf("%.2f", val)), size = 2.8, colour = "grey10") +
    scale_fill_gradient2(low = "#d73027", mid = "grey95", high = "#1a9850",
                         midpoint = 0, name = "P(\u03b1)\u22120.5", guide = "none") +
    scale_x_discrete(position = "top") +
    scale_y_discrete(labels = binom_labels) +
    labs(x = NULL, y = NULL) +
    theme_bw(base_size = 11) +
    theme(axis.text.y  = element_text(size = 8),
          axis.text.x  = element_text(face = "bold"),
          panel.grid   = element_blank())

  p_model <- ggplot(d_model, aes(x = model_size, y = binom, fill = abspref_result)) +
    geom_tile(colour = "white", linewidth = 0.6) +
    geom_text(aes(label = sprintf("%.2f", pref_prob),
                  colour = abspref_result), size = 2.8, fontface = "bold") +
    scale_fill_manual(values = c("AbsPref correct" = "#a8ddb5", "AbsPref wrong" = "#fc8d59"),
                      name = NULL) +
    scale_colour_manual(values = c("AbsPref correct" = "#084081", "AbsPref wrong" = "#7f0000"),
                        guide = "none") +
    scale_x_discrete(position = "top") +
    labs(x = NULL, y = NULL) +
    theme_bw(base_size = 11) +
    theme(axis.text.y  = element_blank(),
          axis.ticks.y = element_blank(),
          axis.text.x  = element_text(face = "bold"),
          panel.grid   = element_blank(),
          legend.position = "bottom")

  p_combined <- p_refs + p_model +
    patchwork::plot_layout(widths = c(2, 3)) +
    patchwork::plot_annotation(
      title    = sprintf("%s \u2014 AbsPref direction agreement by model size", corp),
      subtitle = "Left: AbsPref and RelFreq log-odds (green = prefer \u03b1, red = prefer \u03b2). Right: model preference; green = AbsPref correct, red = wrong.",
      theme    = theme(plot.title    = element_text(face = "bold", size = 12),
                       plot.subtitle = element_text(size = 10))
    )

  print(p_combined)
  cat("\n\n")
}
```


## Full replication table

```{r disagree-replicate-table}
replicate_summary <- dat_final |>
  semi_join(disagree_125m, by = c("corpus", "binom")) |>
  group_by(corpus, binom) |>
  summarise(
    n_sizes_abspref_ok = sum(abspref_ok),
    n_sizes_relfreq_ok = sum(relfreq_ok),
    n_sizes            = n_distinct(model_size),
    pref_125m          = pref_prob[model_size == "125M"][1],
    pref_350m          = pref_prob[model_size == "350M"][1],
    pref_13b           = pref_prob[model_size == "1.3B"][1],
    .groups            = "drop"
  ) |>
  left_join(items_125m |> select(corpus, binom, GenPref_centered,
                                  freq_prob_c, signed_gap_125m),
            by = c("corpus", "binom")) |>
  arrange(corpus, desc(abs(signed_gap_125m)))

replicate_summary |>
  group_by(corpus) |>
  slice_max(abs(signed_gap_125m), n = 15) |>
  ungroup() |>
  mutate(across(c(pref_125m, pref_350m, pref_13b, freq_prob_c,
                   GenPref_centered), \(x) round(x, 2))) |>
  select(corpus, binom, freq_prob_c, GenPref_centered,
         pref_125m, pref_350m, pref_13b,
         n_sizes_abspref_ok, n_sizes_relfreq_ok, n_sizes) |>
  rename(
    Corpus = corpus, Binomial = binom,
    `RelFreq\nP(a)-0.5` = freq_prob_c,
    `AbsPref\nP(a)-0.5` = GenPref_centered,
    `Pref\n125M` = pref_125m, `Pref\n350M` = pref_350m, `Pref\n1.3B` = pref_13b,
    `N sizes:\nAbsPref OK` = n_sizes_abspref_ok,
    `N sizes:\nRelFreq OK` = n_sizes_relfreq_ok,
    `N sizes\ntotal` = n_sizes
  ) |>
  kbl(caption = paste(
    "Top 15 items per corpus (by |signed gap| in 125M) where 125M disagrees with RelFreq,",
    "traced across model sizes. All preference values on P(\u03b1)\u22120.5 scale; + = prefers \u03b1 form.")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 11)
```

------------------------------------------------------------------------

# Section 1b: AbsPref vs RelFreq — Where They Diverge Most

These plots are independent of model behaviour: they select binomials purely on
the basis of the **gap between AbsPref and RelFreq**, then show where each model
lands alongside both reference values.

All three quantities are on the same **P(α) − 0.5** scale (range −0.5 to 0.5):
- `AbsPref` = `GenPref_centered` (already on this scale)
- `RelFreq` = `freq_prob_c` = `plogis(log(α_seen/β_seen)) − 0.5`
- Model preference = `plogis(preference) − 0.5`

Each row is a binomial. Markers show:

- **Purple triangle** = AbsPref (`GenPref_centered`, P(α) − 0.5)
- **Gold diamond** = corpus RelFreq (`freq_prob_c`, P(α) − 0.5)
- **Coloured circles** = each model's preference (`plogis(pref) − 0.5`), by size

```{r abspref-relfreq-helper}
# Shared helper: build long-format data and produce the comparison plot per corpus.
# items_df must have columns: corpus, binom, sort_val, log_total (from items_125m).
plot_abspref_vs_relfreq <- function(items_df, title_suffix, subtitle) {
  # items_df is derived from items_125m and already has GenPref_centered (AbsPref,
  # P(alpha)-0.5), freq_prob_c (RelFreq, P(alpha)-0.5), and log_total.
  # Model preference is pref_prob = plogis(preference) - 0.5, from dat_final.
  d_refs <- items_df |>
    pivot_longer(c(GenPref_centered, freq_prob_c),
                 names_to = "source", values_to = "x") |>
    mutate(source = recode(source,
      "GenPref_centered" = "AbsPref",
      "freq_prob_c"      = "RelFreq"
    ))

  d_models <- dat_final |>
    semi_join(items_df, by = c("corpus", "binom")) |>
    left_join(items_df |> select(corpus, binom, sort_val, log_total_ref = log_total),
              by = c("corpus", "binom")) |>
    transmute(corpus, binom, sort_val, log_total = log_total_ref,
              source = paste0("Model ", as.character(model_size)),
              x      = pref_prob)

  d_long <- bind_rows(
    d_refs |> select(corpus, binom, sort_val, log_total, source, x),
    d_models
  ) |>
    mutate(
      binom  = fct_reorder(binom, log_total),   # high overall freq at top
      source = factor(source, levels = c("AbsPref", "RelFreq",
                                          "Model 125M", "Model 350M", "Model 1.3B"))
    )

  colour_vals <- c(
    "AbsPref"    = "#9e3d8e",
    "RelFreq"    = "#cc8800",
    "Model 125M" = SIZE_COLOURS[["125M"]],
    "Model 350M" = SIZE_COLOURS[["350M"]],
    "Model 1.3B" = SIZE_COLOURS[["1.3B"]]
  )
  shape_vals <- c(
    "AbsPref"    = 17,   # filled triangle
    "RelFreq"    = 18,   # filled diamond
    "Model 125M" = 21,
    "Model 350M" = 22,
    "Model 1.3B" = 24
  )
  size_vals <- c(
    "AbsPref" = 4.5, "RelFreq" = 4.5,
    "Model 125M" = 3.0, "Model 350M" = 3.0, "Model 1.3B" = 3.0
  )

  for (corp in c("BabyLM", "C4")) {
    d <- filter(d_long, corpus == corp)
    if (nrow(d) == 0) next
    d_ann <- d |> distinct(binom, log_total) |>
      mutate(total_label = format(round(exp(log_total)), big.mark = ",", scientific = FALSE))
    p <- ggplot(d, aes(y = binom, x = x, colour = source, shape = source, size = source)) +
      geom_vline(xintercept = 0, colour = "grey60", linewidth = 0.4) +
      geom_point(alpha = 0.9, stroke = 0.4) +
      geom_text(data = d_ann,
                aes(y = binom, x = 0.53, label = total_label),
                inherit.aes = FALSE, hjust = 0, size = 2.6, colour = "grey40") +
      scale_colour_manual(values = colour_vals, name = NULL) +
      scale_shape_manual(values = shape_vals, name = NULL) +
      scale_size_manual(values = size_vals, name = NULL, guide = "none") +
      scale_x_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form; all values on the same scale)") +
      coord_cartesian(xlim = c(-0.5, 0.5), clip = "off") +
      labs(title    = sprintf("%s \u2014 %s", corp, title_suffix),
           subtitle = paste(subtitle, "(rows ordered high\u2192low overall freq; right labels = total corpus count)")) +
      theme_bw(base_size = 11) +
      theme(axis.title.y    = element_blank(),
            legend.position = "bottom",
            plot.title      = element_text(face = "bold"),
            plot.margin     = margin(5, 55, 5, 5, "pt"))
    print(p)
    cat("\n\n")
  }
}
```

## Plot 1b-i: Largest absolute gap between AbsPref and RelFreq

Top 15 binomials per corpus where `|GenPref_centered − freq_prob_c|` is largest
(both on P(α) − 0.5 scale), regardless of whether they agree or disagree in direction.

```{r plot-abspref-relfreq-gap, fig.height=8, results='asis'}
items_gap <- items_125m |>
  mutate(sort_val = abs(GenPref_centered - freq_prob_c)) |>
  group_by(corpus) |>
  slice_max(sort_val, n = 15) |>
  ungroup()

plot_abspref_vs_relfreq(
  items_gap,
  title_suffix = "Largest |AbsPref \u2212 RelFreq| gap",
  subtitle     = "Top 15 per corpus. Purple \u25b2 = AbsPref; gold \u25c6 = RelFreq; circles = model preference by size."
)
```

## Plot 1b-ii: AbsPref and RelFreq point in opposite directions (cross zero)

Top 15 binomials per corpus where AbsPref and RelFreq **disagree in direction**
(one positive, one negative), ranked by `|GenPref_centered| + |freq_prob_c|`
(both on P(α) − 0.5 scale) so that the strongest directional disagreements appear first.

```{r plot-abspref-relfreq-crosszero, fig.height=8, results='asis'}
items_cross <- items_125m |>
  filter(sign(GenPref_centered) != sign(freq_prob_c),
         GenPref_centered != 0, freq_prob_c != 0) |>
  mutate(sort_val = abs(GenPref_centered) + abs(freq_prob_c)) |>
  group_by(corpus) |>
  slice_max(sort_val, n = 15) |>
  ungroup()

plot_abspref_vs_relfreq(
  items_cross,
  title_suffix = "AbsPref and RelFreq point in opposite directions",
  subtitle     = paste(
    "Top 15 per corpus by |AbsPref| + |RelFreq|.",
    "Purple \u25b2 = AbsPref; gold \u25c6 = RelFreq; circles = model preference by size.",
    "AbsPref and RelFreq always on opposite sides of zero."
  )
)
```

## Plot 1b-iii: Items with extreme AbsPref (|AbsPref| > 0.4)

Binomials where abstract preference is very strong (`|GenPref_centered| > 0.4`,
i.e. more than 90% of the time appearing in one order in human preference data).
Same plot style as Section 1 — shows whether model preference aligns with the
strong abstract preference signal, and where RelFreq falls.

```{r plot-abspref-extreme, fig.height=10, results='asis'}
extreme_125m <- items_125m |>
  filter(abs(GenPref_centered) > 0.4) |>
  group_by(corpus) |>
  slice_max(abs(GenPref_centered), n = 15) |>
  ungroup()

cat(sprintf("Extreme AbsPref items (|AbsPref| > 0.4): BabyLM = %d, C4 = %d\n",
  sum(extreme_125m$corpus == "BabyLM"), sum(extreme_125m$corpus == "C4")))

# dat_final is final-checkpoint only; join log_total from items_125m for ordering
dat_extreme_trace <- dat_final |>
  semi_join(extreme_125m, by = c("corpus", "binom")) |>
  left_join(extreme_125m |> select(corpus, binom, log_total_ref = log_total),
            by = c("corpus", "binom")) |>
  mutate(binom = fct_reorder(binom, log_total_ref))  # high overall freq at top

for (corp in c("BabyLM", "C4")) {
  d <- filter(dat_extreme_trace, corpus == corp)
  if (nrow(d) == 0) next

  d_freq    <- d |> distinct(binom, freq_prob_c, GenPref_centered, log_total_ref)
  d_abspref <- d_freq
  d_ann     <- d_freq |>
    mutate(total_label = format(round(exp(log_total_ref)), big.mark = ",", scientific = FALSE))

  p <- ggplot(d, aes(y = binom, x = pref_prob, colour = model_size, shape = model_size)) +
    geom_vline(xintercept = 0, linetype = "solid", colour = "grey60", linewidth = 0.4) +
    geom_segment(data = d_freq,
                 aes(y = binom, yend = binom, x = 0, xend = freq_prob_c),
                 colour = "#cc8800", linewidth = 0.9, linetype = "dashed",
                 inherit.aes = FALSE) +
    geom_segment(data = d_abspref,
                 aes(y = binom, yend = binom, x = 0, xend = GenPref_centered),
                 colour = "#9e3d8e", linewidth = 0.9, linetype = "dashed",
                 inherit.aes = FALSE) +
    geom_point(size = 3.5, stroke = 0.8) +
    geom_text(data = d_ann,
              aes(y = binom, x = 0.53, label = total_label),
              inherit.aes = FALSE, hjust = 0, size = 2.6, colour = "grey40") +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_shape_manual(values = SIZE_SHAPES, name = "Model size") +
    scale_x_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form)") +
    coord_cartesian(xlim = c(-0.5, 0.5), clip = "off") +
    labs(
      title    = sprintf("%s \u2014 Extreme AbsPref items (|AbsPref| > 0.4)", corp),
      subtitle = "Gold dashed = RelFreq; purple dashed = AbsPref; dots = model pref per size; right labels = total corpus count. Ordered high\u2192low overall freq."
    ) +
    theme_bw(base_size = 11) +
    theme(axis.title.y = element_blank(), legend.position = "bottom",
          plot.margin  = margin(5, 55, 5, 5, "pt"))
  print(p)
  cat("\n\n")
}
```

## Plot 1b-iv: Extreme RelFreq vs. extreme AbsPref (contrast)

Two panels: left = items where RelFreq is extreme but AbsPref is not
(`|freq_prob_c| > 0.4`, `|GenPref_centered| ≤ 0.4`); right = items where
AbsPref is extreme but RelFreq is not (`|GenPref_centered| > 0.4`,
`|freq_prob_c| ≤ 0.4`). Top 15 per corpus per panel by signal strength.

```{r plot-extreme-contrast, fig.height=10, results='asis'}
items_relfreq_ext <- items_125m |>
  filter(abs(freq_prob_c) > 0.4, abs(GenPref_centered) <= 0.4) |>
  group_by(corpus) |>
  slice_max(abs(freq_prob_c), n = 15) |>
  ungroup() |>
  mutate(condition = "Extreme RelFreq\n(|RelFreq|>0.4, |AbsPref|\u22640.4)")

items_abspref_ext2 <- items_125m |>
  filter(abs(GenPref_centered) > 0.4, abs(freq_prob_c) <= 0.4) |>
  group_by(corpus) |>
  slice_max(abs(GenPref_centered), n = 15) |>
  ungroup() |>
  mutate(condition = "Extreme AbsPref\n(|AbsPref|>0.4, |RelFreq|\u22640.4)")

items_contrast <- bind_rows(items_relfreq_ext, items_abspref_ext2) |>
  mutate(binom_ord = fct_reorder(paste(binom, condition, sep = "___"), log_total))

dat_contrast <- dat_final |>
  inner_join(items_contrast |> select(corpus, binom, condition, binom_ord,
                                       log_total_ref = log_total),
             by = c("corpus", "binom"))

d_refs_contrast <- dat_contrast |>
  distinct(corpus, binom, binom_ord, condition, freq_prob_c, GenPref_centered)

d_ann_contrast <- dat_contrast |>
  distinct(corpus, binom, binom_ord, condition, log_total_ref) |>
  mutate(total_label = format(round(exp(log_total_ref)), big.mark = ",", scientific = FALSE))

for (corp in c("BabyLM", "C4")) {
  d     <- filter(dat_contrast,    corpus == corp)
  d_ref <- filter(d_refs_contrast, corpus == corp)
  d_a   <- filter(d_ann_contrast,  corpus == corp)
  if (nrow(d) == 0) next

  p <- ggplot(d, aes(y = binom_ord, x = pref_prob, colour = model_size, shape = model_size)) +
    geom_vline(xintercept = 0, linetype = "solid", colour = "grey60", linewidth = 0.4) +
    geom_segment(data = d_ref,
                 aes(y = binom_ord, yend = binom_ord, x = 0, xend = freq_prob_c),
                 colour = "#cc8800", linewidth = 0.9, linetype = "dashed",
                 inherit.aes = FALSE) +
    geom_segment(data = d_ref,
                 aes(y = binom_ord, yend = binom_ord, x = 0, xend = GenPref_centered),
                 colour = "#9e3d8e", linewidth = 0.9, linetype = "dashed",
                 inherit.aes = FALSE) +
    geom_point(size = 3.5, stroke = 0.8) +
    geom_text(data = d_a,
              aes(y = binom_ord, x = 0.53, label = total_label),
              inherit.aes = FALSE, hjust = 0, size = 2.6, colour = "grey40") +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_shape_manual(values = SIZE_SHAPES, name = "Model size") +
    scale_y_discrete(labels = function(x) sub("___.*", "", x)) +
    scale_x_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form)") +
    coord_cartesian(xlim = c(-0.5, 0.5), clip = "off") +
    facet_wrap(~ condition, scales = "free_y", ncol = 2) +
    labs(
      title    = sprintf("%s \u2014 Extreme signal contrast", corp),
      subtitle = "Gold dashed = RelFreq; purple dashed = AbsPref; dots = model pref per size.\nOrdered high\u2192low overall freq within each panel; right labels = total corpus count."
    ) +
    theme_bw(base_size = 11) +
    theme(axis.title.y    = element_blank(),
          legend.position = "bottom",
          strip.text      = element_text(face = "bold"),
          plot.margin     = margin(5, 55, 5, 5, "pt"))
  print(p)
  cat("\n\n")
}
```

------------------------------------------------------------------------

# Section 2: Near-Flat RelFreq Items (Identified in 125M)

Items where the corpus RelFreq is near-chance in the 125M model
(`|log(α/β)| <` `r FLAT_THRESHOLD`). These are the same items in 350M and
1.3B (same training corpus). When the frequency signal is flat, how do
AbsPref and OverallFreq jointly determine preferences — and does this
replicate across model sizes?

```{r flat-selection}
flat_125m <- items_125m |>
  filter(abs(log_freq_ratio) < FLAT_THRESHOLD)

cat(sprintf("Near-flat items in 125M: BabyLM = %d, C4 = %d\n",
  sum(flat_125m$corpus == "BabyLM"), sum(flat_125m$corpus == "C4")))

dat_flat <- dat_final |>
  semi_join(flat_125m, by = c("corpus", "binom")) |>
  filter(!is.na(GenPref_centered), !is.na(preference), !is.na(log_total))
```

## Tug-of-war: where AbsPref, RelFreq, and model preferences fall

For all items where **AbsPref and RelFreq point in opposite directions** (all models,
final checkpoint), we orient everything so AbsPref is always positive and RelFreq
always negative (using `dat_conflict` computed above). This lets us ask: when the
two signals conflict, where does each model's preference land?

**Plot A** shows the *distributions* of all three signals on the same axis — you can
see the spread and central tendency of AbsPref (always right), RelFreq (always left),
and each model's preference (in between). **Plot B** shows model preference as a
function of AbsPref strength, coloured by how strongly RelFreq opposes.

```{r plot-tug-of-war-dist, fig.height=5}
# Build long-format: one row per signal per item
# AbsPref and RelFreq are item-level; model pref is model×item level
conflict_refs <- dat_conflict |>
  distinct(corpus, binom, abspref_or, relfreq_or) |>
  pivot_longer(c(abspref_or, relfreq_or), names_to = "signal", values_to = "value_or") |>
  mutate(signal = recode(signal, "abspref_or" = "AbsPref", "relfreq_or" = "RelFreq"))

conflict_prefs <- dat_conflict |>
  transmute(corpus, binom,
            signal = paste0("Model ", as.character(model_size)),
            value_or = pref_or)

conflict_long <- bind_rows(conflict_refs, conflict_prefs) |>
  mutate(signal = factor(signal,
    levels = c("RelFreq", "Model 125M", "Model 350M", "Model 1.3B", "AbsPref")))

signal_colours <- c(
  "AbsPref"    = "#9e3d8e",
  "RelFreq"    = "#cc8800",
  "Model 125M" = SIZE_COLOURS[["125M"]],
  "Model 350M" = SIZE_COLOURS[["350M"]],
  "Model 1.3B" = SIZE_COLOURS[["1.3B"]]
)

ggplot(conflict_long, aes(x = value_or, y = signal, fill = signal, colour = signal)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_violin(alpha = 0.35, linewidth = 0.5, adjust = 1.2) +
  geom_boxplot(width = 0.18, outlier.size = 0.5, outlier.alpha = 0.3,
               colour = "grey20", fill = NA, linewidth = 0.4) +
  scale_fill_manual(values   = signal_colours, guide = "none") +
  scale_colour_manual(values = signal_colours, guide = "none") +
  scale_x_continuous("P(\u03b1) \u2212 0.5 (oriented: + = AbsPref direction, \u2212 = RelFreq direction)",
                     limits = c(-0.5, 0.5)) +
  facet_wrap(~ corpus) +
  labs(
    title    = "Tug-of-war: distributions of AbsPref, RelFreq, and model preference",
    subtitle = "All items where AbsPref and RelFreq conflict; oriented so AbsPref > 0, RelFreq < 0 (all models, final checkpoint)",
    y        = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), panel.grid.minor = element_blank())
```

```{r abspref-relfreq-cor}
items_125m |>
  group_by(corpus) |>
  summarise(
    n         = n(),
    r         = cor(GenPref_centered, freq_prob_c, use = "complete.obs"),
    r_sq      = r^2,
    .groups   = "drop"
  ) |>
  mutate(across(c(r, r_sq), \(x) round(x, 3))) |>
  rename(Corpus = corpus, N = n, `r` = r, `r²` = r_sq) |>
  kbl(caption = "Pearson correlation between AbsPref (GenPref_centered) and RelFreq (freq_prob_c) across all binomials, per corpus.") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

```{r plot-tug-of-war-dist-all, fig.height=5}
# Same violin style as plot-tug-of-war-dist but for ALL binomials (unoriented)
all_refs <- dat_final |>
  distinct(corpus, binom, GenPref_centered, freq_prob_c) |>
  pivot_longer(c(GenPref_centered, freq_prob_c),
               names_to = "signal", values_to = "value") |>
  mutate(signal = recode(signal,
                         "GenPref_centered" = "AbsPref",
                         "freq_prob_c"      = "RelFreq"))

all_prefs <- dat_final |>
  transmute(corpus, binom,
            signal = paste0("Model ", as.character(model_size)),
            value  = pref_prob)

all_long <- bind_rows(all_refs, all_prefs) |>
  mutate(signal = factor(signal,
    levels = c("RelFreq", "Model 125M", "Model 350M", "Model 1.3B", "AbsPref")))

ggplot(all_long, aes(x = value, y = signal, fill = signal, colour = signal)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_violin(alpha = 0.35, linewidth = 0.5, adjust = 1.2) +
  geom_boxplot(width = 0.18, outlier.size = 0.5, outlier.alpha = 0.3,
               colour = "grey20", fill = NA, linewidth = 0.4) +
  scale_fill_manual(values   = signal_colours, guide = "none") +
  scale_colour_manual(values = signal_colours, guide = "none") +
  scale_x_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form)",
                     limits = c(-0.5, 0.5)) +
  facet_wrap(~ corpus) +
  labs(
    title    = "Distributions of AbsPref, RelFreq, and model preference — all binomials",
    subtitle = "All items, final checkpoint, all model sizes. Values on P(\u03b1)\u22120.5 scale (not oriented).",
    y        = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), panel.grid.minor = element_blank())
```

```{r plot-tug-of-war-dist-neutral, fig.height=6}
# Violin plot split by which signal is extreme vs. neutral
# Extreme threshold: |value| > 0.25; neutral: |value| <= 0.25
neutral_long <- dat_final |>
  filter(
    (abs(GenPref_centered) <= 0.25 & abs(freq_prob_c) > 0.25) |
    (abs(GenPref_centered) > 0.25  & abs(freq_prob_c) <= 0.25)
  ) |>
  mutate(condition = if_else(
    abs(freq_prob_c) > 0.25,
    "Neutral AbsPref, Extreme RelFreq\n(|AbsPref|\u22640.25, |RelFreq|>0.25)",
    "Extreme AbsPref, Neutral RelFreq\n(|AbsPref|>0.25, |RelFreq|\u22640.25)"
  )) |>
  (\(d) bind_rows(
    d |> distinct(corpus, binom, condition, GenPref_centered, freq_prob_c) |>
      pivot_longer(c(GenPref_centered, freq_prob_c),
                   names_to = "signal", values_to = "value") |>
      mutate(signal = recode(signal,
                             "GenPref_centered" = "AbsPref",
                             "freq_prob_c"      = "RelFreq")),
    d |> transmute(corpus, binom, condition,
                   signal = paste0("Model ", as.character(model_size)),
                   value  = pref_prob)
  ))() |>
  mutate(signal = factor(signal,
    levels = c("RelFreq", "Model 125M", "Model 350M", "Model 1.3B", "AbsPref")))

ggplot(neutral_long, aes(x = value, y = signal, fill = signal, colour = signal)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_violin(alpha = 0.35, linewidth = 0.5, adjust = 1.2) +
  geom_boxplot(width = 0.18, outlier.size = 0.5, outlier.alpha = 0.3,
               colour = "grey20", fill = NA, linewidth = 0.4) +
  scale_fill_manual(values   = signal_colours, guide = "none") +
  scale_colour_manual(values = signal_colours, guide = "none") +
  scale_x_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form)",
                     limits = c(-0.5, 0.5)) +
  facet_grid(condition ~ corpus) +
  labs(
    title    = "Distributions when one signal is extreme and the other is neutral",
    subtitle = "Extreme = |value| > 0.25; neutral = |value| \u2264 0.25. Final checkpoint, all model sizes.",
    y        = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), panel.grid.minor = element_blank())
```

```{r plot-tug-of-war-dist-neutral-freq, fig.height=8, results='asis'}
# Filter to extreme/neutral condition items, then bin by frequency within corpus
CONDITIONS <- c(
  "Neutral AbsPref, Extreme RelFreq\n(|AbsPref|\u22640.25, |RelFreq|>0.25)",
  "Extreme AbsPref, Neutral RelFreq\n(|AbsPref|>0.25, |RelFreq|\u22640.25)"
)

dat_cond_freq <- dat_final |>
  filter(
    (abs(GenPref_centered) <= 0.25 & abs(freq_prob_c) > 0.25) |
    (abs(GenPref_centered) > 0.25  & abs(freq_prob_c) <= 0.25)
  ) |>
  mutate(condition = if_else(
    abs(freq_prob_c) > 0.25, CONDITIONS[1], CONDITIONS[2]
  )) |>
  group_by(corpus) |>
  mutate(
    freq_decile = ntile(log_total, 10),
    freq_grp    = case_when(
      freq_decile == 1                    ~ "Low (decile 1)",
      freq_decile == 10                   ~ "High (decile 10)",
      freq_decile >= 3 & freq_decile <= 8 ~ "Mid (deciles 3-8)",
      TRUE                                ~ NA_character_
    )
  ) |>
  ungroup() |>
  filter(!is.na(freq_grp)) |>
  mutate(freq_grp = factor(freq_grp,
    levels = c("Low (decile 1)", "Mid (deciles 3-8)", "High (decile 10)")))

# Mean corpus count per corpus × condition × freq_grp (for annotation)
freq_ann <- dat_cond_freq |>
  distinct(corpus, binom, condition, freq_grp, log_total) |>
  group_by(corpus, condition, freq_grp) |>
  summarise(
    mean_n  = mean(exp(log_total), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    label  = sprintf("mean n = %s", format(round(mean_n), big.mark = ",")),
    signal = "AbsPref",   # top row of violin y-axis
    value  = -0.48
  )

for (cond in CONDITIONS) {
  d <- dat_cond_freq |> filter(condition == cond)
  ann <- freq_ann      |> filter(condition == cond)

  d_long <- bind_rows(
    d |> distinct(corpus, binom, freq_grp, GenPref_centered, freq_prob_c) |>
      pivot_longer(c(GenPref_centered, freq_prob_c),
                   names_to = "signal", values_to = "value") |>
      mutate(signal = recode(signal,
                             "GenPref_centered" = "AbsPref",
                             "freq_prob_c"      = "RelFreq")),
    d |> transmute(corpus, binom, freq_grp,
                   signal = paste0("Model ", as.character(model_size)),
                   value  = pref_prob)
  ) |>
    mutate(signal = factor(signal,
      levels = c("RelFreq", "Model 125M", "Model 350M", "Model 1.3B", "AbsPref")))

  p <- ggplot(d_long, aes(x = value, y = signal, fill = signal, colour = signal)) +
    geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.4) +
    geom_violin(alpha = 0.35, linewidth = 0.5, adjust = 1.2) +
    geom_boxplot(width = 0.18, outlier.size = 0.5, outlier.alpha = 0.3,
                 colour = "grey20", fill = NA, linewidth = 0.4) +
    geom_text(data = ann,
              aes(x = value, y = signal, label = label),
              inherit.aes = FALSE, hjust = 0, vjust = -0.4,
              size = 2.8, colour = "grey30") +
    scale_fill_manual(values   = signal_colours, guide = "none") +
    scale_colour_manual(values = signal_colours, guide = "none") +
    scale_x_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form)",
                       limits = c(-0.5, 0.5)) +
    facet_grid(freq_grp ~ corpus) +
    labs(
      title    = sub("\n.*", "", cond),
      subtitle = sub(".*\n", "", cond),
      y        = NULL
    ) +
    theme_bw(base_size = 11) +
    theme(strip.text = element_text(face = "bold"), panel.grid.minor = element_blank())
  print(p)
  cat("\n\n")
}
```

```{r plot-tug-of-war-dist-freq, fig.height=7}
# Violin plot split into tertiles of overall frequency (log_total)
# Tertile breaks computed per corpus so high/mid/low are relative within each corpus
freq_long <- dat_final |>
  group_by(corpus) |>
  mutate(
    freq_decile = ntile(log_total, 10),
    freq_grp    = case_when(
      freq_decile == 1                    ~ "Low freq (decile 1)",
      freq_decile == 10                   ~ "High freq (decile 10)",
      freq_decile >= 3 & freq_decile <= 8 ~ "Mid freq (deciles 3-8)",
      TRUE                                ~ NA_character_
    )
  ) |>
  ungroup() |>
  filter(!is.na(freq_grp)) |>
  (\(d) bind_rows(
    d |> distinct(corpus, binom, freq_grp, GenPref_centered, freq_prob_c) |>
      pivot_longer(c(GenPref_centered, freq_prob_c),
                   names_to = "signal", values_to = "value") |>
      mutate(signal = recode(signal,
                             "GenPref_centered" = "AbsPref",
                             "freq_prob_c"      = "RelFreq")),
    d |> transmute(corpus, binom, freq_grp,
                   signal = paste0("Model ", as.character(model_size)),
                   value  = pref_prob)
  ))() |>
  mutate(
    signal   = factor(signal,
      levels = c("RelFreq", "Model 125M", "Model 350M", "Model 1.3B", "AbsPref")),
    freq_grp = factor(freq_grp, levels = c("Low freq (decile 1)",
                                            "Mid freq (deciles 3-8)",
                                            "High freq (decile 10)"))
  )

ggplot(freq_long, aes(x = value, y = signal, fill = signal, colour = signal)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_violin(alpha = 0.35, linewidth = 0.5, adjust = 1.2) +
  geom_boxplot(width = 0.18, outlier.size = 0.5, outlier.alpha = 0.3,
               colour = "grey20", fill = NA, linewidth = 0.4) +
  scale_fill_manual(values   = signal_colours, guide = "none") +
  scale_colour_manual(values = signal_colours, guide = "none") +
  scale_x_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form)",
                     limits = c(-0.5, 0.5)) +
  facet_grid(freq_grp ~ corpus) +
  labs(
    title    = "Distributions of AbsPref, RelFreq, and model preference by overall frequency",
    subtitle = "Tertiles of log_total within each corpus. Final checkpoint, all model sizes.",
    y        = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), panel.grid.minor = element_blank())
```

```{r plot-pref-by-signal-all, fig.height=7}
# Scatter: model pref ~ AbsPref (left facet) and model pref ~ RelFreq (right facet)
# All binomials, all model sizes, final checkpoint
dat_pref_by_signal <- dat_final |>
  pivot_longer(c(GenPref_centered, freq_prob_c),
               names_to = "signal", values_to = "signal_val") |>
  mutate(signal = recode(signal,
                         "GenPref_centered" = "AbsPref",
                         "freq_prob_c"      = "RelFreq"))

ggplot(dat_pref_by_signal, aes(x = signal_val, y = pref_prob,
                                colour = model_size, fill = model_size)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  geom_point(alpha = 0.12, size = 0.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, alpha = 0.15) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size",
                      aesthetics = c("colour", "fill")) +
  facet_grid(corpus ~ signal) +
  labs(
    title    = "Model preference as a function of AbsPref and RelFreq — all binomials",
    subtitle = "Each point = one binomial \u00d7 model size. Lines = per-model-size linear smooth.",
    x        = "Signal value  P(\u03b1)\u22120.5",
    y        = "Model preference  P(\u03b1)\u22120.5"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "bottom")
```

```{r plot-pref-both-signals-all, fig.height=5}
# Scatter of RelFreq ~ AbsPref for all binomials
# Dashed y=x line shows where the two signals are equal
dat_refs_all <- dat_final |>
  distinct(corpus, binom, GenPref_centered, freq_prob_c)

ggplot(dat_refs_all, aes(x = GenPref_centered, y = freq_prob_c)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40", linewidth = 0.7, alpha = 0.7) +
  geom_point(alpha = 0.3, size = 1.2, colour = "grey40") +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, colour = "grey20") +
  facet_wrap(~ corpus) +
  labs(
    title    = "RelFreq vs. AbsPref — all binomials",
    subtitle = "Dashed line = AbsPref equals RelFreq (y = x). Each point = one binomial.",
    x = "AbsPref  P(\u03b1)\u22120.5  (GenPref_centered)",
    y = "RelFreq  P(\u03b1)\u22120.5  (freq_prob_c)"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

```{r plot-abspref-relfreq-pref-heatmap, fig.height=7}
# Binned heatmap: AbsPref (x) × RelFreq (y), fill = mean model preference
ggplot(dat_final, aes(x = GenPref_centered, y = freq_prob_c, z = pref_prob)) +
  stat_summary_2d(fun = mean, bins = 20) +
  scale_fill_gradient2(low = "#cc8800", mid = "grey90", high = "#9e3d8e",
                       midpoint = 0, name = "Mean model\npref P(\u03b1)\u22120.5") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  facet_grid(corpus ~ model_size) +
  labs(
    title    = "Mean model preference across AbsPref \u00d7 RelFreq space — heatmap",
    subtitle = "Each cell = mean pref_prob of all binomials in that bin. Purple = follows AbsPref direction; gold = follows RelFreq direction.\nDashed line = AbsPref equals RelFreq (y = x).",
    x = "AbsPref  P(\u03b1)\u22120.5  (GenPref_centered)",
    y = "RelFreq  P(\u03b1)\u22120.5  (freq_prob_c)"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "right")
```

```{r plot-abspref-relfreq-pref-scatter, fig.height=7}
# Colored scatter: AbsPref (x) × RelFreq (y), colour = model preference
ggplot(dat_final, aes(x = GenPref_centered, y = freq_prob_c, colour = pref_prob)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40", linewidth = 0.6) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  geom_point(alpha = 0.4, size = 1.2) +
  scale_colour_gradient2(low = "#cc8800", mid = "grey80", high = "#9e3d8e",
                         midpoint = 0, name = "Model pref\nP(\u03b1)\u22120.5") +
  facet_grid(corpus ~ model_size) +
  labs(
    title    = "Model preference across AbsPref \u00d7 RelFreq space — scatter",
    subtitle = "Each point = one binomial. Purple = follows AbsPref direction; gold = follows RelFreq direction.\nDashed line = AbsPref equals RelFreq (y = x).",
    x = "AbsPref  P(\u03b1)\u22120.5  (GenPref_centered)",
    y = "RelFreq  P(\u03b1)\u22120.5  (freq_prob_c)"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "right")
```

```{r plot-tug-of-war-scatter, fig.height=7}
# Scatter: model preference ~ AbsPref strength, coloured by RelFreq opposition
# abspref_or always > 0 (oriented); relfreq_or always < 0; abs(relfreq_or) = opposition strength

# Reference data: one row per binomial (no model replication) for RelFreq line
dat_conflict_ref <- dat_conflict |>
  distinct(corpus, binom, abspref_or, relfreq_or)

ggplot(dat_conflict,
       aes(x = abspref_or, y = pref_or, colour = abs(relfreq_or))) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  # Perfect AbsPref tracking: y = x
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "#9e3d8e", alpha = 0.5, linewidth = 0.7) +
  annotate("text", x = 0.48, y = 0.50, label = "= AbsPref", hjust = 1,
           colour = "#9e3d8e", size = 3, fontface = "italic") +
  # Perfect RelFreq tracking: smooth of relfreq_or ~ abspref_or
  geom_smooth(data = dat_conflict_ref,
              aes(x = abspref_or, y = relfreq_or),
              method = "lm", se = FALSE, inherit.aes = FALSE,
              colour = "#cc8800", linetype = "dashed", linewidth = 0.7, alpha = 0.8) +
  annotate("text", x = 0.48, y = -0.36, label = "= RelFreq", hjust = 1,
           colour = "#cc8800", size = 3, fontface = "italic") +
  geom_point(alpha = 0.35, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, colour = "grey20") +
  scale_colour_gradient(low = "grey75", high = "#cc8800",
                        name = "|RelFreq|\nopposing\nstrength") +
  facet_grid(corpus ~ model_size) +
  labs(
    title    = "Model preference as a function of AbsPref (conflict items only)",
    subtitle = "Colour = how strongly RelFreq opposes (darker = stronger RelFreq pull).\nPurple dashed = perfect AbsPref tracking (y=x); gold dashed = perfect RelFreq tracking.",
    x = "AbsPref  P(\u03b1)\u22120.5 (oriented +)",
    y = "Model preference  P(\u03b1)\u22120.5 (oriented: + = follows AbsPref)"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "right")
```

## Tug-of-war across all binomials

The x-axis is `GenPref_centered − freq_prob_c` (both on P(α)−0.5 scale):
positive = AbsPref pulls harder toward α than RelFreq does; negative = RelFreq
pulls harder; zero = signals are equal. The y-axis is model preference. This
includes *all* binomials — not just conflict items — so we can see whether the
*magnitude of the difference* between the two signals predicts model behaviour
even when they agree in direction.

```{r plot-tug-of-war-all, fig.height=7}
dat_tug_all <- dat_final |>
  mutate(tug = GenPref_centered - freq_prob_c)

ggplot(dat_tug_all, aes(x = tug, y = pref_prob)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50", linewidth = 0.4) +
  geom_point(alpha = 0.15, size = 1, colour = "grey40") +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, colour = "grey20") +
  facet_grid(corpus ~ model_size) +
  labs(
    title    = "Tug-of-war across all binomials",
    subtitle = "x = AbsPref \u2212 RelFreq (both P(\u03b1)\u22120.5); + = AbsPref pulls harder toward \u03b1, \u2212 = RelFreq pulls harder.\nIncludes cases where signals agree in direction but differ in magnitude.",
    x = "Tug  (GenPref_centered \u2212 freq_prob_c)",
    y = "Model preference  P(\u03b1)\u22120.5"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))
```

```{r plot-tug-of-war-all-large, fig.height=6}
# Restrict to items where the two signals differ by >= 0.3
dat_tug_large <- dat_tug_all |>
  filter(abs(tug) >= 0.3) |>
  mutate(tug_dir = if_else(tug > 0, "AbsPref stronger", "RelFreq stronger"))

# Build long-format: AbsPref and RelFreq (item-level) + model prefs
tug_large_refs <- dat_tug_large |>
  distinct(corpus, binom, tug_dir, GenPref_centered, freq_prob_c) |>
  pivot_longer(c(GenPref_centered, freq_prob_c),
               names_to = "signal", values_to = "value") |>
  mutate(signal = recode(signal,
                         "GenPref_centered" = "AbsPref",
                         "freq_prob_c"      = "RelFreq"))

tug_large_prefs <- dat_tug_large |>
  transmute(corpus, binom, tug_dir,
            signal = paste0("Model ", as.character(model_size)),
            value  = pref_prob)

tug_large_long <- bind_rows(tug_large_refs, tug_large_prefs) |>
  mutate(signal = factor(signal,
    levels = c("RelFreq", "Model 125M", "Model 350M", "Model 1.3B", "AbsPref")))

ggplot(tug_large_long, aes(x = value, y = signal, fill = signal, colour = signal)) +
  geom_vline(xintercept = 0, colour = "grey50", linewidth = 0.4) +
  geom_violin(alpha = 0.35, linewidth = 0.5, adjust = 1.2) +
  geom_boxplot(width = 0.18, outlier.size = 0.5, outlier.alpha = 0.3,
               colour = "grey20", fill = NA, linewidth = 0.4) +
  scale_fill_manual(values = signal_colours, guide = "none") +
  scale_colour_manual(values = signal_colours, guide = "none") +
  scale_x_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form)",
                     limits = c(-0.5, 0.5)) +
  facet_grid(tug_dir ~ corpus) +
  labs(
    title    = "Tug-of-war: large signal differences only (|AbsPref \u2212 RelFreq| \u2265 0.3)",
    subtitle = "Rows split by which signal is stronger. All values on P(\u03b1)\u22120.5 scale (final checkpoint, all model sizes).",
    y = NULL
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), panel.grid.minor = element_blank())
```

## AbsPref × OverallFreq — faceted by model size

```{r plot-flat-interaction, fig.height=8}
dat_flat_grp <- dat_flat |>
  group_by(corpus) |>
  mutate(
    exposure_grp = if_else(
      log_total > median(log_total, na.rm = TRUE), "High overall freq", "Low overall freq"
    )
  ) |>
  ungroup()

ggplot(dat_flat_grp, aes(x = GenPref_centered, y = pref_prob,
                          colour = exposure_grp, linetype = exposure_grp)) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.9, alpha = 0.12) +
  geom_point(aes(shape = model_size), size = 1.8, alpha = 0.5) +
  scale_colour_manual(values = c("High overall freq" = "#154733", "Low overall freq" = "#74add1"),
                      name = "Overall freq") +
  scale_linetype_manual(values = c("High overall freq" = "solid", "Low overall freq" = "dashed"),
                        name = "Overall freq") +
  scale_shape_manual(values = SIZE_SHAPES, name = "Model size") +
  facet_grid(corpus ~ model_size) +
  labs(
    title    = "AbsPref \u00d7 OverallFreq for near-flat RelFreq items",
    subtitle = "Solid = above-median overall freq; dashed = below. Same items in each model-size panel.",
    x = "AbsPref  P(\u03b1) \u2212 0.5", y = "Model preference  P(\u03b1) \u2212 0.5"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"), legend.position = "bottom")
```

## OLS coefficients per corpus × model size

```{r flat-ols-table}
flat_ols <- dat_flat |>
  group_by(corpus, model_size) |>
  group_modify(function(d, k) {
    fit <- lm(pref_prob ~ GenPref_centered * log_total_c, data = d)
    cf  <- coef(summary(fit))
    as_tibble(cf, rownames = "term") |>
      rename(estimate = Estimate, se = `Std. Error`, t = `t value`, p = `Pr(>|t|)`) |>
      filter(term != "(Intercept)") |>
      mutate(
        term = recode(term,
          "GenPref_centered"             = "AbsPref",
          "log_total_c"                  = "OverallFreq (log, z)",
          "GenPref_centered:log_total_c" = "AbsPref \u00d7 OverallFreq"
        ),
        R2 = round(summary(fit)$r.squared, 3),
        n  = nrow(d)
      )
  }) |>
  ungroup() |>
  mutate(
    estimate = round(estimate, 3),
    se       = round(se, 3),
    p_label  = case_when(p < .001 ~ "< .001", p < .01 ~ "< .01",
                         p < .05 ~ "< .05",   TRUE ~ sprintf("%.3f", p))
  )

flat_ols |>
  select(corpus, model_size, term, estimate, se, p_label, R2, n) |>
  rename(Corpus = corpus, Size = model_size, Term = term,
         Estimate = estimate, SE = se, p = p_label) |>
  kbl(caption = paste(
    "OLS: pref_prob ~ AbsPref * OverallFreq, near-flat RelFreq items only.",
    "Outcome and AbsPref are both on P(\u03b1)\u22120.5 scale.",
    "Same items used across 125M/350M/1.3B within each corpus.")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

## OLS: human preference ~ model preference × overall frequency (final checkpoint, all items)

Does the relationship between model preference and human abstract preference depend on
how frequent the binomial is? Frequent binomials may be memorised directly from the
corpus, whereas rare ones require generalisation. An interaction of model preference
with overall frequency tests this.

```{r human-from-model-ols}
human_from_model_ols <- dat_final |>
  mutate(OverallFreq_log_c = as.numeric(scale(OverallFreq_log))) |>
  group_by(corpus, model_size) |>
  group_modify(function(d, k) {
    fit <- lm(GenPref_centered ~ pref_prob * OverallFreq_log_c, data = d)
    cf  <- coef(summary(fit))
    as_tibble(cf, rownames = "term") |>
      rename(estimate = Estimate, se = `Std. Error`, t = `t value`, p = `Pr(>|t|)`) |>
      filter(term != "(Intercept)") |>
      mutate(
        term = recode(term,
          "pref_prob"                       = "Model preference",
          "OverallFreq_log_c"               = "OverallFreq (log, z)",
          "pref_prob:OverallFreq_log_c"     = "Model preference \u00d7 OverallFreq"
        ),
        R2 = round(summary(fit)$r.squared, 3),
        n  = nrow(d)
      )
  }) |>
  ungroup() |>
  mutate(
    estimate = round(estimate, 3),
    se       = round(se, 3),
    p_label  = case_when(p < .001 ~ "< .001", p < .01 ~ "< .01",
                         p < .05  ~ "< .05",  TRUE    ~ sprintf("%.3f", p))
  )

human_from_model_ols |>
  select(corpus, model_size, term, estimate, se, p_label, R2, n) |>
  rename(Corpus = corpus, Size = model_size, Term = term,
         Estimate = estimate, SE = se, p = p_label) |>
  kbl(caption = paste(
    "OLS: GenPref_centered ~ model preference * OverallFreq (final checkpoint, all items).",
    "Outcome = human abstract preference P(\u03b1)\u22120.5.",
    "Model preference = plogis(preference) \u2212 0.5.",
    "OverallFreq log-transformed and z-scored.")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE)
```

------------------------------------------------------------------------

## ΔLL vs. Perplexity (Oh & Schuler–style plot)

Inspired by Oh & Schuler (2023): does higher model quality (lower perplexity) predict
better fit to human ordering preferences? Following their approach, we use individual
human trial data (binary choice: α ordering or not) and fit logistic regressions. For
each checkpoint of each model, ΔLL is the improvement from adding the model's ordering
preference (`preference`, log-odds) to a baseline logistic regression on corpus RelFreq.
Per-checkpoint **training** perplexity is derived from training loss curves as $\exp(\text{loss})$.
A second version of this plot using **validation** perplexity (computed by
`compute_checkpoint_val_perplexity.py` on the corpus-matched held-out sets) follows
immediately below.

```{r delta-ll-perplexity}
LOSS_CURVES_CSV  <- normalizePath(
  file.path(dirname(DATA_DIR), "training_quality/loss_curves.csv"),
  mustWork = FALSE
)
HUMAN_DATA_CSV <- normalizePath(
  file.path(dirname(DATA_DIR), "all_human_data.csv"),
  mustWork = FALSE
)

# Per-checkpoint perplexity from training loss: ppl = exp(loss)
loss_curves <- read_csv(LOSS_CURVES_CSV, show_col_types = FALSE) |>
  rename(model_size = params) |>
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    perplexity = exp(loss)
  ) |>
  select(corpus, model_size, step, perplexity)

# Individual human trial data: resp = "alpha" | "nonalpha"
human_trials <- read_csv(HUMAN_DATA_CSV, show_col_types = FALSE) |>
  mutate(
    binom      = Alpha,                          # "Word1 and Word2"
    resp_alpha = as.integer(resp == "alpha")     # 1 = chose α ordering
  ) |>
  select(binom, resp_alpha, RelFreq)

# Load exposures for all sampled checkpoints (ck_meta has 20 per model)
message("Fetching exposures for all sampled checkpoints (ΔLL trajectory)...")
t0 <- proc.time()
freq_all_ck <- batch_dyn_freq(ck_meta)
message(sprintf("  Done in %.1f s (%d rows)", (proc.time() - t0)[["elapsed"]], nrow(freq_all_ck)))

# Item-level model preferences per checkpoint
model_prefs_ck <- train |>
  inner_join(ck_meta |> select(model, step, ck_idx), by = c("model", "step")) |>
  attach_freq(freq_all_ck) |>
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS)
  ) |>
  filter(!is.na(freq_prob_c), !is.na(preference)) |>
  select(corpus, model_size, step, binom, preference, freq_prob_c)

# Join individual trials with per-checkpoint model preference + dynamic freq
# freq_prob_c is the checkpoint-specific training-exposure frequency (plogis(log RelFreq) - 0.5),
# so it varies across training steps, unlike the static RelFreq in all_human_data.csv.
dat_trials_ck <- human_trials |>
  select(binom, resp_alpha) |>
  inner_join(model_prefs_ck, by = "binom")

# ΔLL per checkpoint, Oh & Schuler (2023):
#   outcome  = binary human choice (resp_alpha; analog of reading times)
#   baseline = logistic(resp_alpha ~ freq_prob_c)         [dynamic, checkpoint-specific]
#   full     = logistic(resp_alpha ~ freq_prob_c + model preference)
#   ΔLL = logLik(full) - logLik(baseline)
delta_ll_traj <- dat_trials_ck |>
  group_by(corpus, model_size, step) |>
  group_modify(function(d, k) {
    baseline_mod <- glm(resp_alpha ~ freq_prob_c,              family = binomial, data = d)
    full_mod     <- glm(resp_alpha ~ freq_prob_c + preference, family = binomial, data = d)
    tibble(
      delta_ll = as.numeric(logLik(full_mod)) - as.numeric(logLik(baseline_mod)),
      n_trials = nrow(d),
      n_items  = n_distinct(d$binom)
    )
  }) |>
  ungroup()

# Join perplexity via nearest step in loss_curves for each (corpus, model_size, step)
delta_ll_traj <- delta_ll_traj |>
  left_join(
    loss_curves |> rename(lc_step = step),
    by = c("corpus", "model_size"),
    relationship = "many-to-many"
  ) |>
  mutate(step_dist = abs(step - lc_step)) |>
  group_by(corpus, model_size, step) |>
  slice_min(step_dist, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(-lc_step, -step_dist)

size_colours <- c("125M" = "#2166ac", "350M" = "#f4a582", "1.3B" = "#b2182b")
size_shapes  <- c("125M" = 16L,      "350M" = 17L,      "1.3B" = 15L)

dat_plot <- delta_ll_traj |> filter(!is.na(perplexity))

# ── Slope significance tests (log2 perplexity ~ delta_ll, per corpus × model_size) ──
# Even though we plot a loess smoother, we report the linear slope test so the
# direction and significance of the perplexity→ΔLL relationship can be stated.
slope_tests <- dat_plot |>
  group_by(corpus, model_size) |>
  group_modify(function(d, k) {
    if (nrow(d) < 3) return(tibble(slope = NA_real_, se = NA_real_, t = NA_real_, p = NA_real_, n = nrow(d)))
    fit <- lm(delta_ll ~ log2(perplexity), data = d)
    cf  <- summary(fit)$coefficients
    tibble(
      slope = cf["log2(perplexity)", "Estimate"],
      se    = cf["log2(perplexity)", "Std. Error"],
      t     = cf["log2(perplexity)", "t value"],
      p     = cf["log2(perplexity)", "Pr(>|t|)"],
      n     = nrow(d)
    )
  }) |>
  ungroup() |>
  mutate(
    sig = case_when(p < .001 ~ "***", p < .01 ~ "**", p < .05 ~ "*", p < .10 ~ ".", TRUE ~ "ns"),
    label = sprintf("β = %.3f, t(%d) = %.2f, p %s %s",
                    slope, n - 2L, t,
                    ifelse(p < .001, "<", "="),
                    ifelse(p < .001, ".001", sprintf("%.3f", p)),
                    sig)
  )

cat("\nSlope tests: log2(perplexity) → ΔLL\n")
print(slope_tests |> select(corpus, model_size, slope, se, t, p, sig, n), n = Inf)

# ── Between-size comparison: do model sizes have significantly different ΔLL curves? ──
# lm(delta_ll ~ log2(perplexity) * model_size) per corpus tests:
#   - main effect of model_size: overall level differences across sizes
#   - interaction log2(perplexity):model_size: slope differences across sizes
#   - pairwise contrasts: which specific pairs differ (Holm-corrected)
cat("\n── Between-size comparison (interaction model) per corpus ──\n")
if (!requireNamespace("emmeans", quietly = TRUE)) {
  message("  \u26a0  emmeans not installed \u2014 skipping pairwise contrasts (install.packages('emmeans'))")
} else {
  size_comparison <- dat_plot |>
    group_by(corpus) |>
    group_modify(function(d, k) {
      if (n_distinct(d$model_size) < 2)
        return(tibble(contrast = NA_character_, estimate = NA_real_, SE = NA_real_,
                      t.ratio = NA_real_, p.value = NA_real_))
      d <- d |> mutate(model_size = factor(model_size))
      fit  <- lm(delta_ll ~ log2(perplexity) * model_size, data = d)
      cat(sprintf("\n  %s \u2014 ANOVA:\n", k$corpus))
      aov_tbl <- anova(fit)
      print(aov_tbl)
      emm      <- emmeans::emmeans(fit, ~ model_size)
      pairs_df <- as.data.frame(pairs(emm, adjust = "holm"))
      pairs_df
    }) |>
    ungroup()
  cat("\nPairwise contrasts (model_size, Holm-adjusted, at mean log2 perplexity):\n")
  print(size_comparison |> select(corpus, contrast, estimate, SE, t.ratio, p.value), n = Inf)
}

# Label the best-trained (lowest-perplexity) checkpoint per (corpus, model_size)
label_pts <- dat_plot |>
  group_by(corpus, model_size) |>
  slice_min(perplexity, n = 1, with_ties = FALSE) |>
  ungroup()

ggplot(dat_plot,
       aes(x = perplexity, y = delta_ll,
           colour = model_size, shape = model_size)) +
  facet_wrap(~ corpus, scales = "free_x") +
  geom_point(size = 2, alpha = 0.45) +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
              linetype = "dashed", linewidth = 0.9, alpha = 0.12) +
  geom_text(data = label_pts,
            aes(label = model_size),
            hjust = -0.15, vjust = -0.4,
            size  = 3.8, fontface = "bold",
            show.legend = FALSE) +
  scale_x_continuous(
    "Perplexity (lower \u2192 better trained)",
    trans  = "log2",
    labels = scales::trans_format("log2", scales::math_format(2^.x))
  ) +
  scale_y_continuous("\u0394LL") +
  scale_colour_manual(values = size_colours, name = "Model size") +
  scale_shape_manual(values  = size_shapes,  name = "Model size") +
  guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)),
         shape  = guide_legend(override.aes = list(size = 3, alpha = 1))) +
  labs(
    title    = "\u0394LL vs. perplexity across training checkpoints",
    subtitle = paste0(
      "\u0394LL = logLik(resp_alpha \u223c freq_prob_c + model pref) \u2212 logLik(resp_alpha \u223c freq_prob_c). ",
      "Each point = one checkpoint. Dashed lines = loess smoother."
    )
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position   = "none",
    strip.background  = element_blank(),
    strip.text        = element_text(face = "bold", size = 13),
    panel.grid.major  = element_line(colour = "grey92", linewidth = 0.4),
    axis.line         = element_line(colour = "grey40"),
    axis.ticks        = element_line(colour = "grey40")
  )
```

## ΔLL vs. Validation Perplexity

Same analysis as above but the x-axis is **validation** perplexity, computed by
`compute_checkpoint_val_perplexity.py` on the corpus-matched held-out sets
(BabyLM: `train[:5%]` of `znhoughton/babylm-150m-v3`;
C4: `validation` split of `znhoughton/c4-subset-10B-tokens`).
This removes the optimistic bias of training perplexity and gives a true picture
of model quality at each checkpoint.

```{r delta-ll-val-perplexity}
VAL_PPL_CSV <- normalizePath(
  file.path(dirname(DATA_DIR), "training_quality", "val_perplexity_curves.csv"),
  mustWork = FALSE
)

if (!file.exists(VAL_PPL_CSV)) {
  message("val_perplexity_curves.csv not found — run compute_checkpoint_val_perplexity.py first.")
} else {
  val_ppl <- read_csv(VAL_PPL_CSV, show_col_types = FALSE) |>
    rename(model_size = params) |>
    mutate(model_size = factor(model_size, levels = SIZE_LEVELS))

  # Join validation perplexity to delta_ll_traj via nearest step
  delta_ll_val <- delta_ll_traj |>
    select(-perplexity) |>        # drop training perplexity
    left_join(
      val_ppl |> rename(lc_step = step),
      by = c("corpus", "model_size"),
      relationship = "many-to-many"
    ) |>
    mutate(step_dist = abs(step - lc_step)) |>
    group_by(corpus, model_size, step) |>
    slice_min(step_dist, n = 1, with_ties = FALSE) |>
    ungroup() |>
    select(-lc_step, -step_dist) |>
    filter(!is.na(val_perplexity))

  # ── Slope significance tests (validation perplexity) ─────────────────────
  slope_tests_val <- delta_ll_val |>
    group_by(corpus, model_size) |>
    group_modify(function(d, k) {
      if (nrow(d) < 3) return(tibble(slope = NA_real_, se = NA_real_, t = NA_real_, p = NA_real_, n = nrow(d)))
      fit <- lm(delta_ll ~ log2(val_perplexity), data = d)
      cf  <- summary(fit)$coefficients
      tibble(
        slope = cf["log2(val_perplexity)", "Estimate"],
        se    = cf["log2(val_perplexity)", "Std. Error"],
        t     = cf["log2(val_perplexity)", "t value"],
        p     = cf["log2(val_perplexity)", "Pr(>|t|)"],
        n     = nrow(d)
      )
    }) |>
    ungroup() |>
    mutate(
      sig = case_when(p < .001 ~ "***", p < .01 ~ "**", p < .05 ~ "*", p < .10 ~ ".", TRUE ~ "ns"),
      label = sprintf("β = %.3f, t(%d) = %.2f, p %s %s",
                      slope, n - 2L, t,
                      ifelse(p < .001, "<", "="),
                      ifelse(p < .001, ".001", sprintf("%.3f", p)),
                      sig)
    )

  cat("\nSlope tests (validation perplexity): log2(val_perplexity) → ΔLL\n")
  print(slope_tests_val |> select(corpus, model_size, slope, se, t, p, sig, n), n = Inf)

  label_pts_val <- delta_ll_val |>
    group_by(corpus, model_size) |>
    slice_min(val_perplexity, n = 1, with_ties = FALSE) |>
    ungroup()

  print(
    ggplot(delta_ll_val,
           aes(x = val_perplexity, y = delta_ll,
               colour = model_size, shape = model_size)) +
      facet_wrap(~ corpus, scales = "free_x") +
      geom_point(size = 2, alpha = 0.45) +
      geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
                  linetype = "dashed", linewidth = 0.9, alpha = 0.12) +
      geom_text(data = label_pts_val,
                aes(label = model_size),
                hjust = -0.15, vjust = -0.4,
                size  = 3.8, fontface = "bold",
                show.legend = FALSE) +
      scale_x_continuous(
        "Validation perplexity (lower \u2192 better trained)",
        trans  = "log2",
        labels = scales::trans_format("log2", scales::math_format(2^.x))
      ) +
      scale_y_continuous("\u0394LL") +
      scale_colour_manual(values = size_colours, name = "Model size") +
      scale_shape_manual(values  = size_shapes,  name = "Model size") +
      guides(colour = guide_legend(override.aes = list(size = 3, alpha = 1)),
             shape  = guide_legend(override.aes = list(size = 3, alpha = 1))) +
      labs(
        title    = "\u0394LL vs. validation perplexity across training checkpoints",
        subtitle = paste0(
          "\u0394LL = logLik(resp_alpha \u223c freq_prob_c + model pref) \u2212 logLik(resp_alpha \u223c freq_prob_c). ",
          "Each point = one checkpoint. Dashed lines = loess smoother."
        )
      ) +
      theme_classic(base_size = 13) +
      theme(
        legend.position   = "none",
        strip.background  = element_blank(),
        strip.text        = element_text(face = "bold", size = 13),
        panel.grid.major  = element_line(colour = "grey92", linewidth = 0.4),
        axis.line         = element_line(colour = "grey40"),
        axis.ticks        = element_line(colour = "grey40")
      )
  )
}
```

## Model vs. Human α-Ordering Proportions

For each binomial, we compare two proportions:

- **Human proportion α**: fraction of human participants who chose the α ordering (e.g., "salt and pepper" over "pepper and salt"), computed across all individual trials in `all_human_data.csv`.
- **Model proportion α**: out of the 51 sentence-frame prompts used to score each binomial, the fraction where the model's log-odds preference was positive (i.e., `preference > 0`), computed from the raw per-prompt checkpoint CSVs, **separately per corpus × model size**.

Each point is one binomial. Plot 1 uses the final training checkpoint; Plot 2 shows how the model–human correlation (Pearson *r*) evolves across training.

```{r model-vs-human-alpha-prop}
CKPT_DIR <- normalizePath(
  file.path(dirname(DATA_DIR), "checkpoint_results"),
  mustWork = FALSE
)
N_PROMPTS <- 51L   # number of sentence-frame prompts in model-prefs-all-ckpts.py

# Helper: load per-prompt preferences for one (model, step) checkpoint CSV
load_ckpt_raw <- function(model_name, step) {
  short <- sub("znhoughton/", "", model_name)
  path  <- file.path(CKPT_DIR, sprintf("%s_step-%d.csv", short, step))
  if (!file.exists(path)) {
    warning("Checkpoint CSV not found: ", path)
    return(NULL)
  }
  read_csv(path, show_col_types = FALSE, col_select = c(binom, preference))
}

# Human proportion α per binomial (pooled across all participants)
human_alpha_prop <- human_trials |>
  group_by(binom) |>
  summarise(prop_human_alpha = mean(resp_alpha), n_humans = n(), .groups = "drop")

# ── Plot 1: final checkpoint ──────────────────────────────────────────────────
message("Loading final-checkpoint CSVs for model-vs-human plot ...")
raw_final <- final_ck_meta |>
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS)
  ) |>
  filter(!is.na(corpus), !is.na(model_size)) |>
  rowwise() |>
  reframe({
    d <- load_ckpt_raw(model, step)
    if (is.null(d)) tibble() else
      mutate(d, corpus = corpus, model_size = model_size, step = step)
  })

prop_final <- raw_final |>
  filter(!is.na(preference)) |>
  group_by(corpus, model_size, binom) |>
  summarise(
    prop_model_alpha = mean(preference > 0),
    n_prompts        = n(),
    .groups          = "drop"
  ) |>
  inner_join(human_alpha_prop, by = "binom")

cors_final <- prop_final |>
  group_by(corpus, model_size) |>
  summarise(r = cor(prop_human_alpha, prop_model_alpha),
            lbl = sprintf("r = %.2f", r), .groups = "drop")

prop_plot_theme <- list(
  facet_grid(model_size ~ corpus),
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey50"),
  geom_point(size = 1.5, alpha = 0.40, colour = "#2166ac"),
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
              linetype = "dashed", linewidth = 0.9,
              colour = "#b2182b", fill = "#b2182b", alpha = 0.12),
  scale_x_continuous("Proportion of humans choosing \u03b1 ordering",
                     limits = c(0, 1), breaks = seq(0, 1, 0.25)),
  scale_y_continuous(sprintf("Proportion of %d prompts where model prefers \u03b1", N_PROMPTS),
                     limits = c(0, 1), breaks = seq(0, 1, 0.25)),
  theme_classic(base_size = 12),
  theme(strip.background = element_blank(),
        strip.text       = element_text(face = "bold", size = 11),
        panel.grid.major = element_line(colour = "grey92", linewidth = 0.4),
        axis.line        = element_line(colour = "grey40"),
        axis.ticks       = element_line(colour = "grey40"))
)

ggplot(prop_final, aes(x = prop_human_alpha, y = prop_model_alpha)) +
  prop_plot_theme +
  geom_text(data = cors_final, aes(label = lbl),
            x = 0.05, y = 0.97, hjust = 0, vjust = 1,
            size = 3.2, fontface = "italic", inherit.aes = FALSE) +
  labs(
    title    = "Model vs. human \u03b1-ordering proportions \u2014 final checkpoint",
    subtitle = "Each point = one binomial. Dotted = perfect agreement. Dashed = loess fit."
  )

# ── Plot 2: trajectory of model–human correlation over training ───────────────
message("Loading sampled-checkpoint CSVs for trajectory plot (may take a moment) ...")
raw_traj <- ck_meta |>
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS)
  ) |>
  filter(!is.na(corpus), !is.na(model_size)) |>
  rowwise() |>
  reframe({
    d <- load_ckpt_raw(model, step)
    if (is.null(d)) tibble() else
      mutate(d, corpus = corpus, model_size = model_size, step = step)
  })

r_traj <- raw_traj |>
  filter(!is.na(preference)) |>
  group_by(corpus, model_size, step, binom) |>
  summarise(prop_model_alpha = mean(preference > 0), .groups = "drop") |>
  inner_join(human_alpha_prop, by = "binom") |>
  group_by(corpus, model_size, step) |>
  summarise(r = cor(prop_human_alpha, prop_model_alpha), .groups = "drop") |>
  left_join(
    loss_curves |> rename(lc_step = step),
    by = c("corpus", "model_size"),
    relationship = "many-to-many"
  ) |>
  mutate(step_dist = abs(step - lc_step)) |>
  group_by(corpus, model_size, step) |>
  slice_min(step_dist, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(-lc_step, -step_dist)

ggplot(r_traj,
       aes(x = perplexity, y = r,
           colour = model_size, shape = model_size)) +
  facet_wrap(~ corpus, scales = "free_x") +
  geom_line(linewidth = 0.7, alpha = 0.8) +
  geom_point(size = 2, alpha = 0.8) +
  scale_x_continuous(
    "Training perplexity (lower \u2192 better trained)",
    trans  = "log2",
    labels = scales::trans_format("log2", scales::math_format(2^.x))
  ) +
  scale_y_continuous("Pearson r (model prop \u03b1 vs human prop \u03b1)") +
  scale_colour_manual(values = size_colours, name = "Model size") +
  scale_shape_manual(values  = size_shapes,  name = "Model size") +
  labs(
    title    = "Model\u2013human \u03b1-ordering agreement over training",
    subtitle = sprintf(
      "r = Pearson correlation between model prop \u03b1 (%d prompts) and human prop \u03b1, per checkpoint.",
      N_PROMPTS
    )
  ) +
  theme_classic(base_size = 13) +
  theme(
    strip.background = element_blank(),
    strip.text       = element_text(face = "bold", size = 13),
    panel.grid.major = element_line(colour = "grey92", linewidth = 0.4),
    axis.line        = element_line(colour = "grey40"),
    axis.ticks       = element_line(colour = "grey40"),
    legend.position  = "right"
  )
```

------------------------------------------------------------------------

# Section 3: Case Studies — Full Training Trajectories

Take the top **`r N_CASE_BINOMS`** 125M-disagreement items per corpus (largest
`|signed_gap|` in 125M, AbsPref predicts correct direction) and plot their
full training trajectories for all three model sizes.

All values are on the **P(α) − 0.5** scale (0 = no preference; +0.5 = always α; −0.5 = always β).

For each binomial:
- **Coloured solid lines** = model preference (`plogis(pref) − 0.5`) across training, one per model size
- **Dashed gold line** = corpus RelFreq (`freq_prob_c = plogis(log(α/β)) − 0.5`; same for all sizes)
- **Horizontal purple line** = AbsPref (`GenPref_centered`, constant, not training-dependent)

```{r select-case-binoms}
case_binoms <- disagree_125m |>
  # disagree_125m already filtered to abspref_ok_125m & opposite directions
  group_by(corpus) |>
  slice_max(abs(signed_gap_125m), n = N_CASE_BINOMS) |>
  ungroup()

cat(sprintf("Case study binomials: %d BabyLM, %d C4\n",
  sum(case_binoms$corpus == "BabyLM"), sum(case_binoms$corpus == "C4")))
print(case_binoms |>
  transmute(corpus, binom,
            relfreq_prob = freq_prob_c,      # P(alpha) - 0.5
            abspref_prob = GenPref_centered,  # P(alpha) - 0.5
            signed_gap_125m))
```

```{r load-case-traj}
case_binom_vec <- unique(case_binoms$binom)
message(sprintf("Loading trajectories for %d binomials (single DuckDB scan)...", length(case_binom_vec)))

# Single scan across ALL training checkpoints, filtered to case binomials only.
t0 <- proc.time()
freq_traj <- batch_dyn_freq(ck_meta_full, binom_filter = case_binom_vec)
message(sprintf("  DuckDB fetch done in %.1f s (%d rows)", (proc.time() - t0)[["elapsed"]], nrow(freq_traj)))

dat_traj <- train |>
  filter(binom %in% case_binom_vec) |>
  inner_join(ck_meta_full, by = c("model", "step")) |>
  attach_freq(freq_traj) |>
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS),
    pref_prob  = plogis(preference) - 0.5   # P(alpha) - 0.5 scale
  ) |>
  filter(!is.na(tokens), !is.na(log_freq_ratio))

message(sprintf("Trajectory dataset: %d rows", nrow(dat_traj)))
```

### Notable case: "days and years" in C4

In C4, **"years and days"** (β ordering) is more common than **"days and years"** (α ordering):
455 vs. 795 occurrences at end of training. This reflects the natural English tendency
to order temporal units from largest to smallest (years → days). Here are 5 example
sentences from the C4 corpus containing "years and days":

> 1. "The reduction is based on the length of time (in **years and days**) that you retire early."
>
> 2. "...counting the **years and days** till retirement (the 'Reginald Perrin syndrome')."
>
> 3. "Every culture must choose a time in history at which to begin its count of **years and days**."
>
> 4. "Extended term is term life insurance purchased in advance by your cash values for a
>    predetermined number of **years and days**."
>
> 5. "The EAU is a page of **years and days** to read with the best Creating Inclusive School
>    Communities in the l of science." *(garbled OCR)*

The consistent pattern across examples 1–4 is that "years and days" appears as a
**duration span** (coarse-to-fine ordering), which conflicts with the abstract preference
for "days and years" captured in AbsPref.

```{r plot-case-traj, fig.height=5, results='asis'}
for (corp in c("BabyLM", "C4")) {
  these_binoms <- case_binoms |> filter(corpus == corp) |> pull(binom)

  for (binom_i in these_binoms) {
    d <- filter(dat_traj, corpus == corp, binom == binom_i, !is.na(log_freq_ratio))
    if (nrow(d) == 0) {
      cat(sprintf("\n*No data for %s / %s.*\n\n", corp, binom_i))
      next
    }

    abspref_val <- d$GenPref_centered[!is.na(d$GenPref_centered)][1]  # already P(alpha)-0.5

    d_smallmodel <- d |> filter(model_size == first(levels(d$model_size))) |>
      distinct(tokens, freq_prob_c)
    relfreq_final_val <- d_smallmodel |> arrange(desc(tokens)) |> slice(1) |> pull(freq_prob_c)

    # Corpus-level final counts: take the checkpoint with the most tokens in this corpus
    # (alpha_seen/beta_seen are cumulative, so max tokens = fullest count)
    final_counts <- d |>
      filter(tokens == max(tokens, na.rm = TRUE)) |>
      slice(1)
    alpha_n <- formatC(final_counts$alpha_seen, format = "d", big.mark = ",")
    beta_n  <- formatC(final_counts$beta_seen,  format = "d", big.mark = ",")

    p <- ggplot(d, aes(x = tokens, y = pref_prob,
                        colour = model_size, fill = model_size, shape = model_size)) +
      # Zero line
      geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.35) +
      # AbsPref reference (constant, P(alpha)-0.5 scale)
      geom_hline(yintercept = abspref_val, colour = "#9e3d8e",
                 linetype = "dotdash", linewidth = 0.8, alpha = 0.85) +
      annotate("text", x = -Inf, y = abspref_val, hjust = -0.05, vjust = -0.45,
               label = sprintf("AbsPref = %.2f", abspref_val), colour = "#9e3d8e", size = 3.2) +
      # Corpus RelFreq trajectory (same for all sizes — use smallest model's data; P(alpha)-0.5)
      geom_line(data = d_smallmodel,
                aes(x = tokens, y = freq_prob_c), colour = "#cc8800",
                linetype = "dashed", linewidth = 0.9, inherit.aes = FALSE) +
      annotate("text", x = Inf, y = relfreq_final_val,
               hjust = 1.05, vjust = -0.4, label = "RelFreq", colour = "#cc8800", size = 3.2) +
      # Model preference per size
      geom_line(linewidth = 0.85) +
      geom_point(size = 2.0, stroke = 0.5, colour = "grey25") +
      scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
      scale_fill_manual(values = SIZE_COLOURS, name = "Model size") +
      scale_shape_manual(values = SIZE_SHAPES, name = "Model size") +
      scale_x_continuous("Training tokens",
                         labels = label_number(scale_cut = cut_short_scale()),
                         expand = expansion(mult = c(0.02, 0.05))) +
      coord_cartesian(xlim = c(0, NA)) +
      scale_y_continuous("P(\u03b1) \u2212 0.5  (+ = prefer \u03b1 form)", limits = c(-0.5, 0.5)) +
      labs(
        title    = sprintf('%s \u2014 "%s"', corp, binom_i),
        subtitle = sprintf(
          "RelFreq P(\u03b1)\u22120.5 = %.2f (predicts %s); AbsPref P(\u03b1)\u22120.5 = %.2f (predicts %s)",
          mean(d$freq_prob_c, na.rm = TRUE),
          if_else(mean(d$freq_prob_c, na.rm = TRUE) > 0, "\u03b1", "\u03b2"),
          abspref_val,
          if_else(abspref_val > 0, "\u03b1", "\u03b2")),
        caption  = sprintf("Corpus counts at end of training: \u03b1 ordering = %s, \u03b2 ordering = %s",
                           alpha_n, beta_n)
      ) +
      theme_bw(base_size = 12) +
      theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

    print(p)
    cat("\n\n")
  }
}
```

------------------------------------------------------------------------

# Summary

```{r summary-table}
tibble(
  Analysis = c(
    "BabyLM: AbsPref vs RelFreq conflict items (125M follows AbsPref)",
    "  \u2514 Replicates in 350M/1.3B (AbsPref ok in all 3 sizes)",
    "C4: AbsPref vs RelFreq conflict items (125M follows AbsPref)",
    "  \u2514 Replicates in 350M/1.3B (AbsPref ok in all 3 sizes)",
    "BabyLM: near-flat RelFreq items",
    "C4: near-flat RelFreq items"
  ),
  N = c(
    sum(disagree_125m$corpus == "BabyLM"),
    {
      r <- dat_final |>
        semi_join(filter(disagree_125m, corpus == "BabyLM"), by = c("corpus","binom")) |>
        group_by(binom) |>
        summarise(all_ok = all(abspref_ok))
      sum(r$all_ok)
    },
    sum(disagree_125m$corpus == "C4"),
    {
      r <- dat_final |>
        semi_join(filter(disagree_125m, corpus == "C4"), by = c("corpus","binom")) |>
        group_by(binom) |>
        summarise(all_ok = all(abspref_ok))
      sum(r$all_ok)
    },
    n_distinct(dat_flat$binom[dat_flat$corpus == "BabyLM"]),
    n_distinct(dat_flat$binom[dat_flat$corpus == "C4"])
  )
) |>
  kbl() |>
  kable_styling(bootstrap_options = c("striped"), full_width = FALSE)
```
