---
title: "Analysis 2: Human Preferences and Individual Constraints"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 7, dpi = 150,
                      out.width = "100%")
```

# Overview

**Before running this script**, make sure `prepare_data.py` has been run to generate `training_attested.csv`, `human_responses_attested.csv`, and `attested_agg.csv` in `../Data/processed/`.

After the fitting section finishes, run `prepare_results.R` to extract coefficients into CSVs for use in `writeup.qmd`.

------------------------------------------------------------------------

**Analysis 2a — Human ordering preference \~ model preference**

For each model × checkpoint, a Bernoulli logistic regression tests whether the model's preference score for a binomial predicts whether a human participant chose the alphabetical ordering:

$$\text{resp\_binary} \sim \text{preference} + (1 \mid \text{binom}), \quad
\text{family} = \text{Bernoulli}$$

**Analysis 2b — Model preference \~ individual constraints**

For each model × checkpoint, the single `AbsPref` predictor from the main analysis is replaced by all ten individual phonological/semantic constraints simultaneously:

$$\text{preference} \sim (\text{Form} + \text{Percept} + \cdots + \text{BStress})
\times \ln(\text{OverallFreq}) + \text{RelFreq} \times \ln(\text{OverallFreq})
+ (1 \mid \text{binom})$$

**Analysis 2c — Prediction accuracy by overall frequency**

No model fitting required. Accuracy (model preference sign matches human majority sign) is computed from `attested_agg.csv` for each model × checkpoint × frequency quintile.

**Analysis 2g — Model preference \~ VIF-filtered individual constraints**

Model selection variant of 2b. `Lapse` and `Percept` are dropped a priori, then a per-corpus VIF analysis (using a plain OLS on item-level data) removes any remaining predictor with VIF \> 3. Because BabyLM and C4 use different binomial sets, VIF is computed once per corpus; the resulting formula is applied to all models within that corpus across 20 checkpoints.

**Analysis 2h — Preference deviation \~ relative frequency + bigram log-odds**

Analogous to 2f but for preference deviation (signed gap): does bigram positional tendency predict deviation from frequency-predicted ordering above and beyond relative frequency?

$$\text{signed\_gap} \sim \text{freq\_prob\_c} + \text{bigram\_logodds\_c} + (1 \mid \text{binom})$$

------------------------------------------------------------------------

```{r libraries}
library(tidyverse)
library(brms)
library(patchwork)
library(future)
library(furrr)
library(duckdb)
```

```{r config}
SCRIPT_DIR <- {
  knitr_input <- tryCatch(knitr::current_input(), error = function(e) NULL)
  if (!is.null(knitr_input) && nchar(knitr_input) > 0) {
    dirname(normalizePath(knitr_input, mustWork = FALSE))
  } else if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
    dirname(normalizePath(rstudioapi::getActiveDocumentContext()$path, mustWork = FALSE))
  } else {
    normalizePath(".", mustWork = FALSE)
  }
}

MODELS_DIR  <- file.path(SCRIPT_DIR, "models")
DATA_DIR    <- normalizePath(file.path(SCRIPT_DIR, "../Data/processed"), mustWork = FALSE)
TRAIN_CSV   <- file.path(DATA_DIR, "training_attested.csv")
HUMAN_CSV   <- file.path(DATA_DIR, "human_responses_attested.csv")
AGG_CSV     <- file.path(DATA_DIR, "attested_agg.csv")
BLM_BIGRAM_CSV <- normalizePath(file.path(SCRIPT_DIR, "../Data/babylm_bigram_exposures.csv"), mustWork = FALSE)
C4_BIGRAM_CSV  <- normalizePath(file.path(SCRIPT_DIR, "../Data/c4_bigram_exposures.csv"),  mustWork = FALSE)
EXPOSURES_CSV     <- normalizePath(file.path(DATA_DIR, "checkpoint_results_with_exposures.csv"), mustWork = FALSE)
EXPOSURES_CSV_FWD <- gsub("\\\\", "/", EXPOSURES_CSV)  # forward-slash path for DuckDB

dir.create(MODELS_DIR, showWarnings = FALSE, recursive = TRUE)

N_TRAJ_CHECKPOINTS <- 20

CONSTRAINT_COLS <- c("Form", "Percept", "Culture", "Power", "Intense",
                     "Icon", "Freq", "Len", "Lapse", "BStress")

SIZE_LEVELS  <- c("125M", "350M", "1.3B")
SIZE_COLOURS <- c("125M" = "#F0B323", "350M" = "#007030", "1.3B" = "#154733")

PRIORS_GAUSSIAN <- c(
  prior(normal(0, 1), class = b),
  prior(normal(0, 1), class = Intercept),
  prior(exponential(1), class = sd),
  prior(exponential(1), class = sigma)
)
PRIORS_BERNOULLI <- c(
  prior(normal(0, 1), class = b),
  prior(normal(0, 1), class = Intercept),
  prior(exponential(1), class = sd)
)

ITER   <- 6000
WARMUP <- 3000

N_CORES           <- parallel::detectCores(logical = FALSE)
USABLE_CORES      <- floor(N_CORES * 3 / 4)
N_PARALLEL_MODELS <- 6
CHAINS            <- min(floor(USABLE_CORES / N_PARALLEL_MODELS), 8)
CORES             <- CHAINS

message(sprintf(
  "Cores: %d total | %d usable (3/4) | %d parallel models | %d chains/cores per model",
  N_CORES, USABLE_CORES, N_PARALLEL_MODELS, CHAINS
))
```

```{r helpers}
make_slug <- function(x) tolower(gsub("[^a-z0-9]", "_", x))

parse_corpus <- function(x) case_when(
  grepl("babylm", x, TRUE) ~ "BabyLM",
  grepl("opt.c4", x, TRUE) ~ "C4",
  TRUE ~ NA_character_
)
parse_size <- function(x) case_when(
  grepl("125m",    x, TRUE) ~ "125M",
  grepl("350m",    x, TRUE) ~ "350M",
  grepl("1[._]3b", x, TRUE) ~ "1.3B",
  TRUE ~ NA_character_
)

sample_steps <- function(model_df, n = N_TRAJ_CHECKPOINTS) {
  steps <- sort(unique(model_df$step))
  idx   <- unique(round(seq(1, length(steps), length.out = min(n, length(steps)))))
  steps[idx]
}

# Join checkpoint-specific freq_prob_c, log_freq_ratio, signed_gap, log_total_c
# onto a single-checkpoint data slice.  Opens a per-call DuckDB connection so
# that no large global is needed (mirrors load_single_checkpoint in analysis.Rmd).
add_dyn_freq <- function(cdat) {
  m <- unique(cdat$model)
  s <- unique(cdat$step)
  if (length(m) != 1 || length(s) != 1)
    stop("add_dyn_freq: cdat must be a single model \u00d7 step slice")
  m_safe <- gsub("'", "''", m)

  con_local <- dbConnect(duckdb::duckdb())
  on.exit(dbDisconnect(con_local, shutdown = TRUE))
  dbExecute(con_local, sprintf(
    "CREATE VIEW exp AS SELECT * FROM read_csv_auto('%s')", EXPOSURES_CSV_FWD
  ))
  freq_raw <- dbGetQuery(con_local, sprintf("
    SELECT binom,
           ANY_VALUE(alpha_seen) AS alpha_seen,
           ANY_VALUE(beta_seen)  AS beta_seen
    FROM exp
    WHERE model = '%s' AND step = %d
    GROUP BY binom
  ", m_safe, s))

  freq_dat <- as_tibble(freq_raw) |>
    mutate(
      log_freq_ratio = if_else(alpha_seen > 0 & beta_seen > 0,
                               log(alpha_seen / beta_seen), NA_real_),
      freq_prob_c    = plogis(log_freq_ratio) - 0.5,
      log_total      = if_else(alpha_seen + beta_seen > 0,
                               log(alpha_seen + beta_seen), NA_real_)
    ) |>
    select(-alpha_seen, -beta_seen)

  cdat |>
    left_join(freq_dat, by = "binom") |>
    mutate(
      signed_gap  = preference - log_freq_ratio,
      log_total_c = as.numeric(scale(log_total))
    )
}

list_fits <- function(prefix) {
  pat <- paste0("^", prefix, ".+\\.rds$")
  list.files(MODELS_DIR, pattern = pat, full.names = TRUE)
}

extract_coefs <- function(fit, model_name, term_levels, term_labels) {
  fe <- as.data.frame(fixef(fit))
  fe$term <- rownames(fe)
  fe |>
    as_tibble() |>
    rename(estimate = Estimate, se = Est.Error, lo = Q2.5, hi = Q97.5) |>
    filter(term != "Intercept") |>
    mutate(
      term       = factor(term, levels = term_levels, labels = term_labels),
      p_gt0      = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE),
      model_name = model_name,
      corpus     = parse_corpus(model_name),
      model_size = factor(parse_size(model_name), levels = SIZE_LEVELS)
    )
}

# Load trajectory results from fitted .rds files, attaching token counts from TRAJ_META
load_traj_coefs <- function(prefix, term_levels, term_labels) {
  paths <- list_fits(prefix)
  if (length(paths) == 0) {
    warning(sprintf("No .rds files for prefix '%s'", prefix))
    return(tibble())
  }
  map_dfr(paths, function(path) {
    fname_base <- sub("\\.rds$", "", basename(path))
    slug       <- sub(paste0("^", prefix), "", fname_base)
    slug       <- sub("_ck[0-9]+$", "", slug)
    ck_idx     <- as.integer(sub("^.*_ck([0-9]+)$", "\\1", fname_base))
    row        <- if (!is.null(TRAJ_META)) {
      filter(TRAJ_META, .data$slug == slug, .data$ck_idx == !!ck_idx)
    } else tibble()
    ck_tokens  <- if (nrow(row) == 1) row$tokens else NA_real_
    fit        <- readRDS(path)
    result     <- extract_coefs(fit, slug, term_levels, term_labels) |>
      mutate(tokens = ck_tokens, ck_idx = ck_idx)
    rm(fit); gc()
    result
  })
}
```

```{r traj-meta}
# Recover token counts for x-axes.
# Prefer the pre-computed traj_meta.csv from prepare_results.R (fast);
# fall back to querying DuckDB directly only if that CSV doesn't exist.
traj_meta_csv <- file.path(DATA_DIR, "results", "traj_meta.csv")
if (file.exists(traj_meta_csv)) {
  TRAJ_META <- read_csv(traj_meta_csv, show_col_types = FALSE)
  message(sprintf("Loaded traj_meta.csv (%d rows)", nrow(TRAJ_META)))
} else if (file.exists(file.path(DATA_DIR, "checkpoint_results_with_exposures.csv"))) {
  message("traj_meta.csv not found — querying DuckDB (slow)...")
  library(duckdb)
  CSV_FWD <- gsub("\\\\", "/",
    normalizePath(file.path(DATA_DIR, "checkpoint_results_with_exposures.csv")))
  con <- dbConnect(duckdb::duckdb())
  dbExecute(con, sprintf("CREATE VIEW ckpt AS SELECT * FROM read_csv_auto('%s')", CSV_FWD))
  models_meta <- sort(dbGetQuery(con, "SELECT DISTINCT model FROM ckpt")$model)
  TRAJ_META <- bind_rows(lapply(models_meta, function(m) {
    slug      <- make_slug(m)
    all_ckpts <- dbGetQuery(con, sprintf(
      "SELECT DISTINCT tokens FROM ckpt WHERE model = '%s' ORDER BY tokens",
      gsub("'", "''", m)))$tokens
    n   <- min(N_TRAJ_CHECKPOINTS, length(all_ckpts))
    idx <- unique(round(seq(1, length(all_ckpts), length.out = n)))
    tibble(slug = slug, ck_idx = seq_along(idx), tokens = all_ckpts[idx])
  }))
  dbDisconnect(con, shutdown = TRUE)
} else {
  TRAJ_META <- NULL
  warning("Neither traj_meta.csv nor checkpoint_results_with_exposures.csv found — x-axis will be NA")
}
```

------------------------------------------------------------------------

# Results

## 2c: Prediction accuracy by overall frequency

```{r plot-acc-freq, fig.height=7}
if (file.exists(AGG_CSV)) {
  agg <- read_csv(AGG_CSV, show_col_types = FALSE) |>
    mutate(
      correct       = (preference > 0.5) == (hum_pref > 0.5),
      corpus        = parse_corpus(model),
      model_size    = factor(parse_size(model), levels = SIZE_LEVELS),
      freq_quintile = cut(log(OverallFreq), breaks = 5,
                          labels = c("Q1 (lowest)", "Q2", "Q3", "Q4", "Q5 (highest)"))
    ) |>
    filter(!is.na(corpus), !is.na(model_size), !is.na(freq_quintile))

  agg_sampled <- agg |>
    group_by(model) |>
    mutate(step_rank = dense_rank(tokens)) |>
    filter(step_rank %in% unique(round(seq(1, max(step_rank), length.out = 20)))) |>
    ungroup()

  acc_by_freq <- agg_sampled |>
    group_by(corpus, model_size, tokens, freq_quintile) |>
    summarise(accuracy = mean(correct, na.rm = TRUE), .groups = "drop")

  ggplot(acc_by_freq, aes(x = freq_quintile, y = accuracy,
                           colour = tokens / 1e6, group = tokens)) +
    geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 0.6, alpha = 0.8) +
    geom_point(size = 1.8) +
    scale_colour_gradient(low = "#c7e9b4", high = "#154733",
                          name = "Training tokens\n(millions)") +
    scale_y_continuous("Accuracy", limits = c(0.3, 1),
                       labels = scales::percent_format()) +
    scale_x_discrete("Overall frequency quintile") +
    facet_grid(corpus ~ model_size) +
    labs(title = "Prediction accuracy by overall frequency quintile") +
    theme_bw(base_size = 13) +
    theme(panel.grid.minor = element_blank(),
          plot.title   = element_text(face = "bold", size = 14),
          strip.text   = element_text(face = "bold", size = 13),
          axis.text.x  = element_text(size = 10),
          axis.text.y  = element_text(colour = "black", size = 11),
          axis.title   = element_text(colour = "black", size = 13),
          legend.text  = element_text(size = 11),
          legend.title = element_text(size = 12))
} else {
  message("attested_agg.csv not found — run prepare_data.py first.")
}
```

```{r plot-cor-freq, fig.height=7}
if (exists("agg_sampled")) {
  cor_by_freq <- agg_sampled |>
    group_by(corpus, model_size, tokens, freq_quintile) |>
    summarise(r = cor(preference, hum_pref, use = "complete.obs"),
              .groups = "drop")

  ggplot(cor_by_freq, aes(x = freq_quintile, y = r,
                           colour = tokens / 1e6, group = tokens)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 0.6, alpha = 0.8) +
    geom_point(size = 1.8) +
    scale_colour_gradient(low = "#c7e9b4", high = "#154733",
                          name = "Training tokens\n(millions)") +
    scale_y_continuous("Pearson r (model pref ~ human pref)", limits = c(-0.5, 1)) +
    scale_x_discrete("Overall frequency quintile") +
    facet_grid(corpus ~ model_size) +
    labs(title = "Correlation with human preference by overall frequency quintile") +
    theme_bw(base_size = 13) +
    theme(panel.grid.minor = element_blank(),
          plot.title   = element_text(face = "bold", size = 14),
          strip.text   = element_text(face = "bold", size = 13),
          axis.text.x  = element_text(size = 10),
          axis.text.y  = element_text(colour = "black", size = 11),
          axis.title   = element_text(colour = "black", size = 13),
          legend.text  = element_text(size = 11),
          legend.title = element_text(size = 12))
} else {
  message("attested_agg.csv not found — run prepare_data.py first.")
}
```

------------------------------------------------------------------------

## Correlation between model preference and relative frequency over training

```{r plot-cor-relfreq-traj, fig.height=5}
if (exists("agg_sampled")) {
  cor_traj_freq <- agg_sampled |>
    group_by(corpus, model_size, tokens) |>
    summarise(r = cor(preference, RelFreq, use = "complete.obs"),
              .groups = "drop")

  ggplot(cor_traj_freq, aes(x = tokens / 1e6, y = r, colour = model_size)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous("Training tokens (millions)") +
    scale_y_continuous("Pearson r (model pref ~ relative freq)", limits = c(-1, 1)) +
    facet_wrap(~ corpus) +
    labs(title = "Correlation of model preference with relative frequency over training") +
    theme_bw(base_size = 13) +
    theme(panel.grid.minor = element_blank(),
          plot.title      = element_text(face = "bold", size = 14),
          strip.text      = element_text(face = "bold", size = 13),
          axis.text       = element_text(colour = "black", size = 11),
          axis.title      = element_text(colour = "black", size = 13),
          legend.position = "bottom")
} else {
  message("Run the 2c chunk first to load agg_sampled.")
}
```

```{r plot-cor-relfreq-by-freq, fig.height=7}
if (exists("agg_sampled")) {
  cor_relfreq_by_freq <- agg_sampled |>
    group_by(corpus, model_size, tokens, freq_quintile) |>
    summarise(r = cor(preference, RelFreq, use = "complete.obs"),
              .groups = "drop")

  ggplot(cor_relfreq_by_freq, aes(x = freq_quintile, y = r,
                                   colour = tokens / 1e6, group = tokens)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 0.6, alpha = 0.8) +
    geom_point(size = 1.8) +
    scale_colour_gradient(low = "#c7e9b4", high = "#154733",
                          name = "Training tokens\n(millions)") +
    scale_y_continuous("Pearson r (model pref ~ relative freq)", limits = c(-1, 1)) +
    scale_x_discrete("Overall frequency quintile") +
    facet_grid(corpus ~ model_size) +
    labs(title = "Correlation of model preference with relative frequency by overall frequency quintile") +
    theme_bw(base_size = 13) +
    theme(panel.grid.minor = element_blank(),
          plot.title   = element_text(face = "bold", size = 14),
          strip.text   = element_text(face = "bold", size = 13),
          axis.text.x  = element_text(size = 10),
          axis.text.y  = element_text(colour = "black", size = 11),
          axis.title   = element_text(colour = "black", size = 13),
          legend.text  = element_text(size = 11),
          legend.title = element_text(size = 12))
} else {
  message("Run the 2c chunk first to load agg_sampled.")
}
```

------------------------------------------------------------------------

## 2a raw data: correlation between model preference and human preference over training

```{r plot-2a-raw, fig.height=7}
if (exists("agg_sampled")) {
  cor_traj_2a <- agg_sampled |>
    group_by(corpus, model_size, tokens) |>
    summarise(r = cor(preference, hum_pref, use = "complete.obs"),
              .groups = "drop")

  ggplot(cor_traj_2a, aes(x = tokens / 1e6, y = r, colour = model_size)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous("Training tokens (millions)") +
    scale_y_continuous("Pearson r (model pref ~ human pref)", limits = c(-0.1, 1)) +
    facet_wrap(~ corpus) +
    labs(title = "Raw trajectory: correlation of model preference with human preference") +
    theme_bw(base_size = 13) +
    theme(panel.grid.minor = element_blank(),
          plot.title      = element_text(face = "bold", size = 14),
          strip.text      = element_text(face = "bold", size = 13),
          axis.text       = element_text(colour = "black", size = 11),
          axis.title      = element_text(colour = "black", size = 13),
          legend.position = "bottom")
} else {
  message("Run the 2c chunk first to load agg_sampled.")
}
```

------------------------------------------------------------------------

## 2b raw data: correlation between model preference and each constraint over training

```{r plot-2b-raw, fig.height=10}
if (exists("agg_sampled")) {
  cor_traj_2b <- agg_sampled |>
    group_by(corpus, model_size, tokens) |>
    summarise(across(all_of(c(CONSTRAINT_COLS, "RelFreq")),
                     ~ cor(preference, .x, use = "complete.obs")),
              .groups = "drop") |>
    pivot_longer(cols = all_of(c(CONSTRAINT_COLS, "RelFreq")),
                 names_to = "constraint", values_to = "r") |>
    mutate(constraint = factor(constraint, levels = c(CONSTRAINT_COLS, "RelFreq")))

  ggplot(cor_traj_2b, aes(x = tokens / 1e6, y = r, colour = model_size)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.5) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous("Training tokens (millions)") +
    scale_y_continuous("Pearson r (model pref ~ constraint)", limits = c(-1, 1)) +
    facet_grid(constraint ~ corpus) +
    labs(title = "Raw trajectories: correlation of model preference with each constraint") +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          plot.title      = element_text(face = "bold", size = 12),
          strip.text.x    = element_text(face = "bold", size = 11),
          strip.text.y    = element_text(size = 9),
          axis.text       = element_text(colour = "black", size = 9),
          axis.title      = element_text(colour = "black", size = 11),
          legend.position = "bottom")
} else {
  message("Run the 2c chunk first to load agg_sampled.")
}
```

------------------------------------------------------------------------

## 2b raw data: correlation between preference deviation and each constraint over training

```{r plot-2b-raw-prefdev, fig.height=10}
if (exists("agg_sampled")) {
  cor_traj_prefdev <- agg_sampled |>
    mutate(pref_dev = preference - RelFreq) |>
    group_by(corpus, model_size, tokens) |>
    summarise(across(all_of(CONSTRAINT_COLS),
                     ~ cor(pref_dev, .x, use = "complete.obs")),
              .groups = "drop") |>
    pivot_longer(cols = all_of(CONSTRAINT_COLS),
                 names_to = "constraint", values_to = "r") |>
    mutate(constraint = factor(constraint, levels = CONSTRAINT_COLS))

  ggplot(cor_traj_prefdev, aes(x = tokens / 1e6, y = r, colour = model_size)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_line(linewidth = 0.7) +
    geom_point(size = 1.5) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous("Training tokens (millions)") +
    scale_y_continuous("Pearson r (pref deviation ~ constraint)", limits = c(-1, 1)) +
    facet_grid(constraint ~ corpus) +
    labs(title = "Raw trajectories: correlation of preference deviation with each constraint",
         subtitle = "Preference deviation = model preference \u2212 RelFreq") +
    theme_bw(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          plot.title      = element_text(face = "bold", size = 12),
          plot.subtitle   = element_text(size = 10, colour = "grey40"),
          strip.text.x    = element_text(face = "bold", size = 11),
          strip.text.y    = element_text(size = 9),
          axis.text       = element_text(colour = "black", size = 9),
          axis.title      = element_text(colour = "black", size = 11),
          legend.position = "bottom")
} else {
  message("Run the 2c chunk first to load agg_sampled.")
}
```

------------------------------------------------------------------------

# Fitting

Cells below fit models and save `.rds` files. Already-fitted models are skipped automatically — re-run at any time to fill in missing checkpoints. After fitting, run `prepare_results.R` to extract coefficients into CSVs.

```{r load-data}
message("Loading training data...")
train <- read_csv(TRAIN_CSV, show_col_types = FALSE)

if (!file.exists(EXPOSURES_CSV))
  stop("checkpoint_results_with_exposures.csv not found — cannot compute dynamic frequency values.")

message("Loading human responses...")
human <- read_csv(HUMAN_CSV, show_col_types = FALSE)
# Fix: old CSV generation produced duplicate 'binom' column headers;
# readr renames them to binom...1, binom...2 etc. Normalise to just binom + resp_binary.
if (!"binom" %in% names(human)) {
  names(human)[1] <- "binom"
}
human <- select(human, binom, resp_binary)

MODELS_LIST <- sort(unique(train$model))
MODELS_LIST <- MODELS_LIST[!is.na(parse_corpus(MODELS_LIST))]
message(sprintf("Fitting %d models: %s", length(MODELS_LIST), paste(MODELS_LIST, collapse = ", ")))
```

```{r setup-parallel}
# multicore uses fork — workers inherit the environment without copying.
plan(multicore, workers = N_PARALLEL_MODELS)
message(sprintf("furrr plan: %s with %d workers", class(plan())[1], N_PARALLEL_MODELS))
```

## 2a: `resp_binary ~ preference + (1 | binom)` (Bernoulli)

```{r fit-2a}
future_map(MODELS_LIST, function(model_name) {
  slug  <- make_slug(model_name)
  mdata <- filter(train, model == model_name)
  steps <- sample_steps(mdata)
  message(sprintf("\n=== [2a] %s (%d checkpoints) ===", slug, length(steps)))

  for (ck_idx in seq_along(steps)) {
    s        <- steps[ck_idx]
    ck       <- sprintf("ck%02d", ck_idx)
    cdat     <- filter(mdata, step == s)
    hum_path <- file.path(MODELS_DIR, sprintf("humfit_traj_%s_%s.rds", slug, ck))

    if (!file.exists(hum_path)) {
      hfit_data <- inner_join(human, select(cdat, binom, preference), by = "binom") |>
        filter(!is.na(preference)) |>
        mutate(preference = as.numeric(scale(preference)))
      if (nrow(hfit_data) < 20) {
        warning(sprintf("Skipping %s %s — only %d rows", slug, ck, nrow(hfit_data)))
      } else {
        message(sprintf("  %s %s  step=%d  n=%d", slug, ck, s, nrow(hfit_data)))
        fit <- brm(
          resp_binary ~ preference + (1 | binom),
          data    = hfit_data,
          family  = bernoulli(link = "logit"),
          prior   = PRIORS_BERNOULLI,
          iter    = ITER, warmup = WARMUP, chains = CHAINS, cores = CORES,
          backend = "cmdstanr", silent = 2, refresh = 0
        )
        saveRDS(fit, hum_path); rm(fit); gc()
        message(sprintf("    Saved: %s", basename(hum_path)))
      }
    } else {
      message(sprintf("  %s %s — exists, skipping", slug, ck))
    }
  }
}, .options = furrr_options(seed = TRUE))
message("\n2a fitting done.")
```

## 2b: `preference ~ (constraints) * log_total_c + RelFreq * log_total_c + (1 | binom)` (Gaussian)

```{r fit-2b}
constraint_terms <- paste(CONSTRAINT_COLS, collapse = " + ")
formula_2b <- as.formula(sprintf(
  "preference ~ (%s) * log_total_c + freq_prob_c * log_total_c + (1 | binom)",
  constraint_terms
))

future_map(MODELS_LIST, function(model_name) {
  slug  <- make_slug(model_name)
  mdata <- filter(train, model == model_name)
  steps <- sample_steps(mdata)
  message(sprintf("\n=== [2b] %s (%d checkpoints) ===", slug, length(steps)))

  for (ck_idx in seq_along(steps)) {
    s        <- steps[ck_idx]
    ck       <- sprintf("ck%02d", ck_idx)
    cdat     <- filter(mdata, step == s)
    con_path <- file.path(MODELS_DIR, sprintf("constr_traj_%s_%s.rds", slug, ck))

    if (!file.exists(con_path)) {
      message(sprintf("  %s %s  step=%d", slug, ck, s))
      con_data <- cdat |>
        add_dyn_freq() |>
        filter(!is.na(preference), !is.na(freq_prob_c), !is.na(log_total_c),
               !if_any(all_of(CONSTRAINT_COLS), is.na))
      fit <- brm(
        formula = formula_2b,
        data    = con_data,
        family  = gaussian(),
        prior   = PRIORS_GAUSSIAN,
        iter    = ITER, warmup = WARMUP, chains = CHAINS, cores = CORES,
        backend = "cmdstanr", silent = 2, refresh = 0
      )
      saveRDS(fit, con_path); rm(fit); gc()
      message(sprintf("    Saved: %s", basename(con_path)))
    } else {
      message(sprintf("  %s %s — exists, skipping", slug, ck))
    }
  }
}, .options = furrr_options(seed = TRUE))
message("\n2b fitting done.")
```

------------------------------------------------------------------------

## 2g: VIF-filtered individual constraints trajectory

Model selection variant of 2b: remove `Lapse` and `Percept` a priori, then drop any remaining constraint with a variance inflation factor (VIF) \> 3. Because BabyLM and C4 use different sets of binomials, the VIF analysis is run once per corpus (items are shared across model sizes within each corpus, so repeating it per model is unnecessary). The selected predictors are then used in the same per-checkpoint trajectory formula as 2b.

```{r vif-selection}
library(car)

CONSTRAINT_COLS_REDUCED <- setdiff(CONSTRAINT_COLS, c("Lapse", "Percept"))

# For each corpus, pick the latest checkpoint of the first (alphabetically)
# model in that corpus. Constraints are item-level and don't vary by checkpoint,
# so any snapshot gives the same VIF.
vif_data_for_corpus <- function(corpus_label) {
  corpus_models <- MODELS_LIST[parse_corpus(MODELS_LIST) == corpus_label]
  if (length(corpus_models) == 0) return(NULL)
  m   <- sort(corpus_models)[1]
  dat <- filter(train, model == m) |>
    group_by(binom) |>
    slice_max(step, n = 1) |>
    ungroup() |>
    add_dyn_freq() |>
    select(preference, freq_prob_c, all_of(CONSTRAINT_COLS_REDUCED)) |>
    filter(complete.cases(across(everything())))
  dat
}

compute_vif <- function(dat) {
  pred_cols <- c(CONSTRAINT_COLS_REDUCED, "freq_prob_c")
  fml <- as.formula(paste("preference ~", paste(pred_cols, collapse = " + ")))
  vif(lm(fml, data = dat))
}

vif_blm <- tryCatch(compute_vif(vif_data_for_corpus("BabyLM")), error = function(e) NULL)
vif_c4  <- tryCatch(compute_vif(vif_data_for_corpus("C4")),     error = function(e) NULL)

select_predictors <- function(vif_vals) {
  if (is.null(vif_vals)) return(CONSTRAINT_COLS_REDUCED)
  names(vif_vals)[vif_vals <= 3]
}

VIF_SELECTED_BLM <- select_predictors(vif_blm)
VIF_SELECTED_C4  <- select_predictors(vif_c4)

# Separate into constraint predictors and freq_prob_c (always kept)
vif_constr_blm <- setdiff(VIF_SELECTED_BLM, "freq_prob_c")
vif_constr_c4  <- setdiff(VIF_SELECTED_C4,  "freq_prob_c")

message("=== VIF selection results ===")
message(sprintf("BabyLM — dropped (VIF > 3): %s",
  paste(setdiff(CONSTRAINT_COLS_REDUCED, vif_constr_blm), collapse = ", ") |>
    (\(x) if (nchar(x) == 0) "(none)" else x)()))
message(sprintf("BabyLM — retained: %s", paste(vif_constr_blm, collapse = ", ")))
message(sprintf("C4 — dropped (VIF > 3): %s",
  paste(setdiff(CONSTRAINT_COLS_REDUCED, vif_constr_c4), collapse = ", ") |>
    (\(x) if (nchar(x) == 0) "(none)" else x)()))
message(sprintf("C4 — retained: %s", paste(vif_constr_c4, collapse = ", ")))
```

```{r vif-table, results="asis"}
# Print VIF tables for both corpora
vif_tbl <- function(vif_vals, corpus_label) {
  if (is.null(vif_vals)) { message(sprintf("No VIF data for %s", corpus_label)); return() }
  df <- data.frame(
    Predictor = names(vif_vals),
    VIF       = round(vif_vals, 2),
    Retained  = ifelse(vif_vals <= 5, "Yes", "No (VIF > 3)"),
    row.names = NULL
  )
  cat(sprintf("\n**%s corpus VIF**\n\n", corpus_label))
  print(knitr::kable(df, format = "markdown"))
}
vif_tbl(vif_blm, "BabyLM")
vif_tbl(vif_c4,  "C4")
```

```{r fit-2g}
# Build per-corpus formulas using VIF-selected constraint predictors
make_vif_formula <- function(constr_cols) {
  terms <- paste(constr_cols, collapse = " + ")
  as.formula(sprintf(
    "preference ~ (%s) * log_total_c + freq_prob_c * log_total_c + (1 | binom)",
    terms
  ))
}

formula_2g_blm <- make_vif_formula(vif_constr_blm)
formula_2g_c4  <- make_vif_formula(vif_constr_c4)

future_map(MODELS_LIST, function(model_name) {
  slug   <- make_slug(model_name)
  corpus <- parse_corpus(model_name)
  mdata  <- filter(train, model == model_name)
  steps  <- sample_steps(mdata)
  formula_2g <- if (corpus == "BabyLM") formula_2g_blm else formula_2g_c4
  vif_constr <- if (corpus == "BabyLM") vif_constr_blm  else vif_constr_c4
  message(sprintf("\n=== [2g] %s (%d checkpoints) ===", slug, length(steps)))

  for (ck_idx in seq_along(steps)) {
    s        <- steps[ck_idx]
    ck       <- sprintf("ck%02d", ck_idx)
    out_path <- file.path(MODELS_DIR, sprintf("constr_vif_traj_%s_%s.rds", slug, ck))
    if (file.exists(out_path)) {
      message(sprintf("  %s %s — exists, skipping", slug, ck)); next
    }
    con_data <- filter(mdata, step == s) |>
      add_dyn_freq() |>
      filter(!is.na(preference), !is.na(freq_prob_c), !is.na(log_total_c),
             !if_any(all_of(vif_constr), is.na))
    if (nrow(con_data) < 20) {
      warning(sprintf("Skipping %s %s — insufficient data (%d rows)", slug, ck, nrow(con_data))); next
    }
    message(sprintf("  %s %s  step=%d  n=%d", slug, ck, s, nrow(con_data)))
    fit <- brm(
      formula = formula_2g,
      data    = con_data,
      family  = gaussian(),
      prior   = PRIORS_GAUSSIAN,
      iter    = ITER, warmup = WARMUP, chains = CHAINS, cores = CORES,
      backend = "cmdstanr", silent = 2, refresh = 0
    )
    saveRDS(fit, out_path); rm(fit); gc()
    message(sprintf("    Saved: %s", basename(out_path)))
  }
}, .options = furrr_options(seed = TRUE))
message("\n2g fitting done.")
```

------------------------------------------------------------------------

## 2d/2e: Trajectory models + bigram log-odds

Replicates the per-checkpoint trajectory analyses for **preference** (Analysis 2d, mirroring Analysis 5 in the main script) and **preference deviation / signed gap** (Analysis 2e, mirroring Analysis 2 in the main script), adding corpus-specific bigram log-odds as an additional predictor.

**Bigram log-odds** captures the positional tendency of each word in "X and Y" frames from cumulative corpus exposure up to each checkpoint. The intuition is that we are computing the ratio of two bigram-chain probabilities — how likely is the sequence "W1 → and → W2" versus "W2 → and → W1", based solely on each word's individual tendency to precede or follow "and":

$$\text{bigram\_logodds} = \log \frac{P(\text{and} \mid W_1) \times P(W_2 \mid \text{and})}{P(\text{and} \mid W_2) \times P(W_1 \mid \text{and})}$$

This treats each word's positional tendency as independent of what the other word in the pair is — in contrast to `freq_prob_c` (RelFreq), which directly observes the specific co-occurrence "W1 and W2". The bigram predictor asks whether the ordering is predictable from each word's general positional habits; RelFreq asks how often the specific pair appeared in each order.

**Derivation to the count formula.** Estimating each conditional probability by its relative frequency ($W_i$ = total corpus count of word $W_i$; $\text{and}$ = total count of "and"):

$$P(\text{and} \mid W_1) \approx \frac{w_1\text{\_and}}{W_1}, \quad P(W_2 \mid \text{and}) \approx \frac{\text{and\_}w_2}{\text{and}}, \quad P(\text{and} \mid W_2) \approx \frac{w_2\text{\_and}}{W_2}, \quad P(W_1 \mid \text{and}) \approx \frac{\text{and\_}w_1}{\text{and}}$$

Substituting into the full ratio:

$$\frac{P(\text{and} \mid W_1) \times P(W_2 \mid \text{and})}{P(\text{and} \mid W_2) \times P(W_1 \mid \text{and})} = \frac{\dfrac{w_1\text{\_and}}{W_1} \times \dfrac{\text{and\_}w_2}{\text{and}}}{\dfrac{w_2\text{\_and}}{W_2} \times \dfrac{\text{and\_}w_1}{\text{and}}}$$

A fraction divided by a fraction equals the first multiplied by the reciprocal of the second $\left(\frac{a/b}{c/d} = \frac{a}{b} \times \frac{d}{c}\right)$, so the numerator fraction gets multiplied by the flipped denominator fraction:

$$= \frac{w_1\text{\_and} \times \text{and\_}w_2}{W_1 \times \text{and}} \times \frac{W_2 \times \text{and}}{w_2\text{\_and} \times \text{and\_}w_1}$$

The $\text{and}$ terms cancel:

$$= \frac{w_1\text{\_and} \times \text{and\_}w_2 \times W_2}{w_2\text{\_and} \times \text{and\_}w_1 \times W_1}$$

Taking the log:

$$\log \frac{w_1\text{\_and} \times \text{and\_}w_2 \times W_2}{w_2\text{\_and} \times \text{and\_}w_1 \times W_1}$$

Taking the log and including the $W_1$/$W_2$ correction gives the full formula actually used:

$$\text{bigram\_logodds} = \log(w_1\text{\_and} + 0.5) + \log(\text{and\_}w_2 + 0.5) - \log(w_2\text{\_and} + 0.5) - \log(\text{and\_}w_1 + 0.5) + \log(W_2 + 0.5) - \log(W_1 + 0.5)$$

where $W_i$ is the total corpus count of word $W_i$ and 0.5 is a pseudocount added before logging to handle zero counts. The $\log(W_2/W_1)$ correction term accounts for the fact that a more frequent word will appear more in pre-"and" position purely by chance; without it, the formula would conflate positional bias with overall word frequency. Here `w1_and` = cumulative count of times W1 appeared immediately before "and" (in any "X and Y" context), `and_w2` = times W2 appeared immediately after "and", etc. Counts come from `count_bigram_exposures.py`; `bigram_logodds_c` is z-scored within each checkpoint.

**Formulas** (fit independently at each of 20 sampled checkpoints):

$$\text{preference} \sim \text{genpref\_c} \times \text{log\_total\_c} + \text{freq\_prob\_c} \times \text{log\_total\_c} + \text{bigram\_logodds\_c} + (1 \mid \text{binom})$$ $$\text{signed\_gap} \sim \text{genpref\_c} \times \text{log\_total\_c} + \text{freq\_prob\_c} \times \text{log\_total\_c} + \text{bigram\_logodds\_c} + (1 \mid \text{binom})$$

Requires `babylm_bigram_exposures.csv` (and optionally `c4_bigram_exposures.csv`) — run `count_bigram_exposures.py` first.

```{r load-bigram-data}
message("Loading bigram exposure counts...")
bigram_blm <- if (file.exists(BLM_BIGRAM_CSV)) {
  read_csv(BLM_BIGRAM_CSV, show_col_types = FALSE)
} else {
  message("  babylm_bigram_exposures.csv not found — run count_bigram_exposures.py first.")
  NULL
}
bigram_c4 <- if (file.exists(C4_BIGRAM_CSV)) {
  read_csv(C4_BIGRAM_CSV, show_col_types = FALSE)
} else {
  message("  c4_bigram_exposures.csv not found — run count_bigram_exposures.py (--skip-babylm) first.")
  NULL
}
has_bigram <- !is.null(bigram_blm) || !is.null(bigram_c4)
message(sprintf("Bigram data available: BabyLM=%s, C4=%s", !is.null(bigram_blm), !is.null(bigram_c4)))
```

```{r fit-bigram-helper, include=FALSE}
# Given a cdat (single-checkpoint slice of train), a corpus string ("BabyLM"/"C4"),
# and the pre-loaded bigram data frames, join bigram counts and return an augmented
# data frame ready for fitting.  Returns NULL if no bigram data for this corpus.
make_bigram_fitdata <- function(cdat, corpus) {
  bg_df <- switch(corpus,
    "BabyLM" = bigram_blm,
    "C4"     = bigram_c4,
    NULL
  )
  if (is.null(bg_df)) return(NULL)

  ck_tokens <- unique(cdat$tokens)
  if (length(ck_tokens) != 1) warning("make_bigram_fitdata: multiple token values in cdat")
  bg_ck <- filter(bg_df, tokens == ck_tokens[1])

  cdat |>
    add_dyn_freq() |>
    mutate(genpref_c = GenPref_centered) |>
    left_join(select(bg_ck, binom, w1_and, and_w1, w2_and, and_w2, w1_total, w2_total), by = "binom") |>
    mutate(
      bigram_logodds   = (log(w1_and + 0.5) + log(and_w2 + 0.5)) -
                         (log(w2_and + 0.5) + log(and_w1 + 0.5)) +
                         log(w2_total + 0.5) - log(w1_total + 0.5),
      bigram_logodds_c = as.numeric(scale(bigram_logodds))
    ) |>
    filter(!is.na(genpref_c), !is.na(freq_prob_c), !is.na(log_total_c),
           !is.na(bigram_logodds_c), !is.na(preference), !is.na(signed_gap))
}
```

```{r fit-2d}
# 2d: preference ~ genpref_c * log_total_c + freq_prob_c * log_total_c + bigram_logodds_c + (1|binom)
if (!has_bigram) {
  message("Skipping 2d — no bigram exposure files found.")
} else {
  formula_pref_bigram <- preference ~ genpref_c * log_total_c +
                                      freq_prob_c * log_total_c +
                                      bigram_logodds_c + (1 | binom)

  future_map(MODELS_LIST, function(model_name) {
    slug   <- make_slug(model_name)
    corpus <- parse_corpus(model_name)
    mdata  <- filter(train, model == model_name)
    steps  <- sample_steps(mdata)
    message(sprintf("\n=== [2d pref] %s (%d checkpoints) ===", slug, length(steps)))

    for (ck_idx in seq_along(steps)) {
      s        <- steps[ck_idx]
      ck       <- sprintf("ck%02d", ck_idx)
      out_path <- file.path(MODELS_DIR, sprintf("pref_btraj_%s_%s.rds", slug, ck))
      if (file.exists(out_path)) {
        message(sprintf("  %s %s — exists, skipping", slug, ck)); next
      }
      fit_data <- make_bigram_fitdata(filter(mdata, step == s), corpus)
      if (is.null(fit_data) || nrow(fit_data) < 20) {
        warning(sprintf("Skipping %s %s — insufficient data", slug, ck)); next
      }
      message(sprintf("  %s %s  step=%d  n=%d", slug, ck, s, nrow(fit_data)))
      fit <- brm(formula = formula_pref_bigram, data = fit_data,
                 family = gaussian(), prior = PRIORS_GAUSSIAN,
                 iter = ITER, warmup = WARMUP, chains = CHAINS, cores = CORES,
                 backend = "cmdstanr", silent = 2, refresh = 0)
      saveRDS(fit, out_path); rm(fit); gc()
      message(sprintf("    Saved: %s", basename(out_path)))
    }
  }, .options = furrr_options(seed = TRUE))
  message("\n2d (pref) fitting done.")
}
```

```{r fit-2e}
# 2e: signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + bigram_logodds_c + (1|binom)
if (!has_bigram) {
  message("Skipping 2e — no bigram exposure files found.")
} else {
  formula_sg_bigram <- signed_gap ~ genpref_c * log_total_c +
                                    freq_prob_c * log_total_c +
                                    bigram_logodds_c + (1 | binom)

  future_map(MODELS_LIST, function(model_name) {
    slug   <- make_slug(model_name)
    corpus <- parse_corpus(model_name)
    mdata  <- filter(train, model == model_name)
    steps  <- sample_steps(mdata)
    message(sprintf("\n=== [2e sg] %s (%d checkpoints) ===", slug, length(steps)))

    for (ck_idx in seq_along(steps)) {
      s        <- steps[ck_idx]
      ck       <- sprintf("ck%02d", ck_idx)
      out_path <- file.path(MODELS_DIR, sprintf("sg_btraj_%s_%s.rds", slug, ck))
      if (file.exists(out_path)) {
        message(sprintf("  %s %s — exists, skipping", slug, ck)); next
      }
      fit_data <- make_bigram_fitdata(filter(mdata, step == s), corpus)
      if (is.null(fit_data) || nrow(fit_data) < 20) {
        warning(sprintf("Skipping %s %s — insufficient data", slug, ck)); next
      }
      message(sprintf("  %s %s  step=%d  n=%d", slug, ck, s, nrow(fit_data)))
      fit <- brm(formula = formula_sg_bigram, data = fit_data,
                 family = gaussian(), prior = PRIORS_GAUSSIAN,
                 iter = ITER, warmup = WARMUP, chains = CHAINS, cores = CORES,
                 backend = "cmdstanr", silent = 2, refresh = 0)
      saveRDS(fit, out_path); rm(fit); gc()
      message(sprintf("    Saved: %s", basename(out_path)))
    }
  }, .options = furrr_options(seed = TRUE))
  message("\n2e (signed gap) fitting done.")
}
```

```{r fit-2f}
# 2f: preference ~ freq_prob_c + bigram_logodds_c + (1 | binom)
# Fit at the final checkpoint only.
if (!has_bigram) {
  message("Skipping 2f — no bigram exposure files found.")
} else {
  formula_pref_bigrel <- preference ~ freq_prob_c + bigram_logodds_c + (1 | binom)

  future_map(MODELS_LIST, function(model_name) {
    slug     <- make_slug(model_name)
    corpus   <- parse_corpus(model_name)
    mdata    <- filter(train, model == model_name)
    s        <- max(mdata$step)
    out_path <- file.path(MODELS_DIR, sprintf("pref_brel_final_%s.rds", slug))
    if (file.exists(out_path)) {
      message(sprintf("  %s — exists, skipping", slug)); return(NULL)
    }
    fit_data <- make_bigram_fitdata(filter(mdata, step == s), corpus)
    if (is.null(fit_data) || nrow(fit_data) < 20) {
      warning(sprintf("Skipping %s — insufficient data", slug)); return(NULL)
    }
    message(sprintf("  %s  step=%d  n=%d", slug, s, nrow(fit_data)))
    fit <- brm(formula = formula_pref_bigrel, data = fit_data,
               family = gaussian(), prior = PRIORS_GAUSSIAN,
               iter = ITER, warmup = WARMUP, chains = CHAINS, cores = CORES,
               backend = "cmdstanr", silent = 2, refresh = 0)
    saveRDS(fit, out_path); rm(fit); gc()
    message(sprintf("    Saved: %s", basename(out_path)))
  }, .options = furrr_options(seed = TRUE))
  message("\n2f fitting done.")
}
```

```{r fit-2h}
# 2h: signed_gap ~ freq_prob_c + bigram_logodds_c + (1 | binom)
# Analogous to 2f but for preference deviation. Fit at the final checkpoint only.
if (!has_bigram) {
  message("Skipping 2h — no bigram exposure files found.")
} else {
  formula_sg_bigrel <- signed_gap ~ freq_prob_c + bigram_logodds_c + (1 | binom)

  future_map(MODELS_LIST, function(model_name) {
    slug     <- make_slug(model_name)
    corpus   <- parse_corpus(model_name)
    mdata    <- filter(train, model == model_name)
    s        <- max(mdata$step)
    out_path <- file.path(MODELS_DIR, sprintf("sg_brel_final_%s.rds", slug))
    if (file.exists(out_path)) {
      message(sprintf("  %s — exists, skipping", slug)); return(NULL)
    }
    fit_data <- make_bigram_fitdata(filter(mdata, step == s), corpus)
    if (is.null(fit_data) || nrow(fit_data) < 20) {
      warning(sprintf("Skipping %s — insufficient data", slug)); return(NULL)
    }
    message(sprintf("  %s  step=%d  n=%d", slug, s, nrow(fit_data)))
    fit <- brm(formula = formula_sg_bigrel, data = fit_data,
               family = gaussian(), prior = PRIORS_GAUSSIAN,
               iter = ITER, warmup = WARMUP, chains = CHAINS, cores = CORES,
               backend = "cmdstanr", silent = 2, refresh = 0)
    saveRDS(fit, out_path); rm(fit); gc()
    message(sprintf("    Saved: %s", basename(out_path)))
  }, .options = furrr_options(seed = TRUE))
  message("\n2h fitting done.")
}
```

------------------------------------------------------------------------

# Results (brms models)

## 2a: Human preference \~ model preference

```{r load-humfit}
TERM_LEVELS_HUM <- "preference"
TERM_LABELS_HUM <- "Model preference"
humfit_coefs <- load_traj_coefs("humfit_traj_", TERM_LEVELS_HUM, TERM_LABELS_HUM)
```

```{r plot-humfit-traj, fig.height=9}
if (nrow(humfit_coefs) > 0) {
  use_tokens <- !all(is.na(humfit_coefs$tokens))
  df <- humfit_coefs |>
    mutate(p_gt0     = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE),
           pct_label = sprintf("%.0f%%", p_gt0 * 100),
           x_val     = if (use_tokens) tokens / 1e6 else ck_idx) |>
    group_by(corpus, model_size) |>
    arrange(x_val, .by_group = TRUE) |>
    mutate(pt_idx = row_number()) |>
    ungroup()
  df_labels <- filter(df, pt_idx %% 2 == 1)
  x_label <- if (use_tokens) "Training tokens (millions)" else "Checkpoint index"

  ggplot(df, aes(x = x_val, y = estimate)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_linerange(aes(ymin = lo, ymax = hi, colour = p_gt0),
                   linewidth = 0.7, alpha = 0.7) +
    geom_point(aes(fill = p_gt0), shape = 21,
               size = 3.2, colour = "grey25", stroke = 0.35) +
    geom_text(data = df_labels, aes(label = pct_label, y = hi),
              size = 2.8, vjust = -0.5, colour = "black") +
    scale_colour_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                           midpoint = 0.5, limits = c(0, 1), guide = "none") +
    scale_fill_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                         midpoint = 0.5, limits = c(0, 1),
                         name = "P(\u03b2 > 0)",
                         labels = scales::percent_format(accuracy = 1)) +
    scale_x_continuous(x_label) +
    scale_y_continuous("Model preference coefficient (95% CI)",
                       expand = expansion(mult = c(0.05, 0.2))) +
    facet_grid(corpus ~ model_size) +
    coord_cartesian(clip = "off") +
    labs(title = "Human ordering choice ~ model preference (logistic)") +
    theme_bw(base_size = 13) +
    theme(legend.position = "right", panel.grid.minor = element_blank(),
          plot.title   = element_text(face = "bold", size = 14),
          axis.text    = element_text(colour = "black", size = 13),
          axis.title   = element_text(colour = "black", size = 15),
          strip.text   = element_text(colour = "black", face = "bold", size = 14))
} else {
  message("No humfit_traj data — fit models first.")
}
```

------------------------------------------------------------------------

## 2b: Model preference \~ individual constraints

```{r load-constr}
TERM_LEVELS_CONSTR <- c(
  CONSTRAINT_COLS, "log_total_c", "freq_prob_c",
  paste0(CONSTRAINT_COLS, ":log_total_c"), "freq_prob_c:log_total_c"
)
TERM_LABELS_CONSTR <- c(
  CONSTRAINT_COLS, "Overall freq\n(ln total)", "RelFreq\n(P(\u03b1)\u22120.5)",
  paste0(CONSTRAINT_COLS, " \u00d7\nOverall freq"), "RelFreq \u00d7\nOverall freq"
)
constr_coefs <- load_traj_coefs("constr_traj_", TERM_LEVELS_CONSTR, TERM_LABELS_CONSTR)
```

```{r plot-constr-main, fig.height=10}
# Main effects only — one plot per model
if (nrow(constr_coefs) > 0) {
  use_tokens <- !all(is.na(constr_coefs$tokens))
  constr_main <- constr_coefs |>
    filter(as.character(term) %in% CONSTRAINT_COLS)

  df <- constr_main |>
    mutate(p_gt0     = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE),
           pct_label = sprintf("%.0f%%", p_gt0 * 100),
           x_val     = if (use_tokens) tokens / 1e6 else ck_idx) |>
    group_by(corpus, model_size, term) |>
    arrange(x_val, .by_group = TRUE) |>
    mutate(pt_idx = row_number()) |>
    ungroup()
  x_label <- if (use_tokens) "Training tokens (millions)" else "Checkpoint index"

  models_list <- df |> distinct(corpus, model_size) |> arrange(corpus, model_size)
  for (i in seq_len(nrow(models_list))) {
    corp <- models_list$corpus[i]
    msize <- models_list$model_size[i]
    df_mod <- filter(df, corpus == corp, model_size == msize)
    df_labels <- filter(df_mod, pt_idx %% 2 == 1)
    p <- ggplot(df_mod, aes(x = x_val, y = estimate)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_linerange(aes(ymin = lo, ymax = hi, colour = p_gt0),
                     linewidth = 0.7, alpha = 0.7) +
      geom_point(aes(fill = p_gt0), shape = 21,
                 size = 3.0, colour = "grey25", stroke = 0.35) +
      geom_text(data = df_labels, aes(label = pct_label, y = hi),
                size = 2.5, vjust = -0.5, colour = "black") +
      scale_colour_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                             midpoint = 0.5, limits = c(0, 1), guide = "none") +
      scale_fill_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                           midpoint = 0.5, limits = c(0, 1),
                           name = "P(\u03b2 > 0)",
                           labels = scales::percent_format(accuracy = 1)) +
      scale_x_continuous(x_label) +
      scale_y_continuous("Coefficient (95% CI)",
                         expand = expansion(mult = c(0.05, 0.2))) +
      facet_wrap(~ term, ncol = 3) +
      coord_cartesian(clip = "off") +
      labs(title = sprintf("Constraint main effects — %s %s", corp, msize)) +
      theme_bw(base_size = 12) +
      theme(legend.position = "right", panel.grid.minor = element_blank(),
            plot.title   = element_text(face = "bold", size = 13),
            axis.text    = element_text(colour = "black", size = 11),
            axis.title   = element_text(colour = "black", size = 13),
            strip.text   = element_text(colour = "black", face = "bold", size = 11))
    print(p)
  }
} else {
  message("No constr_traj data — fit models first.")
}
```

------------------------------------------------------------------------

## 2g: VIF-filtered constraints trajectory

```{r load-constr-vif}
# Build term level vectors from the VIF-selected predictors.
# Use the union across both corpora so load_traj_coefs can handle both.
vif_constr_union <- union(vif_constr_blm, vif_constr_c4)

TERM_LEVELS_VIF <- c(
  vif_constr_union, "log_total_c", "freq_prob_c",
  paste0(vif_constr_union, ":log_total_c"), "freq_prob_c:log_total_c"
)
TERM_LABELS_VIF <- c(
  vif_constr_union, "Overall freq\n(ln total)", "RelFreq\n(P(\u03b1)\u22120.5)",
  paste0(vif_constr_union, " \u00d7\nOverall freq"), "RelFreq \u00d7\nOverall freq"
)
constr_vif_coefs <- load_traj_coefs("constr_vif_traj_", TERM_LEVELS_VIF, TERM_LABELS_VIF)
```

```{r plot-constr-vif-main, fig.height=10}
# Main effects only — one plot per model
if (nrow(constr_vif_coefs) > 0) {
  use_tokens  <- !all(is.na(constr_vif_coefs$tokens))
  vif_main    <- constr_vif_coefs |>
    filter(as.character(term) %in% vif_constr_union)

  df <- vif_main |>
    mutate(p_gt0     = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE),
           pct_label = sprintf("%.0f%%", p_gt0 * 100),
           x_val     = if (use_tokens) tokens / 1e6 else ck_idx) |>
    group_by(corpus, model_size, term) |>
    arrange(x_val, .by_group = TRUE) |>
    mutate(pt_idx = row_number()) |>
    ungroup()
  x_label <- if (use_tokens) "Training tokens (millions)" else "Checkpoint index"

  models_list <- df |> distinct(corpus, model_size) |> arrange(corpus, model_size)
  for (i in seq_len(nrow(models_list))) {
    corp  <- models_list$corpus[i]
    msize <- models_list$model_size[i]
    df_mod    <- filter(df, corpus == corp, model_size == msize)
    df_labels <- filter(df_mod, pt_idx %% 2 == 1)
    p <- ggplot(df_mod, aes(x = x_val, y = estimate)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_linerange(aes(ymin = lo, ymax = hi, colour = p_gt0),
                     linewidth = 0.7, alpha = 0.7) +
      geom_point(aes(fill = p_gt0), shape = 21,
                 size = 3.0, colour = "grey25", stroke = 0.35) +
      geom_text(data = df_labels, aes(label = pct_label, y = hi),
                size = 2.5, vjust = -0.5, colour = "black") +
      scale_colour_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                             midpoint = 0.5, limits = c(0, 1), guide = "none") +
      scale_fill_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                           midpoint = 0.5, limits = c(0, 1),
                           name = "P(\u03b2 > 0)",
                           labels = scales::percent_format(accuracy = 1)) +
      scale_x_continuous(x_label) +
      scale_y_continuous("Coefficient (95% CI)",
                         expand = expansion(mult = c(0.05, 0.2))) +
      facet_wrap(~ term, ncol = 3) +
      coord_cartesian(clip = "off") +
      labs(title = sprintf("VIF-filtered constraint main effects — %s %s", corp, msize)) +
      theme_bw(base_size = 12) +
      theme(legend.position = "right", panel.grid.minor = element_blank(),
            plot.title   = element_text(face = "bold", size = 13),
            axis.text    = element_text(colour = "black", size = 11),
            axis.title   = element_text(colour = "black", size = 13),
            strip.text   = element_text(colour = "black", face = "bold", size = 11))
    print(p)
  }
} else {
  message("No constr_vif_traj data — fit models first (fit-2g chunk).")
}
```

```{r plot-constr-vif-interactions, fig.height=10}
# Interaction terms — one plot per model
if (nrow(constr_vif_coefs) > 0) {
  use_tokens        <- !all(is.na(constr_vif_coefs$tokens))
  interaction_labels <- paste0(vif_constr_union, " \u00d7\nOverall freq")
  constr_vif_int <- constr_vif_coefs |>
    filter(as.character(term) %in% interaction_labels) |>
    mutate(term = gsub(" \u00d7\nOverall freq", "", as.character(term)),
           term = factor(term, levels = vif_constr_union))

  df <- constr_vif_int |>
    mutate(p_gt0     = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE),
           pct_label = sprintf("%.0f%%", p_gt0 * 100),
           x_val     = if (use_tokens) tokens / 1e6 else ck_idx) |>
    group_by(corpus, model_size, term) |>
    arrange(x_val, .by_group = TRUE) |>
    mutate(pt_idx = row_number()) |>
    ungroup()
  x_label <- if (use_tokens) "Training tokens (millions)" else "Checkpoint index"

  models_list <- df |> distinct(corpus, model_size) |> arrange(corpus, model_size)
  for (i in seq_len(nrow(models_list))) {
    corp  <- models_list$corpus[i]
    msize <- models_list$model_size[i]
    df_mod    <- filter(df, corpus == corp, model_size == msize)
    df_labels <- filter(df_mod, pt_idx %% 2 == 1)
    p <- ggplot(df_mod, aes(x = x_val, y = estimate)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_linerange(aes(ymin = lo, ymax = hi, colour = p_gt0),
                     linewidth = 0.7, alpha = 0.7) +
      geom_point(aes(fill = p_gt0), shape = 21,
                 size = 3.0, colour = "grey25", stroke = 0.35) +
      geom_text(data = df_labels, aes(label = pct_label, y = hi),
                size = 2.5, vjust = -0.5, colour = "black") +
      scale_colour_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                             midpoint = 0.5, limits = c(0, 1), guide = "none") +
      scale_fill_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                           midpoint = 0.5, limits = c(0, 1),
                           name = "P(\u03b2 > 0)",
                           labels = scales::percent_format(accuracy = 1)) +
      scale_x_continuous(x_label) +
      scale_y_continuous("Coefficient (95% CI)",
                         expand = expansion(mult = c(0.05, 0.2))) +
      facet_wrap(~ term, ncol = 3) +
      coord_cartesian(clip = "off") +
      labs(title = sprintf("VIF-filtered constraint \u00d7 Overall freq — %s %s", corp, msize)) +
      theme_bw(base_size = 12) +
      theme(legend.position = "right", panel.grid.minor = element_blank(),
            plot.title   = element_text(face = "bold", size = 13),
            axis.text    = element_text(colour = "black", size = 11),
            axis.title   = element_text(colour = "black", size = 13),
            strip.text   = element_text(colour = "black", face = "bold", size = 11))
    print(p)
  }
} else {
  message("No constr_vif_traj data — fit models first (fit-2g chunk).")
}
```

------------------------------------------------------------------------

## 2d: Preference + bigram log-odds trajectory

```{r load-pref-btraj}
TERM_LEVELS_BTRAJ <- c(
  "genpref_c", "freq_prob_c", "bigram_logodds_c", "log_total_c",
  "genpref_c:log_total_c", "freq_prob_c:log_total_c"
)
TERM_LABELS_BTRAJ <- c(
  "AbsPref", "RelFreq", "BigramLogOdds",
  "Overall freq\n(ln, z)", "AbsPref \u00d7\nOverall freq", "RelFreq \u00d7\nOverall freq"
)
pref_btraj_coefs <- load_traj_coefs("pref_btraj_", TERM_LEVELS_BTRAJ, TERM_LABELS_BTRAJ)
```

```{r plot-pref-btraj, fig.width=16, fig.height=7}
if (nrow(pref_btraj_coefs) > 0) {
  use_tokens <- !all(is.na(pref_btraj_coefs$tokens))
  df <- pref_btraj_coefs |>
    mutate(p_gt0     = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE),
           pct_label = sprintf("%.0f%%", p_gt0 * 100),
           x_val     = if (use_tokens) tokens / 1e9 else ck_idx) |>
    group_by(corpus, model_size, term) |>
    arrange(x_val, .by_group = TRUE) |>
    mutate(pt_idx = row_number()) |>
    ungroup()
  df_labels <- filter(df, pt_idx %% 4 == 1, pt_idx > 1)
  x_label <- if (use_tokens) "Training tokens (billions)" else "Checkpoint index"

  ggplot(df, aes(x = x_val, y = estimate)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_linerange(aes(ymin = lo, ymax = hi, colour = p_gt0),
                   linewidth = 0.7, alpha = 0.7) +
    geom_point(aes(fill = p_gt0), shape = 21,
               size = 3.0, colour = "grey25", stroke = 0.35) +
    geom_text(data = df_labels, aes(label = pct_label, y = hi),
              size = 2.8, vjust = -0.5, colour = "black") +
    scale_colour_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                           midpoint = 0.5, limits = c(0, 1), guide = "none") +
    scale_fill_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                         midpoint = 0.5, limits = c(0, 1),
                         name = "P(\u03b2 > 0)",
                         labels = scales::percent_format(accuracy = 1)) +
    scale_x_continuous(x_label) +
    scale_y_continuous("Coefficient (95% CI)", expand = expansion(mult = c(0.05, 0.2))) +
    facet_grid(corpus ~ model_size + term) +
    coord_cartesian(clip = "off") +
    labs(title = "Preference + bigram log-odds — per-checkpoint trajectories") +
    theme_bw(base_size = 12) +
    theme(legend.position = "right", panel.grid.minor = element_blank(),
          plot.title  = element_text(face = "bold", size = 13),
          axis.text   = element_text(colour = "black", size = 11),
          axis.title  = element_text(colour = "black", size = 13),
          strip.text  = element_text(colour = "black", face = "bold", size = 11))
} else {
  message("No pref_btraj data — fit models first (fit-2d chunk).")
}
```

------------------------------------------------------------------------

## 2e: Preference deviation (signed gap) + bigram log-odds trajectory

```{r load-sg-btraj}
sg_btraj_coefs <- load_traj_coefs("sg_btraj_", TERM_LEVELS_BTRAJ, TERM_LABELS_BTRAJ)
```

```{r plot-sg-btraj, fig.width=16, fig.height=7}
if (nrow(sg_btraj_coefs) > 0) {
  use_tokens <- !all(is.na(sg_btraj_coefs$tokens))
  df <- sg_btraj_coefs |>
    mutate(p_gt0     = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE),
           pct_label = sprintf("%.0f%%", p_gt0 * 100),
           x_val     = if (use_tokens) tokens / 1e9 else ck_idx) |>
    group_by(corpus, model_size, term) |>
    arrange(x_val, .by_group = TRUE) |>
    mutate(pt_idx = row_number()) |>
    ungroup()
  df_labels <- filter(df, pt_idx %% 4 == 1, pt_idx > 1)
  x_label <- if (use_tokens) "Training tokens (billions)" else "Checkpoint index"

  ggplot(df, aes(x = x_val, y = estimate)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_linerange(aes(ymin = lo, ymax = hi, colour = p_gt0),
                   linewidth = 0.7, alpha = 0.7) +
    geom_point(aes(fill = p_gt0), shape = 21,
               size = 3.0, colour = "grey25", stroke = 0.35) +
    geom_text(data = df_labels, aes(label = pct_label, y = hi),
              size = 2.8, vjust = -0.5, colour = "black") +
    scale_colour_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                           midpoint = 0.5, limits = c(0, 1), guide = "none") +
    scale_fill_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                         midpoint = 0.5, limits = c(0, 1),
                         name = "P(\u03b2 > 0)",
                         labels = scales::percent_format(accuracy = 1)) +
    scale_x_continuous(x_label) +
    scale_y_continuous("Coefficient (95% CI)", expand = expansion(mult = c(0.05, 0.2))) +
    facet_grid(corpus ~ model_size + term) +
    coord_cartesian(clip = "off") +
    labs(title = "Preference deviation (signed gap) + bigram log-odds — per-checkpoint trajectories") +
    theme_bw(base_size = 12) +
    theme(legend.position = "right", panel.grid.minor = element_blank(),
          plot.title  = element_text(face = "bold", size = 13),
          axis.text   = element_text(colour = "black", size = 11),
          axis.title  = element_text(colour = "black", size = 13),
          strip.text  = element_text(colour = "black", face = "bold", size = 11))
} else {
  message("No sg_btraj data — fit models first (fit-2e chunk).")
}
```

------------------------------------------------------------------------

## 2f: Model preference \~ relative frequency + bigram log-odds (final checkpoint)

Simple model asking whether bigram positional tendency predicts model preference above and beyond relative frequency, fit at the final training checkpoint.

$$\text{preference} \sim \text{freq\_prob\_c} + \text{bigram\_logodds\_c} + (1 \mid \text{binom})$$

```{r load-pref-brel-final}
TERM_LEVELS_BREL <- c("freq_prob_c", "bigram_logodds_c")
TERM_LABELS_BREL <- c("RelFreq", "BigramLogOdds")

load_final_coefs <- function(prefix, term_levels, term_labels) {
  paths <- list.files(MODELS_DIR, pattern = paste0("^", prefix, ".+\\.rds$"),
                      full.names = TRUE)
  if (length(paths) == 0) { warning(sprintf("No files for prefix '%s'", prefix)); return(tibble()) }
  map_dfr(paths, function(path) {
    slug <- sub(paste0("^", prefix), "", sub("\\.rds$", "", basename(path)))
    fit  <- readRDS(path)
    out  <- extract_coefs(fit, slug, term_levels, term_labels)
    rm(fit); gc()
    out
  })
}

pref_brel_coefs <- load_final_coefs("pref_brel_final_", TERM_LEVELS_BREL, TERM_LABELS_BREL)
```

```{r plot-pref-brel-final, fig.width=10, fig.height=5}
if (nrow(pref_brel_coefs) > 0) {
  df <- pref_brel_coefs |>
    mutate(p_gt0 = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE))

  ggplot(df, aes(x = term, y = estimate, colour = p_gt0, fill = p_gt0)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_linerange(aes(ymin = lo, ymax = hi), linewidth = 0.7, alpha = 0.7) +
    geom_point(shape = 21, size = 3.0, colour = "grey25", stroke = 0.35) +
    scale_colour_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                           midpoint = 0.5, limits = c(0, 1), guide = "none") +
    scale_fill_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                         midpoint = 0.5, limits = c(0, 1),
                         name = "P(\u03b2 > 0)",
                         labels = scales::percent_format(accuracy = 1)) +
    scale_y_continuous("Coefficient (95% CI)", expand = expansion(mult = c(0.05, 0.15))) +
    facet_grid(corpus ~ model_size) +
    coord_flip() +
    labs(title = "Preference ~ RelFreq + BigramLogOdds — final checkpoint") +
    theme_bw(base_size = 12) +
    theme(legend.position = "right", panel.grid.minor = element_blank(),
          plot.title  = element_text(face = "bold", size = 13),
          axis.text   = element_text(colour = "black", size = 11),
          axis.title  = element_text(colour = "black", size = 13),
          strip.text  = element_text(colour = "black", face = "bold", size = 11))
} else {
  message("No pref_brel_final data — fit models first (fit-2f chunk).")
}
```

```{r tbl-pref-brel-final, results="asis"}
if (nrow(pref_brel_coefs) > 0) {
  pref_brel_coefs |>
    mutate(p_gt0 = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE)) |>
    select(Corpus = corpus, Size = model_size, Term = term,
           Mean = estimate, SD = se, `2.5%` = lo, `97.5%` = hi,
           `P(β>0)` = p_gt0) |>
    arrange(Corpus, Size, Term) |>
    mutate(across(where(is.numeric), ~ round(.x, 3))) |>
    knitr::kable(format = "markdown")
} else {
  message("No pref_brel_final data.")
}
```

------------------------------------------------------------------------

## 2h: Preference deviation \~ relative frequency + bigram log-odds (final checkpoint)

Analogous to 2f but for preference deviation (signed gap), fit at the final training checkpoint.

$$\text{signed\_gap} \sim \text{freq\_prob\_c} + \text{bigram\_logodds\_c} + (1 \mid \text{binom})$$

```{r load-sg-brel-final}
sg_brel_coefs <- load_final_coefs("sg_brel_final_", TERM_LEVELS_BREL, TERM_LABELS_BREL)
```

```{r plot-sg-brel-final, fig.width=10, fig.height=5}
if (nrow(sg_brel_coefs) > 0) {
  df <- sg_brel_coefs |>
    mutate(p_gt0 = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE))

  ggplot(df, aes(x = term, y = estimate, colour = p_gt0, fill = p_gt0)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_linerange(aes(ymin = lo, ymax = hi), linewidth = 0.7, alpha = 0.7) +
    geom_point(shape = 21, size = 3.0, colour = "grey25", stroke = 0.35) +
    scale_colour_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                           midpoint = 0.5, limits = c(0, 1), guide = "none") +
    scale_fill_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                         midpoint = 0.5, limits = c(0, 1),
                         name = "P(\u03b2 > 0)",
                         labels = scales::percent_format(accuracy = 1)) +
    scale_y_continuous("Coefficient (95% CI)", expand = expansion(mult = c(0.05, 0.15))) +
    facet_grid(corpus ~ model_size) +
    coord_flip() +
    labs(title = "Preference deviation ~ RelFreq + BigramLogOdds — final checkpoint") +
    theme_bw(base_size = 12) +
    theme(legend.position = "right", panel.grid.minor = element_blank(),
          plot.title  = element_text(face = "bold", size = 13),
          axis.text   = element_text(colour = "black", size = 11),
          axis.title  = element_text(colour = "black", size = 13),
          strip.text  = element_text(colour = "black", face = "bold", size = 11))
} else {
  message("No sg_brel_final data — fit models first (fit-2h chunk).")
}
```

```{r tbl-sg-brel-final, results="asis"}
if (nrow(sg_brel_coefs) > 0) {
  sg_brel_coefs |>
    mutate(p_gt0 = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE)) |>
    select(Corpus = corpus, Size = model_size, Term = term,
           Mean = estimate, SD = se, `2.5%` = lo, `97.5%` = hi,
           `P(β>0)` = p_gt0) |>
    arrange(Corpus, Size, Term) |>
    mutate(across(where(is.numeric), ~ round(.x, 3))) |>
    knitr::kable(format = "markdown")
} else {
  message("No sg_brel_final data.")
}
```

------------------------------------------------------------------------

```{r plot-constr-interactions, fig.height=10}

# Constraint × OverallFreq interactions — one plot per model
if (nrow(constr_coefs) > 0) {
  use_tokens <- !all(is.na(constr_coefs$tokens))
  interaction_labels <- paste0(CONSTRAINT_COLS, " \u00d7\nOverall freq")
  constr_int <- constr_coefs |>
    filter(as.character(term) %in% interaction_labels) |>
    mutate(term = gsub(" \u00d7\nOverall freq", "", as.character(term)),
           term = factor(term, levels = CONSTRAINT_COLS))

  df <- constr_int |>
    mutate(p_gt0     = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE),
           pct_label = sprintf("%.0f%%", p_gt0 * 100),
           x_val     = if (use_tokens) tokens / 1e6 else ck_idx) |>
    group_by(corpus, model_size, term) |>
    arrange(x_val, .by_group = TRUE) |>
    mutate(pt_idx = row_number()) |>
    ungroup()
  x_label <- if (use_tokens) "Training tokens (millions)" else "Checkpoint index"

  models_list <- df |> distinct(corpus, model_size) |> arrange(corpus, model_size)
  for (i in seq_len(nrow(models_list))) {
    corp <- models_list$corpus[i]
    msize <- models_list$model_size[i]
    df_mod <- filter(df, corpus == corp, model_size == msize)
    df_labels <- filter(df_mod, pt_idx %% 2 == 1)
    p <- ggplot(df_mod, aes(x = x_val, y = estimate)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_linerange(aes(ymin = lo, ymax = hi, colour = p_gt0),
                     linewidth = 0.7, alpha = 0.7) +
      geom_point(aes(fill = p_gt0), shape = 21,
                 size = 3.0, colour = "grey25", stroke = 0.35) +
      geom_text(data = df_labels, aes(label = pct_label, y = hi),
                size = 2.5, vjust = -0.5, colour = "black") +
      scale_colour_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                             midpoint = 0.5, limits = c(0, 1), guide = "none") +
      scale_fill_gradient2(low = "#F0B323", mid = "grey93", high = "#154733",
                           midpoint = 0.5, limits = c(0, 1),
                           name = "P(\u03b2 > 0)",
                           labels = scales::percent_format(accuracy = 1)) +
      scale_x_continuous(x_label) +
      scale_y_continuous("Coefficient (95% CI)",
                         expand = expansion(mult = c(0.05, 0.2))) +
      facet_wrap(~ term, ncol = 3) +
      coord_cartesian(clip = "off") +
      labs(title = sprintf("Constraint \u00d7 Overall freq interactions — %s %s", corp, msize)) +
      theme_bw(base_size = 12) +
      theme(legend.position = "right", panel.grid.minor = element_blank(),
            plot.title   = element_text(face = "bold", size = 13),
            axis.text    = element_text(colour = "black", size = 11),
            axis.title   = element_text(colour = "black", size = 13),
            strip.text   = element_text(colour = "black", face = "bold", size = 11))
    print(p)
  }
} else {
  message("No constr_traj data — fit models first.")
}
```
