---
title: "Constraint-Level Alignment Over Training"
subtitle: "What proportion of model ordering preferences follow each individual constraint?"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 6, dpi = 150,
                      out.width = "100%")
```

# Overview

`GenPref` (AbsPref) is a composite of individual constraints on binomial ordering.
This document asks: as models train, **how much do their ordering preferences align
with each individual constraint?**

For each constraint at each training checkpoint, we compute:

- **Model alignment**: proportion of items where the model's preferred ordering
  matches what that constraint would predict
- **Corpus RelFreq baseline** (dynamic): same proportion, but using corpus relative
  frequency as the "preference" — shows what you'd expect if the model were just
  tracking training corpus frequency
- **Human baseline** (static): proportion where human preference data agrees with
  each constraint — an upper bound on what a well-calibrated model should reach

A constraint that the model learns *beyond* what corpus frequency teaches will show
model alignment exceeding the corpus RelFreq baseline.

The **constraints** (all signed: positive = predicts α ordering):

| Constraint | Description |
|------------|-------------|
| `Freq`     | Log frequency ratio of the two words (external reference corpus) |
| `Len`      | Length difference |
| `Lapse`    | Temporal distance / recency |
| `BStress`  | Beginning stress |
| `Form`     | Formal/phonological properties |
| `Percept`  | Perceptual salience |
| `Culture`  | Cultural ordering tendencies |
| `Power`    | Power/animacy |
| `Intense`  | Intensity |
| `Icon`     | Iconicity |

---

```{r libraries}
library(tidyverse)
library(duckdb)
library(furrr)
library(patchwork)
library(scales)
library(kableExtra)
```

```{r config}
SCRIPT_DIR <- {
  knitr_input <- tryCatch(knitr::current_input(), error = function(e) NULL)
  if (!is.null(knitr_input) && nchar(knitr_input) > 0) {
    dirname(normalizePath(knitr_input, mustWork = FALSE))
  } else if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
    dirname(normalizePath(rstudioapi::getActiveDocumentContext()$path, mustWork = FALSE))
  } else {
    normalizePath(".", mustWork = FALSE)
  }
}

DATA_DIR          <- normalizePath(file.path(SCRIPT_DIR, "../Data/processed"), mustWork = FALSE)
TRAIN_CSV         <- file.path(DATA_DIR, "training_attested.csv")
EXPOSURES_CSV     <- normalizePath(file.path(DATA_DIR, "checkpoint_results_with_exposures.csv"),
                                   mustWork = FALSE)
EXPOSURES_CSV_FWD <- gsub("\\\\", "/", EXPOSURES_CSV)
EXPOSURES_PARQUET     <- sub("\\.csv$", ".parquet", EXPOSURES_CSV)
EXPOSURES_PARQUET_FWD <- gsub("\\\\", "/", EXPOSURES_PARQUET)

N_CHECKPOINTS <- 20   # number of evenly-spaced checkpoints to sample per model
SIZE_LEVELS   <- c("125M", "350M", "1.3B")
SIZE_COLOURS  <- c("125M" = "#74add1", "350M" = "#007030", "1.3B" = "#154733")

CONSTRAINTS <- c("Freq", "Len", "Lapse", "BStress",
                 "Form", "Percept", "Culture", "Power", "Intense", "Icon")

N_CORES <- parallel::detectCores(logical = FALSE)
plan(multicore, workers = N_CORES)

stopifnot(file.exists(TRAIN_CSV), file.exists(EXPOSURES_CSV))
```

```{r helpers}
parse_corpus <- function(x) case_when(
  grepl("babylm", x, TRUE) ~ "BabyLM",
  grepl("opt.c4", x, TRUE) ~ "C4",
  TRUE ~ NA_character_
)
parse_size <- function(x) case_when(
  grepl("125m",    x, TRUE) ~ "125M",
  grepl("350m",    x, TRUE) ~ "350M",
  grepl("1[._]3b", x, TRUE) ~ "1.3B",
  TRUE ~ NA_character_
)

sample_steps <- function(model_df, n = N_CHECKPOINTS) {
  steps <- sort(unique(model_df$step))
  idx   <- unique(round(seq(1, length(steps), length.out = min(n, length(steps)))))
  steps[idx]
}

# Parquet cache (reuse if already exists from casestudy run)
ensure_parquet <- function() {
  if (file.exists(EXPOSURES_PARQUET)) return(invisible(NULL))
  message("Converting exposures CSV to Parquet (one-time)...")
  con <- dbConnect(duckdb::duckdb())
  on.exit(dbDisconnect(con, shutdown = TRUE))
  dbExecute(con, sprintf("SET threads = %d", N_CORES))
  dbExecute(con, sprintf(
    "COPY (SELECT * FROM read_csv_auto('%s')) TO '%s' (FORMAT PARQUET, COMPRESSION ZSTD)",
    EXPOSURES_CSV_FWD, EXPOSURES_PARQUET_FWD
  ))
}

# Fetch corpus RelFreq (alpha_seen / beta_seen) for a given set of model × step pairs
batch_dyn_freq <- function(meta_df) {
  stopifnot(all(c("model", "step") %in% names(meta_df)))
  con <- dbConnect(duckdb::duckdb())
  on.exit(dbDisconnect(con, shutdown = TRUE))
  dbExecute(con, sprintf("SET threads = %d", N_CORES))

  src <- if (file.exists(EXPOSURES_PARQUET))
    sprintf("read_parquet('%s')", EXPOSURES_PARQUET_FWD)
  else
    sprintf("read_csv_auto('%s')", EXPOSURES_CSV_FWD)
  dbExecute(con, sprintf("CREATE VIEW exp AS SELECT * FROM %s", src))
  dbWriteTable(con, "pairs", meta_df |> distinct(model, step), overwrite = TRUE)

  freq_raw <- dbGetQuery(con, "
    SELECT e.model, e.step, e.binom,
           ANY_VALUE(e.alpha_seen) AS alpha_seen,
           ANY_VALUE(e.beta_seen)  AS beta_seen
    FROM exp e
    INNER JOIN pairs p ON e.model = p.model AND e.step = p.step
    GROUP BY e.model, e.step, e.binom
  ")

  as_tibble(freq_raw) |>
    mutate(
      freq_prob_c = if_else(
        alpha_seen > 0 & beta_seen > 0,
        plogis(log(alpha_seen / beta_seen)) - 0.5,
        NA_real_
      ),
      log_total = if_else(
        alpha_seen + beta_seen > 0,
        log(alpha_seen + beta_seen),
        NA_real_
      )
    )
}

# Compute per-constraint alignment rates for a data frame that has:
#   preference (model log-odds), freq_prob_c (corpus RelFreq centred),
#   hum_pref (human P(alpha)), and all CONSTRAINTS columns.
# Returns one row per constraint with: model_align, relfreq_align, human_align.
compute_alignment <- function(df) {
  map_dfr(CONSTRAINTS, function(cname) {
    vals <- df[[cname]]
    # Drop ties (constraint == 0) so sign is well-defined
    ok <- !is.na(vals) & vals != 0 &
          !is.na(df$preference) & !is.na(df$freq_prob_c) & !is.na(df$hum_pref)
    d <- df[ok, ]
    if (nrow(d) == 0) return(tibble(constraint = cname,
                                    model_align   = NA_real_,
                                    relfreq_align = NA_real_,
                                    human_align   = NA_real_,
                                    n             = 0L))
    c_sign   <- sign(d[[cname]])
    tibble(
      constraint    = cname,
      model_align   = mean(sign(d$preference) == c_sign),
      relfreq_align = mean(sign(d$freq_prob_c) == c_sign),
      human_align   = mean(sign(d$hum_pref - 0.5) == c_sign),
      n             = nrow(d)
    )
  })
}
```

---

# Data Loading

```{r parquet-cache}
ensure_parquet()
```

```{r load-train}
message("Loading training data...")
train <- read_csv(TRAIN_CSV, show_col_types = FALSE)

MODELS_LIST <- sort(unique(train$model))
MODELS_LIST <- MODELS_LIST[!is.na(parse_corpus(MODELS_LIST))]
message(sprintf("%d models: %s", length(MODELS_LIST), paste(MODELS_LIST, collapse = ", ")))
```

```{r ck-index}
# Sampled checkpoint index: N_CHECKPOINTS evenly-spaced steps per model
ck_meta <- map_dfr(MODELS_LIST, function(m) {
  steps <- sample_steps(filter(train, model == m))
  tibble(model = m, ck_idx = seq_along(steps), step = steps)
})
message(sprintf("Sampled %d total checkpoint × model rows", nrow(ck_meta)))
```

```{r load-exposures}
message("Fetching corpus RelFreq at sampled checkpoints (single DuckDB scan)...")
t0 <- proc.time()
freq_sampled <- batch_dyn_freq(ck_meta)
message(sprintf("  Done in %.1f s (%d rows)", (proc.time() - t0)[["elapsed"]], nrow(freq_sampled)))
```

```{r join-data}
# Merge training data (preferences + constraints) with corpus RelFreq at each checkpoint.
# Constraint columns are item-level constants; take them from the training data directly.
dat <- train |>
  inner_join(ck_meta |> select(model, step, ck_idx), by = c("model", "step")) |>
  left_join(freq_sampled, by = c("model", "step", "binom")) |>
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS)
  ) |>
  filter(!is.na(corpus), !is.na(model_size))

# Attach training token count: use the maximum tokens seen at that checkpoint
# (tokens column in the training CSV is checkpoint-specific)
message(sprintf("Joined dataset: %d rows, %d models, %d unique binomials, %d unique steps",
  nrow(dat), n_distinct(dat$model), n_distinct(dat$binom), n_distinct(dat$step)))
```

---

# Alignment Over Training

For each sampled checkpoint we compute, per corpus × model size:
- **Model alignment**: proportion of items where `sign(preference) == sign(constraint)`
- **Corpus RelFreq baseline**: proportion of items where `sign(freq_prob_c) == sign(constraint)`
- **Human baseline**: proportion of items where `sign(hum_pref − 0.5) == sign(constraint)` (constant)

Items where the constraint value is exactly zero are excluded (no directional prediction).

```{r compute-alignment}
message("Computing constraint alignment at each sampled checkpoint...")
t0 <- proc.time()

align_traj <- dat |>
  group_by(corpus, model_size, model, step, ck_idx, tokens) |>
  group_modify(~ compute_alignment(.x)) |>
  ungroup()

message(sprintf("  Done in %.1f s", (proc.time() - t0)[["elapsed"]]))
```

```{r summarise-alignment}
# Summarise across models within corpus × size × checkpoint
# (multiple models may share the same size × corpus combination)
align_summary <- align_traj |>
  group_by(corpus, model_size, ck_idx, tokens, constraint) |>
  summarise(
    model_align   = mean(model_align,   na.rm = TRUE),
    relfreq_align = mean(relfreq_align, na.rm = TRUE),
    human_align   = mean(human_align,   na.rm = TRUE),
    n             = sum(n),
    .groups       = "drop"
  ) |>
  # Constraint as ordered factor for consistent facet ordering
  mutate(constraint = factor(constraint, levels = CONSTRAINTS))
```

## Trajectory plots

Each panel shows one constraint. Solid coloured lines = model alignment rate (per size).
Dashed gold line = what you'd expect if the model used corpus RelFreq as its preference
(dynamic baseline that changes as the model sees more data).
Dotted purple line = human alignment rate (static upper reference).

```{r plot-align-traj, fig.height=14}
for (corp in c("BabyLM", "C4")) {
  d <- filter(align_summary, corpus == corp)
  if (nrow(d) == 0) next

  # One row per constraint × model_size with all three lines
  d_model   <- d
  d_relfreq <- d |> distinct(corpus, ck_idx, tokens, constraint, relfreq_align)
  d_human   <- d |> distinct(corpus, constraint, human_align)

  p <- ggplot(d_model, aes(x = tokens, y = model_align,
                            colour = model_size, group = model_size)) +
    # Human baseline (static horizontal dotted)
    geom_hline(data = d_human,
               aes(yintercept = human_align),
               colour = "#9e3d8e", linetype = "dotted", linewidth = 0.7,
               inherit.aes = FALSE) +
    # RelFreq baseline (dynamic dashed gold; same across model sizes so use any)
    geom_line(data = d_relfreq |>
                filter(!is.na(relfreq_align)) |>
                distinct(ck_idx, tokens, constraint, relfreq_align),
              aes(x = tokens, y = relfreq_align),
              colour = "#cc8800", linetype = "dashed", linewidth = 0.8,
              inherit.aes = FALSE) +
    # Model alignment trajectories
    geom_line(linewidth = 0.85) +
    geom_point(aes(shape = model_size), size = 1.8, stroke = 0.5) +
    geom_hline(yintercept = 0.5, colour = "grey60", linetype = "solid",
               linewidth = 0.35) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_shape_manual(
      values = c("125M" = 21, "350M" = 22, "1.3B" = 24),
      name   = "Model size"
    ) +
    scale_x_continuous("Training tokens",
                       labels = label_number(scale_cut = cut_short_scale())) +
    scale_y_continuous("Proportion following constraint",
                       limits = c(0, 1), breaks = seq(0, 1, 0.25),
                       labels = scales::percent_format(accuracy = 1)) +
    facet_wrap(~ constraint, ncol = 2) +
    labs(
      title    = sprintf("%s — constraint alignment over training", corp),
      subtitle = paste(
        "Solid coloured lines = model alignment per size.",
        "Gold dashed = corpus RelFreq baseline.",
        "Purple dotted = human preference baseline.",
        "Grey solid = 50% (chance)."
      )
    ) +
    theme_bw(base_size = 11) +
    theme(
      strip.text      = element_text(face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
  print(p)
  cat("\n\n")
}
```

---

# Final Checkpoint Summary

At the final training checkpoint, how does each model size's alignment compare to the
RelFreq baseline and human baseline, per constraint and corpus?

```{r final-align}
final_align <- align_traj |>
  group_by(corpus, model_size, model, constraint) |>
  filter(ck_idx == max(ck_idx)) |>
  ungroup() |>
  group_by(corpus, model_size, constraint) |>
  summarise(
    model_align   = mean(model_align,   na.rm = TRUE),
    relfreq_align = mean(relfreq_align, na.rm = TRUE),
    human_align   = mean(human_align,   na.rm = TRUE),
    .groups       = "drop"
  ) |>
  mutate(
    above_relfreq = model_align > relfreq_align,
    constraint    = factor(constraint, levels = CONSTRAINTS)
  )
```

```{r plot-final-bar, fig.height=8}
for (corp in c("BabyLM", "C4")) {
  d <- filter(final_align, corpus == corp)
  if (nrow(d) == 0) next

  d_human <- d |> distinct(constraint, human_align)

  p <- ggplot(d, aes(x = constraint, y = model_align, fill = model_size)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65) +
    # RelFreq baseline: one point per constraint (same across model sizes)
    geom_point(
      data = d |> distinct(constraint, relfreq_align),
      aes(x = constraint, y = relfreq_align),
      inherit.aes = FALSE,
      shape = 18, size = 4, colour = "#cc8800"
    ) +
    # Human baseline
    geom_point(
      data = d_human,
      aes(x = constraint, y = human_align),
      inherit.aes = FALSE,
      shape = 17, size = 3.5, colour = "#9e3d8e"
    ) +
    geom_hline(yintercept = 0.5, linetype = "dashed",
               colour = "grey50", linewidth = 0.5) +
    scale_fill_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_y_continuous("Proportion following constraint",
                       limits = c(0, 1), breaks = seq(0, 1, 0.25),
                       labels = scales::percent_format(accuracy = 1)) +
    labs(
      title    = sprintf("%s — constraint alignment at final checkpoint", corp),
      subtitle = paste(
        "Bars = model alignment per size.",
        "Gold diamond = corpus RelFreq baseline.",
        "Purple triangle = human preference baseline.",
        "Dashed = 50% chance."
      ),
      x = NULL
    ) +
    theme_bw(base_size = 11) +
    theme(
      axis.text.x     = element_text(angle = 30, hjust = 1, face = "bold"),
      legend.position = "bottom"
    )
  print(p)
  cat("\n\n")
}
```

## Table: final checkpoint alignment rates

```{r final-table}
final_align |>
  mutate(across(c(model_align, relfreq_align, human_align), \(x) round(x, 3)),
         above_relfreq = if_else(above_relfreq, "Yes", "No")) |>
  rename(
    Corpus        = corpus,
    Size          = model_size,
    Constraint    = constraint,
    `Model align` = model_align,
    `RelFreq base` = relfreq_align,
    `Human base`  = human_align,
    `Above RelFreq?` = above_relfreq
  ) |>
  kbl(caption = paste(
    "Constraint alignment at final training checkpoint.",
    "Model align = proportion of items where model preference direction matches constraint.",
    "RelFreq base = same proportion using corpus RelFreq as the preference.",
    "Human base = same using human preference data."
  )) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE, font_size = 11) |>
  collapse_rows(columns = 1:2, valign = "middle")
```

---

# Model Alignment vs. RelFreq Baseline: Gap Over Training

How much does each model *exceed* (or fall below) the RelFreq baseline for each
constraint, and does this gap grow over training?

`gap = model_align − relfreq_align`

Positive gap means the model follows this constraint more than corpus frequency alone
would predict.

```{r plot-gap-traj, fig.height=14}
gap_summary <- align_summary |>
  mutate(gap = model_align - relfreq_align)

for (corp in c("BabyLM", "C4")) {
  d <- filter(gap_summary, corpus == corp)
  if (nrow(d) == 0) next

  p <- ggplot(d, aes(x = tokens, y = gap,
                      colour = model_size, group = model_size)) +
    geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.5) +
    geom_line(linewidth = 0.85) +
    geom_point(aes(shape = model_size), size = 1.8, stroke = 0.5) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_shape_manual(
      values = c("125M" = 21, "350M" = 22, "1.3B" = 24),
      name   = "Model size"
    ) +
    scale_x_continuous("Training tokens",
                       labels = label_number(scale_cut = cut_short_scale())) +
    scale_y_continuous("Alignment gap (model − RelFreq baseline)",
                       labels = scales::percent_format(accuracy = 1)) +
    facet_wrap(~ constraint, ncol = 2) +
    labs(
      title    = sprintf("%s — gap between model alignment and RelFreq baseline", corp),
      subtitle = "Positive = model follows constraint more than corpus frequency alone predicts. Zero line = no difference from RelFreq."
    ) +
    theme_bw(base_size = 11) +
    theme(
      strip.text       = element_text(face = "bold"),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
  print(p)
  cat("\n\n")
}
```

---

# Alignment by Corpus Exposure

Does constraint alignment depend on how many times the model has seen a given
binomial in training? A binomial seen 1,000 times gives the model much stronger
evidence than one seen 5 times. If models learn constraints from repeated exposure,
we expect alignment to be higher for high-exposure binomials — especially for
constraints that go *beyond* corpus frequency.

We use `log(α_seen + β_seen)` at each checkpoint as the exposure measure.

## Setup: exposure classification

Classify each binomial as **low / mid / high** exposure based on tertiles of
`log_total` at the **final checkpoint**, computed within corpus. This gives a
stable label that reflects a binomial's overall presence in the training corpus.

```{r exposure-classification}
# Final-checkpoint exposure per model × corpus × binom
final_exposure <- dat |>
  group_by(corpus, model_size, model, binom) |>
  filter(ck_idx == max(ck_idx)) |>
  ungroup() |>
  select(corpus, model_size, model, binom, log_total) |>
  filter(!is.na(log_total))

# Tertile cutpoints within corpus (pooled across model sizes — corpus is shared)
exposure_cuts <- final_exposure |>
  group_by(corpus) |>
  summarise(
    q33 = quantile(log_total, 1/3, na.rm = TRUE),
    q67 = quantile(log_total, 2/3, na.rm = TRUE),
    .groups = "drop"
  )

final_exposure <- final_exposure |>
  left_join(exposure_cuts, by = "corpus") |>
  mutate(
    exposure_grp = case_when(
      log_total <= q33 ~ "Low exposure",
      log_total <= q67 ~ "Mid exposure",
      TRUE             ~ "High exposure"
    ),
    exposure_grp = factor(exposure_grp,
                          levels = c("Low exposure", "Mid exposure", "High exposure"))
  ) |>
  select(corpus, model_size, model, binom, log_total, exposure_grp)

# Print tertile boundaries
exposure_cuts |>
  mutate(across(c(q33, q67), \(x) round(x, 2))) |>
  rename(Corpus = corpus,
         `log_total at 33rd pct` = q33,
         `log_total at 67th pct` = q67) |>
  kbl(caption = "Exposure tertile boundaries (log(α_seen + β_seen) at final checkpoint, pooled across model sizes).") |>
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## Alignment at final checkpoint by exposure group

For each constraint, compare alignment across the three exposure groups.
If models learn a constraint from repeated exposure, alignment should increase
from low → mid → high. If alignment is already high at low exposure, the model
may be using a generalised rule rather than item-specific memory.

```{r align-by-exposure-final}
# Join exposure classification onto the full dataset, then filter to final checkpoint
dat_with_exp <- dat |>
  left_join(final_exposure |> select(corpus, model_size, model, binom, exposure_grp),
            by = c("corpus", "model_size", "model", "binom")) |>
  filter(!is.na(exposure_grp))

align_exp_final <- dat_with_exp |>
  group_by(corpus, model_size, model, exposure_grp) |>
  filter(ck_idx == max(ck_idx)) |>
  group_modify(~ compute_alignment(.x)) |>
  ungroup() |>
  group_by(corpus, model_size, exposure_grp, constraint) |>
  summarise(
    model_align   = mean(model_align,   na.rm = TRUE),
    relfreq_align = mean(relfreq_align, na.rm = TRUE),
    human_align   = mean(human_align,   na.rm = TRUE),
    n_valid       = sum(n,              na.rm = TRUE),
    .groups       = "drop"
  ) |>
  # NaN arises when all constituent values were NA (n_valid == 0); replace with NA
  mutate(
    model_align   = if_else(n_valid == 0, NA_real_, model_align),
    relfreq_align = if_else(n_valid == 0, NA_real_, relfreq_align),
    human_align   = if_else(n_valid == 0, NA_real_, human_align),
    constraint    = factor(constraint, levels = CONSTRAINTS)
  )
```

```{r plot-align-by-exposure, fig.height=10}
for (corp in c("BabyLM", "C4")) {
  d <- filter(align_exp_final, corpus == corp) |>
    complete(exposure_grp, model_size, constraint,
             fill = list(n_valid = 0L, model_align = NA_real_,
                         relfreq_align = NA_real_, human_align = NA_real_)) |>
    mutate(has_data = !is.na(model_align) & coalesce(n_valid, 0L) > 0)
  if (nrow(d) == 0) next

  d_relfreq <- d |> filter(has_data) |> distinct(constraint, exposure_grp, relfreq_align)
  d_human   <- d |> filter(has_data) |> distinct(constraint, human_align)
  d_nodata  <- d |> filter(!has_data)

  p <- ggplot(d, aes(x = exposure_grp, y = model_align, fill = model_size)) +
    # No-data placeholder: grey outline bar
    geom_col(data = d_nodata,
             aes(x = exposure_grp, y = 0.01),
             position = position_dodge(width = 0.75), width = 0.65,
             fill = NA, colour = "grey70", linetype = "dashed",
             inherit.aes = FALSE) +
    geom_text(data = d_nodata,
              aes(x = exposure_grp, y = 0.005, group = model_size, label = "n=0"),
              position = position_dodge(width = 0.75),
              size = 2.2, colour = "grey60", inherit.aes = FALSE) +
    geom_col(data = filter(d, has_data),
             position = position_dodge(width = 0.75), width = 0.65, alpha = 0.9) +
    # RelFreq baseline per exposure group
    geom_point(
      data = d_relfreq,
      aes(x = exposure_grp, y = relfreq_align),
      inherit.aes = FALSE,
      shape = 18, size = 3.5, colour = "#cc8800"
    ) +
    # Human baseline (constant across exposure groups)
    geom_hline(data = d_human,
               aes(yintercept = human_align),
               colour = "#9e3d8e", linetype = "dotted", linewidth = 0.7,
               inherit.aes = FALSE) +
    geom_hline(yintercept = 0.5, linetype = "dashed",
               colour = "grey50", linewidth = 0.4) +
    scale_fill_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_y_continuous("Proportion following constraint",
                       limits = c(0, 1), breaks = seq(0, 1, 0.25),
                       labels = scales::percent_format(accuracy = 1)) +
    facet_wrap(~ constraint, ncol = 2) +
    labs(
      title    = sprintf("%s — constraint alignment by corpus exposure (final checkpoint)", corp),
      subtitle = paste(
        "Bars = model alignment per size.",
        "Gold diamond = corpus RelFreq baseline (varies by exposure group).",
        "Purple dotted = human baseline. Dashed = 50% chance."
      ),
      x = NULL
    ) +
    theme_bw(base_size = 11) +
    theme(
      strip.text       = element_text(face = "bold"),
      legend.position  = "bottom",
      panel.grid.minor = element_blank(),
      axis.text.x      = element_text(angle = 20, hjust = 1)
    )
  print(p)
  cat("\n\n")
}
```

## Alignment trajectories by exposure group

Does the model reach high constraint alignment earlier (in terms of training tokens)
for high-exposure binomials? Here we split the trajectory by exposure group and
show how model alignment evolves over training for each group separately.

```{r align-traj-by-exposure}
align_traj_exp <- dat_with_exp |>
  group_by(corpus, model_size, model, ck_idx, tokens, exposure_grp) |>
  group_modify(~ compute_alignment(.x)) |>
  ungroup() |>
  group_by(corpus, model_size, ck_idx, tokens, exposure_grp, constraint) |>
  summarise(
    model_align   = mean(model_align,   na.rm = TRUE),
    relfreq_align = mean(relfreq_align, na.rm = TRUE),
    n_valid       = sum(n,              na.rm = TRUE),
    .groups       = "drop"
  ) |>
  mutate(
    model_align   = if_else(n_valid == 0, NA_real_, model_align),
    relfreq_align = if_else(n_valid == 0, NA_real_, relfreq_align),
    constraint    = factor(constraint, levels = CONSTRAINTS)
  )
```

```{r plot-traj-by-exposure, fig.height=16}
# Facet: constraint (rows) × exposure_grp (columns)
# Lines: model_size (coloured); RelFreq baseline (gold dashed)
for (corp in c("BabyLM", "C4")) {
  d <- filter(align_traj_exp, corpus == corp)
  if (nrow(d) == 0) next

  d_relfreq <- d |>
    distinct(ck_idx, tokens, constraint, exposure_grp, relfreq_align)

  p <- ggplot(d, aes(x = tokens, y = model_align,
                      colour = model_size, group = model_size)) +
    geom_hline(yintercept = 0.5, colour = "grey60", linewidth = 0.35) +
    geom_line(data = d_relfreq,
              aes(x = tokens, y = relfreq_align),
              colour = "#cc8800", linetype = "dashed", linewidth = 0.7,
              inherit.aes = FALSE) +
    geom_line(linewidth = 0.75) +
    geom_point(aes(shape = model_size), size = 1.5, stroke = 0.5) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_shape_manual(
      values = c("125M" = 21, "350M" = 22, "1.3B" = 24),
      name   = "Model size"
    ) +
    scale_x_continuous("Training tokens",
                       labels = label_number(scale_cut = cut_short_scale())) +
    scale_y_continuous("Proportion following constraint",
                       limits = c(0, 1), breaks = c(0.25, 0.5, 0.75, 1),
                       labels = scales::percent_format(accuracy = 1)) +
    facet_grid(constraint ~ exposure_grp) +
    labs(
      title    = sprintf("%s — constraint alignment over training, by exposure group", corp),
      subtitle = "Exposure groups based on final-checkpoint log(α_seen + β_seen) tertiles within corpus. Gold dashed = RelFreq baseline."
    ) +
    theme_bw(base_size = 10) +
    theme(
      strip.text       = element_text(face = "bold", size = 9),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
  print(p)
  cat("\n\n")
}
```

## Gap by exposure group at sampled training checkpoints

Same bar chart as above but at four evenly-spaced checkpoints across training rather
than only the final checkpoint. No averaging is done — each panel shows the gap at
exactly that checkpoint.

```{r plot-gap-exposure-stages, fig.height=12}
# Sample 4 checkpoints per model × corpus nearest to 25/50/75/100% of total tokens
stage_fracs  <- c(0.25, 0.50, 0.75, 1.00)
stage_labels <- c("25%", "50%", "75%", "100%")

ck_stage_lookup <- align_traj_exp |>
  distinct(corpus, model_size, ck_idx, tokens) |>
  group_by(corpus, model_size) |>
  mutate(max_tokens = max(tokens)) |>
  ungroup() |>
  crossing(stage_frac = stage_fracs) |>
  mutate(dist = abs(tokens - stage_frac * max_tokens)) |>
  group_by(corpus, model_size, stage_frac) |>
  slice_min(dist, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(stage_label = stage_labels[match(stage_frac, stage_fracs)])

# Median token count per corpus × stage for display labels (models may differ slightly)
stage_med_tok <- ck_stage_lookup |>
  group_by(corpus, stage_frac) |>
  summarise(med_tok = median(tokens), .groups = "drop") |>
  mutate(
    stage_label     = stage_labels[match(stage_frac, stage_fracs)],
    stage_tok_label = sprintf("%s\n(%s tokens)", stage_label,
                              scales::label_number(scale_cut = scales::cut_short_scale())(med_tok))
  )

gap_stages <- align_traj_exp |>
  inner_join(select(ck_stage_lookup, corpus, model_size, ck_idx, stage_label, stage_frac),
             by = c("corpus", "model_size", "ck_idx")) |>
  left_join(select(stage_med_tok, corpus, stage_frac, stage_tok_label),
            by = c("corpus", "stage_frac")) |>
  mutate(
    stage_tok_label = fct_reorder(stage_tok_label, stage_frac),
    gap             = model_align - relfreq_align,
    constraint      = factor(constraint, levels = CONSTRAINTS)
  )

for (corp in c("BabyLM", "C4")) {
  d <- filter(gap_stages, corpus == corp) |>
    mutate(stage_tok_label = droplevels(stage_tok_label)) |>
    complete(exposure_grp, model_size, constraint, stage_tok_label,
             fill = list(n_valid = 0L, gap = NA_real_)) |>
    mutate(has_data = !is.na(gap) & coalesce(n_valid, 0L) > 0)
  if (nrow(d) == 0) next

  d_nodata <- filter(d, !has_data)

  p <- ggplot(d, aes(x = exposure_grp, y = gap, fill = model_size)) +
    geom_col(data = d_nodata,
             aes(x = exposure_grp, y = 0.005),
             position = position_dodge(width = 0.75), width = 0.65,
             fill = NA, colour = "grey70", linetype = "dashed",
             inherit.aes = FALSE) +
    geom_text(data = d_nodata,
              aes(x = exposure_grp, y = 0.003, group = model_size, label = "n=0"),
              position = position_dodge(width = 0.75),
              size = 1.8, colour = "grey60", inherit.aes = FALSE) +
    geom_col(data = filter(d, has_data),
             position = position_dodge(width = 0.75), width = 0.65, alpha = 0.9) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.5) +
    scale_fill_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_y_continuous("Alignment gap (model \u2212 RelFreq baseline)",
                       labels = scales::percent_format(accuracy = 1)) +
    facet_grid(constraint ~ stage_tok_label) +
    labs(
      title    = sprintf("%s \u2014 alignment gap by exposure group at sampled checkpoints", corp),
      subtitle = "Checkpoints nearest to 25/50/75/100% of total training tokens per model. No averaging.",
      x        = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      strip.text       = element_text(face = "bold", size = 9),
      legend.position  = "bottom",
      panel.grid.minor = element_blank(),
      axis.text.x      = element_text(angle = 20, hjust = 1)
    )
  print(p)
  cat("\n\n")
}
```

## Gap trajectory by exposure group (constraint × model size grid)

Facet grid: constraint (rows) × model size (columns). Lines show how the gap
between model alignment and the RelFreq baseline evolves over training, separately
for each exposure group. Diverging lines indicate that how much of a binomial the
model has seen affects whether it learns the constraint beyond corpus frequency.

```{r plot-gap-traj-by-exposure, fig.height=22}
gap_traj_exp <- align_traj_exp |>
  mutate(gap = model_align - relfreq_align)

exp_colours <- c(
  "Low exposure"  = "#74add1",
  "Mid exposure"  = "#f46d43",
  "High exposure" = "#1a9850"
)

for (corp in c("BabyLM", "C4")) {
  d <- filter(gap_traj_exp, corpus == corp)
  if (nrow(d) == 0) next

  p <- ggplot(d, aes(x = tokens, y = gap,
                      colour = exposure_grp, group = exposure_grp)) +
    geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.5) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.2, stroke = 0.4) +
    scale_colour_manual(values = exp_colours, name = "Exposure group") +
    scale_x_continuous("Training tokens",
                       labels = label_number(scale_cut = cut_short_scale())) +
    scale_y_continuous("Gap (model align \u2212 RelFreq baseline)",
                       labels = scales::percent_format(accuracy = 1)) +
    facet_grid(constraint ~ model_size) +
    labs(
      title    = sprintf("%s \u2014 alignment gap over training by exposure group", corp),
      subtitle = "Each panel = one constraint \u00d7 model size. Lines = exposure tertiles. Zero = model tracks corpus frequency exactly."
    ) +
    theme_bw(base_size = 10) +
    theme(
      strip.text       = element_text(face = "bold", size = 9),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
  print(p)
  cat("\n\n")
}
```

## Gap (model − RelFreq baseline) by exposure group at final checkpoint

A positive gap means the model follows the constraint more than corpus frequency
predicts. Does this gap grow with exposure, suggesting the model has extracted a
generalizable rule from repeated examples?

```{r plot-gap-by-exposure, fig.height=10}
gap_exp_final <- align_exp_final |>
  mutate(gap = model_align - relfreq_align)

for (corp in c("BabyLM", "C4")) {
  d <- filter(gap_exp_final, corpus == corp) |>
    complete(exposure_grp, model_size, constraint,
             fill = list(n_valid = 0L, gap = NA_real_)) |>
    mutate(has_data = !is.na(gap) & coalesce(n_valid, 0L) > 0)
  if (nrow(d) == 0) next

  d_nodata <- filter(d, !has_data)

  p <- ggplot(d, aes(x = exposure_grp, y = gap, fill = model_size)) +
    geom_col(data = d_nodata,
             aes(x = exposure_grp, y = 0.005),
             position = position_dodge(width = 0.75), width = 0.65,
             fill = NA, colour = "grey70", linetype = "dashed",
             inherit.aes = FALSE) +
    geom_text(data = d_nodata,
              aes(x = exposure_grp, y = 0.003, group = model_size, label = "n=0"),
              position = position_dodge(width = 0.75),
              size = 2.2, colour = "grey60", inherit.aes = FALSE) +
    geom_col(data = filter(d, has_data),
             position = position_dodge(width = 0.75), width = 0.65, alpha = 0.9) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.5) +
    scale_fill_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_y_continuous("Alignment gap (model − RelFreq baseline)",
                       labels = scales::percent_format(accuracy = 1)) +
    facet_wrap(~ constraint, ncol = 2) +
    labs(
      title    = sprintf("%s — alignment gap by exposure group (final checkpoint)", corp),
      subtitle = "Positive = model follows constraint more than corpus frequency predicts. Zero = no benefit beyond RelFreq.",
      x        = NULL
    ) +
    theme_bw(base_size = 11) +
    theme(
      strip.text       = element_text(face = "bold"),
      legend.position  = "bottom",
      panel.grid.minor = element_blank(),
      axis.text.x      = element_text(angle = 20, hjust = 1)
    )
  print(p)
  cat("\n\n")
}
```

---

# Alternative Gap × Exposure Visualizations

## Option 1: Median split (high vs. low exposure)

Collapsing to two groups removes the ambiguous middle tertile and produces a
cleaner contrast: binomials the model has seen more than average vs. less than
average. Each panel = one constraint × model size; two lines per panel.

```{r binary-exposure-setup}
binary_cuts <- final_exposure |>
  group_by(corpus) |>
  summarise(med = median(log_total, na.rm = TRUE), .groups = "drop")

exposure_binary_labels <- final_exposure |>
  left_join(binary_cuts, by = "corpus") |>
  mutate(
    exposure_binary = factor(
      if_else(log_total > med, "High exposure", "Low exposure"),
      levels = c("Low exposure", "High exposure")
    )
  ) |>
  select(corpus, model_size, model, binom, exposure_binary)

dat_with_binary <- dat_with_exp |>
  left_join(exposure_binary_labels,
            by = c("corpus", "model_size", "model", "binom"))

align_traj_binary <- dat_with_binary |>
  filter(!is.na(exposure_binary)) |>
  group_by(corpus, model_size, model, ck_idx, tokens, exposure_binary) |>
  group_modify(~ compute_alignment(.x)) |>
  ungroup() |>
  group_by(corpus, model_size, ck_idx, tokens, exposure_binary, constraint) |>
  summarise(
    model_align   = mean(model_align,   na.rm = TRUE),
    relfreq_align = mean(relfreq_align, na.rm = TRUE),
    .groups       = "drop"
  ) |>
  mutate(
    gap        = model_align - relfreq_align,
    constraint = factor(constraint, levels = CONSTRAINTS)
  )
```

```{r plot-gap-traj-binary, fig.height=22}
binary_colours <- c("Low exposure" = "#74add1", "High exposure" = "#1a9850")

for (corp in c("BabyLM", "C4")) {
  d <- filter(align_traj_binary, corpus == corp)
  if (nrow(d) == 0) next

  p <- ggplot(d, aes(x = tokens, y = gap,
                      colour = exposure_binary, group = exposure_binary)) +
    geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.5) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 1.3, stroke = 0.4) +
    scale_colour_manual(values = binary_colours, name = "Exposure") +
    scale_x_continuous("Training tokens",
                       labels = label_number(scale_cut = cut_short_scale())) +
    scale_y_continuous("Gap (model align \u2212 RelFreq baseline)",
                       labels = scales::percent_format(accuracy = 1)) +
    facet_grid(constraint ~ model_size) +
    labs(
      title    = sprintf("%s \u2014 alignment gap over training (median split)", corp),
      subtitle = "High = above-median log(total exposure) at final checkpoint. Zero = no difference from corpus frequency."
    ) +
    theme_bw(base_size = 10) +
    theme(
      strip.text       = element_text(face = "bold", size = 9),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
  print(p)
  cat("\n\n")
}
```

## Option 2: Separate plots per exposure group

Each plot shows one exposure group across all constraints (rows) and model sizes
(columns), coloured by model size. Comparing across the three plots shows how
the gap trajectory shifts with exposure without cluttering any single panel.

```{r plot-gap-traj-per-expgrp, fig.height=22}
for (corp in c("BabyLM", "C4")) {
  for (exp_grp in levels(gap_traj_exp$exposure_grp)) {
    d <- filter(gap_traj_exp, corpus == corp, exposure_grp == exp_grp)
    if (nrow(d) == 0) next

    p <- ggplot(d, aes(x = tokens, y = gap,
                        colour = model_size, group = model_size)) +
      geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.5) +
      geom_line(linewidth = 0.9) +
      geom_point(aes(shape = model_size), size = 1.5, stroke = 0.4) +
      scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
      scale_shape_manual(
        values = c("125M" = 21, "350M" = 22, "1.3B" = 24),
        name   = "Model size"
      ) +
      scale_x_continuous("Training tokens",
                         labels = label_number(scale_cut = cut_short_scale())) +
      scale_y_continuous("Gap (model align \u2212 RelFreq baseline)",
                         labels = scales::percent_format(accuracy = 1)) +
      facet_grid(constraint ~ model_size) +
      labs(
        title    = sprintf("%s \u2014 %s", corp, exp_grp),
        subtitle = "Alignment gap over training. Zero = model tracks corpus frequency exactly."
      ) +
      theme_bw(base_size = 10) +
      theme(
        strip.text       = element_text(face = "bold", size = 9),
        legend.position  = "none",
        panel.grid.minor = element_blank()
      )
    print(p)
    cat("\n\n")
  }
}
```

## Option 3: High − low exposure contrast

Subtract the low-exposure gap from the high-exposure gap at each checkpoint.
Positive = model exceeds the RelFreq baseline more for frequently-seen binomials
(exposure-driven learning). Near zero = model generalizes equally regardless of
how often it has seen the item.

```{r gap-contrast-compute}
gap_contrast <- gap_traj_exp |>
  filter(exposure_grp %in% c("Low exposure", "High exposure")) |>
  select(corpus, model_size, ck_idx, tokens, constraint, exposure_grp, gap) |>
  pivot_wider(names_from  = exposure_grp,
              values_from = gap) |>
  rename(gap_high = `High exposure`, gap_low = `Low exposure`) |>
  mutate(
    contrast   = gap_high - gap_low,
    constraint = factor(constraint, levels = CONSTRAINTS)
  )
```

```{r plot-gap-contrast, fig.height=14}
for (corp in c("BabyLM", "C4")) {
  d <- filter(gap_contrast, corpus == corp)
  if (nrow(d) == 0) next

  p <- ggplot(d, aes(x = tokens, y = contrast,
                      colour = model_size, group = model_size)) +
    geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.5) +
    geom_line(linewidth = 0.9) +
    geom_point(aes(shape = model_size), size = 1.8, stroke = 0.5) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_shape_manual(
      values = c("125M" = 21, "350M" = 22, "1.3B" = 24),
      name   = "Model size"
    ) +
    scale_x_continuous("Training tokens",
                       labels = label_number(scale_cut = cut_short_scale())) +
    scale_y_continuous("High \u2212 Low gap contrast",
                       labels = scales::percent_format(accuracy = 1)) +
    facet_wrap(~ constraint, ncol = 2) +
    labs(
      title    = sprintf("%s \u2014 exposure contrast: high \u2212 low alignment gap", corp),
      subtitle = "Positive = model exceeds RelFreq baseline more for high-exposure items (exposure-driven). Near zero = generalizes regardless of exposure."
    ) +
    theme_bw(base_size = 11) +
    theme(
      strip.text       = element_text(face = "bold"),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
  print(p)
  cat("\n\n")
}
```

---

# Gap as a Continuous Function of Exposure

For each binomial at the final checkpoint, we compute a per-item gap:
`+1` if the model follows the constraint but RelFreq would not,
`−1` if RelFreq follows it but the model does not, and
`0` if both agree or both disagree.
A GAM smooth over log-total exposure reveals whether the gap grows, shrinks, or is
flat as a function of how many times the model has seen the binomial.

```{r item-gap-setup}
MIN_GAM_N <- 10L   # GAM s(x, bs="cs") needs at least this many points to fit

dat_final_item <- dat |>
  group_by(model) |>
  filter(ck_idx == max(ck_idx)) |>
  ungroup() |>
  filter(!is.na(log_total), is.finite(log_total))

item_gap_long <- dat_final_item |>
  pivot_longer(all_of(CONSTRAINTS), names_to = "constraint", values_to = "c_val") |>
  filter(!is.na(c_val), c_val != 0,
         !is.na(preference), !is.na(freq_prob_c)) |>
  mutate(
    model_aligns   = as.numeric(sign(preference)  == sign(c_val)),
    relfreq_aligns = as.numeric(sign(freq_prob_c) == sign(c_val)),
    gap            = model_aligns - relfreq_aligns,
    constraint     = factor(constraint, levels = CONSTRAINTS)
  )

# Item counts per corpus × model_size × constraint
item_gap_n <- item_gap_long |> count(corpus, model_size, constraint)

# Report and remove combinations too sparse for GAM
item_gap_dropped <- filter(item_gap_n, n < MIN_GAM_N)
if (nrow(item_gap_dropped) > 0) {
  message(sprintf(
    "GAM smooth omitted (n < %d) for: %s",
    MIN_GAM_N,
    paste(sprintf("%s / %s / %s (n=%d)",
                  item_gap_dropped$corpus, item_gap_dropped$model_size,
                  item_gap_dropped$constraint, item_gap_dropped$n),
          collapse = "; ")
  ))
}
item_gap_long <- semi_join(item_gap_long, filter(item_gap_n, n >= MIN_GAM_N),
                           by = c("corpus", "model_size", "constraint"))

# Diagnostic: item counts per corpus × model_size × constraint
item_gap_n |>
  pivot_wider(names_from = constraint, values_from = n, values_fill = 0L) |>
  knitr::kable(caption = "Item counts per corpus × model size × constraint (after filtering zero/NA values)")
```

```{r plot-gap-smooth-checkpoints, fig.height=22}
# Target token counts for checkpoint columns (nearest available checkpoint per model)
TOK_TARGETS <- c(1e9, 5e9, 10e9)
tok_target_labels <- c("~1B tokens", "~5B tokens", "~10B tokens")

ck_tok_lookup <- dat |>
  distinct(corpus, model_size, model, ck_idx, tokens) |>
  crossing(tok_target = TOK_TARGETS) |>
  mutate(dist = abs(tokens - tok_target)) |>
  group_by(corpus, model_size, model, tok_target) |>
  slice_min(dist, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(tok_label = factor(
    tok_target_labels[match(tok_target, TOK_TARGETS)],
    levels = tok_target_labels
  ))

item_gap_stages <- dat |>
  filter(!is.na(log_total), is.finite(log_total)) |>
  inner_join(select(ck_tok_lookup, corpus, model_size, model, ck_idx, tok_target, tok_label),
             by = c("corpus", "model_size", "model", "ck_idx")) |>
  pivot_longer(all_of(CONSTRAINTS), names_to = "constraint", values_to = "c_val") |>
  filter(!is.na(c_val), c_val != 0,
         !is.na(preference), !is.na(freq_prob_c)) |>
  mutate(
    model_aligns   = as.numeric(sign(preference)  == sign(c_val)),
    relfreq_aligns = as.numeric(sign(freq_prob_c) == sign(c_val)),
    gap            = model_aligns - relfreq_aligns,
    constraint     = factor(constraint, levels = CONSTRAINTS)
  )

# Report and remove combinations too sparse for GAM (per stage)
item_gap_stages_n <- item_gap_stages |>
  count(corpus, model_size, constraint, tok_target)

item_gap_stages_dropped <- filter(item_gap_stages_n, n < MIN_GAM_N)
if (nrow(item_gap_stages_dropped) > 0) {
  message(sprintf(
    "GAM smooth omitted at checkpoint (n < %d) for: %s",
    MIN_GAM_N,
    paste(sprintf("%s / %s / %s / %s (n=%d)",
                  item_gap_stages_dropped$corpus, item_gap_stages_dropped$model_size,
                  item_gap_stages_dropped$constraint,
                  tok_target_labels[match(item_gap_stages_dropped$tok_target, TOK_TARGETS)],
                  item_gap_stages_dropped$n),
          collapse = "; ")
  ))
}
item_gap_stages <- semi_join(item_gap_stages,
                             filter(item_gap_stages_n, n >= MIN_GAM_N),
                             by = c("corpus", "model_size", "constraint", "tok_target"))

for (corp in c("BabyLM", "C4")) {
  d <- filter(item_gap_stages, corpus == corp)
  if (nrow(d) == 0) next

  p <- ggplot(d, aes(x = log_total, y = gap, colour = model_size, fill = model_size)) +
    geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
    geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"),
                alpha = 0.15, linewidth = 0.9) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_fill_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous("log(tokens seen)", breaks = scales::pretty_breaks(4)) +
    scale_y_continuous("Alignment gap (model \u2212 RelFreq)",
                       limits = c(-1, 1),
                       breaks = c(-1, -0.5, 0, 0.5, 1),
                       labels = scales::percent_format(accuracy = 1)) +
    facet_grid(constraint ~ tok_label, scales = "free_x") +
    labs(
      title    = sprintf("%s \u2014 alignment gap as a function of exposure at sampled checkpoints", corp),
      subtitle = "GAM smooth with 95% CI. Columns = checkpoint nearest to 1B / 5B / 10B training tokens."
    ) +
    theme_bw(base_size = 10) +
    theme(
      strip.text       = element_text(face = "bold", size = 9),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
  print(p)
  cat("\n\n")
}
```

```{r plot-gap-smooth, fig.height=22}
for (corp in c("BabyLM", "C4")) {
  d <- filter(item_gap_long, corpus == corp)
  if (nrow(d) == 0) next

  p <- ggplot(d, aes(x = log_total, y = gap, colour = model_size, fill = model_size)) +
    geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.4) +
    geom_smooth(method = "gam", formula = y ~ s(x, bs = "cs"),
                alpha = 0.15, linewidth = 0.9) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_fill_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous("log(tokens seen)", breaks = scales::pretty_breaks(4)) +
    scale_y_continuous("Alignment gap (model \u2212 RelFreq)",
                       limits = c(-1, 1),
                       breaks = c(-1, -0.5, 0, 0.5, 1),
                       labels = scales::percent_format(accuracy = 1)) +
    facet_grid(constraint ~ model_size) +
    labs(
      title    = sprintf("%s \u2014 alignment gap as a function of corpus exposure (final checkpoint)", corp),
      subtitle = paste(
        "GAM smooth with 95% CI.",
        "x = log total times the binomial appeared in the training corpus at the final checkpoint.",
        "Positive gap = model follows constraint more than corpus frequency predicts."
      )
    ) +
    theme_bw(base_size = 10) +
    theme(
      strip.text       = element_text(face = "bold", size = 9),
      legend.position  = "bottom",
      panel.grid.minor = element_blank()
    )
  print(p)
  cat("\n\n")
}
```
