---
title: "Model-Human Alignment: Correlation, Accuracy, and Constraint Analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(knitr)
```

## Overview

This analysis examines three complementary perspectives on how language models develop
binomial ordering preferences during training:

1. **Correlation**: Pearson r between model preference and human preference at each checkpoint
2. **Accuracy**: Proportion of binomials where model and human agree on ordering direction (binary match)
3. **Constraint sensitivity**: How model preferences align with individual linguistic constraints
   (Form, Percept, Culture, etc.) and the composite GenPref measure

All analyses are conducted separately for **nonce** (novel) and **attested** (corpus-found) binomials.

---

## Load and Prepare Data

```{r load-data}
# Load pre-processed data. Run prepare_data.py first if any file is missing.
# Constraint and GenPref regressions are fit below in R (see Analysis 3 & 4).
processed_path <- "../Data/processed"

csv_names <- c(
  "nonce_correlations", "attested_correlations",
  "nonce_accuracy",     "attested_accuracy",
  "final_nonce",        "final_attested",
  "training_nonce",     "training_attested"
)

missing <- csv_names[!file.exists(file.path(processed_path, paste0(csv_names, ".csv")))]
if (length(missing) > 0) {
  stop(sprintf(
    "Missing processed files: %s\nRun prepare_data.py first.",
    paste(missing, collapse = ", ")
  ))
}

for (nm in csv_names) {
  assign(nm, read_csv(file.path(processed_path, paste0(nm, ".csv")),
                      show_col_types = FALSE))
}

cat(sprintf("Nonce correlations:    %d rows\n",  nrow(nonce_correlations)))
cat(sprintf("Attested correlations: %d rows\n",  nrow(attested_correlations)))
cat(sprintf("Final nonce data:      %d rows\n",  nrow(final_nonce)))
cat(sprintf("Final attested data:   %d rows\n",  nrow(final_attested)))
cat(sprintf("Training nonce:        %d rows (%d checkpoints)\n",
            nrow(training_nonce),    n_distinct(training_nonce$checkpoint)))
cat(sprintf("Training attested:     %d rows (%d checkpoints)\n",
            nrow(training_attested), n_distinct(training_attested$checkpoint)))
```

```{r compute-constraints-genpref}
# Fit OLS constraint regressions and GenPref regressions in R at the
# log-sampled checkpoints (stored in training_nonce / training_attested).

CONSTRAINT_COLS <- c("Form", "Percept", "Culture", "Power", "Intense",
                     "Icon", "Freq", "Len", "Lapse", "BStress")

FREQ_COVARIATES <- c("OverallFreq_scaled", "RelFreq_centered")

# Nonce: preference ~ constraints only (matches brms Model B for nonce)
fit_constraint_models_nonce_R <- function(data) {
  data %>%
    group_by(model, checkpoint, step, tokens) %>%
    group_map(function(g, keys) {
      complete <- g %>%
        select(preference, all_of(CONSTRAINT_COLS)) %>%
        drop_na()
      if (nrow(complete) < 15) return(NULL)
      fit <- lm(reformulate(CONSTRAINT_COLS, "preference"), data = complete)
      cf  <- coef(summary(fit))
      tibble(
        model      = keys$model,
        checkpoint = keys$checkpoint,
        step       = keys$step,
        tokens     = keys$tokens,
        constraint  = rownames(cf)[-1],
        coefficient = cf[-1, "Estimate"],
        se          = cf[-1, "Std. Error"],
        p_value     = cf[-1, "Pr(>|t|)"]
      )
    }, .keep = TRUE) %>%
    bind_rows()
}

# Attested: preference ~ constraints + OverallFreq + RelFreq + all constraint×OverallFreq
# interactions (matches brms Model B for attested); extract only constraint main effects
fit_constraint_models_attested_R <- function(data) {
  rhs <- c(
    CONSTRAINT_COLS,
    FREQ_COVARIATES,
    "RelFreq_centered:OverallFreq_scaled",
    paste0(CONSTRAINT_COLS, ":OverallFreq_scaled")
  )
  f <- reformulate(rhs, "preference")

  data %>%
    group_by(model, checkpoint, step, tokens) %>%
    group_map(function(g, keys) {
      complete <- g %>%
        select(preference, all_of(CONSTRAINT_COLS), all_of(FREQ_COVARIATES)) %>%
        drop_na()
      if (nrow(complete) < 15) return(NULL)
      fit <- lm(f, data = complete)
      cf  <- coef(summary(fit))
      cf_constraints <- cf[rownames(cf) %in% CONSTRAINT_COLS, , drop = FALSE]
      tibble(
        model      = keys$model,
        checkpoint = keys$checkpoint,
        step       = keys$step,
        tokens     = keys$tokens,
        constraint  = rownames(cf_constraints),
        coefficient = cf_constraints[, "Estimate"],
        se          = cf_constraints[, "Std. Error"],
        p_value     = cf_constraints[, "Pr(>|t|)"]
      )
    }, .keep = TRUE) %>%
    bind_rows()
}

fit_genpref_models_R <- function(data) {
  data %>%
    group_by(model, checkpoint, step, tokens) %>%
    group_map(function(g, keys) {
      complete <- g %>% select(preference, GenPref) %>% drop_na()
      if (nrow(complete) < 5) return(NULL)
      fit <- lm(preference ~ GenPref, data = complete)
      cf  <- coef(summary(fit))
      tibble(
        model      = keys$model,
        checkpoint = keys$checkpoint,
        step       = keys$step,
        tokens     = keys$tokens,
        coefficient = cf["GenPref", "Estimate"],
        se          = cf["GenPref", "Std. Error"],
        p_value     = cf["GenPref", "Pr(>|t|)"]
      )
    }, .keep = TRUE) %>%
    bind_rows()
}

cat("Fitting constraint models (nonce)...\n")
nonce_constraints    <- fit_constraint_models_nonce_R(training_nonce)
cat("Fitting constraint models (attested, controlling for frequency)...\n")
attested_constraints <- fit_constraint_models_attested_R(training_attested)

cat("Fitting GenPref models (nonce)...\n")
nonce_genpref    <- fit_genpref_models_R(training_nonce)
cat("Fitting GenPref models (attested)...\n")
attested_genpref <- fit_genpref_models_R(training_attested)

cat(sprintf("Constraint results: nonce %d rows, attested %d rows\n",
            nrow(nonce_constraints), nrow(attested_constraints)))
cat(sprintf("GenPref results:    nonce %d rows, attested %d rows\n",
            nrow(nonce_genpref),    nrow(attested_genpref)))
```

```{r model-setup}
# Consistent model ordering: BabyLM → C4 → Pythia → OLMo3, ascending params within family
MODEL_LEVELS <- c(
  "znhoughton/opt-babylm-125m-64eps-seed964",
  "znhoughton/opt-babylm-350m-64eps-seed964",
  "znhoughton/opt-babylm-1.3b-64eps-seed964",
  "znhoughton/opt-c4-125m-seed964",
  "znhoughton/opt-c4-350m-seed964",
  "znhoughton/opt-c4-1.3b-seed964",
  "EleutherAI/pythia-160m",
  "EleutherAI/pythia-410m",
  "EleutherAI/pythia-1.4b",
  "EleutherAI/pythia-2.8b",
  "allenai/Olmo-3-1025-7B",
  "allenai/Olmo-3-1125-32B"
)
MODEL_LABELS <- c(
  "BabyLM 125M", "BabyLM 350M", "BabyLM 1.3B",
  "C4 125M",     "C4 350M",     "C4 1.3B",
  "Pythia 160M", "Pythia 410M", "Pythia 1.4B", "Pythia 2.8B",
  "OLMo3 7B",    "OLMo3 32B"
)
fct_model <- function(x) factor(x, levels = MODEL_LEVELS, labels = MODEL_LABELS)

FAMILY_LEVELS <- c("BabyLM", "C4", "Pythia", "OLMo3")
model_family <- function(x) {
  case_when(
    grepl("babylm", x, ignore.case = TRUE) ~ "BabyLM",
    grepl("opt-c4", x, ignore.case = TRUE) ~ "C4",
    grepl("pythia", x, ignore.case = TRUE) ~ "Pythia",
    grepl("olmo",   x, ignore.case = TRUE) ~ "OLMo3",
    TRUE ~ "Other"
  ) %>% factor(levels = FAMILY_LEVELS)
}

CORPUS_COLORS <- c(
  "BabyLM 125M" = "#1b7837", "BabyLM 350M" = "#5aae61", "BabyLM 1.3B" = "#a6dba0",
  "C4 125M"     = "#762a83", "C4 350M"     = "#af8dc3", "C4 1.3B"     = "#e7d4e8",
  "Pythia 160M" = "#08519c", "Pythia 410M" = "#2171b5",
  "Pythia 1.4B" = "#6baed6", "Pythia 2.8B" = "#bdd7e7",
  "OLMo3 7B"    = "#d94801", "OLMo3 32B"   = "#fd8d3c"
)
```

---

## Analysis 1: Correlation Over Training

Compute Pearson r between model preference and human preference at each checkpoint.

```{r correlation-analysis}
cat("\n=== FINAL CORRELATIONS ===\n\n")
cat("NONCE BINOMIALS:\n")
nonce_correlations %>%
  group_by(model) %>%
  slice_max(tokens, n = 1) %>%
  select(model, tokens, correlation) %>%
  kable(digits = 3)

cat("\nATTESTED BINOMIALS:\n")
attested_correlations %>%
  group_by(model) %>%
  slice_max(tokens, n = 1) %>%
  select(model, tokens, correlation) %>%
  kable(digits = 3)
```

### Plot: Correlation Over Training

```{r plot-correlation, fig.width=16, fig.height=8}
bind_rows(
  nonce_correlations    %>% mutate(item_type = "Nonce"),
  attested_correlations %>% mutate(item_type = "Attested")
) %>%
  mutate(
    model_family = model_family(model),
    model        = fct_model(model)
  ) %>%
  ggplot(aes(x = tokens, y = correlation, color = model)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  facet_grid(item_type ~ model_family, scales = "free_x") +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_manual(values = CORPUS_COLORS) +
  labs(
    x        = "Training tokens",
    y        = "Pearson r (model vs. human preference)",
    title    = "Model-Human Correlation Over Training",
    subtitle = "Nonce (top) vs. Attested (bottom) | columns = model family (note: different x-axis scales)",
    color    = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  ) +
  guides(color = guide_legend(nrow = 3))
```

---

## Analysis 2: Accuracy Over Training

Accuracy = proportion of binomials where model and human agree on ordering direction.

- Model direction: mean preference > 0.5 (prefers alphabetical) vs. ≤ 0.5
- Human direction: hum_pref > 0.5 vs. ≤ 0.5
- Correct if both agree on direction

```{r accuracy-analysis}
cat("\n=== FINAL ACCURACIES ===\n\n")
cat("NONCE BINOMIALS:\n")
nonce_accuracy %>%
  group_by(model) %>%
  slice_max(tokens, n = 1) %>%
  select(model, tokens, accuracy) %>%
  kable(digits = 3)

cat("\nATTESTED BINOMIALS:\n")
attested_accuracy %>%
  group_by(model) %>%
  slice_max(tokens, n = 1) %>%
  select(model, tokens, accuracy) %>%
  kable(digits = 3)
```

### Plot: Accuracy Over Training

```{r plot-accuracy, fig.width=16, fig.height=8}
bind_rows(
  nonce_accuracy    %>% mutate(item_type = "Nonce"),
  attested_accuracy %>% mutate(item_type = "Attested")
) %>%
  mutate(
    model_family = model_family(model),
    model        = fct_model(model)
  ) %>%
  ggplot(aes(x = tokens, y = accuracy, color = model)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray60") +
  facet_grid(item_type ~ model_family, scales = "free_x") +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_color_manual(values = CORPUS_COLORS) +
  labs(
    x        = "Training tokens",
    y        = "Accuracy (proportion direction match)",
    title    = "Model-Human Direction Agreement Over Training",
    subtitle = "Nonce (top) vs. Attested (bottom) | columns = model family (note: different x-axis scales); dashed = chance (50%)",
    color    = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  ) +
  guides(color = guide_legend(nrow = 3))
```

---

## Analysis 3: Individual Constraint Effects Over Training

For each checkpoint, fit:
`lm(preference ~ Form + Percept + Culture + Power + Intense + Icon + Freq + Len + Lapse + BStress)`

```{r constraint-analysis}
cat("\n=== FINAL CONSTRAINT COEFFICIENTS ===\n\n")
cat("NONCE BINOMIALS (final checkpoint):\n")
nonce_constraints %>%
  group_by(model, constraint) %>%
  slice_max(tokens, n = 1) %>%
  select(model, constraint, coefficient) %>%
  pivot_wider(names_from = constraint, values_from = coefficient) %>%
  kable(digits = 3)

cat("\nATTESTED BINOMIALS (final checkpoint):\n")
attested_constraints %>%
  group_by(model, constraint) %>%
  slice_max(tokens, n = 1) %>%
  select(model, constraint, coefficient) %>%
  pivot_wider(names_from = constraint, values_from = coefficient) %>%
  kable(digits = 3)
```

### Plot: Individual Constraint Coefficients Over Training (Nonce)

```{r plot-constraints-nonce, fig.width=20, fig.height=12}
nonce_constraints %>%
  mutate(model = fct_model(model)) %>%
  ggplot(aes(x = tokens, y = coefficient, color = constraint)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  facet_wrap(~ model, scales = "free_x", ncol = 4) +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_viridis_d() +
  labs(
    x        = "Training tokens",
    y        = "Coefficient",
    title    = "Individual Constraint Effects Over Training — NONCE Binomials",
    subtitle = "OLS: preference ~ constraints | columns ordered by parameter size within family",
    color    = "Constraint"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(color = guide_legend(nrow = 2))
```

### Plot: Individual Constraint Coefficients Over Training (Attested)

```{r plot-constraints-attested, fig.width=20, fig.height=12}
attested_constraints %>%
  mutate(model = fct_model(model)) %>%
  ggplot(aes(x = tokens, y = coefficient, color = constraint)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  facet_wrap(~ model, scales = "free_x", ncol = 4) +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_viridis_d() +
  labs(
    x        = "Training tokens",
    y        = "Coefficient",
    title    = "Individual Constraint Effects Over Training — ATTESTED Binomials",
    subtitle = "OLS: preference ~ constraints + frequency covariates | columns ordered by parameter size within family",
    color    = "Constraint"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(color = guide_legend(nrow = 2))
```

---

## Analysis 4: GenPref Effect Over Training

Fit `lm(preference ~ GenPref)` at each checkpoint to track overall sensitivity
to the composite measure.

```{r genpref-analysis}
cat("\n=== FINAL GENPREF COEFFICIENTS ===\n\n")
cat("NONCE BINOMIALS:\n")
nonce_genpref %>%
  group_by(model) %>%
  slice_max(tokens, n = 1) %>%
  select(model, tokens, coefficient) %>%
  kable(digits = 3)

cat("\nATTESTED BINOMIALS:\n")
attested_genpref %>%
  group_by(model) %>%
  slice_max(tokens, n = 1) %>%
  select(model, tokens, coefficient) %>%
  kable(digits = 3)
```

### Plot: GenPref Coefficient Over Training

```{r plot-genpref, fig.width=16, fig.height=8}
bind_rows(
  nonce_genpref    %>% mutate(item_type = "Nonce"),
  attested_genpref %>% mutate(item_type = "Attested")
) %>%
  mutate(
    model_family = model_family(model),
    model        = fct_model(model)
  ) %>%
  ggplot(aes(x = tokens, y = coefficient, color = model)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  facet_grid(item_type ~ model_family, scales = "free_x") +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_manual(values = CORPUS_COLORS) +
  labs(
    x        = "Training tokens",
    y        = "GenPref coefficient",
    title    = "GenPref Effect Over Training (OLS)",
    subtitle = "Nonce (top) vs. Attested (bottom) | columns = model family (note: different x-axis scales)",
    color    = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  ) +
  guides(color = guide_legend(nrow = 3))
```

---

## Combined Visualization: Correlation vs. Accuracy

```{r plot-corr-acc-combined, fig.width=16, fig.height=10}
bind_rows(
  nonce_correlations    %>% mutate(item_type = "Nonce",    metric = "Correlation", value = correlation),
  attested_correlations %>% mutate(item_type = "Attested", metric = "Correlation", value = correlation),
  nonce_accuracy        %>% mutate(item_type = "Nonce",    metric = "Accuracy",    value = accuracy),
  attested_accuracy     %>% mutate(item_type = "Attested", metric = "Accuracy",    value = accuracy)
) %>%
  select(model, tokens, item_type, metric, value) %>%
  mutate(
    model_family = model_family(model),
    model        = fct_model(model),
    item_type    = factor(item_type, levels = c("Nonce", "Attested")),
    metric       = factor(metric,    levels = c("Correlation", "Accuracy"))
  ) %>%
  ggplot(aes(x = tokens, y = value, color = model, linetype = item_type)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray60", alpha = 0.5) +
  facet_grid(metric ~ model_family, scales = "free_x") +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_manual(values = CORPUS_COLORS) +
  scale_linetype_manual(values = c("Nonce" = "solid", "Attested" = "dashed")) +
  labs(
    x        = "Training tokens",
    y        = "Value",
    title    = "Model-Human Alignment: Correlation vs. Accuracy",
    subtitle = "Correlation (top) vs. Accuracy (bottom) | columns = model family | solid = Nonce, dashed = Attested",
    color    = "Model", linetype = "Binomial type"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(
    color    = guide_legend(nrow = 3),
    linetype = guide_legend(nrow = 1)
  )
```

---

## Summary Statistics

```{r summary-stats}
cat("\n=== SUMMARY: INITIAL vs. FINAL VALUES ===\n\n")

get_init_final <- function(data, value_col) {
  data %>%
    group_by(model) %>%
    arrange(tokens) %>%
    summarize(
      initial = first({{ value_col }}),
      final   = last({{ value_col }}),
      change  = final - initial,
      .groups = "drop"
    )
}

cat("NONCE - Correlation:\n")
get_init_final(nonce_correlations, correlation) %>% kable(digits = 3)

cat("\nNONCE - Accuracy:\n")
get_init_final(nonce_accuracy, accuracy) %>% kable(digits = 3)

cat("\nATTESTED - Correlation:\n")
get_init_final(attested_correlations, correlation) %>% kable(digits = 3)

cat("\nATTESTED - Accuracy:\n")
get_init_final(attested_accuracy, accuracy) %>% kable(digits = 3)

cat("\nNONCE - GenPref Coefficient:\n")
get_init_final(nonce_genpref, coefficient) %>% kable(digits = 3)

cat("\nATTESTED - GenPref Coefficient:\n")
get_init_final(attested_genpref, coefficient) %>% kable(digits = 3)
```

---

## Analysis 5: Bayesian Mixed-Effects Models

```{r setup-brms}
library(brms)

dir.create("brms_cache", showWarnings = FALSE)

# Weakly informative priors centered at zero
bayes_priors <- c(
  prior(normal(0, 0.5), class = b),
  prior(exponential(1), class = sd),
  prior(exponential(1), class = sigma)
)
```

---

### Part A: Final Checkpoint Analysis

Fit models on the final (fully trained) checkpoint for each model.

```{r final-checkpoint-data}
cat(sprintf(
  "Final checkpoint data: nonce %d rows, attested %d rows, across %d models\n",
  nrow(final_nonce), nrow(final_attested),
  n_distinct(c(final_nonce$model, final_attested$model))
))
```

#### Nonce Binomials: Final Checkpoint

```{r brms-nonce-final}
# fit → extract → discard within each iteration
nonce_final_results <- map(
  setNames(unique(final_nonce$model), unique(final_nonce$model)),
  function(model_name) {
    cat(sprintf("\n=== %s (NONCE - FINAL) ===\n", model_name))
    md <- final_nonce %>% filter(model == model_name)
    cat(sprintf("  %d rows | NAs — preference: %d, GenPref_centered: %d, binom: %d\n",
                nrow(md),
                sum(is.na(md$preference)),
                sum(is.na(md$GenPref_centered)),
                sum(is.na(md$binom))))

    result <- list()

    # ── Model A: GenPref effect ──────────────────────────────────────────────
    md_a <- md %>%
      filter(!is.na(preference), !is.na(GenPref_centered), !is.na(binom))

    if (nrow(md_a) < 10) {
      message("  Skipping Model A: < 10 complete cases")
    } else {
      cat("  → Model A: preference ~ GenPref_centered + (1 | binom)\n")
      fit_a <- brm(
        preference ~ GenPref_centered + (1 | binom),
        data    = md_a, family = gaussian(), prior = bayes_priors,
        chains  = 4, cores = 4, iter = 12000, warmup = 6000, seed = 964,
        silent  = 2,
        file    = sprintf("brms_cache/final_nonce_%s_modelA",
                          gsub("[/:-]", "_", model_name)),
        control = list(adapt_delta = 0.98)
      )
      print(round(summary(fit_a)$fixed, 3))
      rstan::check_hmc_diagnostics(fit_a$fit)

      # Extract what we need, then discard the model object
      fe_a <- fixef(fit_a)
      result$model_a_fixed <- as.data.frame(fe_a) %>%
        rownames_to_column("parameter") %>%
        mutate(model = model_name, item_type = "Nonce", checkpoint = "final")

      rm(fit_a); gc()
    }

    # ── Model B: Individual constraints ─────────────────────────────────────
    md_b <- md %>%
      filter(!is.na(preference), !is.na(binom),
             !is.na(Form), !is.na(Percept), !is.na(Culture), !is.na(Power),
             !is.na(Intense), !is.na(Icon), !is.na(Freq), !is.na(Len),
             !is.na(Lapse), !is.na(BStress))

    if (nrow(md_b) < 15) {
      message("  Skipping Model B: < 15 complete cases")
    } else {
      cat("  → Model B: preference ~ individual constraints + (1 | binom)\n")
      fit_b <- brm(
        preference ~ Form + Percept + Culture + Power + Intense + Icon +
                     Freq + Len + Lapse + BStress + (1 | binom),
        data    = md_b, family = gaussian(), prior = bayes_priors,
        chains  = 4, cores = 4, iter = 12000, warmup = 6000, seed = 964,
        silent  = 2,
        file    = sprintf("brms_cache/final_nonce_%s_modelB",
                          gsub("[/:-]", "_", model_name)),
        control = list(adapt_delta = 0.98)
      )
      print(round(summary(fit_b)$fixed, 3))
      rstan::check_hmc_diagnostics(fit_b$fit)

      fe_b <- fixef(fit_b)
      result$model_b_fixed <- as.data.frame(fe_b) %>%
        rownames_to_column("parameter") %>%
        mutate(model = model_name, item_type = "Nonce", checkpoint = "final")

      rm(fit_b); gc()
    }

    result
  }
)
```

```{r nonce-final-summaries}
cat("\n=== NONCE BINOMIALS (FINAL CHECKPOINT): FIXED EFFECTS SUMMARIES ===\n\n")
for (model_name in names(nonce_final_results)) {
  cat(sprintf("\n### %s - NONCE (FINAL)\n", model_name))
  cat("--- Model A: GenPref Effect ---\n")
  print(kable(nonce_final_results[[model_name]]$model_a_fixed, digits = 3))
  cat("\n--- Model B: Individual Constraints ---\n")
  print(kable(nonce_final_results[[model_name]]$model_b_fixed, digits = 3))
}
```

#### Attested Binomials: Final Checkpoint

```{r brms-attested-final}
attested_final_results <- map(
  setNames(unique(final_attested$model), unique(final_attested$model)),
  function(model_name) {
    cat(sprintf("\n=== %s (ATTESTED - FINAL) ===\n", model_name))
    md <- final_attested %>% filter(model == model_name)

    result <- list()

    # ── Model A: GenPref + frequency interactions ────────────────────────────
    cat("  → Model A: preference ~ GenPref_centered * OverallFreq + RelFreq * OverallFreq + (1 | binom)\n")
    fit_a <- brm(
      preference ~ GenPref_centered * OverallFreq_scaled +
                   RelFreq_centered * OverallFreq_scaled + (1 | binom),
      data    = md, family = gaussian(), prior = bayes_priors,
      chains  = 4, cores = 4, iter = 12000, warmup = 6000, seed = 964,
      silent  = 2,
      file    = sprintf("brms_cache/final_attested_%s_modelA",
                        gsub("[/:-]", "_", model_name)),
      control = list(adapt_delta = 0.98)
    )
    print(round(summary(fit_a)$fixed, 3))
    rstan::check_hmc_diagnostics(fit_a$fit)

    fe_a <- fixef(fit_a)
    result$model_a_fixed <- as.data.frame(fe_a) %>%
      rownames_to_column("parameter") %>%
      mutate(model = model_name, item_type = "Attested", checkpoint = "final")

    rm(fit_a); gc()

    # ── Model B: Individual constraints + frequency interactions ─────────────
    cat("  → Model B: preference ~ constraints + frequencies + interactions + (1 | binom)\n")
    fit_b <- brm(
      preference ~ Form + Percept + Culture + Power + Intense + Icon +
                   Freq + Len + Lapse + BStress +
                   OverallFreq_scaled + RelFreq_centered +
                   RelFreq_centered:OverallFreq_scaled +
                   Form:OverallFreq_scaled     + Percept:OverallFreq_scaled  +
                   Culture:OverallFreq_scaled  + Power:OverallFreq_scaled    +
                   Intense:OverallFreq_scaled  + Icon:OverallFreq_scaled     +
                   Freq:OverallFreq_scaled     + Len:OverallFreq_scaled      +
                   Lapse:OverallFreq_scaled    + BStress:OverallFreq_scaled  +
                   (1 | binom),
      data    = md, family = gaussian(), prior = bayes_priors,
      chains  = 4, cores = 4, iter = 12000, warmup = 6000, seed = 964,
      silent  = 2,
      file    = sprintf("brms_cache/final_attested_%s_modelB",
                        gsub("[/:-]", "_", model_name)),
      control = list(adapt_delta = 0.98, max_treedepth = 12)
    )
    print(round(summary(fit_b)$fixed, 3))
    rstan::check_hmc_diagnostics(fit_b$fit)

    fe_b <- fixef(fit_b)
    result$model_b_fixed <- as.data.frame(fe_b) %>%
      rownames_to_column("parameter") %>%
      mutate(model = model_name, item_type = "Attested", checkpoint = "final")

    rm(fit_b); gc()

    result
  }
)
```

```{r attested-final-summaries}
cat("\n=== ATTESTED BINOMIALS (FINAL CHECKPOINT): FIXED EFFECTS SUMMARIES ===\n\n")
for (model_name in names(attested_final_results)) {
  cat(sprintf("\n### %s - ATTESTED (FINAL)\n", model_name))
  cat("--- Model A: GenPref + Frequency Interactions ---\n")
  print(kable(attested_final_results[[model_name]]$model_a_fixed, digits = 3))
  cat("\n--- Model B: Individual Constraints + Frequency Interactions ---\n")
  print(kable(attested_final_results[[model_name]]$model_b_fixed, digits = 3))
}
```

#### Final Checkpoint: Key Effects Summary (95% Credible Intervals)

```{r brms-final-effects-summary}
cat("\n=== KEY EFFECTS AT FINAL CHECKPOINT: 95% CREDIBLE INTERVALS ===\n\n")

cat("NONCE BINOMIALS - GenPref Effects:\n")
map_dfr(names(nonce_final_results), function(mn) {
  nonce_final_results[[mn]]$model_a_fixed %>%
    filter(parameter == "GenPref_centered")
}) %>%
  select(model, parameter, Estimate, Est.Error, Q2.5, Q97.5) %>%
  kable(digits = 3)

cat("\n\nATTESTED BINOMIALS - GenPref and Frequency Effects:\n")
map_dfr(names(attested_final_results), function(mn) {
  attested_final_results[[mn]]$model_a_fixed %>%
    filter(grepl("GenPref_centered|OverallFreq|RelFreq", parameter))
}) %>%
  select(model, parameter, Estimate, Est.Error, Q2.5, Q97.5) %>%
  kable(digits = 3)
```

#### Final Checkpoint: Plots

```{r extract-final-effects}
# GenPref effect at final checkpoint
nonce_genpref_final <- map_dfr(names(nonce_final_results), function(mn) {
  nonce_final_results[[mn]]$model_a_fixed %>%
    filter(parameter == "GenPref_centered") %>%
    transmute(
      model     = mn,
      item_type = "Nonce",
      estimate  = Estimate,
      lower     = Q2.5,
      upper     = Q97.5
    )
})

attested_genpref_final <- map_dfr(names(attested_final_results), function(mn) {
  attested_final_results[[mn]]$model_a_fixed %>%
    filter(parameter == "GenPref_centered") %>%
    transmute(
      model     = mn,
      item_type = "Attested",
      estimate  = Estimate,
      lower     = Q2.5,
      upper     = Q97.5
    )
})

# Individual constraint effects at final checkpoint
nonce_constraint_final <- map_dfr(names(nonce_final_results), function(mn) {
  nonce_final_results[[mn]]$model_b_fixed %>%
    filter(parameter %in% CONSTRAINT_COLS) %>%
    transmute(
      model      = mn,
      item_type  = "Nonce",
      constraint = parameter,
      estimate   = Estimate,
      lower      = Q2.5,
      upper      = Q97.5
    )
})

attested_constraint_final <- map_dfr(names(attested_final_results), function(mn) {
  attested_final_results[[mn]]$model_b_fixed %>%
    filter(parameter %in% CONSTRAINT_COLS) %>%
    transmute(
      model      = mn,
      item_type  = "Attested",
      constraint = parameter,
      estimate   = Estimate,
      lower      = Q2.5,
      upper      = Q97.5
    )
})
```

```{r plot-final-genpref, fig.width=10, fig.height=5}
bind_rows(nonce_genpref_final, attested_genpref_final) %>%
  mutate(
    model     = fct_model(model),
    item_type = factor(item_type, levels = c("Nonce", "Attested"))
  ) %>%
  ggplot(aes(x = model, y = estimate, color = model, shape = item_type)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.25,
                position = position_dodge(0.5), linewidth = 0.8) +
  geom_point(size = 3.5, position = position_dodge(0.5)) +
  scale_color_manual(values = CORPUS_COLORS, guide = "none") +
  scale_shape_manual(values = c("Nonce" = 16, "Attested" = 17)) +
  labs(
    x        = NULL,
    y        = "GenPref coefficient (posterior mean ± 95% CrI)",
    title    = "GenPref Effect at Final Checkpoint",
    subtitle = "Bayesian: preference ~ GenPref_centered + (1 | binom)",
    shape    = "Binomial type"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x      = element_text(angle = 30, hjust = 1),
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  )
```

```{r plot-final-constraints, fig.width=16, fig.height=10}
bind_rows(nonce_constraint_final, attested_constraint_final) %>%
  mutate(
    model      = fct_model(model),
    item_type  = factor(item_type, levels = c("Nonce", "Attested")),
    constraint = factor(constraint, levels = CONSTRAINT_COLS)
  ) %>%
  ggplot(aes(x = model, y = estimate, color = model, shape = item_type)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.25,
                position = position_dodge(0.5), linewidth = 0.7) +
  geom_point(size = 2.5, position = position_dodge(0.5)) +
  facet_wrap(~ constraint, scales = "free_y", ncol = 5) +
  scale_color_manual(values = CORPUS_COLORS, guide = "none") +
  scale_shape_manual(values = c("Nonce" = 16, "Attested" = 17)) +
  labs(
    x        = NULL,
    y        = "Coefficient (posterior mean ± 95% CrI)",
    title    = "Individual Constraint Effects at Final Checkpoint",
    subtitle = "Nonce: preference ~ constraints + (1|binom); Attested: preference ~ constraints × frequency + (1|binom)",
    shape    = "Binomial type"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 8),
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "right"
  )
```

---

### Part B: Effects Over Training

Fit Bayesian models at logarithmically-sampled checkpoints using a
**fit → extract → discard** strategy: model objects are never accumulated in
memory. Extracted coefficients are saved to CSV after each item-type loop so
the notebook is resumable if interrupted (brms will skip refitting from cache,
and the CSVs will already hold any completed extractions).

```{r training-checkpoint-data}
cat(sprintf(
  "Training data: nonce %d rows (%d checkpoints), attested %d rows (%d checkpoints)\n",
  nrow(training_nonce),    n_distinct(training_nonce$checkpoint),
  nrow(training_attested), n_distinct(training_attested$checkpoint)
))

# Paths for cached trajectory CSVs
traj_cache <- list(
  nonce_genpref    = file.path(processed_path, "traj_nonce_genpref_bayes.csv"),
  attested_genpref = file.path(processed_path, "traj_attested_genpref_bayes.csv"),
  attested_freq    = file.path(processed_path, "traj_attested_freq_bayes.csv")
)
```

#### Nonce Binomials: Training Trajectory

```{r brms-nonce-training}
# ── Load cache; fit only models not yet present ───────────────────────────────
if (file.exists(traj_cache$nonce_genpref)) {
  nonce_genpref_traj <- read_csv(traj_cache$nonce_genpref, show_col_types = FALSE)
  missing_models     <- setdiff(unique(training_nonce$model),
                                unique(nonce_genpref_traj$model))
  if (length(missing_models) == 0) {
    message("Nonce GenPref trajectory fully cached.")
  } else {
    message(sprintf("Fitting %d model(s) missing from nonce cache: %s",
                    length(missing_models), paste(missing_models, collapse = ", ")))
  }
} else {
  nonce_genpref_traj <- NULL
  missing_models     <- unique(training_nonce$model)
}

if (length(missing_models) > 0) {
  nonce_traj_rows <- list()

  for (model_name in missing_models) {
    model_df <- training_nonce %>% filter(model == model_name)

    for (ckpt in unique(model_df$checkpoint)) {
      cat(sprintf("\n=== %s | checkpoint %s (NONCE) ===\n", model_name, ckpt))
      ckpt_data <- model_df %>% filter(checkpoint == ckpt)

      fit_a <- brm(
        preference ~ GenPref_centered + (1 | binom),
        data    = ckpt_data, family = gaussian(), prior = bayes_priors,
        chains  = 4, cores = 4, iter = 12000, warmup = 6000, seed = 964,
        silent  = 2,
        file    = sprintf("brms_cache/training_nonce_%s_%s_modelA",
                          gsub("[/:-]", "_", model_name), ckpt),
        control = list(adapt_delta = 0.98)
      )
      print(round(summary(fit_a)$fixed, 3))
      rstan::check_hmc_diagnostics(fit_a$fit)

      coefs <- fixef(fit_a)
      nonce_traj_rows[[length(nonce_traj_rows) + 1]] <- tibble(
        model      = model_name,
        checkpoint = ckpt,
        tokens     = unique(ckpt_data$tokens),
        step       = unique(ckpt_data$step),
        item_type  = "Nonce",
        estimate   = coefs["GenPref_centered", "Estimate"],
        lower      = coefs["GenPref_centered", "Q2.5"],
        upper      = coefs["GenPref_centered", "Q97.5"]
      )

      rm(fit_a); gc()
    }
  }

  nonce_genpref_traj <- bind_rows(nonce_genpref_traj, bind_rows(nonce_traj_rows))
  write_csv(nonce_genpref_traj, traj_cache$nonce_genpref)
  message("Updated nonce GenPref trajectory cache: ", traj_cache$nonce_genpref)
}
```

#### Attested Binomials: Training Trajectory

```{r brms-attested-training}
# ── Frequency parameters to track ───────────────────────────────────────────
FREQ_PARAMS <- c(
  "OverallFreq_scaled",
  "RelFreq_centered",
  "GenPref_centered:OverallFreq_scaled",
  "RelFreq_centered:OverallFreq_scaled"
)

# ── Load cache; fit only models not yet present ───────────────────────────────
if (file.exists(traj_cache$attested_genpref) && file.exists(traj_cache$attested_freq)) {
  attested_genpref_traj <- read_csv(traj_cache$attested_genpref, show_col_types = FALSE)
  attested_freq_traj    <- read_csv(traj_cache$attested_freq,    show_col_types = FALSE)
  missing_models <- setdiff(unique(training_attested$model),
                            unique(attested_genpref_traj$model))
  if (length(missing_models) == 0) {
    message("Attested trajectories fully cached.")
  } else {
    message(sprintf("Fitting %d model(s) missing from attested cache: %s",
                    length(missing_models), paste(missing_models, collapse = ", ")))
  }
} else {
  attested_genpref_traj <- NULL
  attested_freq_traj    <- NULL
  missing_models        <- unique(training_attested$model)
}

if (length(missing_models) > 0) {
  attested_genpref_rows <- list()
  attested_freq_rows    <- list()

  for (model_name in missing_models) {
    model_df <- training_attested %>% filter(model == model_name)

    for (ckpt in unique(model_df$checkpoint)) {
      cat(sprintf("\n=== %s | checkpoint %s (ATTESTED) ===\n", model_name, ckpt))
      ckpt_data <- model_df %>% filter(checkpoint == ckpt)

      fit_a <- brm(
        preference ~ GenPref_centered * OverallFreq_scaled +
                     RelFreq_centered * OverallFreq_scaled + (1 | binom),
        data    = ckpt_data, family = gaussian(), prior = bayes_priors,
        chains  = 4, cores = 4, iter = 12000, warmup = 6000, seed = 964,
        silent  = 2,
        file    = sprintf("brms_cache/training_attested_%s_%s_modelA",
                          gsub("[/:-]", "_", model_name), ckpt),
        control = list(adapt_delta = 0.98)
      )
      print(round(summary(fit_a)$fixed, 3))
      rstan::check_hmc_diagnostics(fit_a$fit)

      coefs  <- fixef(fit_a)
      tokens <- unique(ckpt_data$tokens)
      step   <- unique(ckpt_data$step)

      attested_genpref_rows[[length(attested_genpref_rows) + 1]] <- tibble(
        model      = model_name,
        checkpoint = ckpt,
        tokens     = tokens,
        step       = step,
        item_type  = "Attested",
        estimate   = coefs["GenPref_centered", "Estimate"],
        lower      = coefs["GenPref_centered", "Q2.5"],
        upper      = coefs["GenPref_centered", "Q97.5"]
      )

      for (param in FREQ_PARAMS) {
        if (param %in% rownames(coefs)) {
          attested_freq_rows[[length(attested_freq_rows) + 1]] <- tibble(
            model      = model_name,
            checkpoint = ckpt,
            tokens     = tokens,
            step       = step,
            parameter  = param,
            estimate   = coefs[param, "Estimate"],
            lower      = coefs[param, "Q2.5"],
            upper      = coefs[param, "Q97.5"]
          )
        }
      }

      rm(fit_a); gc()
    }
  }

  attested_genpref_traj <- bind_rows(attested_genpref_traj, bind_rows(attested_genpref_rows))
  attested_freq_traj    <- bind_rows(attested_freq_traj,    bind_rows(attested_freq_rows))

  write_csv(attested_genpref_traj, traj_cache$attested_genpref)
  write_csv(attested_freq_traj,    traj_cache$attested_freq)
  message("Updated attested trajectory cache.")
}
```

#### Training Trajectory: Plots

```{r plot-genpref-trajectory, fig.width=16, fig.height=8}
# Columns = model family (each with its own token scale); rows = item type
bind_rows(nonce_genpref_traj, attested_genpref_traj) %>%
  mutate(
    model_family = model_family(model),
    model        = fct_model(model)
  ) %>%
  ggplot(aes(x = tokens, y = estimate, color = model, fill = model)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  facet_grid(item_type ~ model_family, scales = "free_x") +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_manual(values = CORPUS_COLORS) +
  scale_fill_manual(values  = CORPUS_COLORS) +
  labs(
    x        = "Training tokens",
    y        = "GenPref coefficient (posterior mean ± 95% CrI)",
    title    = "Bayesian GenPref Effect Over Training",
    subtitle = "Nonce (top) vs. Attested (bottom) | columns = model family (note: different x-axis scales)",
    color    = "Model", fill = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  ) +
  guides(color = guide_legend(nrow = 3), fill = guide_legend(nrow = 3))
```

```{r plot-freq-trajectory, fig.width=16, fig.height=10}
# Columns = model family; rows = frequency parameter
attested_freq_traj %>%
  mutate(
    model_family = model_family(model),
    model        = fct_model(model),
    parameter    = factor(
      parameter,
      levels = FREQ_PARAMS,
      labels = c("OverallFreq", "RelFreq",
                 "GenPref × OverallFreq", "RelFreq × OverallFreq")
    )
  ) %>%
  ggplot(aes(x = tokens, y = estimate, color = model, fill = model)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  facet_grid(parameter ~ model_family, scales = "free") +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_manual(values = CORPUS_COLORS) +
  scale_fill_manual(values  = CORPUS_COLORS) +
  labs(
    x        = "Training tokens",
    y        = "Coefficient (posterior mean ± 95% CrI)",
    title    = "Bayesian Frequency Effects Over Training (Attested Binomials)",
    subtitle = "Columns = model family (note: different x-axis scales)",
    color    = "Model", fill = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    panel.grid.minor = element_blank()
  ) +
  guides(color = guide_legend(nrow = 3), fill = guide_legend(nrow = 3))
```

#### Training Trajectory: Individual Constraint Effects (OLS)

```{r plot-constraint-trajectory-nonce, fig.width=20, fig.height=20}
nonce_constraints %>%
  mutate(
    model_family = model_family(model),
    model        = fct_model(model),
    constraint   = factor(constraint, levels = CONSTRAINT_COLS)
  ) %>%
  ggplot(aes(x = tokens, y = coefficient, color = model)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  facet_grid(constraint ~ model_family, scales = "free_x") +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_manual(values = CORPUS_COLORS) +
  labs(
    x        = "Training tokens",
    y        = "OLS coefficient",
    title    = "Individual Constraint Effects Over Training — NONCE Binomials",
    subtitle = "OLS at ~20 log-sampled checkpoints per model family",
    color    = "Model"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(color = guide_legend(nrow = 3))
```

```{r plot-constraint-trajectory-attested, fig.width=20, fig.height=20}
attested_constraints %>%
  mutate(
    model_family = model_family(model),
    model        = fct_model(model),
    constraint   = factor(constraint, levels = CONSTRAINT_COLS)
  ) %>%
  ggplot(aes(x = tokens, y = coefficient, color = model)) +
  geom_line(linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  facet_grid(constraint ~ model_family, scales = "free_x") +
  scale_x_continuous(labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  scale_color_manual(values = CORPUS_COLORS) +
  labs(
    x        = "Training tokens",
    y        = "OLS coefficient",
    title    = "Individual Constraint Effects Over Training — ATTESTED Binomials",
    subtitle = "OLS controlling for OverallFreq & RelFreq at ~20 log-sampled checkpoints per model family",
    color    = "Model"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text       = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  ) +
  guides(color = guide_legend(nrow = 3))
```

