---
title: "Corpus Exposure vs. Model Preference"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Overview

This analysis examines whether each model's **continuous preference score**
(`alpha_logprob - nonalpha_logprob`) tracks the training frequency ratio for
each binomial at each checkpoint.

The key idea: if a model has seen "bread and butter" 60 times and "butter and
bread" 40 times, we'd expect its log-probability difference to lean positive in
proportion to `log(60/40)`. Concretely:

- **Observed preference**: `preference = alpha_logprob - nonalpha_logprob`
  (continuous; positive = model favours alpha ordering)
- **Expected log frequency ratio**: `log(alpha_seen / beta_seen)`
  (positive = alpha ordering was more frequent in training)

If the model perfectly tracks relative frequency, the two should be positively
correlated and the relationship should be linear. We also ask how this
relationship changes as a function of **total frequency** (`alpha_seen +
beta_seen`) and over the course of **training**.

---

## Setup

```{r libraries}
library(tidyverse)
library(duckdb)
library(scales)
library(patchwork)
library(lme4)
library(brms)
```

```{r brms-config}
# ── Shared brms settings ──────────────────────────────────────────────────────
MODELS_DIR  <- "models"
dir.create(MODELS_DIR, showWarnings = FALSE, recursive = TRUE)

BRMS_ITER   <- 6000
BRMS_WARMUP <- 3000
BRMS_CHAINS <- 4
BRMS_CORES  <- 4
BRMS_SEED   <- 42

# Uninformative priors centred at 0 for continuous outcomes
BRMS_PRIORS <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(0, 1), class = "Intercept"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "sigma")
)

# Same but without sigma (for bernoulli/binomial families)
BRMS_PRIORS_LOGISTIC <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(0, 1), class = "Intercept"),
  prior(exponential(1), class = "sd")
)
```

```{r helpers}
parse_corpus <- function(x) {
  case_when(
    grepl("babylm", x, ignore.case = TRUE) ~ "BabyLM",
    grepl("opt-c4", x, ignore.case = TRUE) ~ "C4",
    TRUE ~ NA_character_
  )
}

parse_size <- function(x) {
  case_when(
    grepl("125m", x, ignore.case = TRUE) ~ "125M",
    grepl("350m", x, ignore.case = TRUE) ~ "350M",
    grepl("1\\.3b", x, ignore.case = TRUE) ~ "1.3B",
    TRUE ~ NA_character_
  )
}

SIZE_LEVELS  <- c("125M", "350M", "1.3B")
SIZE_COLOURS <- c("125M" = "#1b7837", "350M" = "#762a83", "1.3B" = "#b2182b")

# Two-category semantic colours (corpus/frequency reference vs. model preference)
COL_CORPUS <- "#ff7f0e"  # warm orange: corpus / frequency-expectation reference
COL_MODEL  <- "#1f77b4"  # cool blue:  model preference
COL_OVER   <- "#d62728"  # red:        over-prediction / positive-direction indicator
```

---

## Load and Aggregate Data

DuckDB scans the ~28 GB CSV and aggregates to one row per
model x checkpoint x binomial, computing the **mean preference score** across
the 51 prompts. The log frequency ratio is computed here as well.

```{r load-duckdb, cache=TRUE}
csv_path <- normalizePath(
  "../Data/processed/checkpoint_results_with_exposures.csv",
  mustWork = TRUE
)
csv_path_fwd <- gsub("\\\\", "/", csv_path)

con <- dbConnect(duckdb::duckdb())
dbExecute(con, sprintf(
  "CREATE VIEW ckpt AS SELECT * FROM read_csv_auto('%s')",
  csv_path_fwd
))

message("Aggregating via DuckDB ...")
agg <- dbGetQuery(con, "
  SELECT
    model,
    tokens,
    binom,
    ROUND(alpha_seen) AS alpha_seen,
    ROUND(beta_seen)  AS beta_seen,
    ROUND(alpha_seen) + ROUND(beta_seen)         AS total_seen,
    -- log frequency ratio: positive = alpha was more frequent in training
    CASE WHEN alpha_seen > 0 AND beta_seen > 0
         THEN LN(alpha_seen / beta_seen)
         ELSE NULL END                           AS log_freq_ratio,
    AVG(preference)                              AS mean_preference,
    COUNT(*)                                     AS n_prompts
  FROM ckpt
  GROUP BY model, tokens, binom, alpha_seen, beta_seen
  ORDER BY model, tokens, binom
")

dbDisconnect(con, shutdown = TRUE)
message(sprintf("Aggregated to %s rows.", format(nrow(agg), big.mark = ",")))
```

```{r metadata}
agg <- agg %>%
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS),
    total_seen = as.numeric(total_seen)
  )

cat(sprintf("Rows with total_seen > 0:           %s (%.1f%%)\n",
            format(sum(agg$total_seen > 0), big.mark = ","),
            100 * mean(agg$total_seen > 0)))
cat(sprintf("Rows with both orderings seen (log ratio defined): %s (%.1f%%)\n",
            format(sum(!is.na(agg$log_freq_ratio)), big.mark = ","),
            100 * mean(!is.na(agg$log_freq_ratio))))
```

---

## Generative Preference (GenPref)

**GenPref** is a linguistic/generative preference score for each binomial derived
from a logistic regression over phonological and semantic constraint scores
(Form, Percept, Culture, Power, Intense, Icon, Freq, Len, Lapse, BStress).
It ranges from roughly -0.5 to 0.5, where positive = alpha ordering is
linguistically preferred, negative = non-alpha ordering is preferred.

The hypothesis: even when a model's training corpus diverges from what GenPref
predicts, the model may still be *pulled* toward GenPref  -  reflecting an
inductive bias toward canonical (natural) orderings.

```{r load-genpref}
human_path <- normalizePath(
  "../Data/all_human_data.csv",
  mustWork = TRUE
)

# Constraint-score logistic formula for recomputing GenPref where missing
GENPREF_FORMULA <- function(Form, Percept, Culture, Power, Intense,
                            Icon, Freq, Len, Lapse, BStress) {
  y <- 0.02191943 +
       0.23925834 * Form    +
       0.24889543 * Percept +
       0.41836997 * Culture +
       0.25967334 * Power   +
       0.01867604 * Intense +
       1.30365980 * Icon    +
       0.08553552 * Freq    +
       0.15241566 * Len     -
       0.19381657 * Lapse   +
       0.36019221 * BStress
  (1 / (1 + exp(-y))) - 0.5
}

genpref_raw <- read_csv(human_path, show_col_types = FALSE) %>%
  rename(BStress = `*BStress`) %>%          # column header has leading *
  select(Alpha, Form, Percept, Culture, Power, Intense,
         Icon, Freq, Len, Lapse, BStress, GenPref) %>%
  # One row per unique binomial (participants x items -> items)
  group_by(Alpha) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    # The formula outputs plogis(y) - 0.5, which is on [-0.5, 0.5].
    # Stored GenPref values may be on [0, 1] (raw human proportion preferring
    # alpha). Normalise stored values to [-0.5, 0.5] by subtracting 0.5 so
    # that all genpref values are on a consistent scale throughout the analysis.
    GenPref_norm = case_when(
      is.na(GenPref)  ~ NA_real_,
      GenPref > 0.5   ~ GenPref - 0.5,   # [0, 1] -> [-0.5, 0.5]
      GenPref < -0.5  ~ GenPref,          # already outside [0,1]; leave as-is and flag below
      TRUE            ~ GenPref - 0.5     # [0, 1] -> [-0.5, 0.5]
    ),
    genpref = if_else(
      is.na(GenPref),
      GENPREF_FORMULA(Form, Percept, Culture, Power, Intense,
                      Icon, Freq, Len, Lapse, BStress),
      GenPref_norm
    ),
    binom_lower = tolower(Alpha)
  ) %>%
  select(binom_lower, genpref)

# ── Diagnostic: confirm genpref is on [-0.5, 0.5] ────────────────────────────
gp_range <- range(genpref_raw$genpref, na.rm = TRUE)
cat(sprintf("Unique binomials with GenPref: %d  (NA remaining: %d)\n",
            sum(!is.na(genpref_raw$genpref)),
            sum( is.na(genpref_raw$genpref))))
cat(sprintf("GenPref range after normalisation: [%.4f, %.4f]\n",
            gp_range[1], gp_range[2]))
if (gp_range[1] < -0.5 - 1e-6 || gp_range[2] > 0.5 + 1e-6) {
  warning(sprintf(
    "GenPref values outside [-0.5, 0.5]: min = %.4f, max = %.4f. ",
    gp_range[1], gp_range[2]
  ))
} else {
  cat("OK: all GenPref values are within [-0.5, 0.5].\n")
}
```

```{r join-genpref}
# Collapse agg to one row per binomial x corpus (average over checkpoints)
# then join GenPref
agg_gp <- agg %>%
  mutate(binom_lower = tolower(binom)) %>%
  left_join(genpref_raw, by = "binom_lower")

cat(sprintf("Rows with GenPref after join: %s / %s  (%.1f%%)\n",
            format(sum(!is.na(agg_gp$genpref)), big.mark = ","),
            format(nrow(agg_gp),                big.mark = ","),
            100 * mean(!is.na(agg_gp$genpref))))
```

---

## Plot 1: Does the Log Frequency Ratio Predict the Preference Score?

For each model x checkpoint x binomial where both orderings appeared at least
once, we ask whether `log(alpha_seen / beta_seen)` predicts the preference score.

If the model tracks relative frequency, the relationship should be monotonically
increasing and pass through zero at x = 0 (equal exposure -> no preference).

Two versions are shown: (1a) raw scatter with a LOESS smoother per model size,
and (1b) binned summary with 95% CI ribbons.

```{r plot1-raw-data}
p1_raw <- agg %>%
  filter(!is.na(log_freq_ratio), !is.na(model_size))
```

### Plot 1a: Raw scatter + LOESS smoother

```{r plot1a-scatter, fig.width = 10, fig.height = 5}
ggplot(p1_raw, aes(x = log_freq_ratio, y = mean_preference,
                   colour = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.04, size = 0.4, show.legend = FALSE) +
  geom_smooth(method = "loess", span = 0.4, se = TRUE, linewidth = 1,
              aes(fill = model_size)) +
  scale_x_continuous(name = "log(alpha_seen / beta_seen)  [training frequency log-ratio]") +
  scale_y_continuous(name = "Mean preference score  (alpha_logprob - nonalpha_logprob)") +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus) +
  labs(
    title    = "Does training frequency log-ratio predict model preference score?",
    subtitle = "Each point = one model x checkpoint x binomial  -  LOESS smoother 95% CI per model size"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

### Plot 1b: Binned summary (raw points in background)

```{r plot1b-binned, fig.width = 10, fig.height = 5}
p1_data <- p1_raw %>%
  mutate(
    lrat_bin = cut(log_freq_ratio,
                   breaks = c(-Inf, -3, -2, -1, -0.5, 0,
                                0.5,  1,  2,  3,  Inf),
                   labels = c("<-3", "-3--2", "-2--1", "-1--.5",
                              "-.5-0", "0-.5", ".5-1", "1-2",
                              "2-3", ">3"))
  ) %>%
  group_by(corpus, model_size, lrat_bin) %>%
  summarize(
    mean_pref = mean(mean_preference, na.rm = TRUE),
    se        = sd(mean_preference, na.rm = TRUE) / sqrt(n()),
    midpoint  = mean(log_freq_ratio, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  )

ggplot(p1_data, aes(x = midpoint, y = mean_pref,
                    colour = model_size, fill = model_size,
                    group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(data = p1_raw,
             aes(x = log_freq_ratio, y = mean_preference, colour = model_size),
             alpha = 0.04, size = 0.4, inherit.aes = FALSE, show.legend = FALSE) +
  geom_ribbon(aes(ymin = mean_pref - 1.96 * se, ymax = mean_pref + 1.96 * se),
              alpha = 0.25, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_x_continuous(name = "log(alpha_seen / beta_seen)  [training frequency log-ratio]") +
  scale_y_continuous(name = "Mean preference score  (alpha_logprob - nonalpha_logprob)") +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus) +
  labs(
    title    = "Does training frequency log-ratio predict model preference score?",
    subtitle = "Grey cloud = raw data  -  Lines = bin means 95% CI  -  Dashed lines at x = 0, y = 0"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Plot 2: Correlation Between Preference and Log Frequency Ratio Over Training

Does the model's preference score come to track training frequency more closely
as training progresses? For each checkpoint we compute the Pearson r across all
binomials where both orderings were seen.

```{r plot3-corr-over-training, fig.width = 11, fig.height = 5}
p3_data <- agg %>%
  filter(!is.na(log_freq_ratio)) %>%
  group_by(corpus, model_size, tokens) %>%
  summarize(
    r        = cor(mean_preference, log_freq_ratio, use = "complete.obs"),
    n_binoms = n(),
    .groups  = "drop"
  )

ggplot(p3_data, aes(x = tokens / 1e9, y = r,
                    colour = model_size, fill = model_size,
                    group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(
    name   = "Pearson r  (preference ~ log frequency ratio)",
    limits = c(-0.1, 1)
  ) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus, scales = "free_x") +
  labs(
    title    = "Preference-frequency correlation over training",
    subtitle = "r computed across binomials where both orderings appear in training at that checkpoint"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r plot3-n-binoms, fig.width = 11, fig.height = 4}
# Companion: how many binomials contribute at each point?
ggplot(p3_data, aes(x = tokens / 1e9, y = n_binoms,
                    colour = model_size, group = model_size)) +
  geom_line(linewidth = 0.7) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "# binomials with both orderings seen", labels = comma) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus, scales = "free_x") +
  labs(title = "Number of binomials contributing to the correlation at each checkpoint") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Summary Table

```{r summary}
agg %>%
  filter(!is.na(log_freq_ratio)) %>%
  group_by(corpus, model_size) %>%
  summarize(
    overall_r      = cor(mean_preference, log_freq_ratio, use = "complete.obs"),
    n_observations = n(),
    .groups        = "drop"
  ) %>%
  arrange(corpus, model_size) %>%
  mutate(overall_r = round(overall_r, 3)) %>%
  knitr::kable(
    col.names = c("Corpus", "Model size", "Overall r", "N observations"),
    caption   = paste("Pearson r between mean preference score and log frequency ratio,",
                      "pooled across all checkpoints.")
  )
```

---

## Plot 3: GenPref vs. Mean Preference Score

Does the model's preference score correlate with the linguistic/generative
preference, independent of relative frequency? We use only the final checkpoint
for each model and fit a Bayesian quadratic regression.

```{r plot5-brms-setup, cache=TRUE, cache.lazy=FALSE}
# Filter to final checkpoint only
p5_final <- agg_gp %>%
  filter(!is.na(genpref)) %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup() %>%
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    # Mean-centre genpref
    genpref_c  = genpref - mean(genpref, na.rm = TRUE)
  )

plot5_fit <- brm(
  mean_preference ~ genpref_c + I(genpref_c^2) + (1 | binom),
  data    = p5_final,
  family  = gaussian(),
  prior   = BRMS_PRIORS,
  iter    = BRMS_ITER,
  warmup  = BRMS_WARMUP,
  chains  = BRMS_CHAINS,
  cores   = BRMS_CORES,
  file    = file.path(MODELS_DIR, "plot5_fit"),
  seed    = BRMS_SEED
)

print(summary(plot5_fit))
```

```{r plot5-genpref-vs-pref, fig.width=10, fig.height=5}
# Population-level predictions over a grid, replicated per corpus for faceting
gp_seq <- seq(min(p5_final$genpref_c, na.rm = TRUE),
              max(p5_final$genpref_c, na.rm = TRUE), length.out = 100)

gp_grid <- purrr::map_dfr(unique(p5_final$corpus), function(corp) {
  grid <- data.frame(genpref_c = gp_seq, corpus = corp)
  pred <- fitted(plot5_fit, newdata = grid, re_formula = NA)
  grid$estimate <- pred[, "Estimate"]
  grid$lo       <- pred[, "Q2.5"]
  grid$hi       <- pred[, "Q97.5"]
  grid$genpref  <- grid$genpref_c + mean(p5_final$genpref, na.rm = TRUE)
  grid
})

ggplot(data = gp_grid, aes(x = genpref)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(data = p5_final,
             aes(x = genpref, y = mean_preference, colour = model_size),
             alpha = 0.05, size = 0.5, show.legend = FALSE) +
  geom_ribbon(aes(ymin = lo, ymax = hi, group = corpus),
              fill = "grey40", alpha = 0.25) +
  geom_line(aes(y = estimate, group = corpus),
            colour = "grey20", linewidth = 1) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_continuous(name = "GenPref  (linguistic/generative preference score)") +
  scale_y_continuous(name = "Mean preference score  (alpha_logprob - nonalpha_logprob)") +
  facet_wrap(~corpus) +
  labs(
    title    = "Does linguistic preference (GenPref) predict model preference?",
    subtitle = "Bayesian polynomial regression (degree 2)  -  95% credible ribbon  -  Final checkpoint"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Regression: Does GenPref Predict Preference Beyond Corpus Frequency?

A single regression at each checkpoint tests whether linguistic structure
(`genpref`) and its interaction with overall exposure (`log_total`) explain
any preference beyond what relative frequency alone predicts.

**Model**: `mean_preference ~ log_freq_ratio + genpref + log_total_c + genpref:log_total_c`

- `log_total_c` is `log(total_seen)` **centred within each checkpoint**,
  so the main effects are interpretable at average frequency.
- A positive `genpref` coefficient means GenPref pushes preference toward the
  alpha order even after controlling for relative frequency.
- A negative `genpref:log_total_c` interaction would mean the GenPref pull
  is stronger for rare binomials (consistent with the fill-in hypothesis).

```{r regression-setup, cache=TRUE}
reg_data <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(genpref), !is.na(total_seen),
         total_seen > 0) %>%
  mutate(log_total = log(total_seen))

reg_coefs <- reg_data %>%
  group_by(corpus, model_size, tokens) %>%
  group_map(function(g, keys) {
    if (nrow(g) < 10) return(NULL)
    sd_lfr <- sd(g$log_freq_ratio, na.rm = TRUE)
    sd_gp  <- sd(g$genpref,        na.rm = TRUE)
    sd_lt  <- sd(g$log_total,    na.rm = TRUE)
    if (is.na(sd_lfr) || sd_lfr == 0 ||
        is.na(sd_gp)  || sd_gp  == 0 ||
        is.na(sd_lt)  || sd_lt  == 0) return(NULL)
    g2 <- g %>% mutate(log_total_c = as.numeric(scale(log_total)))
    fit <- lm(mean_preference ~ log_freq_ratio * log_total_c +
                genpref * log_total_c,
              data = g2)
    cf <- coef(summary(fit))
    terms <- c("log_freq_ratio", "genpref", "log_total_c",
               "log_freq_ratio:log_total_c", "genpref:log_total_c")
    purrr::map_dfr(terms, function(tm) {
      if (!tm %in% rownames(cf)) return(NULL)
      tibble(
        corpus     = keys$corpus,
        model_size = keys$model_size,
        tokens     = keys$tokens,
        term       = tm,
        estimate   = cf[tm, "Estimate"],
        se         = cf[tm, "Std. Error"],
        t_value    = cf[tm, "t value"],
        p_value    = cf[tm, "Pr(>|t|)"],
        n          = nrow(g2)
      )
    })
  }, .keep = TRUE) %>%
  bind_rows() %>%
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    term = factor(term,
                  levels = c("log_freq_ratio", "genpref", "log_total_c",
                             "log_freq_ratio:log_total_c", "genpref:log_total_c"),
                  labels = c("RelFreq\n(log ratio)",
                             "GenPref",
                             "Overall frequency\n(ln total)",
                             "RelFreq x\nOverall frequency",
                             "GenPref x\nOverall frequency"))
  )

cat(sprintf("Regression coefficients: %s rows across %d checkpoints\n",
            format(nrow(reg_coefs), big.mark = ","),
            n_distinct(reg_coefs$tokens)))
```

---

### Plot 4a: Regression Coefficients  -  All Checkpoints Pooled

Each (binomial x checkpoint) pair is one observation. The regression is run
once per corpus x model size using the full dataset, so rarer binomials
contribute their early-checkpoint observations alongside frequent ones.

```{r plot11a-reg-pooled, fig.width = 11, fig.height = 5}
TERM_LEVELS <- c("log_freq_ratio", "genpref", "log_total_c",
                 "log_freq_ratio:log_total_c", "genpref:log_total_c")
TERM_LABELS <- c("RelFreq\n(log ratio)", "GenPref",
                 "Overall frequency\n(ln total)",
                 "RelFreq x\nOverall frequency",
                 "GenPref x\nOverall frequency")

reg_pooled <- reg_data %>%
  group_by(corpus, model_size) %>%
  group_map(function(g, keys) {
    sd_lfr <- sd(g$log_freq_ratio, na.rm = TRUE)
    sd_gp  <- sd(g$genpref,        na.rm = TRUE)
    sd_lt  <- sd(g$log_total,    na.rm = TRUE)
    if (is.na(sd_lfr) || sd_lfr == 0 ||
        is.na(sd_gp)  || sd_gp  == 0 ||
        is.na(sd_lt)  || sd_lt  == 0) return(NULL)
    g2 <- g %>% mutate(log_total_c = as.numeric(scale(log_total)))
    fit <- lmer(mean_preference ~ log_freq_ratio * log_total_c +
                  genpref * log_total_c + (1 | binom),
                data    = g2,
                REML    = FALSE,
                control = lmerControl(optimizer = "bobyqa"))
    cf <- coef(summary(fit))
    # Normalise interaction names: R orders by variable index, not formula order,
    # so genpref:log_total_c may be stored as log_total_c:genpref
    rownames(cf) <- dplyr::recode(rownames(cf),
      "log_total_c:genpref"        = "genpref:log_total_c",
      "log_total_c:log_freq_ratio" = "log_freq_ratio:log_total_c"
    )
    purrr::map_dfr(TERM_LEVELS, function(tm) {
      if (!tm %in% rownames(cf)) return(NULL)
      tibble(
        corpus     = keys$corpus,
        model_size = keys$model_size,
        term       = tm,
        estimate   = cf[tm, "Estimate"],
        se         = cf[tm, "Std. Error"],
        n          = nrow(g2)
      )
    })
  }, .keep = TRUE) %>%
  bind_rows() %>%
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    term       = factor(term, levels = TERM_LEVELS, labels = TERM_LABELS)
  )

ggplot(reg_pooled,
       aes(x = term, y = estimate,
           colour = model_size, group = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
                width = 0.2, position = position_dodge(0.5),
                linewidth = 0.7) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(name = "Mixed-effects coefficient  (95% CI)") +
  facet_wrap(~corpus) +
  labs(
    title    = "Regression: mean_preference ~ RelFreq x overall freq + GenPref x overall freq",
    subtitle = "All checkpoints pooled  -  Random intercept by binomial  -  95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 9))
```

### Plot 4b: GenPref vs. Preference (final checkpoint)

Direct scatterplot of GenPref against model preference at the final
checkpoint, pooled across model sizes within each corpus. Each point is
one binomial. A degree-2 polynomial smooth shows the overall shape of
the relationship after controlling for nothing else.

```{r plot11c-genpref-scatter, fig.width = 11, fig.height = 5}
p11c_data <- reg_data %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup()

ggplot(p11c_data, aes(x = genpref, y = mean_preference,
                      colour = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.07, size = 0.6, show.legend = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE,
              linewidth = 1, aes(fill = model_size)) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  scale_x_continuous(name = "GenPref  (linguistic preference, -0.5 to +0.5)") +
  scale_y_continuous(name = "Mean preference score  (log-odds)") +
  facet_wrap(~corpus) +
  labs(
    title    = "GenPref vs. model preference at final checkpoint",
    subtitle = "Each point = one binomial  -  Degree-2 polynomial 95% CI per model size"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

### Plot 4c: GenPref x Overall Frequency interaction (final checkpoint)

To visualise the `genpref:log_total_c` interaction, binomials are split
into three GenPref terciles and the relationship between overall frequency
and model preference is shown for each. If the interaction is real, the
slope of the frequency-preference curve should differ across GenPref levels.

```{r plot11d-genpref-overall-interaction, fig.width = 12, fig.height = 5}
p11d_data <- reg_data %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup() %>%
  group_by(corpus) %>%
  mutate(genpref_tercile = cut(genpref,
    breaks = quantile(genpref, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE),
    labels = c("Low GenPref", "Mid GenPref", "High GenPref"),
    include.lowest = TRUE)) %>%
  ungroup() %>%
  filter(!is.na(genpref_tercile))

ggplot(p11d_data, aes(x = log_total, y = mean_preference,
                      colour = genpref_tercile)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE,
              linewidth = 1, aes(fill = genpref_tercile), alpha = 0.12) +
  scale_colour_manual(values = c("Low GenPref"  = "#d73027",
                                 "Mid GenPref"  = "#999999",
                                 "High GenPref" = "#1a9850"),
                      name = NULL) +
  scale_fill_manual(values = c("Low GenPref"  = "#d73027",
                               "Mid GenPref"  = "#999999",
                               "High GenPref" = "#1a9850"),
                    name = NULL) +
  scale_x_continuous(name   = "Overall frequency  (ln total seen)",
                     labels = function(x) scales::comma(round(exp(x)))) +
  scale_y_continuous(name = "Mean preference score  (log-odds)") +
  facet_wrap(~corpus) +
  labs(
    title    = "GenPref x Overall frequency interaction at final checkpoint",
    subtitle = "Binomials split by GenPref tercile  -  if interaction exists, slopes differ  -  Degree-2 polynomial 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

### Plot 4d: Regression Coefficients Over Training (per-checkpoint)

```{r plot11b-reg-traj, cache=TRUE, fig.width = 13, fig.height = 9}
ggplot(reg_coefs,
       aes(x = tokens / 1e9, y = estimate,
           colour = model_size, fill = model_size,
           group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "OLS coefficient  (95% CI ribbon)") +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_grid(term ~ corpus, scales = "free_y") +
  labs(
    title    = "Regression coefficients over training",
    subtitle = paste("Rows = predictor  -  RelFreq + GenPref + overall + interaction",
                     "  -  95% CI ribbon")
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text.y     = element_text(size = 9, angle = 0, hjust = 0)
  )
```

---

# Section A: Systematic Divergence by GenPref Strength

## Plot 5: Signed Deviation by GenPref Quartile Over Training

If GenPref pulls model preferences systematically, binomials in the top
GenPref quartile should show positive signed deviations from frequency
expectation, and bottom-quartile binomials should show negative deviations.

```{r plot19-setup, cache=TRUE}
gp_quartiles <- agg_gp %>%
  filter(!is.na(genpref)) %>%
  group_by(binom) %>%
  summarise(genpref = first(genpref), .groups = "drop") %>%
  mutate(
    gp_quartile = ntile(genpref, 4),
    gp_label    = paste0("Q", gp_quartile, "  (",
                         c("low", "mid-low", "mid-high", "high")[gp_quartile],
                         " GenPref)")
  )

gp_dev_traj <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(genpref)) %>%
  inner_join(gp_quartiles %>% select(binom, gp_quartile, gp_label), by = "binom") %>%
  mutate(signed_gap = mean_preference - log_freq_ratio) %>%
  group_by(corpus, model_size, tokens, gp_quartile, gp_label) %>%
  summarise(
    mean_dev = mean(signed_gap, na.rm = TRUE),
    se_dev   = sd(signed_gap,   na.rm = TRUE) / sqrt(n()),
    n        = n(),
    .groups  = "drop"
  ) %>%
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    gp_label   = factor(gp_label, levels = unique(gp_label[order(gp_quartile)]))
  )

cat(sprintf("GenPref quartile trajectories: %d rows\n", nrow(gp_dev_traj)))
```

```{r plot19-gp-quartile, fig.width = 13, fig.height = 8}
ggplot(gp_dev_traj,
       aes(x = tokens / 1e9, y = mean_dev,
           colour = gp_label, fill = gp_label, group = gp_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_ribbon(aes(ymin = mean_dev - 1.96 * se_dev, ymax = mean_dev + 1.96 * se_dev),
              alpha = 0.12, colour = NA) +
  geom_line(linewidth = 0.9) +
  scale_colour_brewer(palette = "RdBu", direction = -1, name = "GenPref quartile") +
  scale_fill_brewer(palette   = "RdBu", direction = -1, name = "GenPref quartile") +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "Mean signed gap  (preference - log_freq_ratio)") +
  facet_grid(model_size ~ corpus, scales = "free_x") +
  labs(
    title    = "Signed deviation from frequency expectation by GenPref quartile over training",
    subtitle = "If GenPref pulls: Q4 (high) stays positive, Q1 (low) stays negative  -  95% CI ribbon"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 9))
```

---

# Section B: Mixed-Effects Model and Variance Decomposition

How much variance in model preference is explained by relative frequency alone,
GenPref alone, both together, and a full model with interactions? The
mixed-effects version adds a random intercept per binomial to account for
the repeated-measures structure across checkpoints.

```{r vardecomp-setup, cache=TRUE, cache.lazy=FALSE}
library(lme4)

# ── Centre predictors ─────────────────────────────────────────────────────────
vd_raw <- reg_data %>% filter(!is.na(genpref))

# genpref is already on [-0.5, 0.5] after normalisation in load-genpref.
# Centre by subtracting its mean (should be near 0 if binomials are balanced).
vd_data <- vd_raw %>%
  mutate(
    freq_prob_c = plogis(log_freq_ratio) - 0.5,
    genpref_c   = genpref - mean(genpref, na.rm = TRUE),
    log_total_c = as.numeric(scale(log_total))
  )

cat(sprintf("Predictors centred: genpref mean = %.4f  (range: [%.4f, %.4f])\n",
            mean(vd_raw$genpref, na.rm = TRUE),
            min(vd_raw$genpref,  na.rm = TRUE),
            max(vd_raw$genpref,  na.rm = TRUE)))

y      <- vd_data$mean_preference
ss_tot <- sum((y - mean(y))^2)

# ── Pre-compute fold assignments ONCE  -  shared by every model ─────────────────
# Folds are at the binomial level: ALL checkpoints for a binomial land in the
# same fold. Using identical folds for OLS and lmer makes Q^2 directly comparable.
K <- 5
set.seed(42)
unique_binoms <- unique(vd_data$binom)
fold_assign   <- setNames(
  sample(rep(1:K, length.out = length(unique_binoms))),
  unique_binoms
)

# ── CV function  -  5-fold binomial-grouped holdout ────────────────────────────
# Returns per-observation squared errors (one entry per row of data).
# Each observation belongs to exactly one fold and is predicted exactly once.
lobo_errors <- function(formula, data, fit_fn) {
  binom_vals <- data$binom
  sq_err     <- numeric(nrow(data))
  for (fold in seq_len(K)) {
    held              <- names(fold_assign)[fold_assign == fold]
    in_train          <- !binom_vals %in% held
    fit_k             <- fit_fn(formula, data[in_train, ])
    pred              <- predict(fit_k,
                                 newdata          = data[!in_train, ],
                                 allow.new.levels = TRUE)
    sq_err[!in_train] <- (data$mean_preference[!in_train] - pred)^2
  }
  sq_err
}

lmer_fn <- function(frm, dat) lmer(frm, data = dat, REML = FALSE,
                                   control = lmerControl(optimizer = "bobyqa"))

# ── Model specifications (mixed-effects only) ─────────────────────────────────
mixed_specs <- list(
  "Null (intercept only)"                    = mean_preference ~ 1                                                              + (1 | binom),
  "Frequency"                                = mean_preference ~ freq_prob_c                                                    + (1 | binom),
  "GenPref"                                  = mean_preference ~ genpref_c                                                      + (1 | binom),
  "Overall frequency"                        = mean_preference ~ log_total_c                                                  + (1 | binom),
  "Frequency + GenPref"                      = mean_preference ~ freq_prob_c + genpref_c                                        + (1 | binom),
  "Frequency + Overall"                      = mean_preference ~ freq_prob_c + log_total_c                                    + (1 | binom),
  "GenPref + Overall"                        = mean_preference ~ genpref_c + log_total_c                                      + (1 | binom),
  "Frequency + GenPref + Overall"            = mean_preference ~ freq_prob_c + genpref_c + log_total_c                        + (1 | binom),
  "Frequency x Overall"                      = mean_preference ~ freq_prob_c * log_total_c                                    + (1 | binom),
  "GenPref x Overall"                        = mean_preference ~ genpref_c * log_total_c                                      + (1 | binom),
  "Frequency x Overall + GenPref"            = mean_preference ~ freq_prob_c * log_total_c + genpref_c                        + (1 | binom),
  "Frequency + GenPref x Overall"            = mean_preference ~ freq_prob_c + genpref_c * log_total_c                        + (1 | binom),
  "Frequency x Overall + GenPref x Overall"  = mean_preference ~ freq_prob_c * log_total_c + genpref_c * log_total_c        + (1 | binom)
)

# ── Run 5-fold binomial-grouped LOO-CV for every mixed model ──────────────────
cat(sprintf("Running %d-fold binomial-grouped CV for %d mixed models...\n",
            K, length(mixed_specs)))

cv_results <- purrr::imap(mixed_specs, function(frm, name) {
  cat(sprintf("  Mixed: %s\n", name))
  errs    <- tryCatch(lobo_errors(frm, vd_data, lmer_fn),
                      error = function(e) { warning(e); NULL })
  fit_mix <- tryCatch(lmer_fn(frm, vd_data), error = function(e) NULL)
  list(
    errors   = errs,
    q2       = if (!is.null(errs)) 1 - sum(errs) / ss_tot else NA_real_,
    aic      = if (!is.null(fit_mix)) round(AIC(fit_mix), 1) else NA_real_,
    n_params = if (!is.null(fit_mix)) length(fixef(fit_mix)) + 2L else NA_integer_
  )
})

cv_errors <- purrr::map(cv_results, "errors")   # named list of per-obs squared error vectors

cv_table <- purrr::imap_dfr(cv_results, function(res, name) {
  tibble(
    Model    = name,
    LOBO_Q2  = round(res$q2, 4),
    AIC      = res$aic,
    n_params = res$n_params
  )
})

print(cv_table %>% select(Model, LOBO_Q2, AIC, n_params))
```

```{r vardecomp-summary}
# ── 1. Ranked table ───────────────────────────────────────────────────────────
cat("=== Models ranked by LOBO-CV Q^2 (best first) ===\n")
cv_ranked <- cv_table %>%
  arrange(desc(LOBO_Q2)) %>%
  mutate(rank = row_number(),
         delta_Q2_from_null = round(LOBO_Q2 - LOBO_Q2[Model == "Null (intercept only)"], 4))
print(cv_ranked %>% select(rank, Model, LOBO_Q2, delta_Q2_from_null, AIC, n_params))

# ── 2. Marginal contribution of each component ────────────────────────────────
# Compare pairs of nested models that differ by exactly one component
null_q2  <- cv_table$LOBO_Q2[cv_table$Model == "Null (intercept only)"]
freq_q2  <- cv_table$LOBO_Q2[cv_table$Model == "Frequency"]
gp_q2    <- cv_table$LOBO_Q2[cv_table$Model == "GenPref"]
ovr_q2   <- cv_table$LOBO_Q2[cv_table$Model == "Overall frequency"]
fgp_q2   <- cv_table$LOBO_Q2[cv_table$Model == "Frequency + GenPref"]
fovr_q2  <- cv_table$LOBO_Q2[cv_table$Model == "Frequency + Overall"]
all3_q2  <- cv_table$LOBO_Q2[cv_table$Model == "Frequency + GenPref + Overall"]
fxovr_q2 <- cv_table$LOBO_Q2[cv_table$Model == "Frequency x Overall + GenPref x Overall"]

marginal <- tribble(
  ~Comparison,                                          ~delta_Q2,
  "Null -> Frequency (main effect)",                    round(freq_q2  - null_q2, 4),
  "Null -> GenPref (main effect)",                      round(gp_q2    - null_q2, 4),
  "Null -> Overall frequency (main effect)",            round(ovr_q2   - null_q2, 4),
  "Frequency -> Frequency + GenPref",                   round(fgp_q2   - freq_q2, 4),
  "Frequency -> Frequency + Overall",                   round(fovr_q2  - freq_q2, 4),
  "Freq + GenPref + Overall -> full interactions",      round(fxovr_q2 - all3_q2, 4)
)
cat("\n=== Marginal Q^2 gain from adding each component ===\n")
print(marginal)

# ── 3. Best model: full fixed-effect summary ──────────────────────────────────
best_name <- cv_ranked$Model[1]
best_frm  <- mixed_specs[[best_name]]
cat(sprintf("\n=== Best model: '%s'  (Q^2 = %.4f) ===\n", best_name,
            cv_ranked$LOBO_Q2[1]))
best_fit <- lmer(best_frm, data = vd_data, REML = FALSE,
                 control = lmerControl(optimizer = "bobyqa"))
print(summary(best_fit), correlation = FALSE)

cat("\n=== Random-effects variance decomposition ===\n")
vc <- as.data.frame(VarCorr(best_fit))
vc$pct <- round(100 * vc$vcov / sum(vc$vcov), 1)
print(vc %>% select(grp, var1, vcov, pct))

# ── 4. Significance testing — paired CV t-test on per-observation errors ───────
# For each adjacent pair in the Q^2 ranking, test whether Model A's squared
# prediction errors are systematically smaller than Model B's.  Because both
# models are evaluated on the *same* held-out observations (identical fold
# assignment), the errors are paired — this is valid for both nested and
# non-nested model pairs.
cat("\n=== Significance testing: paired CV t-test (adjacent models by Q^2) ===\n")
cat("(Paired on per-observation squared prediction errors; two-sided t-test)\n\n")

n_models <- nrow(cv_ranked)
cv_paired <- purrr::map2_dfr(
  seq_len(n_models - 1),
  seq(2, n_models),
  function(i, j) {
    nm_a  <- cv_ranked$Model[i]
    nm_b  <- cv_ranked$Model[j]
    err_a <- cv_errors[[nm_a]]
    err_b <- cv_errors[[nm_b]]
    if (is.null(err_a) || is.null(err_b)) return(NULL)
    tt <- t.test(err_a, err_b, paired = TRUE)
    tibble(
      Model_A  = nm_a,
      Model_B  = nm_b,
      delta_Q2 = round(cv_ranked$LOBO_Q2[i] - cv_ranked$LOBO_Q2[j], 4),
      t_stat   = round(tt$statistic, 3),
      p_value  = round(tt$p.value, 4),
      sig      = dplyr::case_when(
        tt$p.value < 0.001 ~ "***",
        tt$p.value < 0.01  ~ "**",
        tt$p.value < 0.05  ~ "*",
        TRUE               ~ "ns"
      )
    )
  }
)
print(cv_paired)

# Determine winners: start from top and keep models until one is
# significantly worse than the previous (alpha = 0.05)
winner_models <- cv_ranked$Model[1]  # top model always wins
for (i in seq_len(nrow(cv_paired))) {
  if (!is.na(cv_paired$p_value[i]) && cv_paired$p_value[i] < 0.05) break
  winner_models <- c(winner_models, cv_ranked$Model[i + 1])
}
winner_params <- sapply(winner_models, function(nm) cv_ranked$n_params[cv_ranked$Model == nm])
simplest_winner <- winner_models[which.min(winner_params)]

cat(sprintf("\n=== WINNERS (no significant difference among these) ===\n"))
cat(paste(" -", winner_models, "\n"), sep = "")
cat(sprintf("\nSimplest winner: '%s'  (%d fixed-effect parameters)\n",
            simplest_winner, min(winner_params)))
```

```{r vardecomp-plot, fig.width = 11, fig.height = 7}
cv_table %>%
  arrange(LOBO_Q2) %>%                        # ascending so best lands at top of y-axis
  mutate(
    Model    = factor(Model, levels = Model),  # factor order = plot order (bottom = worst)
    bar_fill = case_when(
      Model %in% winner_models ~ "#2c7bb6",   # winning models: blue
      TRUE                     ~ "#cccccc"    # eliminated models: grey
    )
  ) %>%
  ggplot(aes(x = LOBO_Q2, y = Model, fill = bar_fill)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.4f", LOBO_Q2)), hjust = -0.1, size = 2.8) +
  geom_vline(xintercept = 0, colour = "grey30", linewidth = 0.5) +
  scale_x_continuous(
    name   = "LOBO-CV Q^2  (leave-one-binomial-out predictive R^2)",
    expand = expansion(mult = c(0.05, 0.2))
  ) +
  scale_y_discrete(name = NULL) +
  scale_fill_identity() +
  labs(
    title    = "Out-of-sample predictive accuracy  -  mixed-effects models",
    subtitle = paste("5-fold binomial-grouped CV  -",
                     "Holdout unit = entire binomial (all checkpoints)  -",
                     "Freq = P(alpha) - 0.5  -  GenPref on [-0.5, 0.5]  -",
                     "All models include (1|binom)  -",
                     "Blue = winner (not significantly different from best)  -  Grey = eliminated")
  ) +
  theme_bw(base_size = 11) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none", axis.text.y = element_text(size = 8))
```

---

# Section C: What Determines the Direction of the Signed Gap?

When the model diverges from relative frequency, is that divergence explained by
GenPref, overall frequency, corpus-predicted frequency, or interactions among
these? Here `signed_gap` (`mean_preference - log_freq_ratio`) is the explicit
outcome, so every coefficient directly describes a predictor of the gap rather
than of raw preference.

A positive signed gap means the model is *more* alpha-preferring than the
corpus predicts; a negative gap means *less*.

```{r gap-direction-setup, cache=TRUE, cache.lazy=FALSE}
# Re-use centred predictors from vd_data (same centering as Section E)
gd_data <- vd_data %>%
  mutate(
    signed_gap   = mean_preference - log_freq_ratio,
    gap_positive = as.integer(signed_gap > 0)
  )

cat(sprintf("Gap-direction data: %d observations, %d unique binomials\n",
            nrow(gd_data), n_distinct(gd_data$binom)))

# ── Final-checkpoint subset (feasible for brms) ────────────────────────────────
gd_final <- gd_data %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup()

cat(sprintf("Final-checkpoint subset: %d observations, %d unique binomials\n",
            nrow(gd_final), n_distinct(gd_final$binom)))

# ── Bayesian linear mixed model: signed gap ~ predictors + (1|binom) ──────────
# freq_prob_c  = P(alpha in corpus) - 0.5  (what corpus predicts, centred)
# genpref_c    = GenPref - mean            (linguistic preference, centred)
# log_total_c  = scale(log(total_seen))    (overall exposure, z-scored)
gd_fit <- brm(
  signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
  data   = gd_final,
  family = gaussian(),
  prior  = BRMS_PRIORS,
  iter   = BRMS_ITER,
  warmup = BRMS_WARMUP,
  chains = BRMS_CHAINS,
  cores  = BRMS_CORES,
  file   = file.path(MODELS_DIR, "gd_fit"),
  seed   = BRMS_SEED
)
cat("\n=== Bayesian linear mixed model (final checkpoint): signed_gap ===\n")
print(summary(gd_fit))

# ── Bayesian logistic mixed model: P(gap > 0) ─────────────────────────────────
gd_logit <- brm(
  gap_positive ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
  data   = gd_final,
  family = bernoulli(),
  prior  = BRMS_PRIORS_LOGISTIC,
  iter   = BRMS_ITER,
  warmup = BRMS_WARMUP,
  chains = BRMS_CHAINS,
  cores  = BRMS_CORES,
  file   = file.path(MODELS_DIR, "gd_logit"),
  seed   = BRMS_SEED + 1L
)
cat("\n=== Bayesian logistic mixed model (final checkpoint): P(gap > 0) ===\n")
print(summary(gd_logit))

# ── Coefficient table for plotting ────────────────────────────────────────────
extract_coefs_brms <- function(fit, model_label) {
  fe        <- as.data.frame(fixef(fit))
  fe$term   <- rownames(fe)
  fe$model  <- model_label
  fe
}

TERM_LEVELS_GD <- c("freq_prob_c", "genpref_c", "log_total_c",
                    "freq_prob_c:log_total_c", "genpref_c:log_total_c")
TERM_LABELS_GD <- c("RelFreq\n(P(alpha)-0.5)", "GenPref",
                    "Overall freq\n(ln total)",
                    "RelFreq x\nOverall freq", "GenPref x\nOverall freq")

gd_coefs <- bind_rows(
  extract_coefs_brms(gd_fit,   "Linear (signed gap)"),
  extract_coefs_brms(gd_logit, "Logistic (gap > 0)")
) %>%
  rename(estimate = Estimate, se = Est.Error, lo = Q2.5, hi = Q97.5) %>%
  filter(term != "Intercept") %>%
  mutate(
    term = factor(term, levels = TERM_LEVELS_GD, labels = TERM_LABELS_GD),
    sig  = if_else(lo > 0 | hi < 0, "*", "")
  )

# ── Per-corpus brms models (final checkpoint) ──────────────────────────────────
CORPORA <- sort(unique(gd_final$corpus))

corpus_slug <- function(corp) tolower(gsub("[^a-z0-9]", "", corp))

gd_fit_corpus <- setNames(lapply(CORPORA, function(corp) {
  brm(
    signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
    data   = filter(gd_final, corpus == corp),
    family = gaussian(),
    prior  = BRMS_PRIORS,
    iter   = BRMS_ITER,
    warmup = BRMS_WARMUP,
    chains = BRMS_CHAINS,
    cores  = BRMS_CORES,
    file   = file.path(MODELS_DIR, paste0("gd_fit_", corpus_slug(corp))),
    seed   = BRMS_SEED
  )
}), CORPORA)

gd_logit_corpus <- setNames(lapply(CORPORA, function(corp) {
  brm(
    gap_positive ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
    data   = filter(gd_final, corpus == corp),
    family = bernoulli(),
    prior  = BRMS_PRIORS_LOGISTIC,
    iter   = BRMS_ITER,
    warmup = BRMS_WARMUP,
    chains = BRMS_CHAINS,
    cores  = BRMS_CORES,
    file   = file.path(MODELS_DIR, paste0("gd_logit_", corpus_slug(corp))),
    seed   = BRMS_SEED + 1L
  )
}), CORPORA)

for (corp in CORPORA) {
  cat(sprintf("\n=== brms linear (final checkpoint, %s): signed_gap ===\n", corp))
  print(summary(gd_fit_corpus[[corp]]))
  cat(sprintf("\n=== brms logistic (final checkpoint, %s): P(gap > 0) ===\n", corp))
  print(summary(gd_logit_corpus[[corp]]))
}

gd_coefs_by_corpus <- bind_rows(
  purrr::imap_dfr(gd_fit_corpus,   function(fit, corp)
    extract_coefs_brms(fit, "Linear (signed gap)") %>% mutate(corpus = corp)),
  purrr::imap_dfr(gd_logit_corpus, function(fit, corp)
    extract_coefs_brms(fit, "Logistic (gap > 0)") %>% mutate(corpus = corp))
) %>%
  rename(estimate = Estimate, se = Est.Error, lo = Q2.5, hi = Q97.5) %>%
  filter(term != "Intercept") %>%
  mutate(
    term = factor(term, levels = TERM_LEVELS_GD, labels = TERM_LABELS_GD),
    sig  = if_else(lo > 0 | hi < 0, "*", "")
  )
```

```{r gap-direction-pooled, cache=TRUE}
# ── lmer: all checkpoints pooled, linear ──────────────────────────────────────
gd_pooled_fit <- lmer(
  signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
  data    = gd_data,
  REML    = FALSE,
  control = lmerControl(optimizer = "bobyqa")
)
cat("=== lmer (all checkpoints pooled): signed_gap ===\n")
print(summary(gd_pooled_fit))

# ── glmer: all checkpoints pooled, logistic ────────────────────────────────────
gd_pooled_logit <- glmer(
  gap_positive ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
  data    = gd_data,
  family  = binomial(),
  control = glmerControl(optimizer = "bobyqa")
)
cat("\n=== glmer (all checkpoints pooled): P(gap > 0) ===\n")
print(summary(gd_pooled_logit))

# ── Extract coefficients for plotting ─────────────────────────────────────────
extract_coefs_lmer <- function(fit, model_label) {
  cf <- coef(summary(fit))
  tibble(
    term     = rownames(cf),
    estimate = cf[, "Estimate"],
    se       = cf[, "Std. Error"],
    lo       = cf[, "Estimate"] - 1.96 * cf[, "Std. Error"],
    hi       = cf[, "Estimate"] + 1.96 * cf[, "Std. Error"],
    model    = model_label
  ) %>%
    # Normalise interaction names to match TERM_LEVELS_GD regardless of lmer's ordering
    mutate(term = recode(term,
      "log_total_c:freq_prob_c" = "freq_prob_c:log_total_c",
      "log_total_c:genpref_c"   = "genpref_c:log_total_c",
      "(Intercept)"             = "Intercept"
    ))
}

gd_pooled_coefs <- bind_rows(
  extract_coefs_lmer(gd_pooled_fit,   "Linear (signed gap)"),
  extract_coefs_lmer(gd_pooled_logit, "Logistic (gap > 0)")
) %>%
  filter(term != "Intercept") %>%
  mutate(
    term = factor(term, levels = TERM_LEVELS_GD, labels = TERM_LABELS_GD),
    sig  = if_else(lo > 0 | hi < 0, "*", "")
  )

# ── Per-corpus lmer / glmer (all checkpoints pooled) ──────────────────────────
gd_pooled_coefs_by_corpus <- bind_rows(
  purrr::map_dfr(CORPORA, function(corp) {
    fit <- lmer(
      signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
      data    = filter(gd_data, corpus == corp),
      REML    = FALSE,
      control = lmerControl(optimizer = "bobyqa")
    )
    cat(sprintf("=== lmer pooled (%s): signed_gap ===\n", corp))
    print(summary(fit))
    extract_coefs_lmer(fit, "Linear (signed gap)") %>% mutate(corpus = corp)
  }),
  purrr::map_dfr(CORPORA, function(corp) {
    fit <- glmer(
      gap_positive ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
      data    = filter(gd_data, corpus == corp),
      family  = binomial(),
      control = glmerControl(optimizer = "bobyqa")
    )
    cat(sprintf("=== glmer pooled (%s): P(gap > 0) ===\n", corp))
    print(summary(fit))
    extract_coefs_lmer(fit, "Logistic (gap > 0)") %>% mutate(corpus = corp)
  })
) %>%
  filter(term != "Intercept") %>%
  mutate(
    term = factor(term, levels = TERM_LEVELS_GD, labels = TERM_LABELS_GD),
    sig  = if_else(lo > 0 | hi < 0, "*", "")
  )
```

```{r gap-direction-pooled-plot, fig.width = 12, fig.height = 5}
ggplot(gd_pooled_coefs,
       aes(x = term, y = estimate,
           colour = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = lo, ymax = hi),
                width = 0.2, position = position_dodge(0.5), linewidth = 0.8) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_text(aes(label = sig, y = hi + 0.02),
            position = position_dodge(0.5), size = 4, vjust = 0) +
  scale_colour_manual(values = c("Linear (signed gap)" = COL_MODEL,
                                 "Logistic (gap > 0)"  = COL_CORPUS),
                      name = NULL) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(name = "Coefficient  (95% CI)") +
  facet_wrap(~model, scales = "free_y") +
  labs(
    title    = "What predicts the signed gap? (all checkpoints pooled)",
    subtitle = paste0(
      "Linear outcome: signed_gap  -  Logistic outcome: P(gap > 0)\n",
      "All predictors centred  -  Random intercept by binomial  -  ",
      "lmer / glmer  -  Wald 95% CI  -  * = CI excludes zero")
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 10),
        axis.text.x = element_text(size = 9))
```

```{r gap-direction-plot, fig.width = 12, fig.height = 5}
ggplot(gd_coefs,
       aes(x = term, y = estimate,
           colour = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = lo, ymax = hi),
                width = 0.2, position = position_dodge(0.5), linewidth = 0.8) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_text(aes(label = sig, y = hi + 0.02),
            position = position_dodge(0.5), size = 4, vjust = 0) +
  scale_colour_manual(values = c("Linear (signed gap)" = COL_MODEL,
                                 "Logistic (gap > 0)"  = COL_CORPUS),
                      name = NULL) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(name = "Coefficient  (95% credible interval)") +
  facet_wrap(~model, scales = "free_y") +
  labs(
    title    = "What predicts the direction and magnitude of the signed gap?",
    subtitle = paste0(
      "Linear model outcome: signed_gap = mean_preference - log_freq_ratio  -  ",
      "Logistic outcome: P(gap > 0)\n",
      "All predictors centred  -  Random intercept by binomial  -  Final checkpoint  -  ",
      "Bayesian (brms)  -  * = 95% CI excludes zero")
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 10),
        axis.text.x = element_text(size = 9))
```

```{r gap-direction-by-corpus-plot, fig.width = 13, fig.height = 7}
ggplot(gd_coefs_by_corpus,
       aes(x = term, y = estimate,
           colour = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = lo, ymax = hi),
                width = 0.2, position = position_dodge(0.5), linewidth = 0.8) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_text(aes(label = sig, y = hi + 0.02),
            position = position_dodge(0.5), size = 4, vjust = 0) +
  scale_colour_manual(values = c("Linear (signed gap)" = COL_MODEL,
                                 "Logistic (gap > 0)"  = COL_CORPUS),
                      name = NULL) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(name = "Coefficient  (95% credible interval)") +
  facet_grid(corpus ~ model, scales = "free_y") +
  labs(
    title    = "What predicts the signed gap? By corpus (final checkpoint)",
    subtitle = paste0(
      "Linear outcome: signed_gap  -  Logistic outcome: P(gap > 0)\n",
      "All predictors centred  -  Random intercept by binomial  -  ",
      "Bayesian (brms)  -  * = 95% CI excludes zero")
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 10),
        axis.text.x = element_text(size = 9))
```

```{r gap-direction-pooled-by-corpus-plot, fig.width = 13, fig.height = 7}
ggplot(gd_pooled_coefs_by_corpus,
       aes(x = term, y = estimate,
           colour = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = lo, ymax = hi),
                width = 0.2, position = position_dodge(0.5), linewidth = 0.8) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_text(aes(label = sig, y = hi + 0.02),
            position = position_dodge(0.5), size = 4, vjust = 0) +
  scale_colour_manual(values = c("Linear (signed gap)" = COL_MODEL,
                                 "Logistic (gap > 0)"  = COL_CORPUS),
                      name = NULL) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(name = "Coefficient  (95% CI)") +
  facet_grid(corpus ~ model, scales = "free_y") +
  labs(
    title    = "What predicts the signed gap? By corpus (all checkpoints pooled)",
    subtitle = paste0(
      "Linear outcome: signed_gap  -  Logistic outcome: P(gap > 0)\n",
      "All predictors centred  -  Random intercept by binomial  -  ",
      "lmer / glmer  -  Wald 95% CI  -  * = CI excludes zero")
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 10),
        axis.text.x = element_text(size = 9))
```

```{r gap-direction-genpref-plot, fig.width = 11, fig.height = 5}
# Visualise the GenPref effect directly: mean signed gap binned by GenPref decile
genpref_bins <- gd_data %>%
  filter(!is.na(genpref_c)) %>%
  mutate(genpref_decile = ntile(genpref_c, 10)) %>%
  group_by(corpus, genpref_decile) %>%
  summarise(
    mean_genpref  = mean(genpref_c + mean(gd_data$genpref, na.rm = TRUE),
                         na.rm = TRUE),
    mean_gap      = mean(signed_gap,      na.rm = TRUE),
    se_gap        = sd(signed_gap, na.rm = TRUE) / sqrt(n()),
    pct_positive  = mean(signed_gap > 0,  na.rm = TRUE),
    n             = n(),
    .groups       = "drop"
  )

ggplot(genpref_bins,
       aes(x = mean_genpref, y = mean_gap, colour = corpus)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = mean_gap - 1.96 * se_gap,
                    ymax = mean_gap + 1.96 * se_gap),
                width = 0.01, linewidth = 0.7) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("BabyLM" = "#1b7837", "C4" = "#762a83"),
                      name = NULL) +
  scale_x_continuous(name = "GenPref  (binned by decile, original scale)") +
  scale_y_continuous(name = "Mean signed gap  (log-odds)") +
  facet_wrap(~corpus) +
  labs(
    title    = "Mean signed gap by GenPref decile",
    subtitle = "Positive gap = model more alpha-preferring than corpus predicts  -  All checkpoints & model sizes pooled  -  95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

---

# Summary Statistics: Over- vs. Under-Prediction Relative to Corpus Frequency

**Positive signed gap** (`mean_preference > log_freq_ratio`) means the model
prefers the alpha order *more* than relative frequency predicts  -  it is
over-predicting the alpha order.

**Negative signed gap** means the model is less alpha-biased than the corpus  -
under-predicting the alpha order.

```{r overpred-setup, cache=TRUE}
# ── All checkpoints pooled ───────────────────────────────────────────────────
gap_base <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(total_seen), total_seen > 0) %>%
  mutate(
    signed_gap  = mean_preference - log_freq_ratio,
    log_total = log(total_seen),
    freq_bin    = cut(log_total,
                      breaks = c(0, log(10), log(100), log(1000), Inf),
                      labels = c("1-10", "10-100", "100-1k", ">1k"),
                      include.lowest = TRUE),
    model_size  = factor(model_size, levels = SIZE_LEVELS)
  )

# ── Overall summary: mean signed gap per corpus x model_size ─────────────────
overall_summary <- gap_base %>%
  group_by(corpus, model_size) %>%
  summarise(
    mean_gap   = mean(signed_gap, na.rm = TRUE),
    median_gap = median(signed_gap, na.rm = TRUE),
    sd_gap     = sd(signed_gap, na.rm = TRUE),
    n          = n(),
    se         = sd_gap / sqrt(n),
    t_stat     = mean_gap / se,
    p_value    = 2 * pt(-abs(t_stat), df = n - 1),
    direction  = if_else(mean_gap > 0, "Over-predicts alpha order", "Under-predicts alpha order"),
    .groups    = "drop"
  ) %>%
  mutate(across(c(mean_gap, median_gap, sd_gap, se), ~round(.x, 4)),
         p_value = signif(p_value, 3))

cat("=== Overall over/under-prediction (signed, all checkpoints pooled) ===\n")
print(overall_summary %>%
        select(corpus, model_size, direction, mean_gap, median_gap, se, p_value, n))

# ── Overall summary: mean ABSOLUTE gap ───────────────────────────────────────
# The signed mean can cancel if some binomials are over- and others under-
# predicted. The mean absolute gap avoids this.
abs_overall_summary <- gap_base %>%
  group_by(corpus, model_size) %>%
  summarise(
    mean_abs_gap   = mean(abs(signed_gap),   na.rm = TRUE),
    median_abs_gap = median(abs(signed_gap), na.rm = TRUE),
    sd_abs_gap     = sd(abs(signed_gap),     na.rm = TRUE),
    n              = n(),
    se             = sd_abs_gap / sqrt(n),
    # One-sample t-test: is mean |gap| significantly greater than 0?
    t_stat         = mean_abs_gap / se,
    p_value        = pt(-abs(t_stat), df = n - 1, lower.tail = TRUE),
    .groups        = "drop"
  ) %>%
  mutate(across(c(mean_abs_gap, median_abs_gap, sd_abs_gap, se), ~round(.x, 4)),
         p_value    = signif(p_value, 3),
         model_size = factor(model_size, levels = SIZE_LEVELS))

cat("\n=== Overall absolute deviation from frequency expectation ===\n")
cat("(Mean |preference - log_freq_ratio|  -  not affected by cancellation)\n")
print(abs_overall_summary %>%
        select(corpus, model_size, mean_abs_gap, median_abs_gap, se, p_value, n))

# ── By frequency bin (signed and absolute) ───────────────────────────────────
freq_summary <- gap_base %>%
  group_by(corpus, model_size, freq_bin) %>%
  summarise(
    mean_gap      = mean(signed_gap,      na.rm = TRUE),
    mean_abs_gap  = mean(abs(signed_gap), na.rm = TRUE),
    se_signed     = sd(signed_gap,        na.rm = TRUE) / sqrt(n()),
    se_abs        = sd(abs(signed_gap),   na.rm = TRUE) / sqrt(n()),
    n             = n(),
    pct_over      = mean(signed_gap > 0,  na.rm = TRUE),
    .groups       = "drop"
  ) %>%
  mutate(across(c(mean_gap, mean_abs_gap, se_signed, se_abs), ~round(.x, 4)),
         pct_over   = round(pct_over, 3),
         model_size = factor(model_size, levels = SIZE_LEVELS))

cat("\n=== Signed and absolute deviation by overall frequency bin ===\n")
print(freq_summary)
```

```{r overpred-overall-plot, fig.width = 10, fig.height = 4}
# Signed gap  -  direction of bias
ggplot(overall_summary,
       aes(x = model_size, y = mean_gap,
           colour = mean_gap > 0, group = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = mean_gap - 1.96 * se,
                    ymax = mean_gap + 1.96 * se),
                width = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("TRUE" = COL_OVER, "FALSE" = COL_MODEL),
                      labels = c("TRUE" = "Over-predicts", "FALSE" = "Under-predicts"),
                      name = NULL) +
  scale_y_continuous(name = "Mean signed gap  (preference - log_freq_ratio)") +
  scale_x_discrete(name = "Model size") +
  facet_wrap(~corpus) +
  labs(title    = "Direction of bias: does the model over- or under-predict the alpha order?",
       subtitle = "All checkpoints pooled  -  Error bars = 95% CI") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r overpred-abs-overall-plot, fig.width = 10, fig.height = 4}
# Absolute gap  -  magnitude of deviation regardless of direction
ggplot(abs_overall_summary,
       aes(x = model_size, y = mean_abs_gap,
           colour = model_size, group = 1)) +
  geom_errorbar(aes(ymin = mean_abs_gap - 1.96 * se,
                    ymax = mean_abs_gap + 1.96 * se),
                width = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_y_continuous(name = "Mean |preference - log_freq_ratio|",
                     limits = c(0, NA)) +
  scale_x_discrete(name = "Model size") +
  facet_wrap(~corpus) +
  labs(
    title    = "Magnitude of deviation from frequency expectation (absolute gap)",
    subtitle = "Not affected by cancellation of opposite errors  -  All checkpoints pooled  -  Error bars = 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

```{r overpred-byfreq-plot, fig.width = 12, fig.height = 9}
# Side-by-side: signed and absolute gap by frequency bin
freq_long <- freq_summary %>%
  pivot_longer(cols      = c(mean_gap, mean_abs_gap),
               names_to  = "metric",
               values_to = "value") %>%
  mutate(
    se_val = if_else(metric == "mean_gap", se_signed, se_abs),
    metric = factor(metric,
                    levels = c("mean_gap", "mean_abs_gap"),
                    labels = c("Signed gap\n(cancellation possible)",
                               "Absolute gap\n(no cancellation)"))
  )

ggplot(freq_long,
       aes(x = freq_bin, y = value,
           colour = model_size, group = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = value - 1.96 * se_val,
                    ymax = value + 1.96 * se_val),
                width = 0.2, position = position_dodge(0.4), linewidth = 0.7) +
  geom_line(linewidth = 0.8, position = position_dodge(0.4)) +
  geom_point(size = 3,   position = position_dodge(0.4)) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_discrete(name = "Times binomial seen in corpus  (total alpha + beta)") +
  scale_y_continuous(name = "Gap (log-odds)") +
  facet_grid(metric ~ corpus, scales = "free_y") +
  labs(
    title    = "Signed vs. absolute deviation from frequency expectation by exposure level",
    subtitle = "All checkpoints pooled  -  Error bars = 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.text.y = element_text(size = 9, angle = 0, hjust = 0))
```

## Reconciling the Frequency-Bin Plot with the Scatter Plot

The frequency-bin plot uses four bins (1-10, 10-100, 100-1k, **>1k**), where the
final bin collapses `total_seen` from 1,000 to 100,000+ into a single average.
If the gap peaks somewhere in the 1k-10k range and then declines at very high
exposure (as the continuous LOESS suggests), the ">1k" bin average will be
pulled downward by the dense right tail, making it appear smaller than the
"100-1k" bin  -  producing an apparent contradiction.

The plot below breaks the ">1k" bin into finer sub-bins to reveal whether the
relationship is non-monotone at high frequency.

```{r overpred-finebins, cache=TRUE, fig.width = 13, fig.height = 9}
fine_summary <- gap_base %>%
  mutate(
    freq_bin_fine = cut(
      log_total,
      breaks = c(0, log(10), log(100), log(1000), log(3000), log(10000), log(30000), log(100000), Inf),
      labels = c("1-10", "10-100", "100-1k",
                 "1k-3k", "3k-10k", "10k-30k", "30k-100k", ">100k"),
      include.lowest = TRUE
    )
  ) %>%
  group_by(corpus, model_size, freq_bin_fine) %>%
  summarise(
    mean_abs_gap = mean(abs(signed_gap), na.rm = TRUE),
    se_abs       = sd(abs(signed_gap),   na.rm = TRUE) / sqrt(n()),
    n            = n(),
    .groups      = "drop"
  ) %>%
  mutate(model_size = factor(model_size, levels = SIZE_LEVELS))

ggplot(fine_summary,
       aes(x = freq_bin_fine, y = mean_abs_gap,
           colour = model_size, group = model_size)) +
  geom_errorbar(aes(ymin = mean_abs_gap - 1.96 * se_abs,
                    ymax = mean_abs_gap + 1.96 * se_abs),
                width = 0.3, position = position_dodge(0.5), linewidth = 0.7) +
  geom_line(linewidth = 0.9, position = position_dodge(0.5)) +
  geom_point(size = 3,    position = position_dodge(0.5)) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_discrete(name = "Times binomial seen in corpus  (total alpha + beta)") +
  scale_y_continuous(name = "|preference - log_freq_ratio|  (log-odds)",
                     limits = c(0, NA)) +
  facet_wrap(~corpus) +
  labs(
    title    = "Absolute gap by fine-grained frequency bin   -   reconciling bin plot vs. scatter LOESS",
    subtitle = paste0(
      "The coarse '>1k' bin hid the high-frequency tail. If gap peaks at 1k-10k\n",
      "then declines, the '>1k' average is pulled down  -  explaining the apparent contradiction.\n",
      "All checkpoints pooled  -  Error bars = 95% CI")
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 35, hjust = 1)
  )
```

```{r overpred-pct-plot, fig.width = 12, fig.height = 6}
# What proportion of binomials are over-predicted at each frequency level?
ggplot(freq_summary,
       aes(x = freq_bin, y = pct_over,
           colour = model_size, group = model_size)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.8, position = position_dodge(0.4)) +
  geom_point(size = 3,   position = position_dodge(0.4)) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_discrete(name = "Times binomial seen in corpus  (total alpha + beta)") +
  scale_y_continuous(name = "Proportion of binomials over-predicted",
                     labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  facet_wrap(~corpus) +
  labs(
    title    = "Proportion of binomials where model is more alpha-preferring than relative frequency predicts",
    subtitle = "Dashed line = 50% (no systematic bias)  -  All checkpoints pooled"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```
