---
title: "Corpus Exposure vs. Model Preference"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Overview

This analysis examines whether each model's **continuous preference score**
(`alpha_logprob - nonalpha_logprob`) tracks the training frequency ratio for
each binomial at each checkpoint.

The key idea: if a model has seen "bread and butter" 60 times and "butter and
bread" 40 times, we'd expect its log-probability difference to lean positive in
proportion to `log(60/40)`. Concretely:

- **Observed preference**: `preference = alpha_logprob - nonalpha_logprob`
  (continuous; positive = model favours alpha ordering)
- **Expected log frequency ratio**: `log(alpha_seen / beta_seen)`
  (positive = alpha ordering was more frequent in training)

If the model perfectly tracks relative frequency, the two should be positively
correlated and the relationship should be linear. We also ask how this
relationship changes as a function of **total frequency** (`alpha_seen +
beta_seen`) and over the course of **training**.

---

## Setup

```{r libraries}
library(tidyverse)
library(duckdb)
library(scales)
library(patchwork)
library(brms)
```

```{r brms-config}
# ── Shared brms settings ──────────────────────────────────────────────────────
MODELS_DIR  <- "models"
dir.create(MODELS_DIR, showWarnings = FALSE, recursive = TRUE)

BRMS_ITER   <- 6000
BRMS_WARMUP <- 3000
BRMS_CHAINS <- 4
BRMS_CORES  <- 4
BRMS_SEED   <- 42

# Uninformative priors centred at 0 for continuous outcomes
BRMS_PRIORS <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(0, 1), class = "Intercept"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "sigma")
)

# Same but without sigma (for bernoulli/binomial families)
BRMS_PRIORS_LOGISTIC <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(0, 1), class = "Intercept"),
  prior(exponential(1), class = "sd")
)
```

```{r helpers}
parse_corpus <- function(x) {
  case_when(
    grepl("babylm", x, ignore.case = TRUE) ~ "BabyLM",
    grepl("opt-c4", x, ignore.case = TRUE) ~ "C4",
    TRUE ~ NA_character_
  )
}

parse_size <- function(x) {
  case_when(
    grepl("125m", x, ignore.case = TRUE) ~ "125M",
    grepl("350m", x, ignore.case = TRUE) ~ "350M",
    grepl("1\\.3b", x, ignore.case = TRUE) ~ "1.3B",
    TRUE ~ NA_character_
  )
}

SIZE_LEVELS  <- c("125M", "350M", "1.3B")
SIZE_COLOURS <- c("125M" = "#1b7837", "350M" = "#762a83", "1.3B" = "#b2182b")
```

---

## Load and Aggregate Data

DuckDB scans the ~28 GB CSV and aggregates to one row per
model x checkpoint x binomial, computing the **mean preference score** across
the 51 prompts. The log frequency ratio is computed here as well.

```{r load-duckdb, cache=TRUE}
csv_path <- normalizePath(
  "../Data/processed/checkpoint_results_with_exposures.csv",
  mustWork = TRUE
)
csv_path_fwd <- gsub("\\\\", "/", csv_path)

con <- dbConnect(duckdb::duckdb())
dbExecute(con, sprintf(
  "CREATE VIEW ckpt AS SELECT * FROM read_csv_auto('%s')",
  csv_path_fwd
))

message("Aggregating via DuckDB ...")
agg <- dbGetQuery(con, "
  SELECT
    model,
    tokens,
    binom,
    ROUND(alpha_seen) AS alpha_seen,
    ROUND(beta_seen)  AS beta_seen,
    ROUND(alpha_seen) + ROUND(beta_seen)         AS total_seen,
    -- log frequency ratio: positive = alpha was more frequent in training
    CASE WHEN alpha_seen > 0 AND beta_seen > 0
         THEN LN(alpha_seen / beta_seen)
         ELSE NULL END                           AS log_freq_ratio,
    AVG(preference)                              AS mean_preference,
    COUNT(*)                                     AS n_prompts
  FROM ckpt
  GROUP BY model, tokens, binom, alpha_seen, beta_seen
  ORDER BY model, tokens, binom
")

dbDisconnect(con, shutdown = TRUE)
message(sprintf("Aggregated to %s rows.", format(nrow(agg), big.mark = ",")))
```

```{r metadata}
agg <- agg %>%
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS),
    total_seen = as.numeric(total_seen)
  )

cat(sprintf("Rows with total_seen > 0:           %s (%.1f%%)\n",
            format(sum(agg$total_seen > 0), big.mark = ","),
            100 * mean(agg$total_seen > 0)))
cat(sprintf("Rows with both orderings seen (log ratio defined): %s (%.1f%%)\n",
            format(sum(!is.na(agg$log_freq_ratio)), big.mark = ","),
            100 * mean(!is.na(agg$log_freq_ratio))))
```

---

## Plot 1: Does the Log Frequency Ratio Predict the Preference Score?

For each model x checkpoint x binomial where both orderings appeared at least
once, we ask whether `log(alpha_seen / beta_seen)` predicts the preference score.

If the model tracks relative frequency, the relationship should be monotonically
increasing and pass through zero at x = 0 (equal exposure -> no preference).

Two versions are shown: (1a) raw scatter with a LOESS smoother per model size,
and (1b) binned summary with +/-1 SE ribbons.

```{r plot1-raw-data}
p1_raw <- agg %>%
  filter(!is.na(log_freq_ratio), !is.na(model_size))
```

### Plot 1a: Raw scatter + LOESS smoother

```{r plot1a-scatter, fig.width = 10, fig.height = 5}
ggplot(p1_raw, aes(x = log_freq_ratio, y = mean_preference,
                   colour = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.04, size = 0.4, show.legend = FALSE) +
  geom_smooth(method = "loess", span = 0.4, se = TRUE, linewidth = 1,
              aes(fill = model_size)) +
  scale_x_continuous(name = "log(alpha_seen / beta_seen)  [training frequency log-ratio]") +
  scale_y_continuous(name = "Mean preference score  (alpha_logprob - nonalpha_logprob)") +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus) +
  labs(
    title    = "Does training frequency log-ratio predict model preference score?",
    subtitle = "Each point = one model x checkpoint x binomial  -  LOESS smoother +/- 95% CI per model size"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

### Plot 1b: Binned summary (raw points in background)

```{r plot1b-binned, fig.width = 10, fig.height = 5}
p1_data <- p1_raw %>%
  mutate(
    lrat_bin = cut(log_freq_ratio,
                   breaks = c(-Inf, -3, -2, -1, -0.5, 0,
                                0.5,  1,  2,  3,  Inf),
                   labels = c("<-3", "-3--2", "-2--1", "-1--.5",
                              "-.5-0", "0-.5", ".5-1", "1-2",
                              "2-3", ">3"))
  ) %>%
  group_by(corpus, model_size, lrat_bin) %>%
  summarize(
    mean_pref = mean(mean_preference, na.rm = TRUE),
    se        = sd(mean_preference, na.rm = TRUE) / sqrt(n()),
    midpoint  = mean(log_freq_ratio, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  )

ggplot(p1_data, aes(x = midpoint, y = mean_pref,
                    colour = model_size, fill = model_size,
                    group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(data = p1_raw,
             aes(x = log_freq_ratio, y = mean_preference, colour = model_size),
             alpha = 0.04, size = 0.4, inherit.aes = FALSE, show.legend = FALSE) +
  geom_ribbon(aes(ymin = mean_pref - se, ymax = mean_pref + se),
              alpha = 0.25, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_x_continuous(name = "log(alpha_seen / beta_seen)  [training frequency log-ratio]") +
  scale_y_continuous(name = "Mean preference score  (alpha_logprob - nonalpha_logprob)") +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus) +
  labs(
    title    = "Does training frequency log-ratio predict model preference score?",
    subtitle = "Grey cloud = raw data  -  Lines = bin means +/- 1 SE  -  Dashed lines at x = 0, y = 0"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Plot 2: Preference-Frequency Correlation vs. Total Exposure

Does seeing more tokens of a binomial strengthen the relationship between
preference and log frequency ratio? We compute the Pearson correlation between
`mean_preference` and `log_freq_ratio` within bins of `log10(total_seen)`.

```{r plot2-corr-vs-freq, fig.width = 10, fig.height = 5}
p2_data <- agg %>%
  filter(!is.na(log_freq_ratio), total_seen > 0) %>%
  mutate(
    log_total = log(total_seen),
    freq_bin  = cut(log_total,
                    breaks = c(0, 0.5, 1, 1.5, 2, 2.5, 3, Inf),
                    labels = c("1-3", "3-10", "10-32", "32-100",
                               "100-316", "316-1k", ">1k"),
                    include.lowest = TRUE)
  ) %>%
  group_by(corpus, model_size, freq_bin) %>%
  summarize(
    r        = cor(mean_preference, log_freq_ratio, use = "complete.obs"),
    midpoint = mean(log_total, na.rm = TRUE),
    n        = n(),
    .groups  = "drop"
  )

ggplot(p2_data, aes(x = midpoint, y = r,
                    colour = model_size, fill = model_size,
                    group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_x_continuous(
    name   = "Total frequency, either ordering (log10 scale)",
    breaks = 0:4,
    labels = c("1", "10", "100", "1k", "10k")
  ) +
  scale_y_continuous(
    name   = "Pearson r  (preference ~ log frequency ratio)",
    limits = c(-0.1, 1)
  ) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus) +
  labs(
    title    = "Preference-frequency correlation vs. total exposure",
    subtitle = "Higher r = model preference score tracks training frequency ratio more closely"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Plot 3: Correlation Between Preference and Log Frequency Ratio Over Training

Does the model's preference score come to track training frequency more closely
as training progresses? For each checkpoint we compute the Pearson r across all
binomials where both orderings were seen.

```{r plot3-corr-over-training, fig.width = 11, fig.height = 5}
p3_data <- agg %>%
  filter(!is.na(log_freq_ratio)) %>%
  group_by(corpus, model_size, tokens) %>%
  summarize(
    r        = cor(mean_preference, log_freq_ratio, use = "complete.obs"),
    n_binoms = n(),
    .groups  = "drop"
  )

ggplot(p3_data, aes(x = tokens / 1e9, y = r,
                    colour = model_size, fill = model_size,
                    group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(
    name   = "Pearson r  (preference ~ log frequency ratio)",
    limits = c(-0.1, 1)
  ) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus, scales = "free_x") +
  labs(
    title    = "Preference-frequency correlation over training",
    subtitle = "r computed across binomials where both orderings appear in training at that checkpoint"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r plot3-n-binoms, fig.width = 11, fig.height = 4}
# Companion: how many binomials contribute at each point?
ggplot(p3_data, aes(x = tokens / 1e9, y = n_binoms,
                    colour = model_size, group = model_size)) +
  geom_line(linewidth = 0.7) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "# binomials with both orderings seen", labels = comma) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus, scales = "free_x") +
  labs(title = "Number of binomials contributing to the correlation at each checkpoint") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Plot 4: Regression Slope  -  How Well Does Preference Scale with Log Frequency Ratio?

Beyond correlation (direction + monotonicity), we also care about **scaling**:
does each unit increase in `log(alpha_seen / beta_seen)` produce a consistent
shift in the preference score? We fit `mean_preference ~ log_freq_ratio` at
each checkpoint and track the slope over training.

A steeper slope means a stronger, more calibrated frequency signal in the
model's preferences.

```{r plot4-slope-over-training, fig.width = 11, fig.height = 5}
p4_data <- agg %>%
  filter(!is.na(log_freq_ratio)) %>%
  group_by(corpus, model_size, tokens) %>%
  group_map(function(g, keys) {
    if (nrow(g) < 5 || sd(g$log_freq_ratio, na.rm = TRUE) == 0) return(NULL)
    fit <- lm(mean_preference ~ log_freq_ratio, data = g)
    cf  <- coef(summary(fit))
    tibble(
      corpus     = keys$corpus,
      model_size = keys$model_size,
      tokens     = keys$tokens,
      slope      = cf["log_freq_ratio", "Estimate"],
      se         = cf["log_freq_ratio", "Std. Error"]
    )
  }, .keep = TRUE) %>%
  bind_rows()

ggplot(p4_data, aes(x = tokens / 1e9, y = slope,
                    colour = model_size, fill = model_size,
                    group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = slope - se, ymax = slope + se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "Regression slope  (preference ~ log frequency ratio)") +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus, scales = "free_x") +
  labs(
    title    = "Slope of preference ~ log frequency ratio over training",
    subtitle = "Steeper slope = stronger, more calibrated frequency signal  -  +/-1 SE ribbon"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Summary Table

```{r summary}
agg %>%
  filter(!is.na(log_freq_ratio)) %>%
  group_by(corpus, model_size) %>%
  summarize(
    overall_r      = cor(mean_preference, log_freq_ratio, use = "complete.obs"),
    n_observations = n(),
    .groups        = "drop"
  ) %>%
  arrange(corpus, model_size) %>%
  mutate(overall_r = round(overall_r, 3)) %>%
  knitr::kable(
    col.names = c("Corpus", "Model size", "Overall r", "N observations"),
    caption   = paste("Pearson r between mean preference score and log frequency ratio,",
                      "pooled across all checkpoints.")
  )
```

---

## Generative Preference (GenPref)

**GenPref** is a linguistic/generative preference score for each binomial derived
from a logistic regression over phonological and semantic constraint scores
(Form, Percept, Culture, Power, Intense, Icon, Freq, Len, Lapse, BStress).
It ranges from roughly -0.5 to 0.5, where positive = alpha ordering is
linguistically preferred, negative = non-alpha ordering is preferred.

The hypothesis: even when a model's training corpus diverges from what GenPref
predicts, the model may still be *pulled* toward GenPref  -  reflecting an
inductive bias toward canonical (natural) orderings.

```{r load-genpref}
human_path <- normalizePath(
  "../Data/all_human_data.csv",
  mustWork = TRUE
)

# Constraint-score logistic formula for recomputing GenPref where missing
GENPREF_FORMULA <- function(Form, Percept, Culture, Power, Intense,
                            Icon, Freq, Len, Lapse, BStress) {
  y <- 0.02191943 +
       0.23925834 * Form    +
       0.24889543 * Percept +
       0.41836997 * Culture +
       0.25967334 * Power   +
       0.01867604 * Intense +
       1.30365980 * Icon    +
       0.08553552 * Freq    +
       0.15241566 * Len     -
       0.19381657 * Lapse   +
       0.36019221 * BStress
  (1 / (1 + exp(-y))) - 0.5
}

genpref_raw <- read_csv(human_path, show_col_types = FALSE) %>%
  rename(BStress = `*BStress`) %>%          # column header has leading *
  select(Alpha, Form, Percept, Culture, Power, Intense,
         Icon, Freq, Len, Lapse, BStress, GenPref) %>%
  # One row per unique binomial (participants x items -> items)
  group_by(Alpha) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    # The formula outputs plogis(y) - 0.5, which is on [-0.5, 0.5].
    # Stored GenPref values may be on [0, 1] (raw human proportion preferring
    # alpha). Normalise stored values to [-0.5, 0.5] by subtracting 0.5 so
    # that all genpref values are on a consistent scale throughout the analysis.
    GenPref_norm = case_when(
      is.na(GenPref)  ~ NA_real_,
      GenPref > 0.5   ~ GenPref - 0.5,   # [0, 1] -> [-0.5, 0.5]
      GenPref < -0.5  ~ GenPref,          # already outside [0,1]; leave as-is and flag below
      TRUE            ~ GenPref - 0.5     # [0, 1] -> [-0.5, 0.5]
    ),
    genpref = if_else(
      is.na(GenPref),
      GENPREF_FORMULA(Form, Percept, Culture, Power, Intense,
                      Icon, Freq, Len, Lapse, BStress),
      GenPref_norm
    ),
    binom_lower = tolower(Alpha)
  ) %>%
  select(binom_lower, genpref)

# ── Diagnostic: confirm genpref is on [-0.5, 0.5] ────────────────────────────
gp_range <- range(genpref_raw$genpref, na.rm = TRUE)
cat(sprintf("Unique binomials with GenPref: %d  (NA remaining: %d)\n",
            sum(!is.na(genpref_raw$genpref)),
            sum( is.na(genpref_raw$genpref))))
cat(sprintf("GenPref range after normalisation: [%.4f, %.4f]\n",
            gp_range[1], gp_range[2]))
if (gp_range[1] < -0.5 - 1e-6 || gp_range[2] > 0.5 + 1e-6) {
  warning(sprintf(
    "GenPref values outside [-0.5, 0.5]: min = %.4f, max = %.4f. ",
    gp_range[1], gp_range[2]
  ))
} else {
  cat("OK: all GenPref values are within [-0.5, 0.5].\n")
}
```

```{r join-genpref}
# Collapse agg to one row per binomial x corpus (average over checkpoints)
# then join GenPref
agg_gp <- agg %>%
  mutate(binom_lower = tolower(binom)) %>%
  left_join(genpref_raw, by = "binom_lower")

cat(sprintf("Rows with GenPref after join: %s / %s  (%.1f%%)\n",
            format(sum(!is.na(agg_gp$genpref)), big.mark = ","),
            format(nrow(agg_gp),                big.mark = ","),
            100 * mean(!is.na(agg_gp$genpref))))
```

---

## Plot 5: GenPref vs. Mean Preference Score

Does the model's preference score correlate with the linguistic/generative
preference, independent of relative frequency? We average across all checkpoints
and frequency conditions, then plot.

```{r plot5-genpref-vs-pref, fig.width = 10, fig.height = 5}
p5_data <- agg_gp %>%
  filter(!is.na(genpref)) %>%
  mutate(
    gp_bin = cut(genpref,
                 breaks = c(-Inf, -0.3, -0.15, -0.05, 0.05, 0.15, 0.3, Inf),
                 labels = c("<-0.3", "-0.3--0.15", "-0.15--0.05",
                            "-0.05-0.05", "0.05-0.15", "0.15-0.3", ">0.3"))
  ) %>%
  group_by(corpus, model_size, gp_bin) %>%
  summarize(
    mean_pref = mean(mean_preference, na.rm = TRUE),
    se        = sd(mean_preference, na.rm = TRUE) / sqrt(n()),
    midpoint  = mean(genpref, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  )

ggplot(p5_data, aes(x = midpoint, y = mean_pref,
                    colour = model_size, fill = model_size,
                    group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = mean_pref - se, ymax = mean_pref + se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_x_continuous(name = "GenPref  (linguistic/generative preference score)") +
  scale_y_continuous(name = "Mean preference score  (alpha_logprob - nonalpha_logprob)") +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus) +
  labs(
    title    = "Does linguistic preference (GenPref) predict model preference?",
    subtitle = "Averaged over all checkpoints  -  +/-1 SE ribbon"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Plot 6: The "Pull"  -  GenPref x Log Frequency Ratio -> Preference

Key question: when relative frequency and GenPref **agree**, does the model show
stronger preference than when they **conflict**? We split binomials into three
GenPref levels (negative / neutral / positive) and plot mean preference as a
function of `log(alpha_seen / beta_seen)`.

If GenPref adds an independent pull, binomials with positive GenPref should sit
*above* binomials with negative GenPref at the same frequency ratio.

```{r plot6-pull, fig.width = 11, fig.height = 6}
p6_data <- agg_gp %>%
  filter(!is.na(genpref), !is.na(log_freq_ratio)) %>%
  mutate(
    gp_level = case_when(
      genpref >  0.05 ~ "GenPref positive (>0.05)",
      genpref < -0.05 ~ "GenPref negative (<-0.05)",
      TRUE            ~ "GenPref neutral (-0.05-0.05)"
    ),
    gp_level = factor(gp_level,
                      levels = c("GenPref negative (<-0.05)",
                                 "GenPref neutral (-0.05-0.05)",
                                 "GenPref positive (>0.05)")),
    lrat_bin = cut(log_freq_ratio,
                   breaks = c(-Inf, -3, -2, -1, -0.5, 0,
                                0.5,  1,  2,  3,  Inf),
                   labels = c("<-3", "-3--2", "-2--1", "-1--.5",
                              "-.5-0", "0-.5", ".5-1", "1-2",
                              "2-3", ">3"))
  ) %>%
  group_by(corpus, model_size, gp_level, lrat_bin) %>%
  summarize(
    mean_pref = mean(mean_preference, na.rm = TRUE),
    se        = sd(mean_preference, na.rm = TRUE) / sqrt(n()),
    midpoint  = mean(log_freq_ratio, na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  )

GP_COLOURS <- c(
  "GenPref negative (<-0.05)"   = "#d73027",
  "GenPref neutral (-0.05-0.05)" = "#999999",
  "GenPref positive (>0.05)"    = "#4575b4"
)

ggplot(p6_data, aes(x = midpoint, y = mean_pref,
                    colour = gp_level, fill = gp_level,
                    group  = gp_level)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = mean_pref - se, ymax = mean_pref + se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous(name = "log(alpha_seen / beta_seen)  [training frequency log-ratio]") +
  scale_y_continuous(name = "Mean preference score  (alpha_logprob - nonalpha_logprob)") +
  scale_colour_manual(values = GP_COLOURS, name = "GenPref level") +
  scale_fill_manual(values   = GP_COLOURS, name = "GenPref level") +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "GenPref 'pull': does linguistic preference shift model output beyond frequency alone?",
    subtitle = "If GenPref adds signal, coloured lines should be vertically separated at the same x"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Plot 7: Conflict Cases  -  Does GenPref Pull Away from Frequency?

Here we isolate **conflict binomials** where relative frequency and GenPref point
in opposite directions (i.e., `sign(log_freq_ratio) != sign(genpref)`). Among
these, we ask: does the strength of GenPref predict how far the model's
preference *deviates* from what frequency alone would predict?

We define:
- **Frequency-predicted preference**: the fitted value from the regression
  `mean_preference ~ log_freq_ratio` (using the final checkpoint only)
- **Residual preference**: `mean_preference - fitted` (positive = pulled toward GenPref)

```{r plot7-conflict, fig.width = 10, fig.height = 5}
# Use the last checkpoint for each model (highest token count)
last_ckpt <- agg_gp %>%
  filter(!is.na(genpref), !is.na(log_freq_ratio)) %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup()

# Fit frequency baseline per model and compute residual
last_ckpt_resid <- last_ckpt %>%
  group_by(corpus, model_size) %>%
  group_map(function(g, keys) {
    if (nrow(g) < 10) return(NULL)
    fit <- lm(mean_preference ~ log_freq_ratio, data = g)
    g %>%
      mutate(
        fitted_by_freq = fitted(fit),
        resid_from_freq = mean_preference - fitted_by_freq,
        corpus     = keys$corpus,
        model_size = keys$model_size
      )
  }, .keep = TRUE) %>%
  bind_rows() %>%
  # Conflict cases only: frequency and GenPref in opposite directions
  filter(
    !is.na(genpref),
    sign(log_freq_ratio) != 0,
    sign(genpref) != 0,
    sign(log_freq_ratio) != sign(genpref)
  ) %>%
  mutate(
    # Flip residual so positive = pulled toward GenPref direction
    pull_toward_genpref = resid_from_freq * sign(genpref),
    abs_genpref_bin = cut(abs(genpref),
                          breaks = c(0, 0.05, 0.1, 0.2, 0.3, Inf),
                          labels = c("0-0.05", "0.05-0.1", "0.1-0.2",
                                     "0.2-0.3", ">0.3"))
  )

p7_data <- last_ckpt_resid %>%
  group_by(corpus, model_size, abs_genpref_bin) %>%
  summarize(
    mean_pull = mean(pull_toward_genpref, na.rm = TRUE),
    se        = sd(pull_toward_genpref,   na.rm = TRUE) / sqrt(n()),
    midpoint  = mean(abs(genpref),        na.rm = TRUE),
    n         = n(),
    .groups   = "drop"
  )

ggplot(p7_data, aes(x = midpoint, y = mean_pull,
                    colour = model_size, fill = model_size,
                    group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = mean_pull - se, ymax = mean_pull + se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_x_continuous(name = "|GenPref|  (strength of linguistic preference)") +
  scale_y_continuous(name = "Mean pull toward GenPref\n(residual from frequency baseline, sign-flipped)") +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_wrap(~corpus) +
  labs(
    title    = "Do conflict binomials get pulled toward GenPref?",
    subtitle = paste("Conflict = frequency and GenPref disagree  -",
                     "Positive y = model leans toward GenPref despite frequency signal  -  +/-1 SE")
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r plot7-n-conflict}
cat("N conflict binomials per corpus x model size (at final checkpoint):\n")
last_ckpt_resid %>%
  count(corpus, model_size) %>%
  knitr::kable(col.names = c("Corpus", "Model size", "N conflict binomials"))
```

---

## Plot 8: GenPref vs. Frequency Signal Over Training, by Exposure Level

This plot brings the three threads together: **training dynamics**, **relative pull
of GenPref vs. relative frequency**, and **total exposure**.

At each checkpoint, binomials are placed into frequency bins based on how many
times the model has seen them *at that point in training* (`total_seen`). Within
each bin we fit:

$$\text{mean\_preference} \sim \beta_{\text{freq}} \cdot z(\log\text{freq\_ratio}) + \beta_{\text{gp}} \cdot z(\text{GenPref})$$

Both predictors are z-scored so that $\beta_{\text{freq}}$ and $\beta_{\text{gp}}$
are directly comparable. We then plot both coefficients over training, split by
exposure level.

If relative frequency gradually takes over from GenPref as the model accumulates
evidence, we expect $\beta_{\text{freq}}$ to rise and $\beta_{\text{gp}}$ to
shrink  -  particularly for the **high-frequency bins** where the frequency signal
is strongest.

```{r plot8-pushpull-over-training, fig.width = 12, fig.height = 8}
FREQ_BREAKS  <- c(0, 1, 2, 3, Inf)
FREQ_LABELS  <- c("1-10 tokens", "10-100 tokens", "100-1k tokens", ">1k tokens")

p8_data <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(genpref), total_seen >= 1) %>%
  mutate(
    freq_bin = cut(log(total_seen),
                   breaks = FREQ_BREAKS,
                   labels = FREQ_LABELS,
                   include.lowest = TRUE)
  ) %>%
  group_by(corpus, model_size, tokens, freq_bin) %>%
  group_map(function(g, keys) {
    if (nrow(g) < 10) return(NULL)
    # Check SD before scaling  -  scale() returns NaN (not 0) when SD = 0,
    # and NaN == 0 evaluates to NA which breaks if()
    sd_lograt  <- sd(g$log_freq_ratio, na.rm = TRUE)
    sd_genpref <- sd(g$genpref,        na.rm = TRUE)
    if (is.na(sd_lograt)  || sd_lograt  == 0) return(NULL)
    if (is.na(sd_genpref) || sd_genpref == 0) return(NULL)
    z_lograt  <- scale(g$log_freq_ratio)[, 1]
    z_genpref <- scale(g$genpref)[, 1]
    fit <- lm(g$mean_preference ~ z_lograt + z_genpref)
    cf  <- coef(summary(fit))
    bind_rows(
      tibble(
        corpus     = keys$corpus,
        model_size = keys$model_size,
        tokens     = keys$tokens,
        freq_bin   = keys$freq_bin,
        predictor  = "RelFreq\n(log freq ratio)",
        slope      = cf["z_lograt",  "Estimate"],
        se         = cf["z_lograt",  "Std. Error"]
      ),
      tibble(
        corpus     = keys$corpus,
        model_size = keys$model_size,
        tokens     = keys$tokens,
        freq_bin   = keys$freq_bin,
        predictor  = "Linguistic preference\n(GenPref)",
        slope      = cf["z_genpref", "Estimate"],
        se         = cf["z_genpref", "Std. Error"]
      )
    )
  }, .keep = TRUE) %>%
  bind_rows() %>%
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    freq_bin   = factor(freq_bin,   levels = FREQ_LABELS)
  )

PRED_COLOURS <- c(
  "RelFreq\n(log freq ratio)" = "#e66101",
  "Linguistic preference\n(GenPref)"   = "#5e3c99"
)

ggplot(p8_data, aes(x = tokens / 1e9, y = slope,
                    colour   = predictor,
                    linetype = model_size,
                    group    = interaction(predictor, model_size))) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.75, alpha = 0.9) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "Standardised regression coefficient\n(preference ~ z(log freq ratio) + z(GenPref))") +
  scale_colour_manual(values = PRED_COLOURS, name = "Predictor") +
  scale_linetype_manual(
    values = c("125M" = "dotted", "350M" = "dashed", "1.3B" = "solid"),
    name   = "Model size"
  ) +
  facet_grid(freq_bin ~ corpus, scales = "free_x") +
  labs(
    title    = "Push-and-pull of relative frequency vs. linguistic preference over training",
    subtitle = paste("Both predictors z-scored  -  coefficients directly comparable  -",
                     "Rows = exposure level at each checkpoint")
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 9)
  )
```

---

## Plot 9: Three Attractors  -  Where GenPref, Frequency, and the Model All Land

The key question made visual: for each binomial, where is the model's preference
relative to its two "attractors"  -  the linguistic preference (GenPref) and the
relative frequency signal?

**How values are made comparable:** each predictor is multiplied by its
pooled OLS regression slope on `mean_preference` (fit across all checkpoints
within corpus x model size), putting GenPref and log frequency ratio on the
same log-probability scale as the actual preference. All three are then
mean-centred so the origin (0) represents the average preference.

Two sub-plots:

- **9a (dumbbell):** At the final checkpoint, binomials are grouped into
  GenPref deciles. Each row shows three dots  -  where frequency expects the
  preference to be, where GenPref expects it to be, and where it actually is  - 
  connected by a line so you can see the "tug-of-war."
- **9b (trajectory):** The three averaged signals over training, showing
  whether the model drifts toward the frequency attractor as it trains.

```{r plot9-setup}
ATTRACTOR_COLS <- c(
  "Frequency expectation" = "#e66101",
  "Actual preference"     = "#404040",
  "GenPref expectation"   = "#5e3c99"
)

# Pooled regression slopes (all checkpoints, within corpus x model_size)
# These act as "exchange rates" to put log_freq_ratio and genpref
# on the same log-probability scale as mean_preference.
attractor_slopes <- agg_gp %>%
  filter(!is.na(genpref), !is.na(log_freq_ratio)) %>%
  group_by(corpus, model_size) %>%
  summarize(
    b_freq = coef(lm(mean_preference ~ log_freq_ratio))["log_freq_ratio"],
    b_gp   = coef(lm(mean_preference ~ genpref))["genpref"],
    .groups = "drop"
  )

# Scale predictors to preference units, then mean-centre all three
attractors <- agg_gp %>%
  filter(!is.na(genpref), !is.na(log_freq_ratio)) %>%
  left_join(attractor_slopes, by = c("corpus", "model_size")) %>%
  mutate(
    freq_scaled = b_freq * log_freq_ratio,
    gp_scaled   = b_gp   * genpref
  ) %>%
  group_by(corpus, model_size) %>%
  mutate(
    actual_c = mean_preference - mean(mean_preference, na.rm = TRUE),
    freq_c   = freq_scaled     - mean(freq_scaled,     na.rm = TRUE),
    gp_c     = gp_scaled       - mean(gp_scaled,       na.rm = TRUE)
  ) %>%
  ungroup()
```

### Plot 9a: Dumbbell at the final checkpoint

```{r plot9a-dumbbell, fig.width = 11, fig.height = 7}
dumbbell_data <- attractors %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup() %>%
  group_by(corpus, model_size) %>%
  mutate(gp_decile = ntile(genpref, 10)) %>%
  ungroup() %>%
  group_by(corpus, model_size, gp_decile) %>%
  summarize(
    gp_mid  = mean(genpref,  na.rm = TRUE),
    actual  = mean(actual_c, na.rm = TRUE),
    freq    = mean(freq_c,   na.rm = TRUE),
    genpref = mean(gp_c,     na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(actual, freq, genpref),
               names_to = "signal", values_to = "value") %>%
  mutate(signal = factor(signal,
                         levels = c("freq", "actual", "genpref"),
                         labels = c("Frequency expectation",
                                    "Actual preference",
                                    "GenPref expectation")))

# Range segment per row (connects min to max of the three dots)
dumbbell_range <- dumbbell_data %>%
  group_by(corpus, model_size, gp_decile, gp_mid) %>%
  summarize(lo = min(value), hi = max(value), .groups = "drop")

ggplot(dumbbell_data,
       aes(x = value,
           y = reorder(factor(gp_decile), gp_mid),
           colour = signal)) +
  geom_segment(data = dumbbell_range,
               aes(x = lo, xend = hi,
                   y = reorder(factor(gp_decile), gp_mid),
                   yend = reorder(factor(gp_decile), gp_mid)),
               colour = "grey70", linewidth = 0.7,
               inherit.aes = FALSE) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3.5) +
  scale_x_continuous(name = "Value (preference-scale units, mean-centred)") +
  scale_y_discrete(name = "GenPref decile  (low <- -> high)") +
  scale_colour_manual(values = ATTRACTOR_COLS, name = NULL) +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "Three attractors at the final checkpoint",
    subtitle = paste("Each row = one GenPref decile bin  -",
                     "Dots: where frequency, GenPref, and the model's actual preference each fall")
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

### Plot 9b: Trajectories over training

```{r plot9b-trajectory, fig.width = 11, fig.height = 5}
traj_data <- attractors %>%
  group_by(corpus, model_size, tokens) %>%
  summarize(
    `Frequency expectation` = mean(freq_c,   na.rm = TRUE),
    `Actual preference`     = mean(actual_c, na.rm = TRUE),
    `GenPref expectation`   = mean(gp_c,     na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = -c(corpus, model_size, tokens),
               names_to = "signal", values_to = "value") %>%
  mutate(signal = factor(signal,
                         levels = c("Frequency expectation",
                                    "Actual preference",
                                    "GenPref expectation")))

ggplot(traj_data,
       aes(x = tokens / 1e9, y = value,
           colour   = signal,
           linetype = model_size,
           group    = interaction(signal, model_size))) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "Mean value (preference-scale units, mean-centred)") +
  scale_colour_manual(values = ATTRACTOR_COLS, name = NULL) +
  scale_linetype_manual(
    values = c("125M" = "dotted", "350M" = "dashed", "1.3B" = "solid"),
    name   = "Model size"
  ) +
  facet_wrap(~corpus, scales = "free_x") +
  labs(
    title    = "Three attractors over training",
    subtitle = paste("Does actual preference track frequency or GenPref as training progresses?",
                     "  -  Averaged across all binomials at each checkpoint")
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

## Plot 10: Three Attractors on the Probability Scale (Real Binomials)

Same idea as Plot 9a, but:

- All values are on the **probability scale** (0 = always prefer non-alpha,
  1 = always prefer alpha, 0.5 = no preference):
  - **GenPref** -> `genpref + 0.5` (GenPref is already a +/-0.5 deviation from 0.5)
  - **Frequency expectation** -> `plogis(log_freq_ratio)` (raw frequency
    log-odds converted to probability, assuming perfect calibration)
  - **Model preference** -> `plogis(mean_preference)` (logistic of the
    log-probability difference, averaged across model sizes)
- Instead of GenPref decile bins, we show **real sampled binomials**,
  stratified by GenPref level x overall frequency so the sample covers
  the full range of both dimensions.
- Binomials with the largest GenPref-frequency conflict are **annotated**.

```{r plot10-prob-dumbbell, fig.width = 13, fig.height = 11}
library(ggrepel)

N_SAMPLE <- 45   # binomials to show per corpus panel
set.seed(964)

# ── Probability-scale values at the final checkpoint ─────────────────────────
prob_final <- agg_gp %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup() %>%
  filter(!is.na(genpref), !is.na(log_freq_ratio)) %>%
  mutate(
    pref_prob = plogis(mean_preference),   # model -> probability
    freq_prob = plogis(log_freq_ratio),    # relative frequency ratio -> probability
    gp_prob   = genpref + 0.5             # GenPref deviation -> probability
  ) %>%
  # Frequency and GenPref are corpus-level constants; average model preference
  # across model sizes for a single "overall model" estimate per binomial.
  group_by(corpus, binom) %>%
  summarize(
    pref_prob  = mean(pref_prob,  na.rm = TRUE),
    freq_prob  = first(freq_prob),
    gp_prob    = first(gp_prob),
    total_seen = mean(total_seen, na.rm = TRUE),
    .groups    = "drop"
  )

# ── Stratified sample: 3 GenPref levels x 3 corpus-frequency levels ──────────
sampled10 <- prob_final %>%
  group_by(corpus) %>%
  mutate(
    gp_str   = ntile(gp_prob,                      3),
    freq_str = ntile(log(pmax(total_seen, 1)), 3)
  ) %>%
  group_by(corpus, gp_str, freq_str) %>%
  slice_sample(n = ceiling(N_SAMPLE / 9), replace = FALSE) %>%
  ungroup()

# ── Pre-compute binom factor ordering (by gp_prob, within corpus) ─────────────
# Must be done before pivot_longer drops gp_prob as a column.
binom_levels10 <- sampled10 %>%
  arrange(corpus, gp_prob) %>%
  pull(binom) %>%
  unique()

sampled10 <- sampled10 %>%
  mutate(binom_f = factor(binom, levels = binom_levels10))

# ── Binomials to annotate: largest |GenPref - frequency| conflict ─────────────
annotate10 <- sampled10 %>%
  group_by(corpus) %>%
  mutate(conflict = abs(gp_prob - freq_prob)) %>%
  slice_max(conflict, n = 8) %>%
  ungroup()

# ── Long format for three-dot layer ──────────────────────────────────────────
long10 <- sampled10 %>%
  pivot_longer(
    cols      = c(freq_prob, pref_prob, gp_prob),
    names_to  = "signal",
    values_to = "probability"
  ) %>%
  mutate(
    signal  = factor(signal,
                     levels = c("freq_prob", "pref_prob", "gp_prob"),
                     labels = c("Frequency expectation",
                                "Model preference",
                                "GenPref")),
    binom_f = factor(binom, levels = binom_levels10)
  )

PROB_COLS <- c(
  "Frequency expectation" = "#e66101",
  "Model preference"      = "#1a1a1a",
  "GenPref"               = "#5e3c99"
)

ggplot() +
  # Grey segment spanning the two attractors (freq ↔ GenPref)
  geom_segment(
    data = sampled10,
    aes(x    = pmin(freq_prob, gp_prob),
        xend = pmax(freq_prob, gp_prob),
        y    = binom_f,
        yend = binom_f),
    colour    = "grey82",
    linewidth = 0.55
  ) +
  # 0.5 = no preference
  geom_vline(xintercept = 0.5, linetype = "dashed",
             colour = "grey55", linewidth = 0.4) +
  # Three coloured dots per binomial
  geom_point(
    data = long10,
    aes(x      = probability,
        y      = binom_f,
        colour = signal,
        shape  = signal),
    size = 2.8
  ) +
  # Text labels on highest-conflict binomials
  geom_text_repel(
    data           = annotate10,
    aes(x = pref_prob, y = binom_f, label = binom),
    colour         = "grey20",
    size           = 2.5,
    nudge_x        = 0.08,
    direction      = "y",
    hjust          = 0,
    segment.colour = "grey60",
    segment.size   = 0.25,
    max.overlaps   = Inf
  ) +
  scale_x_continuous(
    name   = "P(alpha ordering)",
    limits = c(0, 1.4),
    breaks = seq(0, 1, 0.25),
    labels = c("0", ".25", ".50", ".75", "1")
  ) +
  scale_y_discrete(
    name   = "Binomials  (sorted low -> high GenPref)",
    labels = NULL
  ) +
  scale_colour_manual(values = PROB_COLS, name = NULL) +
  scale_shape_manual(
    values = c("Frequency expectation" = 16,
               "Model preference"      = 18,
               "GenPref"               = 15),
    name   = NULL
  ) +
  facet_wrap(~corpus, scales = "free_y", nrow = 1) +
  labs(
    title    = "Three attractors on the probability scale  -  sampled binomials",
    subtitle = sprintf(
      paste("N <= %d binomials per corpus, stratified by GenPref x overall frequency",
            "  -  Sorted bottom -> top by GenPref  -  Labels = highest GenPref-frequency conflict"),
      N_SAMPLE
    )
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position    = "bottom",
    axis.ticks.y       = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

---

## Regression: Does GenPref Predict Preference Beyond Corpus Frequency?

A single regression at each checkpoint tests whether linguistic structure
(`genpref`) and its interaction with overall exposure (`log_total`) explain
any preference beyond what relative frequency alone predicts.

**Model**: `mean_preference ~ log_freq_ratio + genpref + log_total_c + genpref:log_total_c`

- `log_total_c` is `log(total_seen)` **centred within each checkpoint**,
  so the main effects are interpretable at average frequency.
- A positive `genpref` coefficient means GenPref pushes preference toward the
  alpha order even after controlling for relative frequency.
- A negative `genpref:log_total_c` interaction would mean the GenPref pull
  is stronger for rare binomials (consistent with the fill-in hypothesis).

```{r regression-setup, cache=TRUE}
reg_data <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(genpref), !is.na(total_seen),
         total_seen > 0) %>%
  mutate(log_total = log(total_seen))

reg_coefs <- reg_data %>%
  group_by(corpus, model_size, tokens) %>%
  group_map(function(g, keys) {
    if (nrow(g) < 10) return(NULL)
    sd_lfr <- sd(g$log_freq_ratio, na.rm = TRUE)
    sd_gp  <- sd(g$genpref,        na.rm = TRUE)
    sd_lt  <- sd(g$log_total,    na.rm = TRUE)
    if (is.na(sd_lfr) || sd_lfr == 0 ||
        is.na(sd_gp)  || sd_gp  == 0 ||
        is.na(sd_lt)  || sd_lt  == 0) return(NULL)
    g2 <- g %>% mutate(log_total_c = log_total - mean(log_total, na.rm = TRUE))
    fit <- lm(mean_preference ~ log_freq_ratio * log_total_c +
                genpref * log_total_c,
              data = g2)
    cf <- coef(summary(fit))
    terms <- c("log_freq_ratio", "genpref", "log_total_c",
               "log_freq_ratio:log_total_c", "genpref:log_total_c")
    purrr::map_dfr(terms, function(tm) {
      if (!tm %in% rownames(cf)) return(NULL)
      tibble(
        corpus     = keys$corpus,
        model_size = keys$model_size,
        tokens     = keys$tokens,
        term       = tm,
        estimate   = cf[tm, "Estimate"],
        se         = cf[tm, "Std. Error"],
        t_value    = cf[tm, "t value"],
        p_value    = cf[tm, "Pr(>|t|)"],
        n          = nrow(g2)
      )
    })
  }, .keep = TRUE) %>%
  bind_rows() %>%
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    term = factor(term,
                  levels = c("log_freq_ratio", "genpref", "log_total_c",
                             "log_freq_ratio:log_total_c", "genpref:log_total_c"),
                  labels = c("RelFreq\n(log ratio)",
                             "GenPref",
                             "Overall frequency\n(ln total)",
                             "RelFreq x\nOverall frequency",
                             "GenPref x\nOverall frequency"))
  )

cat(sprintf("Regression coefficients: %s rows across %d checkpoints\n",
            format(nrow(reg_coefs), big.mark = ","),
            n_distinct(reg_coefs$tokens)))
```

---

### Plot 11a: Regression Coefficients  -  All Checkpoints Pooled

Each (binomial x checkpoint) pair is one observation. The regression is run
once per corpus x model size using the full dataset, so rarer binomials
contribute their early-checkpoint observations alongside frequent ones.

```{r plot11a-reg-pooled, cache=TRUE, fig.width = 11, fig.height = 5}
TERM_LEVELS <- c("log_freq_ratio", "genpref", "log_total_c",
                 "log_freq_ratio:log_total_c", "genpref:log_total_c")
TERM_LABELS <- c("RelFreq\n(log ratio)", "GenPref",
                 "Overall frequency\n(ln total)",
                 "RelFreq x\nOverall frequency",
                 "GenPref x\nOverall frequency")

reg_pooled <- reg_data %>%
  group_by(corpus, model_size) %>%
  group_map(function(g, keys) {
    sd_lfr <- sd(g$log_freq_ratio, na.rm = TRUE)
    sd_gp  <- sd(g$genpref,        na.rm = TRUE)
    sd_lt  <- sd(g$log_total,    na.rm = TRUE)
    if (is.na(sd_lfr) || sd_lfr == 0 ||
        is.na(sd_gp)  || sd_gp  == 0 ||
        is.na(sd_lt)  || sd_lt  == 0) return(NULL)
    g2 <- g %>% mutate(log_total_c = log_total - mean(log_total, na.rm = TRUE))
    fit <- lm(mean_preference ~ log_freq_ratio * log_total_c +
                genpref * log_total_c,
              data = g2)
    cf <- coef(summary(fit))
    purrr::map_dfr(TERM_LEVELS, function(tm) {
      if (!tm %in% rownames(cf)) return(NULL)
      tibble(
        corpus     = keys$corpus,
        model_size = keys$model_size,
        term       = tm,
        estimate   = cf[tm, "Estimate"],
        se         = cf[tm, "Std. Error"],
        n          = nrow(g2)
      )
    })
  }, .keep = TRUE) %>%
  bind_rows() %>%
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    term       = factor(term, levels = TERM_LEVELS, labels = TERM_LABELS)
  )

ggplot(reg_pooled,
       aes(x = term, y = estimate,
           colour = model_size, group = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = estimate - se, ymax = estimate + se),
                width = 0.2, position = position_dodge(0.5),
                linewidth = 0.7) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(name = "OLS coefficient  (+/-1 SE)") +
  facet_wrap(~corpus) +
  labs(
    title    = "Regression: mean_preference ~ RelFreq x overall freq + GenPref x overall freq",
    subtitle = "All checkpoints pooled  -  Each obs = one binomial x checkpoint  -  +/-1 SE"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 9))
```

---

### Plot 11b: Regression Coefficients Over Training (per-checkpoint)

```{r plot11b-reg-traj, cache=TRUE, fig.width = 13, fig.height = 9}
ggplot(reg_coefs,
       aes(x = tokens / 1e9, y = estimate,
           colour = model_size, fill = model_size,
           group  = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = estimate - se, ymax = estimate + se),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "OLS coefficient  (+/-1 SE ribbon)") +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_grid(term ~ corpus, scales = "free_y") +
  labs(
    title    = "Regression coefficients over training",
    subtitle = paste("Rows = predictor  -  RelFreq + GenPref + overall + interaction",
                     "  -  +/-1 SE ribbon")
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text.y     = element_text(size = 8, angle = 0, hjust = 0)
  )
```

---

## Plot 12: Expected vs. Actual Preference for Sampled Binomials Across Frequencies

For a representative set of binomials spanning different overall corpus
frequencies, this plot shows:

- **Orange dot**  -  expected preference from relative frequency (`plogis(log_freq_ratio)`)
- **Dark dot**  -  actual model preference (`plogis(mean_preference)`)
- **Connecting line**  -  coloured by whether the deviation aligns with GenPref
  (green = model pulled toward GenPref; red = model pulled away)

The x-axis is `log(total_seen)` so we can see whether the gap between
expected and actual varies with how much frequency evidence the model had.

```{r plot12-setup, cache=TRUE}
# ── sample at the final checkpoint ──────────────────────────────────────────
plot12_base <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(genpref), !is.na(total_seen),
         total_seen > 0) %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup() %>%
  mutate(
    freq_prob  = plogis(log_freq_ratio),   # expected from corpus
    pref_prob  = plogis(mean_preference),  # actual model
    gp_prob    = genpref + 0.5,            # GenPref on [0,1]
    deviation  = pref_prob - freq_prob,    # +ve = model prefers alpha more than freq predicts
    gp_direction = sign(genpref),          # +1 = GenPref favours alpha
    # Aligned = deviation goes in same direction as GenPref
    aligned = case_when(
      gp_direction > 0 & deviation > 0  ~ "Toward GenPref",
      gp_direction < 0 & deviation < 0  ~ "Toward GenPref",
      gp_direction == 0                  ~ "GenPref neutral",
      TRUE                               ~ "Away from GenPref"
    ),
    log_total = log(total_seen)
  )

# Stratified sample: 5 freq bins x 3 genpref terciles = 15 cells,
# pick up to 4 binomials per cell per corpus x model_size
set.seed(42)
sampled12 <- plot12_base %>%
  mutate(
    freq_tier = ntile(log_total, 5),
    gp_tier   = ntile(genpref,     3)
  ) %>%
  group_by(corpus, model_size, freq_tier, gp_tier) %>%
  slice_sample(n = 4) %>%
  ungroup() %>%
  # Assign a fixed per-binomial jitter so segment and both dots share the same x
  group_by(binom) %>%
  mutate(jitter_x = log_total + runif(1, -0.08, 0.08)) %>%
  ungroup()

cat(sprintf("Plot 12 sample: %d binomials\n", n_distinct(sampled12$binom)))
```

```{r plot12-dumbbell, fig.width = 13, fig.height = 7}
ggplot(sampled12, aes(x = jitter_x)) +
  # Connecting segment (neutral grey  -  just shows the gap)
  geom_segment(aes(xend = jitter_x,
                   y    = freq_prob,
                   yend = pref_prob),
               colour = "grey55", linewidth = 1, alpha = 0.7) +
  # Expected (relative frequency) dot  -  orange
  geom_point(aes(y = freq_prob),
             shape = 21, fill = "#ff7f0e", colour = "white",
             size = 2.8, stroke = 0.5) +
  # Actual (model preference) dot  -  blue
  geom_point(aes(y = pref_prob),
             shape = 16, colour = "#1f77b4",
             size = 2.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  scale_x_continuous(
    name   = "Overall frequency  (ln total seen)",
    labels = function(x) scales::comma(round(exp(x)))
  ) +
  scale_y_continuous(
    name   = "P(alpha order)",
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "Expected vs. actual model preference across relative frequency levels",
    subtitle = "Orange = expected from relative frequency  -  Blue = actual model preference  -  Points jittered to reduce overlap"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 9)
  )
```

```{r plot12-annotated, fig.width = 14, fig.height = 8}
# Annotated version: label binomials with the largest absolute deviation
top_annotate <- sampled12 %>%
  group_by(corpus, model_size) %>%
  slice_max(abs(deviation), n = 6) %>%
  ungroup()

ggplot(sampled12, aes(x = jitter_x)) +
  geom_segment(aes(xend = jitter_x,
                   y    = freq_prob,
                   yend = pref_prob),
               colour = "grey55", linewidth = 1, alpha = 0.7) +
  geom_point(aes(y = freq_prob),
             shape = 21, fill = "#ff7f0e", colour = "white",
             size = 2.5, stroke = 0.5) +
  geom_point(aes(y = pref_prob),
             shape = 16, colour = "#1f77b4", size = 2.5) +
  ggrepel::geom_label_repel(
    data = top_annotate,
    aes(y = pref_prob, label = binom),
    size = 2.5, max.overlaps = 20,
    box.padding = 0.3, point.padding = 0.2,
    segment.colour = "grey40", segment.size = 0.3,
    fill = alpha("white", 0.8)
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  scale_x_continuous(
    name   = "Overall frequency  (ln total seen)",
    labels = function(x) scales::comma(round(exp(x)))
  ) +
  scale_y_continuous(
    name   = "P(alpha order)",
    limits = c(0, 1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "Expected vs. actual model preference  -  annotated extremes",
    subtitle = "Labels on binomials with largest deviation from frequency expectation"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 9)
  )
```

---

## Plot 13: Gap Between Actual and Expected Preference Over Training

How does the absolute deviation `|pref_prob - freq_prob|` change as training
progresses, and does it differ by how often the binomial was seen in the corpus?

```{r plot13-setup, cache=TRUE}
FREQ_BREAKS13 <- c(0, 1, 2, 3, Inf)
FREQ_LABELS13 <- c("1-10", "10-100", "100-1k", ">1k")

gap_traj <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(total_seen), total_seen > 0) %>%
  mutate(
    freq_prob   = plogis(log_freq_ratio),
    pref_prob   = plogis(mean_preference),
    abs_gap     = abs(pref_prob - freq_prob),
    log_total = log(total_seen),
    freq_bin    = cut(log_total,
                      breaks = FREQ_BREAKS13,
                      labels = FREQ_LABELS13,
                      include.lowest = TRUE)
  ) %>%
  group_by(corpus, model_size, tokens, freq_bin) %>%
  summarise(
    mean_gap = mean(abs_gap, na.rm = TRUE),
    se_gap   = sd(abs_gap,   na.rm = TRUE) / sqrt(n()),
    n        = n(),
    .groups  = "drop"
  ) %>%
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    freq_bin   = factor(freq_bin,   levels = FREQ_LABELS13)
  )
```

```{r plot13-traj, fig.width = 13, fig.height = 8}
ggplot(gap_traj,
       aes(x = tokens / 1e9, y = mean_gap,
           colour = model_size, fill = model_size,
           group  = model_size)) +
  geom_ribbon(aes(ymin = mean_gap - se_gap, ymax = mean_gap + se_gap),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(
    name   = "Mean |P(model) - P(frequency)|",
    labels = scales::percent_format(accuracy = 0.1)
  ) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  facet_grid(freq_bin ~ corpus, scales = "free_x") +
  labs(
    title    = "Gap between actual and expected preference over training, by overall frequency",
    subtitle = "Rows = overall frequency bin  -  +/-1 SE ribbon  -  All binomials (no sampling)"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 9)
  )
```

---

## Plot 14: Gap Between Actual and Expected Preference vs. Binomial Frequency

At the final checkpoint, how does the absolute deviation vary continuously
with `log(total_seen)`? A LOESS smoother reveals the overall trend.

```{r plot14-gap-vs-freq, fig.width = 11, fig.height = 6}
gap_final <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(total_seen), total_seen > 0) %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup() %>%
  mutate(
    freq_prob   = plogis(log_freq_ratio),
    pref_prob   = plogis(mean_preference),
    abs_gap     = abs(pref_prob - freq_prob),
    log_total = log(total_seen),
    model_size  = factor(model_size, levels = SIZE_LEVELS)
  )

ggplot(gap_final, aes(x = log_total, y = abs_gap, colour = model_size)) +
  geom_point(alpha = 0.04, size = 0.6, show.legend = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE,
              aes(fill = model_size), linewidth = 1, alpha = 0.15) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  scale_x_continuous(
    name   = "Overall frequency  (ln total seen)",
    labels = function(x) scales::comma(round(exp(x)))
  ) +
  scale_y_continuous(
    name   = "|P(model) - P(frequency expected)|",
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, NA)
  ) +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "Absolute gap between model and frequency-expected preference vs. overall frequency",
    subtitle = "Final checkpoint  -  Each point = one binomial  -  Degree-2 polynomial +/- 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 9)
  )
```


---

## Plot 15: Gap vs. Frequency Across All Checkpoints

Same as Plot 14 but every checkpoint contributes its own points. Because
`total_seen` grows with training, the same binomial traces a path rightward
across checkpoints  -  each observation is one (binomial x checkpoint) pair.
This lets the LOESS span the full range of exposure levels rather than only
the final snapshot.

```{r plot15-gap-all-checkpoints, cache=TRUE, fig.width = 11, fig.height = 6}
gap_all <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(total_seen), total_seen > 0) %>%
  mutate(
    freq_prob   = plogis(log_freq_ratio),
    pref_prob   = plogis(mean_preference),
    abs_gap     = abs(pref_prob - freq_prob),
    log_total = log(total_seen),
    model_size  = factor(model_size, levels = SIZE_LEVELS)
  )

ggplot(gap_all, aes(x = log_total, y = abs_gap, colour = model_size)) +
  geom_point(alpha = 0.01, size = 0.4, show.legend = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE,
              aes(fill = model_size), linewidth = 1, alpha = 0.15) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  scale_x_continuous(
    name   = "Overall frequency at checkpoint  (ln total seen)",
    labels = function(x) scales::comma(round(exp(x)))
  ) +
  scale_y_continuous(
    name   = "|P(model) - P(frequency expected)|",
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, NA)
  ) +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "Absolute gap vs. overall frequency - all checkpoints pooled",
    subtitle = "Each point = one binomial x checkpoint  .  Same binomial appears at increasing x as training progresses  .  Degree-2 polynomial +/- 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 9)
  )
```

---

## Plot 16: Signed Gap vs. Frequency Across All Checkpoints

Like Plot 15 but using the raw signed difference `P(model) - P(frequency expected)`.
Negative values mean the model prefers the alpha order *less* than corpus
frequency would predict; positive values mean it prefers it *more*.

```{r plot16-signed-gap-all, cache=TRUE, fig.width = 11, fig.height = 6}
signed_all <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(total_seen), total_seen > 0) %>%
  mutate(
    freq_prob   = plogis(log_freq_ratio),
    pref_prob   = plogis(mean_preference),
    signed_gap  = pref_prob - freq_prob,
    log_total = log(total_seen),
    model_size  = factor(model_size, levels = SIZE_LEVELS)
  )

ggplot(signed_all, aes(x = log_total, y = signed_gap, colour = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_point(alpha = 0.01, size = 0.4, show.legend = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE,
              aes(fill = model_size), linewidth = 1, alpha = 0.15) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  scale_x_continuous(
    name   = "Overall frequency at checkpoint  (ln total seen)",
    labels = function(x) scales::comma(round(exp(x)))
  ) +
  scale_y_continuous(
    name   = "P(model) - P(frequency expected)",
    labels = scales::percent_format(accuracy = 1)
  ) +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "Signed gap between model and frequency-expected preference vs. overall frequency  -  all checkpoints pooled",
    subtitle = "Negative = model under-predicts alpha order relative to frequency  .  Positive = over-predicts  .  Degree-2 polynomial +/- 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 9)
  )
```

---

## Plots 15b / 16b: Absolute and Signed Gap on Logit Scale

Rather than converting to probability first, these use `mean_preference` and
`log_freq_ratio` directly (both already in log-odds). The difference is the
gap in log-odds units, which is more symmetric and avoids probability-scale
compression near 0 and 1.

```{r plot15b-abs-logit, cache=TRUE, fig.width = 11, fig.height = 6}
gap_logit <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(total_seen), total_seen > 0) %>%
  mutate(
    abs_gap_logit    = abs(mean_preference - log_freq_ratio),
    signed_gap_logit = mean_preference - log_freq_ratio,
    log_total      = log(total_seen),
    model_size       = factor(model_size, levels = SIZE_LEVELS)
  )

# Plot 15b: absolute gap, logit scale
ggplot(gap_logit, aes(x = log_total, y = abs_gap_logit, colour = model_size)) +
  geom_point(alpha = 0.01, size = 0.4, show.legend = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE,
              aes(fill = model_size), linewidth = 1, alpha = 0.15) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  scale_x_continuous(
    name   = "Overall frequency at checkpoint  (ln total seen)",
    labels = function(x) scales::comma(round(exp(x)))
  ) +
  scale_y_continuous(name = "|preference - log_freq_ratio|  (log-odds)") +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "Absolute gap vs. overall frequency (logit scale)  -  all checkpoints pooled",
    subtitle = "Each point = one binomial x checkpoint  .  Degree-2 polynomial +/- 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 9)
  )
```

```{r plot16b-signed-logit, cache=TRUE, fig.width = 11, fig.height = 6}
# Plot 16b: signed gap, logit scale
ggplot(gap_logit, aes(x = log_total, y = signed_gap_logit, colour = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_point(alpha = 0.01, size = 0.4, show.legend = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = TRUE,
              aes(fill = model_size), linewidth = 1, alpha = 0.15) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
  scale_x_continuous(
    name   = "Overall frequency at checkpoint  (ln total seen)",
    labels = function(x) scales::comma(round(exp(x)))
  ) +
  scale_y_continuous(name = "preference - log_freq_ratio  (log-odds)") +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "Signed gap vs. overall frequency (logit scale)  -  all checkpoints pooled",
    subtitle = "Negative = model under-predicts alpha order relative to frequency  .  Positive = over-predicts  .  Degree-2 polynomial +/- 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "none",
    panel.grid.minor = element_blank(),
    strip.text       = element_text(size = 9)
  )
```

---

# Section A: Calibration  -  How Well Does the Model Copy Corpus Frequency?

## Plot 17: R^2, Slope, and Intercept of preference ~ log_freq_ratio Over Training

R^2 answers "how much variance does relative frequency explain?". The slope
answers "does the model copy frequency faithfully (slope ≈ 1) or compress it
toward neutrality (slope < 1)?". The intercept answers "is there a systematic
alpha-order bias independent of frequency?"

```{r plot17-r2-slope, cache=TRUE}
r2_traj <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(mean_preference)) %>%
  group_by(corpus, model_size, tokens) %>%
  group_map(function(g, keys) {
    if (nrow(g) < 10) return(NULL)
    sd_lfr <- sd(g$log_freq_ratio, na.rm = TRUE)
    if (is.na(sd_lfr) || sd_lfr == 0) return(NULL)
    fit <- lm(mean_preference ~ log_freq_ratio, data = g)
    cf  <- coef(fit)
    tibble(
      corpus     = keys$corpus,
      model_size = keys$model_size,
      tokens     = keys$tokens,
      r2         = summary(fit)$r.squared,
      slope      = cf["log_freq_ratio"],
      intercept  = cf["(Intercept)"],
      n          = nrow(g)
    )
  }, .keep = TRUE) %>%
  bind_rows() %>%
  mutate(model_size = factor(model_size, levels = SIZE_LEVELS))

cat(sprintf("R2 trajectory: %d rows\n", nrow(r2_traj)))
```

```{r plot17a-r2, fig.width = 11, fig.height = 5}
ggplot(r2_traj, aes(x = tokens / 1e9, y = r2,
                    colour = model_size, group = model_size)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "R^2  (preference ~ log_freq_ratio)",
                     limits = c(0, 1)) +
  facet_wrap(~corpus, scales = "free_x") +
  labs(
    title    = "R^2 of model preference ~ RelFreq over training",
    subtitle = "Higher = model preference more explained by corpus distribution"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r plot17b-slope-intercept, fig.width = 12, fig.height = 7}
ref_df <- tibble(
  param = factor(c("Slope  (1 = faithful copy)", "Intercept  (0 = no alpha bias)")),
  ref   = c(1, 0)
)

r2_long <- r2_traj %>%
  pivot_longer(cols = c(slope, intercept),
               names_to = "param", values_to = "value") %>%
  mutate(param = factor(param,
                        levels = c("slope", "intercept"),
                        labels = c("Slope  (1 = faithful copy)",
                                   "Intercept  (0 = no alpha bias)")))

ggplot(r2_long, aes(x = tokens / 1e9, y = value,
                    colour = model_size, group = model_size)) +
  geom_hline(data = ref_df, aes(yintercept = ref),
             linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "Estimate") +
  facet_grid(param ~ corpus, scales = "free") +
  labs(
    title    = "Slope and intercept of preference ~ log_freq_ratio over training",
    subtitle = "Slope < 1 = compression toward neutrality  .  Non-zero intercept = systematic alpha bias"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.text.y = element_text(size = 9, angle = 0, hjust = 0))
```

---

## Plot 18: Calibration Scatter  -  Model vs. Frequency Expectation

If the model is a perfect frequency-copier, points lie on the 45° line.
Points above it = model is more alpha-preferring than frequency predicts;
below = less alpha-preferring.

```{r plot18-calibration, fig.width = 12, fig.height = 7}
calib_data <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(mean_preference),
         !is.na(total_seen), total_seen > 0) %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup() %>%
  mutate(
    freq_bin = cut(log(pmax(total_seen, 1)),
                   breaks = c(0, log(10), log(100), log(1000), Inf),
                   labels = c("1-10", "10-100", "100-1k", ">1k"),
                   include.lowest = TRUE),
    model_size = factor(model_size, levels = SIZE_LEVELS)
  )

lim <- range(c(calib_data$log_freq_ratio, calib_data$mean_preference), na.rm = TRUE)

ggplot(calib_data, aes(x = log_freq_ratio, y = mean_preference,
                       colour = freq_bin)) +
  geom_abline(slope = 1, intercept = 0,
              linetype = "dashed", colour = "grey40", linewidth = 0.6) +
  geom_point(alpha = 0.05, size = 0.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1,
              aes(group = freq_bin)) +
  scale_colour_brewer(palette = "RdYlBu", direction = -1,
                      name = "Total seen") +
  scale_x_continuous(name = "log_freq_ratio  (corpus expectation, log-odds)", limits = lim) +
  scale_y_continuous(name = "mean_preference  (model, log-odds)",             limits = lim) +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "Calibration: model preference vs. RelFreq expectation",
    subtitle = "Dashed line = perfect copying  .  Above = model more alpha-preferring than expected  .  Coloured by frequency bin"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 9))
```

---

# Section B: Systematic Divergence by GenPref Strength

## Plot 19: Signed Deviation by GenPref Quartile Over Training

If GenPref pulls model preferences systematically, binomials in the top
GenPref quartile should show positive signed deviations from frequency
expectation, and bottom-quartile binomials should show negative deviations.

```{r plot19-setup, cache=TRUE}
gp_quartiles <- agg_gp %>%
  filter(!is.na(genpref)) %>%
  group_by(binom) %>%
  summarise(genpref = first(genpref), .groups = "drop") %>%
  mutate(
    gp_quartile = ntile(genpref, 4),
    gp_label    = paste0("Q", gp_quartile, "  (",
                         c("low", "mid-low", "mid-high", "high")[gp_quartile],
                         " GenPref)")
  )

gp_dev_traj <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(genpref)) %>%
  inner_join(gp_quartiles %>% select(binom, gp_quartile, gp_label), by = "binom") %>%
  mutate(signed_gap = mean_preference - log_freq_ratio) %>%
  group_by(corpus, model_size, tokens, gp_quartile, gp_label) %>%
  summarise(
    mean_dev = mean(signed_gap, na.rm = TRUE),
    se_dev   = sd(signed_gap,   na.rm = TRUE) / sqrt(n()),
    n        = n(),
    .groups  = "drop"
  ) %>%
  mutate(
    model_size = factor(model_size, levels = SIZE_LEVELS),
    gp_label   = factor(gp_label, levels = unique(gp_label[order(gp_quartile)]))
  )

cat(sprintf("GenPref quartile trajectories: %d rows\n", nrow(gp_dev_traj)))
```

```{r plot19-gp-quartile, fig.width = 13, fig.height = 8}
ggplot(gp_dev_traj,
       aes(x = tokens / 1e9, y = mean_dev,
           colour = gp_label, fill = gp_label, group = gp_label)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", linewidth = 0.4) +
  geom_ribbon(aes(ymin = mean_dev - se_dev, ymax = mean_dev + se_dev),
              alpha = 0.12, colour = NA) +
  geom_line(linewidth = 0.9) +
  scale_colour_brewer(palette = "RdBu", direction = -1, name = "GenPref quartile") +
  scale_fill_brewer(palette   = "RdBu", direction = -1, name = "GenPref quartile") +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "Mean signed gap  (preference - log_freq_ratio)") +
  facet_grid(model_size ~ corpus, scales = "free_x") +
  labs(
    title    = "Signed deviation from frequency expectation by GenPref quartile over training",
    subtitle = "If GenPref pulls: Q4 (high) stays positive, Q1 (low) stays negative  .  +/-1 SE ribbon"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 9))
```

---

# Section C: GAM Interaction Surface

## Plot 20: GAM te(log_total, tokens)  -  Deviation as a 2D Surface

A tensor-product smooth jointly over overall frequency and training time
captures non-linear interactions that regression misses. The surface shows
whether the deviation from frequency expectation varies with both how much
frequency evidence a binomial has and how far along training is.

```{r plot20-gam-setup, cache=TRUE}
library(mgcv)

gam_input <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(total_seen), total_seen > 0) %>%
  mutate(
    signed_gap  = mean_preference - log_freq_ratio,
    log_total = log(total_seen),
    tokens_b    = tokens / 1e9
  )

set.seed(7)
gam_grids <- gam_input %>%
  group_by(corpus, model_size) %>%
  group_map(function(g, keys) {
    if (nrow(g) < 200) return(NULL)
    g_fit <- if (nrow(g) > 100000) slice_sample(g, n = 100000) else g
    fit <- bam(
      signed_gap ~ te(log_total, tokens_b, k = c(8, 8)),
      data     = g_fit,
      method   = "fREML",
      discrete = TRUE
    )
    grid <- expand.grid(
      log_total = seq(quantile(g$log_total, 0.02, na.rm = TRUE),
                        quantile(g$log_total, 0.98, na.rm = TRUE),
                        length.out = 60),
      tokens_b    = seq(min(g$tokens_b, na.rm = TRUE),
                        max(g$tokens_b, na.rm = TRUE),
                        length.out = 60)
    )
    pred <- predict(fit, newdata = grid, se.fit = TRUE)
    grid %>% mutate(
      fitted     = pred$fit,
      se         = pred$se.fit,
      corpus     = keys$corpus,
      model_size = keys$model_size
    )
  }, .keep = TRUE) %>%
  bind_rows() %>%
  mutate(model_size = factor(model_size, levels = SIZE_LEVELS))

cat(sprintf("GAM grids: %d prediction points\n", nrow(gam_grids)))
```

```{r plot20-gam-surface, fig.width = 14, fig.height = 9}
zlim <- max(abs(gam_grids$fitted), na.rm = TRUE)

ggplot(gam_grids, aes(x = log_total, y = tokens_b, fill = fitted)) +
  geom_tile() +
  geom_contour(aes(z = fitted), colour = "white", linewidth = 0.3,
               breaks = seq(-2, 2, by = 0.5)) +
  scale_fill_distiller(palette = "RdBu", direction = -1,
                       name = "Signed gap\n(log-odds)",
                       limits = c(-zlim, zlim)) +
  scale_x_continuous(name   = "Overall frequency  (ln total seen)",
                     labels = function(x) scales::comma(round(exp(x)))) +
  scale_y_continuous(name = "Training progress  (billions of tokens)") +
  facet_grid(model_size ~ corpus) +
  labs(
    title    = "GAM surface: signed gap as a function of overall frequency x training time",
    subtitle = "te(log_total, tokens)  .  Red = model more alpha-preferring than frequency predicts"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "right", panel.grid = element_blank(),
        strip.text = element_text(size = 9))
```

---

# Section D: Individual Binomial Preference Trajectories

## Plot 21: Sampled Binomials  -  Preference and Frequency Expectation Over Training

Both `mean_preference` and `log_freq_ratio` change as the model trains (more
frequency evidence accumulates). Plotting both as trajectories for a
stratified sample shows whether the model converges to or persistently
diverges from frequency.

```{r plot21-traj-setup, cache=TRUE}
set.seed(99)
traj_binoms <- agg_gp %>%
  filter(!is.na(genpref), !is.na(log_freq_ratio), !is.na(total_seen)) %>%
  group_by(corpus, binom) %>%
  filter(tokens == max(tokens)) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    gp_tier   = ntile(genpref, 3),
    freq_tier = ntile(log(pmax(total_seen, 1)), 3)
  ) %>%
  group_by(corpus, gp_tier, freq_tier) %>%
  slice_sample(n = 3) %>%
  ungroup() %>%
  select(corpus, binom, genpref) %>%
  distinct()

traj_data <- agg_gp %>%
  inner_join(traj_binoms, by = c("corpus", "binom")) %>%
  filter(!is.na(log_freq_ratio)) %>%
  mutate(model_size = factor(model_size, levels = SIZE_LEVELS)) %>%
  pivot_longer(cols      = c(mean_preference, log_freq_ratio),
               names_to  = "measure",
               values_to = "log_odds") %>%
  mutate(measure = recode(measure,
                          mean_preference = "Model preference",
                          log_freq_ratio  = "Corpus expectation"))

cat(sprintf("Trajectory sample: %d unique binomials\n", n_distinct(traj_binoms$binom)))
```

```{r plot21-trajectories, fig.width = 15, fig.height = 10}
ggplot(traj_data %>% filter(model_size == last(SIZE_LEVELS)),
       aes(x      = tokens / 1e9,
           y      = log_odds,
           colour = measure,
           group  = interaction(binom, measure))) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  scale_colour_manual(values = c("Model preference"   = "#1f77b4",
                                 "Corpus expectation" = "#ff7f0e"),
                      name = NULL) +
  scale_x_continuous(name = "Tokens seen (billions)") +
  scale_y_continuous(name = "Log-odds") +
  facet_wrap(corpus ~ binom, scales = "free", ncol = 9) +
  labs(
    title    = "Individual binomial trajectories: model preference vs. corpus expectation over training",
    subtitle = "Largest model only  .  Blue = model preference  .  Orange = RelFreq expectation"
  ) +
  theme_bw(base_size = 8) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 6))
```

---

# Section E: Mixed-Effects Model and Variance Decomposition

How much variance in model preference is explained by relative frequency alone,
GenPref alone, both together, and a full model with interactions? The
mixed-effects version adds a random intercept per binomial to account for
the repeated-measures structure across checkpoints.

```{r vardecomp-setup, cache=TRUE, cache.lazy=FALSE}
library(lme4)

# ── Centre predictors ─────────────────────────────────────────────────────────
vd_raw <- reg_data %>% filter(!is.na(genpref))

# genpref is already on [-0.5, 0.5] after normalisation in load-genpref.
# Centre by subtracting its mean (should be near 0 if binomials are balanced).
vd_data <- vd_raw %>%
  mutate(
    freq_prob_c = plogis(log_freq_ratio) - 0.5,
    genpref_c   = genpref - mean(genpref, na.rm = TRUE),
    log_total_c = log_total - mean(log_total, na.rm = TRUE)
  )

cat(sprintf("Predictors centred: genpref mean = %.4f  (range: [%.4f, %.4f])\n",
            mean(vd_raw$genpref, na.rm = TRUE),
            min(vd_raw$genpref,  na.rm = TRUE),
            max(vd_raw$genpref,  na.rm = TRUE)))

y      <- vd_data$mean_preference
ss_tot <- sum((y - mean(y))^2)

# ── Pre-compute fold assignments ONCE  -  shared by every model ─────────────────
# Folds are at the binomial level: ALL checkpoints for a binomial land in the
# same fold. Using identical folds for OLS and lmer makes Q^2 directly comparable.
K <- 5
set.seed(42)
unique_binoms <- unique(vd_data$binom)
fold_assign   <- setNames(
  sample(rep(1:K, length.out = length(unique_binoms))),
  unique_binoms
)

# ── CV function  -  5-fold binomial-grouped holdout ────────────────────────────
lobo_q2 <- function(formula, data, fit_fn) {
  binom_vals <- data$binom
  lobo_ss    <- 0
  for (fold in seq_len(K)) {
    held     <- names(fold_assign)[fold_assign == fold]
    in_train <- !binom_vals %in% held
    fit_k    <- fit_fn(formula, data[in_train, ])
    pred     <- predict(fit_k,
                        newdata          = data[!in_train, ],
                        allow.new.levels = TRUE)
    lobo_ss  <- lobo_ss + sum((data$mean_preference[!in_train] - pred)^2)
  }
  1 - lobo_ss / ss_tot
}

lmer_fn <- function(frm, dat) lmer(frm, data = dat, REML = FALSE,
                                   control = lmerControl(optimizer = "bobyqa"))

# ── Model specifications (mixed-effects only) ─────────────────────────────────
mixed_specs <- list(
  "Null (intercept only)"                    = mean_preference ~ 1                                                              + (1 | binom),
  "Frequency"                                = mean_preference ~ freq_prob_c                                                    + (1 | binom),
  "GenPref"                                  = mean_preference ~ genpref_c                                                      + (1 | binom),
  "Overall frequency"                        = mean_preference ~ log_total_c                                                  + (1 | binom),
  "Frequency + GenPref"                      = mean_preference ~ freq_prob_c + genpref_c                                        + (1 | binom),
  "Frequency + Overall"                      = mean_preference ~ freq_prob_c + log_total_c                                    + (1 | binom),
  "GenPref + Overall"                        = mean_preference ~ genpref_c + log_total_c                                      + (1 | binom),
  "Frequency + GenPref + Overall"            = mean_preference ~ freq_prob_c + genpref_c + log_total_c                        + (1 | binom),
  "Frequency x Overall"                      = mean_preference ~ freq_prob_c * log_total_c                                    + (1 | binom),
  "GenPref x Overall"                        = mean_preference ~ genpref_c * log_total_c                                      + (1 | binom),
  "Frequency x Overall + GenPref"            = mean_preference ~ freq_prob_c * log_total_c + genpref_c                        + (1 | binom),
  "Frequency + GenPref x Overall"            = mean_preference ~ freq_prob_c + genpref_c * log_total_c                        + (1 | binom),
  "Frequency x Overall + GenPref x Overall"  = mean_preference ~ freq_prob_c * log_total_c + genpref_c * log_total_c        + (1 | binom)
)

# ── Run 5-fold binomial-grouped LOO-CV for every mixed model ──────────────────
cat(sprintf("Running %d-fold binomial-grouped CV for %d mixed models...\n",
            K, length(mixed_specs)))

cv_table <- purrr::imap_dfr(mixed_specs, function(frm, name) {
  cat(sprintf("  Mixed: %s\n", name))
  q2_mix  <- tryCatch(lobo_q2(frm, vd_data, lmer_fn),
                      error = function(e) { warning(e); NA_real_ })
  fit_mix <- tryCatch(lmer_fn(frm, vd_data), error = function(e) NULL)

  tibble(
    Model    = name,
    LOBO_Q2  = round(q2_mix, 4),
    AIC      = if (!is.null(fit_mix)) round(AIC(fit_mix), 1) else NA_real_,
    n_params = if (!is.null(fit_mix)) length(fixef(fit_mix)) + 2L else NA_integer_
  )
})

print(cv_table %>% select(Model, LOBO_Q2, AIC, n_params))
```

```{r vardecomp-summary}
# ── 1. Ranked table ───────────────────────────────────────────────────────────
cat("=== Models ranked by LOBO-CV Q^2 (best first) ===\n")
cv_ranked <- cv_table %>%
  arrange(desc(LOBO_Q2)) %>%
  mutate(rank = row_number(),
         delta_Q2_from_null = round(LOBO_Q2 - LOBO_Q2[Model == "Null (intercept only)"], 4))
print(cv_ranked %>% select(rank, Model, LOBO_Q2, delta_Q2_from_null, AIC, n_params))

# ── 2. Marginal contribution of each component ────────────────────────────────
# Compare pairs of nested models that differ by exactly one component
null_q2  <- cv_table$LOBO_Q2[cv_table$Model == "Null (intercept only)"]
freq_q2  <- cv_table$LOBO_Q2[cv_table$Model == "Frequency"]
gp_q2    <- cv_table$LOBO_Q2[cv_table$Model == "GenPref"]
ovr_q2   <- cv_table$LOBO_Q2[cv_table$Model == "Overall frequency"]
fgp_q2   <- cv_table$LOBO_Q2[cv_table$Model == "Frequency + GenPref"]
fovr_q2  <- cv_table$LOBO_Q2[cv_table$Model == "Frequency + Overall"]
all3_q2  <- cv_table$LOBO_Q2[cv_table$Model == "Frequency + GenPref + Overall"]
fxovr_q2 <- cv_table$LOBO_Q2[cv_table$Model == "Frequency x Overall + GenPref x Overall"]

marginal <- tribble(
  ~Comparison,                                          ~delta_Q2,
  "Null -> Frequency (main effect)",                    round(freq_q2  - null_q2, 4),
  "Null -> GenPref (main effect)",                      round(gp_q2    - null_q2, 4),
  "Null -> Overall frequency (main effect)",            round(ovr_q2   - null_q2, 4),
  "Frequency -> Frequency + GenPref",                   round(fgp_q2   - freq_q2, 4),
  "Frequency -> Frequency + Overall",                   round(fovr_q2  - freq_q2, 4),
  "Freq + GenPref + Overall -> full interactions",      round(fxovr_q2 - all3_q2, 4)
)
cat("\n=== Marginal Q^2 gain from adding each component ===\n")
print(marginal)

# ── 3. Best model: full fixed-effect summary ──────────────────────────────────
best_name <- cv_ranked$Model[1]
best_frm  <- mixed_specs[[best_name]]
cat(sprintf("\n=== Best model: '%s'  (Q^2 = %.4f) ===\n", best_name,
            cv_ranked$LOBO_Q2[1]))
best_fit <- lmer(best_frm, data = vd_data, REML = FALSE,
                 control = lmerControl(optimizer = "bobyqa"))
print(summary(best_fit), correlation = FALSE)

cat("\n=== Random-effects variance decomposition ===\n")
vc <- as.data.frame(VarCorr(best_fit))
vc$pct <- round(100 * vc$vcov / sum(vc$vcov), 1)
print(vc %>% select(grp, var1, vcov, pct))
```

```{r vardecomp-plot, fig.width = 11, fig.height = 7}
cv_table %>%
  mutate(Model = factor(Model, levels = rev(Model))) %>%
  ggplot(aes(x = LOBO_Q2, y = Model)) +
  geom_col(width = 0.7, fill = "#2ca02c") +
  geom_text(aes(label = sprintf("%.4f", LOBO_Q2)), hjust = -0.1, size = 2.8) +
  geom_vline(xintercept = 0, colour = "grey30", linewidth = 0.5) +
  scale_x_continuous(
    name   = "LOBO-CV Q^2  (leave-one-binomial-out predictive R^2)",
    expand = expansion(mult = c(0.05, 0.2))
  ) +
  scale_y_discrete(name = NULL) +
  labs(
    title    = "Out-of-sample predictive accuracy  -  mixed-effects models",
    subtitle = paste("5-fold binomial-grouped CV  .",
                     "Holdout unit = entire binomial (all checkpoints)  .",
                     "Freq = P(alpha) - 0.5  .  GenPref on [-0.5, 0.5]  .",
                     "All models include (1|binom)")
  ) +
  theme_bw(base_size = 10) +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        legend.position = "none", axis.text.y = element_text(size = 8))
```

---

# Section F: What Determines the Direction of the Signed Gap?

When the model diverges from relative frequency, is that divergence explained by
GenPref, overall frequency, corpus-predicted frequency, or interactions among
these? Here `signed_gap` (`mean_preference - log_freq_ratio`) is the explicit
outcome, so every coefficient directly describes a predictor of the gap rather
than of raw preference.

A positive signed gap means the model is *more* alpha-preferring than the
corpus predicts; a negative gap means *less*.

```{r gap-direction-setup, cache=TRUE, cache.lazy=FALSE}
# Re-use centred predictors from vd_data (same centering as Section E)
gd_data <- vd_data %>%
  mutate(
    signed_gap   = mean_preference - log_freq_ratio,
    gap_positive = as.integer(signed_gap > 0)
  )

cat(sprintf("Gap-direction data: %d observations, %d unique binomials\n",
            nrow(gd_data), n_distinct(gd_data$binom)))

# ── Final-checkpoint subset (feasible for brms) ────────────────────────────────
gd_final <- gd_data %>%
  group_by(corpus, model_size, model) %>%
  filter(tokens == max(tokens)) %>%
  ungroup()

cat(sprintf("Final-checkpoint subset: %d observations, %d unique binomials\n",
            nrow(gd_final), n_distinct(gd_final$binom)))

# ── Bayesian linear mixed model: signed gap ~ predictors + (1|binom) ──────────
# freq_prob_c  = P(alpha in corpus) - 0.5  (what corpus predicts, centred)
# genpref_c    = GenPref - mean            (linguistic preference, centred)
# log_total_c  = log(total_seen) - mean    (overall exposure, centred)
gd_fit <- brm(
  signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
  data   = gd_final,
  family = gaussian(),
  prior  = BRMS_PRIORS,
  iter   = BRMS_ITER,
  warmup = BRMS_WARMUP,
  chains = BRMS_CHAINS,
  cores  = BRMS_CORES,
  file   = file.path(MODELS_DIR, "gd_fit"),
  seed   = BRMS_SEED
)
cat("\n=== Bayesian linear mixed model (final checkpoint): signed_gap ===\n")
print(summary(gd_fit))

# ── Bayesian logistic mixed model: P(gap > 0) ─────────────────────────────────
gd_logit <- brm(
  gap_positive ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
  data   = gd_final,
  family = bernoulli(),
  prior  = BRMS_PRIORS_LOGISTIC,
  iter   = BRMS_ITER,
  warmup = BRMS_WARMUP,
  chains = BRMS_CHAINS,
  cores  = BRMS_CORES,
  file   = file.path(MODELS_DIR, "gd_logit"),
  seed   = BRMS_SEED + 1L
)
cat("\n=== Bayesian logistic mixed model (final checkpoint): P(gap > 0) ===\n")
print(summary(gd_logit))

# ── Coefficient table for plotting ────────────────────────────────────────────
extract_coefs_brms <- function(fit, model_label) {
  fe        <- as.data.frame(fixef(fit))
  fe$term   <- rownames(fe)
  fe$model  <- model_label
  fe
}

TERM_LEVELS_GD <- c("freq_prob_c", "genpref_c", "log_total_c",
                    "freq_prob_c:log_total_c", "genpref_c:log_total_c")
TERM_LABELS_GD <- c("RelFreq\n(P(alpha)-0.5)", "GenPref",
                    "Overall freq\n(ln total)",
                    "RelFreq x\nOverall freq", "GenPref x\nOverall freq")

gd_coefs <- bind_rows(
  extract_coefs_brms(gd_fit,   "Linear (signed gap)"),
  extract_coefs_brms(gd_logit, "Logistic (gap > 0)")
) %>%
  rename(estimate = Estimate, se = Est.Error, lo = Q2.5, hi = Q97.5) %>%
  filter(term != "Intercept") %>%
  mutate(
    term = factor(term, levels = TERM_LEVELS_GD, labels = TERM_LABELS_GD),
    # Mark coefficients whose 95% credible interval excludes zero
    sig  = if_else(lo > 0 | hi < 0, "*", "")
  )
```

```{r gap-direction-plot, fig.width = 12, fig.height = 5}
ggplot(gd_coefs,
       aes(x = term, y = estimate,
           colour = model, group = model)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = lo, ymax = hi),
                width = 0.2, position = position_dodge(0.5), linewidth = 0.8) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  geom_text(aes(label = sig, y = hi + 0.02),
            position = position_dodge(0.5), size = 4, vjust = 0) +
  scale_colour_manual(values = c("Linear (signed gap)" = "#1f77b4",
                                 "Logistic (gap > 0)"  = "#ff7f0e"),
                      name = NULL) +
  scale_x_discrete(name = NULL) +
  scale_y_continuous(name = "Coefficient  (95% credible interval)") +
  facet_wrap(~model, scales = "free_y") +
  labs(
    title    = "What predicts the direction and magnitude of the signed gap?",
    subtitle = paste0(
      "Linear model outcome: signed_gap = mean_preference - log_freq_ratio  -  ",
      "Logistic outcome: P(gap > 0)\n",
      "All predictors centred  -  Random intercept by binomial  -  Final checkpoint  -  ",
      "Bayesian (brms)  -  * = 95% CI excludes zero")
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank(),
        strip.text = element_text(size = 10),
        axis.text.x = element_text(size = 9))
```

```{r gap-direction-genpref-plot, fig.width = 11, fig.height = 5}
# Visualise the GenPref effect directly: mean signed gap binned by GenPref decile
genpref_bins <- gd_data %>%
  mutate(genpref_decile = ntile(genpref_c, 10)) %>%
  group_by(corpus, genpref_decile) %>%
  summarise(
    mean_genpref  = mean(genpref_c + mean(gd_data$genpref, na.rm = TRUE),
                         na.rm = TRUE),
    mean_gap      = mean(signed_gap,      na.rm = TRUE),
    se_gap        = sd(signed_gap, na.rm = TRUE) / sqrt(n()),
    pct_positive  = mean(signed_gap > 0,  na.rm = TRUE),
    n             = n(),
    .groups       = "drop"
  )

ggplot(genpref_bins,
       aes(x = mean_genpref, y = mean_gap, colour = corpus)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = mean_gap - 1.96 * se_gap,
                    ymax = mean_gap + 1.96 * se_gap),
                width = 0.01, linewidth = 0.7) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("BabyLM" = "#1b7837", "C4" = "#762a83"),
                      name = NULL) +
  scale_x_continuous(name = "GenPref  (binned by decile, original scale)") +
  scale_y_continuous(name = "Mean signed gap  (log-odds)") +
  facet_wrap(~corpus) +
  labs(
    title    = "Mean signed gap by GenPref decile",
    subtitle = "Positive gap = model more alpha-preferring than corpus predicts  -  All checkpoints & model sizes pooled  -  +/-95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

---

# Section G: Cross-Model-Size Consistency of Deviation

If the deviation from frequency expectation is structural (driven by
properties of the binomial such as GenPref), the direction and magnitude of
deviation should be consistent across model sizes.

```{r consistency-setup, cache=TRUE}
# Long format: one row per corpus x binom x model_size at each model's final checkpoint.
# Average across models of the same size within a corpus (handles multiple seeds/runs).
consistency_long <- agg_gp %>%
  filter(!is.na(log_freq_ratio)) %>%
  group_by(corpus, model_size, binom) %>%
  filter(tokens == max(tokens)) %>%
  summarise(signed_gap = mean(mean_preference - log_freq_ratio, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(model_size = factor(model_size, levels = SIZE_LEVELS))

cat(sprintf("Consistency data: %d rows, %d unique binomials, %d model sizes\n",
            nrow(consistency_long),
            n_distinct(consistency_long$binom),
            n_distinct(consistency_long$model_size)))
```

```{r plot22-consistency, fig.width = 12, fig.height = 5}
size_pairs <- combn(SIZE_LEVELS, 2, simplify = FALSE)

for (pair in size_pairs) {
  # Inner join: only binomials present in BOTH sizes (avoids NA from pivot_wider)
  d <- inner_join(
    consistency_long %>%
      filter(model_size == pair[1]) %>%
      select(corpus, binom, gap_x = signed_gap),
    consistency_long %>%
      filter(model_size == pair[2]) %>%
      select(corpus, binom, gap_y = signed_gap),
    by = c("corpus", "binom")
  )

  if (nrow(d) == 0) next

  r_vals <- d %>%
    group_by(corpus) %>%
    summarise(r    = round(cor(gap_x, gap_y, use = "complete.obs"), 2),
              n    = n(),
              .groups = "drop") %>%
    mutate(label = paste0("r = ", r, "  (n = ", scales::comma(n), ")"))

  p <- ggplot(d, aes(x = gap_x, y = gap_y)) +
    geom_point(alpha = 0.06, size = 0.5, colour = "grey40") +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", colour = "grey60") +
    geom_smooth(method = "lm", se = FALSE, colour = "#d62728", linewidth = 0.9) +
    geom_text(data = r_vals, aes(label = label),
              x = -Inf, y = Inf, hjust = -0.08, vjust = 1.4,
              size = 3, inherit.aes = FALSE) +
    scale_x_continuous(name = paste(pair[1], "deviation (log-odds)")) +
    scale_y_continuous(name = paste(pair[2], "deviation (log-odds)")) +
    facet_wrap(~corpus) +
    labs(title    = paste("Deviation consistency:", pair[1], "vs.", pair[2]),
         subtitle = "Only binomials present in both model sizes  .  Dashed = perfect agreement  .  Final checkpoint") +
    theme_bw(base_size = 10) +
    theme(panel.grid.minor = element_blank(), strip.text = element_text(size = 8))

  print(p)
}
```

---

# Section H: Outlier Characterization

Which binomials show the largest persistent deviation from relative frequency
across all checkpoints? These are the clearest cases where the model is not
simply copying the corpus.

```{r outliers-setup, cache=TRUE}
outliers <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(genpref)) %>%
  mutate(signed_gap = mean_preference - log_freq_ratio) %>%
  group_by(corpus, binom) %>%
  summarise(
    mean_signed_gap  = mean(signed_gap,    na.rm = TRUE),
    abs_mean_gap     = abs(mean(signed_gap, na.rm = TRUE)),
    sd_gap           = sd(signed_gap,      na.rm = TRUE),
    genpref          = first(na.omit(genpref)),
    final_total_seen = total_seen[which.max(tokens)],
    n_checkpoints    = n(),
    .groups          = "drop"
  ) %>%
  group_by(corpus) %>%
  slice_max(abs_mean_gap, n = 20) %>%
  ungroup() %>%
  arrange(corpus, desc(abs_mean_gap))

print(outliers %>% select(corpus, binom, mean_signed_gap, genpref,
                          final_total_seen, n_checkpoints))
```

```{r plot23-outliers, fig.width = 13, fig.height = 7}
outliers_plot <- outliers %>%
  group_by(corpus) %>%
  mutate(binom_f = factor(binom, levels = binom[order(mean_signed_gap)])) %>%
  ungroup()

ggplot(outliers_plot,
       aes(x = mean_signed_gap, y = binom_f, fill = genpref)) +
  geom_col(width = 0.7) +
  geom_vline(xintercept = 0, colour = "grey30", linewidth = 0.5) +
  scale_fill_distiller(palette = "RdBu", direction = -1,
                       name = "GenPref", limits = c(-0.5, 0.5)) +
  scale_x_continuous(
    name = "Mean signed gap  (preference - log_freq_ratio, log-odds)"
  ) +
  scale_y_discrete(name = NULL) +
  facet_wrap(~corpus, scales = "free_y") +
  labs(
    title    = "Top 20 binomials with largest persistent deviation from frequency expectation",
    subtitle = "Fill = GenPref  .  Right = model more alpha-preferring than corpus predicts  .  Left = less"
  ) +
  theme_bw(base_size = 10) +
  theme(legend.position  = "right",
        panel.grid.major.y = element_blank(),
        panel.grid.minor   = element_blank(),
        strip.text         = element_text(size = 9),
        axis.text.y        = element_text(size = 7))
```

---

# Summary Statistics: Over- vs. Under-Prediction Relative to Corpus Frequency

**Positive signed gap** (`mean_preference > log_freq_ratio`) means the model
prefers the alpha order *more* than relative frequency predicts  -  it is
over-predicting the alpha order.

**Negative signed gap** means the model is less alpha-biased than the corpus  - 
under-predicting the alpha order.

```{r overpred-setup, cache=TRUE}
# ── All checkpoints pooled ───────────────────────────────────────────────────
gap_base <- agg_gp %>%
  filter(!is.na(log_freq_ratio), !is.na(total_seen), total_seen > 0) %>%
  mutate(
    signed_gap  = mean_preference - log_freq_ratio,
    log_total = log(total_seen),
    freq_bin    = cut(log_total,
                      breaks = c(0, log(10), log(100), log(1000), Inf),
                      labels = c("1-10", "10-100", "100-1k", ">1k"),
                      include.lowest = TRUE),
    model_size  = factor(model_size, levels = SIZE_LEVELS)
  )

# ── Overall summary: mean signed gap per corpus x model_size ─────────────────
overall_summary <- gap_base %>%
  group_by(corpus, model_size) %>%
  summarise(
    mean_gap   = mean(signed_gap, na.rm = TRUE),
    median_gap = median(signed_gap, na.rm = TRUE),
    sd_gap     = sd(signed_gap, na.rm = TRUE),
    n          = n(),
    se         = sd_gap / sqrt(n),
    t_stat     = mean_gap / se,
    p_value    = 2 * pt(-abs(t_stat), df = n - 1),
    direction  = if_else(mean_gap > 0, "Over-predicts alpha order", "Under-predicts alpha order"),
    .groups    = "drop"
  ) %>%
  mutate(across(c(mean_gap, median_gap, sd_gap, se), ~round(.x, 4)),
         p_value = signif(p_value, 3))

cat("=== Overall over/under-prediction (signed, all checkpoints pooled) ===\n")
print(overall_summary %>%
        select(corpus, model_size, direction, mean_gap, median_gap, se, p_value, n))

# ── Overall summary: mean ABSOLUTE gap ───────────────────────────────────────
# The signed mean can cancel if some binomials are over- and others under-
# predicted. The mean absolute gap avoids this.
abs_overall_summary <- gap_base %>%
  group_by(corpus, model_size) %>%
  summarise(
    mean_abs_gap   = mean(abs(signed_gap),   na.rm = TRUE),
    median_abs_gap = median(abs(signed_gap), na.rm = TRUE),
    sd_abs_gap     = sd(abs(signed_gap),     na.rm = TRUE),
    n              = n(),
    se             = sd_abs_gap / sqrt(n),
    # One-sample t-test: is mean |gap| significantly greater than 0?
    t_stat         = mean_abs_gap / se,
    p_value        = pt(-abs(t_stat), df = n - 1, lower.tail = TRUE),
    .groups        = "drop"
  ) %>%
  mutate(across(c(mean_abs_gap, median_abs_gap, sd_abs_gap, se), ~round(.x, 4)),
         p_value    = signif(p_value, 3),
         model_size = factor(model_size, levels = SIZE_LEVELS))

cat("\n=== Overall absolute deviation from frequency expectation ===\n")
cat("(Mean |preference - log_freq_ratio|  -  not affected by cancellation)\n")
print(abs_overall_summary %>%
        select(corpus, model_size, mean_abs_gap, median_abs_gap, se, p_value, n))

# ── By frequency bin (signed and absolute) ───────────────────────────────────
freq_summary <- gap_base %>%
  group_by(corpus, model_size, freq_bin) %>%
  summarise(
    mean_gap      = mean(signed_gap,      na.rm = TRUE),
    mean_abs_gap  = mean(abs(signed_gap), na.rm = TRUE),
    se_signed     = sd(signed_gap,        na.rm = TRUE) / sqrt(n()),
    se_abs        = sd(abs(signed_gap),   na.rm = TRUE) / sqrt(n()),
    n             = n(),
    pct_over      = mean(signed_gap > 0,  na.rm = TRUE),
    .groups       = "drop"
  ) %>%
  mutate(across(c(mean_gap, mean_abs_gap, se_signed, se_abs), ~round(.x, 4)),
         pct_over   = round(pct_over, 3),
         model_size = factor(model_size, levels = SIZE_LEVELS))

cat("\n=== Signed and absolute deviation by overall frequency bin ===\n")
print(freq_summary)
```

```{r overpred-overall-plot, fig.width = 10, fig.height = 4}
# Signed gap  -  direction of bias
ggplot(overall_summary,
       aes(x = model_size, y = mean_gap,
           colour = mean_gap > 0, group = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = mean_gap - 1.96 * se,
                    ymax = mean_gap + 1.96 * se),
                width = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("TRUE" = "#d62728", "FALSE" = "#1f77b4"),
                      labels = c("TRUE" = "Over-predicts", "FALSE" = "Under-predicts"),
                      name = NULL) +
  scale_y_continuous(name = "Mean signed gap  (preference - log_freq_ratio)") +
  scale_x_discrete(name = "Model size") +
  facet_wrap(~corpus) +
  labs(title    = "Direction of bias: does the model over- or under-predict the alpha order?",
       subtitle = "All checkpoints pooled  .  Error bars = 95% CI") +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

```{r overpred-abs-overall-plot, fig.width = 10, fig.height = 4}
# Absolute gap  -  magnitude of deviation regardless of direction
ggplot(abs_overall_summary,
       aes(x = model_size, y = mean_abs_gap,
           colour = model_size, group = 1)) +
  geom_errorbar(aes(ymin = mean_abs_gap - 1.96 * se,
                    ymax = mean_abs_gap + 1.96 * se),
                width = 0.2, linewidth = 0.8) +
  geom_point(size = 3) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_y_continuous(name = "Mean |preference - log_freq_ratio|",
                     limits = c(0, NA)) +
  scale_x_discrete(name = "Model size") +
  facet_wrap(~corpus) +
  labs(
    title    = "Magnitude of deviation from frequency expectation (absolute gap)",
    subtitle = "Not affected by cancellation of opposite errors  .  All checkpoints pooled  .  Error bars = 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

```{r overpred-byfreq-plot, fig.width = 12, fig.height = 9}
# Side-by-side: signed and absolute gap by frequency bin
freq_long <- freq_summary %>%
  pivot_longer(cols      = c(mean_gap, mean_abs_gap),
               names_to  = "metric",
               values_to = "value") %>%
  mutate(
    se_val = if_else(metric == "mean_gap", se_signed, se_abs),
    metric = factor(metric,
                    levels = c("mean_gap", "mean_abs_gap"),
                    labels = c("Signed gap\n(cancellation possible)",
                               "Absolute gap\n(no cancellation)"))
  )

ggplot(freq_long,
       aes(x = freq_bin, y = value,
           colour = model_size, group = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = value - 1.96 * se_val,
                    ymax = value + 1.96 * se_val),
                width = 0.2, position = position_dodge(0.4), linewidth = 0.7) +
  geom_line(linewidth = 0.8, position = position_dodge(0.4)) +
  geom_point(size = 3,   position = position_dodge(0.4)) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_discrete(name = "Times binomial seen in corpus  (total alpha + beta)") +
  scale_y_continuous(name = "Gap (log-odds)") +
  facet_grid(metric ~ corpus, scales = "free_y") +
  labs(
    title    = "Signed vs. absolute deviation from frequency expectation by exposure level",
    subtitle = "All checkpoints pooled  .  Error bars = 95% CI"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.text.y = element_text(size = 9, angle = 0, hjust = 0))
```

## Reconciling the Frequency-Bin Plot with the Scatter Plot

The frequency-bin plot uses four bins (1-10, 10-100, 100-1k, **>1k**), where the
final bin collapses `total_seen` from 1,000 to 100,000+ into a single average.
If the gap peaks somewhere in the 1k-10k range and then declines at very high
exposure (as the continuous LOESS suggests), the ">1k" bin average will be
pulled downward by the dense right tail, making it appear smaller than the
"100-1k" bin  -  producing an apparent contradiction.

The plot below breaks the ">1k" bin into finer sub-bins to reveal whether the
relationship is non-monotone at high frequency.

```{r overpred-finebins, cache=TRUE, fig.width = 13, fig.height = 9}
fine_summary <- gap_base %>%
  mutate(
    freq_bin_fine = cut(
      log_total,
      breaks = c(0, log(10), log(100), log(1000), log(3000), log(10000), log(30000), log(100000), Inf),
      labels = c("1-10", "10-100", "100-1k",
                 "1k-3k", "3k-10k", "10k-30k", "30k-100k", ">100k"),
      include.lowest = TRUE
    )
  ) %>%
  group_by(corpus, model_size, freq_bin_fine) %>%
  summarise(
    mean_abs_gap = mean(abs(signed_gap), na.rm = TRUE),
    se_abs       = sd(abs(signed_gap),   na.rm = TRUE) / sqrt(n()),
    n            = n(),
    .groups      = "drop"
  ) %>%
  mutate(model_size = factor(model_size, levels = SIZE_LEVELS))

ggplot(fine_summary,
       aes(x = freq_bin_fine, y = mean_abs_gap,
           colour = model_size, group = model_size)) +
  geom_errorbar(aes(ymin = mean_abs_gap - 1.96 * se_abs,
                    ymax = mean_abs_gap + 1.96 * se_abs),
                width = 0.3, position = position_dodge(0.5), linewidth = 0.7) +
  geom_line(linewidth = 0.9, position = position_dodge(0.5)) +
  geom_point(size = 3,    position = position_dodge(0.5)) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_discrete(name = "Times binomial seen in corpus  (total alpha + beta)") +
  scale_y_continuous(name = "|preference - log_freq_ratio|  (log-odds)",
                     limits = c(0, NA)) +
  facet_wrap(~corpus) +
  labs(
    title    = "Absolute gap by fine-grained frequency bin   -   reconciling bin plot vs. scatter LOESS",
    subtitle = paste0(
      "The coarse '>1k' bin hid the high-frequency tail. If gap peaks at 1k-10k\n",
      "then declines, the '>1k' average is pulled down  -  explaining the apparent contradiction.\n",
      "All checkpoints pooled  -  Error bars = 95% CI")
  ) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(angle = 35, hjust = 1)
  )
```

```{r overpred-pct-plot, fig.width = 12, fig.height = 6}
# What proportion of binomials are over-predicted at each frequency level?
ggplot(freq_summary,
       aes(x = freq_bin, y = pct_over,
           colour = model_size, group = model_size)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  geom_line(linewidth = 0.8, position = position_dodge(0.4)) +
  geom_point(size = 3,   position = position_dodge(0.4)) +
  scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
  scale_x_discrete(name = "Times binomial seen in corpus  (total alpha + beta)") +
  scale_y_continuous(name = "Proportion of binomials over-predicted",
                     labels = scales::percent_format(accuracy = 1),
                     limits = c(0, 1)) +
  facet_wrap(~corpus) +
  labs(
    title    = "Proportion of binomials where model is more alpha-preferring than relative frequency predicts",
    subtitle = "Dashed line = 50% (no systematic bias)  .  All checkpoints pooled"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```
