---
title: "GenPref × Signed Gap & Preference — Results Only (No Refitting)"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Overview

This document loads pre-fitted brms models from disk and renders all coefficient
plots and marginal effect plots. No models are refit and no data is loaded from
DuckDB. All six analyses (3 signed-gap × 3 preference) are covered.

---

```{r libraries}
library(tidyverse)
library(brms)
library(patchwork)
```

```{r config}
# Adjust this if your models/ folder is not relative to this .Rmd file.
# Common alternatives:
#   MODELS_DIR <- "models"          # same directory as this .Rmd
#   MODELS_DIR <- "../models"       # one level up
#   MODELS_DIR <- "/absolute/path/to/models"
MODELS_DIR <- "models"

# ── Term levels/labels ────────────────────────────────────────────────────────
TERM_LEVELS <- c("freq_prob_c", "genpref_c", "log_total_c",
                 "freq_prob_c:log_total_c", "genpref_c:log_total_c")
TERM_LABELS <- c("RelFreq\n(P(alpha)−0.5)", "GenPref",
                 "Overall freq\n(ln total)",
                 "RelFreq ×\nOverall freq", "GenPref ×\nOverall freq")

TERM_LEVELS_CK <- c(
  "genpref_c", "freq_prob_c", "log_total_c", "tokens_c",
  "genpref_c:tokens_c", "freq_prob_c:tokens_c", "log_total_c:tokens_c"
)
TERM_LABELS_CK <- c(
  "GenPref", "RelFreq\n(P(alpha)−0.5)", "Overall freq\n(ln total)",
  "Checkpoint\n(tokens, z)",
  "GenPref ×\nCheckpoint", "RelFreq ×\nCheckpoint", "Overall freq ×\nCheckpoint"
)

SIZE_LEVELS  <- c("125M", "350M", "1.3B")
SIZE_COLOURS <- c("125M" = "#1b7837", "350M" = "#762a83", "1.3B" = "#b2182b")

PRED_SPECS <- list(
  list(
    var    = "genpref_c",
    vals   = c(-0.25, 0, 0.25),
    labels = c("Low GenPref (−0.25)", "Neutral (0)", "High GenPref (+0.25)"),
    cols   = c("Low GenPref (−0.25)"  = "#d73027",
               "Neutral (0)"          = "#999999",
               "High GenPref (+0.25)" = "#1a9850"),
    title  = "Effect of GenPref over training"
  ),
  list(
    var    = "freq_prob_c",
    vals   = c(-0.25, 0, 0.25),
    labels = c("Low RelFreq (−0.25)", "Neutral (0)", "High RelFreq (+0.25)"),
    cols   = c("Low RelFreq (−0.25)"  = "#d73027",
               "Neutral (0)"          = "#999999",
               "High RelFreq (+0.25)" = "#1a9850"),
    title  = "Effect of RelFreq over training"
  ),
  list(
    var    = "log_total_c",
    vals   = c(-1, 0, 1),
    labels = c("Low overall freq (−1 SD)", "Mean overall freq", "High overall freq (+1 SD)"),
    cols   = c("Low overall freq (−1 SD)"  = "#d73027",
               "Mean overall freq"          = "#999999",
               "High overall freq (+1 SD)"  = "#1a9850"),
    title  = "Effect of Overall Frequency over training"
  )
)
```

```{r helpers}
parse_corpus <- function(x) {
  case_when(
    grepl("babylm",  x, ignore.case = TRUE) ~ "BabyLM",
    grepl("opt.c4",  x, ignore.case = TRUE) ~ "C4",   # . matches - or _ (slug or original)
    TRUE ~ NA_character_
  )
}

parse_size <- function(x) {
  case_when(
    grepl("125m",    x, ignore.case = TRUE) ~ "125M",
    grepl("350m",    x, ignore.case = TRUE) ~ "350M",
    grepl("1[._]3b", x, ignore.case = TRUE) ~ "1.3B",  # matches 1.3b or 1_3b (slug)
    TRUE ~ NA_character_
  )
}

normalise_terms <- function(term) {
  dplyr::recode(term,
    "tokens_c:genpref_c"      = "genpref_c:tokens_c",
    "tokens_c:freq_prob_c"    = "freq_prob_c:tokens_c",
    "tokens_c:log_total_c"    = "log_total_c:tokens_c",
    "log_total_c:genpref_c"   = "genpref_c:log_total_c",
    "log_total_c:freq_prob_c" = "freq_prob_c:log_total_c"
  )
}

extract_coefs <- function(fit, model_name, term_levels, term_labels) {
  fe <- as.data.frame(fixef(fit))
  fe$term <- normalise_terms(rownames(fe))
  fe %>%
    as_tibble() %>%
    rename(estimate = Estimate, se = Est.Error, lo = Q2.5, hi = Q97.5) %>%
    filter(term != "Intercept") %>%
    mutate(
      term       = factor(term, levels = term_levels, labels = term_labels),
      sig        = if_else(lo > 0 | hi < 0, "*", ""),
      model_name = model_name,
      corpus     = parse_corpus(model_name),
      model_size = factor(parse_size(model_name), levels = SIZE_LEVELS)
    )
}

coef_plot <- function(coef_df, title, subtitle,
                      ylab = "Coefficient  (95% credible interval)") {
  if (nrow(coef_df) == 0) {
    message("No data available for: ", title, " — skipping plot.")
    return(invisible(NULL))
  }
  ggplot(coef_df, aes(x = term, y = estimate, colour = model_size)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_errorbar(aes(ymin = lo, ymax = hi),
                  width = 0.2, linewidth = 0.8,
                  position = position_dodge(0.5)) +
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_text(aes(label = sig, y = hi + 0.02),
              position = position_dodge(0.5), size = 4, vjust = 0) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_discrete(name = NULL) +
    scale_y_continuous(name = ylab) +
    facet_grid(corpus ~ ., scales = "free_y") +
    labs(title = title, subtitle = subtitle) +
    theme_bw(base_size = 11) +
    theme(legend.position  = "bottom",
          panel.grid.minor = element_blank(),
          axis.text.x      = element_text(size = 9))
}

# ── Marginal predictions from a single loaded fit ────────────────────────────
# Uses fitted(..., re_formula = NA) so no group-level draws are needed,
# keeping memory and compute low.
pred_grid_for <- function(predictor, levels_vec, label_vec, fit) {
  tk_seq <- seq(-2, 2, length.out = 30)
  grid <- expand.grid(
    tokens_c    = tk_seq,
    pred_val    = levels_vec,
    genpref_c   = 0,
    freq_prob_c = 0,
    log_total_c = 0
  )
  grid[[predictor]] <- grid$pred_val
  pred <- fitted(fit, newdata = grid, re_formula = NA, summary = TRUE)
  grid %>%
    mutate(
      estimate = pred[, "Estimate"],
      lo       = pred[, "Q2.5"],
      hi       = pred[, "Q97.5"],
      pred_f   = factor(pred_val, levels = levels_vec, labels = label_vec)
    ) %>%
    select(tokens_c, pred_f, estimate, lo, hi)
}

# ── Predictor summary from a loaded fit ───────────────────────────────────────
pred_summary <- function(fit, slug) {
  d    <- fit$data
  vars <- intersect(c("genpref_c", "freq_prob_c", "log_total_c", "tokens_c"), names(d))
  var_labels <- c(
    genpref_c   = "GenPref",
    freq_prob_c = "RelFreq (P(alpha)−0.5)",
    log_total_c = "Overall freq (ln total)",
    tokens_c    = "Checkpoint (tokens, z)"
  )
  purrr::map_dfr(vars, function(v) {
    x <- d[[v]]
    tibble(
      model    = slug,
      variable = var_labels[v],
      n        = sum(!is.na(x)),
      min      = round(min(x,    na.rm = TRUE), 3),
      median   = round(median(x, na.rm = TRUE), 3),
      mean     = round(mean(x,   na.rm = TRUE), 3),
      max      = round(max(x,    na.rm = TRUE), 3)
    )
  })
}

# ── Discover saved models from the models/ directory ─────────────────────────
# Each fit was saved as  <prefix><slug>.rds  where prefix is one of:
#   sg_final_, sg_pooled_, sg_ckpt_, pref_final_, pref_pooled_, pref_ckpt_
# The slug is  tolower(gsub("[^a-z0-9]", "_", model_name)).
# parse_corpus() / parse_size() are designed to match slug patterns directly.
list_fits <- function(prefix) {
  pat   <- paste0("^", prefix, ".+\\.rds$")
  paths <- list.files(MODELS_DIR, pattern = pat, full.names = TRUE)
  if (length(paths) == 0)
    warning(sprintf("No .rds files found for prefix '%s' in '%s'", prefix, MODELS_DIR))
  paths
}
```

---

## Discover fitted models

```{r path-diagnostic}
cat("Working directory:", getwd(), "\n")
cat("MODELS_DIR resolves to:", normalizePath(MODELS_DIR, mustWork = FALSE), "\n")
cat("Directory exists:", dir.exists(MODELS_DIR), "\n\n")

all_rds <- list.files(MODELS_DIR, pattern = "\\.rds$", full.names = FALSE)
cat(sprintf("%d .rds file(s) found in models dir:\n", length(all_rds)))
cat(paste(" ", all_rds, collapse = "\n"), "\n")
```

```{r discover-models}
# We discover what's available rather than hard-coding model names.
# All six prefixes:
PREFIXES <- c(
  sg_final   = "sg_final_",
  sg_pooled  = "sg_pooled_",
  sg_ckpt    = "sg_ckpt_",
  pref_final = "pref_final_",
  pref_pooled= "pref_pooled_",
  pref_ckpt  = "pref_ckpt_"
)

available <- map(PREFIXES, list_fits)
walk2(names(available), available, ~ {
  cat(sprintf("%-12s : %d file(s)\n", .x, length(.y)))
  walk(.y, ~ cat("  ", basename(.x), "\n"))
})
```

---

# Part I — Signed Gap

---

## Analysis 1: Signed Gap — Final Checkpoint

```{r sg-final-coefs, fig.width=13, fig.height=7}
sg_results_final <- map(available$sg_final, function(path) {
  slug <- sub("\\.rds$", "", sub("^sg_final_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS, TERM_LABELS),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
sg_coefs_final <- map_dfr(sg_results_final, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(sg_results_final, "summ"), n = Inf)

coef_plot(
  sg_coefs_final,
  title    = "Predictors of signed gap — final checkpoint",
  subtitle = "signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  * = CI excludes zero"
)
```

---

## Analysis 2: Signed Gap — Checkpoints Pooled

```{r sg-pooled-coefs, fig.width=13, fig.height=7}
sg_results_pooled <- map(available$sg_pooled, function(path) {
  slug <- sub("\\.rds$", "", sub("^sg_pooled_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS, TERM_LABELS),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
sg_coefs_pooled <- map_dfr(sg_results_pooled, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(sg_results_pooled, "summ"), n = Inf)

coef_plot(
  sg_coefs_pooled,
  title    = "Predictors of signed gap — all checkpoints pooled",
  subtitle = "signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  * = CI excludes zero"
)
```

---

## Analysis 3: Signed Gap — Checkpoint-Interacted

### Plot 3a: Coefficients

```{r sg-ckpt-coefs, fig.width=14, fig.height=7}
sg_results_ckpt <- map(available$sg_ckpt, function(path) {
  slug <- sub("\\.rds$", "", sub("^sg_ckpt_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS_CK, TERM_LABELS_CK),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
sg_coefs_ckpt <- map_dfr(sg_results_ckpt, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(sg_results_ckpt, "summ"), n = Inf)

coef_plot(
  sg_coefs_ckpt,
  title    = "Predictors of signed gap — checkpoint as continuous predictor",
  subtitle = "signed_gap ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1|binom)  |  * = CI excludes zero"
)
```

### Plot 3b: Marginal Effects Over Training

```{r sg-ckpt-marginal, fig.width=14, fig.height=10}
# Load each ckpt fit once and compute all three marginal grids, then plot.
sg_marginal_all <- tibble()

for (path in available$sg_ckpt) {
  slug   <- sub("\\.rds$", "", sub("^sg_ckpt_", "", basename(path)))
  message("Marginals: ", basename(path))
  fit    <- readRDS(path)

  for (spec in PRED_SPECS) {
    sg_marginal_all <- bind_rows(
      sg_marginal_all,
      pred_grid_for(spec$var, spec$vals, spec$labels, fit) %>%
        mutate(
          predictor   = spec$var,
          model_label = slug,
          corpus      = parse_corpus(slug),
          model_size  = factor(parse_size(slug), levels = SIZE_LEVELS)
        )
    )
  }
  rm(fit); gc()
}

if (nrow(sg_marginal_all) == 0) {
  message("No sg_ckpt marginal data available — skipping plots.")
} else {
  for (spec in PRED_SPECS) {
    df <- filter(sg_marginal_all, predictor == spec$var)
    if (nrow(df) == 0) { message("No marginal data for: ", spec$title); next }
    p  <- ggplot(df, aes(x = tokens_c, y = estimate,
                         colour = pred_f, fill = pred_f)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
      geom_line(linewidth = 0.9) +
      scale_colour_manual(values = spec$cols, name = NULL) +
      scale_fill_manual(values   = spec$cols, name = NULL) +
      scale_x_continuous(name = "Checkpoint  (tokens, z-scored within model)") +
      scale_y_continuous(name = "Predicted signed gap") +
      facet_wrap(~ model_label, scales = "free_y", ncol = 3) +
      labs(title    = spec$title,
           subtitle = "Other predictors held at 0  |  95% posterior credible ribbon  |  population-level") +
      theme_bw(base_size = 11) +
      theme(legend.position = "bottom", panel.grid.minor = element_blank())
    print(p)
  }
}
```

---

# Part II — Preference

---

## Analysis 4: Preference — Final Checkpoint

```{r pref-final-coefs, fig.width=13, fig.height=7}
pref_results_final <- map(available$pref_final, function(path) {
  slug <- sub("\\.rds$", "", sub("^pref_final_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS, TERM_LABELS),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
pref_coefs_final <- map_dfr(pref_results_final, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(pref_results_final, "summ"), n = Inf)

coef_plot(
  pref_coefs_final,
  title    = "Predictors of preference — final checkpoint",
  subtitle = "preference ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  * = CI excludes zero"
)
```

---

## Analysis 5: Preference — Checkpoints Pooled

```{r pref-pooled-coefs, fig.width=13, fig.height=7}
pref_results_pooled <- map(available$pref_pooled, function(path) {
  slug <- sub("\\.rds$", "", sub("^pref_pooled_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS, TERM_LABELS),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
pref_coefs_pooled <- map_dfr(pref_results_pooled, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(pref_results_pooled, "summ"), n = Inf)

coef_plot(
  pref_coefs_pooled,
  title    = "Predictors of preference — all checkpoints pooled",
  subtitle = "preference ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  * = CI excludes zero"
)
```

---

## Analysis 6: Preference — Checkpoint-Interacted

### Plot 6a: Coefficients

```{r pref-ckpt-coefs, fig.width=14, fig.height=7}
pref_results_ckpt <- map(available$pref_ckpt, function(path) {
  slug <- sub("\\.rds$", "", sub("^pref_ckpt_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS_CK, TERM_LABELS_CK),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
pref_coefs_ckpt <- map_dfr(pref_results_ckpt, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(pref_results_ckpt, "summ"), n = Inf)

coef_plot(
  pref_coefs_ckpt,
  title    = "Predictors of preference — checkpoint as continuous predictor",
  subtitle = "preference ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1|binom)  |  * = CI excludes zero"
)
```

### Plot 6b: Marginal Effects Over Training

```{r pref-ckpt-marginal, fig.width=14, fig.height=10}
pref_marginal_all <- tibble()

for (path in available$pref_ckpt) {
  slug   <- sub("\\.rds$", "", sub("^pref_ckpt_", "", basename(path)))
  message("Marginals: ", basename(path))
  fit    <- readRDS(path)

  for (spec in PRED_SPECS) {
    pref_marginal_all <- bind_rows(
      pref_marginal_all,
      pred_grid_for(spec$var, spec$vals, spec$labels, fit) %>%
        mutate(
          predictor   = spec$var,
          model_label = slug,
          corpus      = parse_corpus(slug),
          model_size  = factor(parse_size(slug), levels = SIZE_LEVELS)
        )
    )
  }
  rm(fit); gc()
}

if (nrow(pref_marginal_all) == 0) {
  message("No pref_ckpt marginal data available — skipping plots.")
} else {
  for (spec in PRED_SPECS) {
    df <- filter(pref_marginal_all, predictor == spec$var)
    if (nrow(df) == 0) { message("No marginal data for: ", spec$title); next }
    p  <- ggplot(df, aes(x = tokens_c, y = estimate,
                         colour = pred_f, fill = pred_f)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
      geom_line(linewidth = 0.9) +
      scale_colour_manual(values = spec$cols, name = NULL) +
      scale_fill_manual(values   = spec$cols, name = NULL) +
      scale_x_continuous(name = "Checkpoint  (tokens, z-scored within model)") +
      scale_y_continuous(name = "Predicted preference") +
      facet_wrap(~ model_label, scales = "free_y", ncol = 3) +
      labs(title    = spec$title,
           subtitle = "Other predictors held at 0  |  95% posterior credible ribbon  |  population-level") +
      theme_bw(base_size = 11) +
      theme(legend.position = "bottom", panel.grid.minor = element_blank())
    print(p)
  }
}
```
