---
title: "GenPref × Signed Gap & Preference — Per-Model Analysis"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Overview

This script uses raw (prompt-level) data — the 51 per-prompt observations are
**not** averaged before modelling.  Six trained models are analysed separately
(3 BabyLM × 3 sizes; 3 C4 × 3 sizes).  Three analyses are run for each model,
first on the **signed gap** outcome and then on **preference** directly:

- **Signed gap**: `preference − log_freq_ratio` — the deviation of the model's
  preference from what corpus frequency alone would predict.
- **Preference**: raw `alpha_logprob − nonalpha_logprob`.

For each outcome, three analyses are run:

1. **Final checkpoint only** – brms at the last checkpoint.
2. **All checkpoints pooled** – same formula, all checkpoints.
3. **Checkpoint as continuous predictor** – `tokens` (scaled) included and
   interacted with every other predictor.

---

## Setup

```{r libraries}
library(tidyverse)
library(duckdb)
library(scales)
library(patchwork)
library(brms)
library(cmdstanr)
```

```{r brms-config}
MODELS_DIR <- "models"
dir.create(MODELS_DIR, showWarnings = FALSE, recursive = TRUE)

N_CORES <- parallel::detectCores(logical = FALSE)
message(sprintf("Detected %d physical cores", N_CORES))

BRMS_ITER      <- 6000
BRMS_WARMUP    <- 3000
BRMS_CHAINS    <- min(N_CORES, 8)
BRMS_CORES     <- BRMS_CHAINS
BRMS_SEED      <- 964
BRMS_BACKEND   <- "cmdstanr"
BRMS_NORMALIZE <- FALSE

BRMS_PRIORS <- c(
  prior(normal(0, 1), class = "b"),
  prior(normal(0, 1), class = "Intercept"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "sigma")
)
```

```{r helpers}
parse_corpus <- function(x) {
  case_when(
    grepl("babylm", x, ignore.case = TRUE) ~ "BabyLM",
    grepl("opt-c4", x, ignore.case = TRUE) ~ "C4",
    TRUE ~ NA_character_
  )
}

parse_size <- function(x) {
  case_when(
    grepl("125m",  x, ignore.case = TRUE) ~ "125M",
    grepl("350m",  x, ignore.case = TRUE) ~ "350M",
    grepl("1\\.3b", x, ignore.case = TRUE) ~ "1.3B",
    TRUE ~ NA_character_
  )
}

SIZE_LEVELS  <- c("125M", "350M", "1.3B")
SIZE_COLOURS <- c("125M" = "#1b7837", "350M" = "#762a83", "1.3B" = "#b2182b")

# Normalise brms interaction term names to a canonical order
normalise_terms <- function(term) {
  dplyr::recode(term,
    "tokens_c:genpref_c"      = "genpref_c:tokens_c",
    "tokens_c:freq_prob_c"    = "freq_prob_c:tokens_c",
    "tokens_c:log_total_c"    = "log_total_c:tokens_c",
    "log_total_c:genpref_c"   = "genpref_c:log_total_c",
    "log_total_c:freq_prob_c" = "freq_prob_c:log_total_c"
  )
}

# Extract fixed-effect coefficients from a brms fit into a tidy tibble
extract_coefs <- function(fit, model_name, term_levels, term_labels) {
  fe <- as.data.frame(fixef(fit))
  fe$term <- normalise_terms(rownames(fe))
  fe %>%
    as_tibble() %>%
    rename(estimate = Estimate, se = Est.Error, lo = Q2.5, hi = Q97.5) %>%
    filter(term != "Intercept") %>%
    mutate(
      term       = factor(term, levels = term_levels, labels = term_labels),
      sig        = if_else(lo > 0 | hi < 0, "*", ""),
      model_name = model_name,
      corpus     = parse_corpus(model_name),
      model_size = factor(parse_size(model_name), levels = SIZE_LEVELS)
    )
}

# Standard coefficient dot-and-whisker plot
coef_plot <- function(coef_df, title, subtitle, ylab = "Coefficient  (95% credible interval)") {
  ggplot(coef_df, aes(x = term, y = estimate, colour = model_size)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_errorbar(aes(ymin = lo, ymax = hi),
                  width = 0.2, linewidth = 0.8,
                  position = position_dodge(0.5)) +
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_text(aes(label = sig, y = hi + 0.02),
              position = position_dodge(0.5), size = 4, vjust = 0) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_discrete(name = NULL) +
    scale_y_continuous(name = ylab) +
    facet_grid(corpus ~ ., scales = "free_y") +
    labs(title = title, subtitle = subtitle) +
    theme_bw(base_size = 11) +
    theme(legend.position  = "bottom",
          panel.grid.minor = element_blank(),
          axis.text.x      = element_text(size = 9))
}

# Marginal effect plots for the checkpoint-interacted model
PRED_SPECS <- list(
  list(
    var    = "genpref_c",
    vals   = c(-0.25, 0, 0.25),
    labels = c("Low GenPref (−0.25)", "Neutral (0)", "High GenPref (+0.25)"),
    cols   = c("Low GenPref (−0.25)"  = "#d73027",
               "Neutral (0)"          = "#999999",
               "High GenPref (+0.25)" = "#1a9850"),
    title  = "Effect of GenPref over training"
  ),
  list(
    var    = "freq_prob_c",
    vals   = c(-0.25, 0, 0.25),
    labels = c("Low RelFreq (−0.25)", "Neutral (0)", "High RelFreq (+0.25)"),
    cols   = c("Low RelFreq (−0.25)"  = "#d73027",
               "Neutral (0)"          = "#999999",
               "High RelFreq (+0.25)" = "#1a9850"),
    title  = "Effect of RelFreq over training"
  ),
  list(
    var    = "log_total_c",
    vals   = c(-1, 0, 1),
    labels = c("Low overall freq (−1 SD)", "Mean overall freq", "High overall freq (+1 SD)"),
    cols   = c("Low overall freq (−1 SD)"  = "#d73027",
               "Mean overall freq"          = "#999999",
               "High overall freq (+1 SD)"  = "#1a9850"),
    title  = "Effect of Overall Frequency over training"
  )
)

pred_grid_for <- function(predictor, levels_vec, label_vec, fit, model_data_m) {
  tk_seq <- seq(
    min(model_data_m$tokens_c, na.rm = TRUE),
    max(model_data_m$tokens_c, na.rm = TRUE),
    length.out = 30
  )
  grid <- expand.grid(
    tokens_c    = tk_seq,
    pred_val    = levels_vec,
    genpref_c   = 0,
    freq_prob_c = 0,
    log_total_c = 0
  )
  grid[[predictor]] <- grid$pred_val
  pred <- fitted(fit, newdata = grid, re_formula = NA, summary = TRUE)
  grid %>%
    mutate(
      estimate  = pred[, "Estimate"],
      lo        = pred[, "Q2.5"],
      hi        = pred[, "Q97.5"],
      pred_f    = factor(pred_val, levels = levels_vec, labels = label_vec),
      predictor = predictor
    ) %>%
    select(tokens_c, pred_f, predictor, estimate, lo, hi)
}

marginal_plots_for <- function(fits_ckpt, model_data_ck, ylab) {
  marginal_all <- purrr::map_dfr(MODELS_LIST, function(m) {
    m_ck  <- filter(model_data_ck, model == m)
    m_lab <- paste0(parse_corpus(m), " ", parse_size(m))
    purrr::map_dfr(PRED_SPECS, function(spec) {
      pred_grid_for(spec$var, spec$vals, spec$labels, fits_ckpt[[m]], m_ck) %>%
        mutate(model_label = m_lab,
               corpus      = parse_corpus(m),
               model_size  = factor(parse_size(m), levels = SIZE_LEVELS))
    })
  })

  purrr::walk(PRED_SPECS, function(spec) {
    df <- filter(marginal_all, predictor == spec$var)
    p <- ggplot(df, aes(x = tokens_c, y = estimate,
                        colour = pred_f, fill = pred_f)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
      geom_line(linewidth = 0.9) +
      scale_colour_manual(values = spec$cols, name = NULL) +
      scale_fill_manual(values   = spec$cols, name = NULL) +
      scale_x_continuous(name = "Checkpoint  (tokens, z-scored within model)") +
      scale_y_continuous(name = ylab) +
      facet_wrap(~ model_label, scales = "free_y", ncol = 3) +
      labs(title    = spec$title,
           subtitle = "Other predictors held at 0  |  95% posterior credible ribbon  |  population-level") +
      theme_bw(base_size = 11) +
      theme(legend.position = "bottom", panel.grid.minor = element_blank())
    print(p)
  })
}
```

---

## Load Data

```{r load-raw, cache=TRUE}
csv_path     <- normalizePath("../Data/processed/checkpoint_results_with_exposures.csv",
                               mustWork = TRUE)
csv_path_fwd <- gsub("\\\\", "/", csv_path)

con <- dbConnect(duckdb::duckdb())
dbExecute(con, sprintf(
  "CREATE VIEW ckpt AS SELECT * FROM read_csv_auto('%s')",
  csv_path_fwd
))

message("Loading prompt-level data via DuckDB (no prompt aggregation) ...")
raw <- dbGetQuery(con, "
  SELECT
    model,
    tokens,
    binom,
    preference,
    ROUND(alpha_seen)                                AS alpha_seen,
    ROUND(beta_seen)                                 AS beta_seen,
    ROUND(alpha_seen) + ROUND(beta_seen)             AS total_seen,
    CASE WHEN alpha_seen > 0 AND beta_seen > 0
         THEN LN(alpha_seen / beta_seen)
         ELSE NULL END                               AS log_freq_ratio
  FROM ckpt
  ORDER BY model, tokens, binom
")
dbDisconnect(con, shutdown = TRUE)

message(sprintf("Loaded %s rows.", format(nrow(raw), big.mark = ",")))
```

```{r metadata}
raw <- raw %>%
  mutate(
    corpus     = parse_corpus(model),
    model_size = factor(parse_size(model), levels = SIZE_LEVELS),
    total_seen = as.numeric(total_seen),
    log_total  = if_else(total_seen > 0, log(total_seen), NA_real_)
  )

MODELS_LIST <- sort(unique(raw$model))
cat(sprintf("Models found: %d\n", length(MODELS_LIST)))
print(MODELS_LIST)
```

```{r load-genpref}
human_path <- normalizePath("../Data/all_human_data.csv", mustWork = TRUE)

GENPREF_FORMULA <- function(Form, Percept, Culture, Power, Intense,
                             Icon, Freq, Len, Lapse, BStress) {
  y <- 0.02191943 +
       0.23925834 * Form    +
       0.24889543 * Percept +
       0.41836997 * Culture +
       0.25967334 * Power   +
       0.01867604 * Intense +
       1.30365980 * Icon    +
       0.08553552 * Freq    +
       0.15241566 * Len     -
       0.19381657 * Lapse   +
       0.36019221 * BStress
  (1 / (1 + exp(-y))) - 0.5
}

genpref_raw <- read_csv(human_path, show_col_types = FALSE) %>%
  rename(BStress = `*BStress`) %>%
  select(Alpha, Form, Percept, Culture, Power, Intense,
         Icon, Freq, Len, Lapse, BStress, GenPref) %>%
  group_by(Alpha) %>%
  slice(1) %>%
  ungroup() %>%
  mutate(
    GenPref_norm = case_when(
      is.na(GenPref) ~ NA_real_,
      GenPref > 0.5  ~ GenPref - 0.5,
      GenPref < -0.5 ~ GenPref,
      TRUE           ~ GenPref - 0.5
    ),
    genpref = if_else(
      is.na(GenPref),
      GENPREF_FORMULA(Form, Percept, Culture, Power, Intense,
                      Icon, Freq, Len, Lapse, BStress),
      GenPref_norm
    ),
    binom_lower = tolower(Alpha)
  ) %>%
  select(binom_lower, genpref)
```

```{r join-centre}
model_data <- raw %>%
  mutate(binom_lower = tolower(binom)) %>%
  left_join(genpref_raw, by = "binom_lower") %>%
  filter(!is.na(genpref), !is.na(log_freq_ratio), !is.na(log_total)) %>%
  mutate(
    signed_gap  = preference - log_freq_ratio,
    freq_prob_c = plogis(log_freq_ratio) - 0.5,
    genpref_c   = genpref - mean(genpref,    na.rm = TRUE),
    log_total_c = log_total - mean(log_total, na.rm = TRUE)
  )

cat(sprintf("Rows after join and filtering: %s\n",
            format(nrow(model_data), big.mark = ",")))
cat(sprintf("Unique binomials with GenPref: %d\n",
            n_distinct(model_data$binom)))
```

```{r ckpt-prep}
# Shared dataset with tokens scaled within each model (used by analyses 3 and 6)
model_data_ck <- model_data %>%
  group_by(model) %>%
  mutate(tokens_c = as.numeric(scale(tokens))) %>%
  ungroup()
```

---

# Part I: Signed Gap Analyses

`signed_gap = preference − log_freq_ratio`

A positive value means the model is *more* alpha-preferring than corpus frequency
alone predicts; negative means *less*.  Predictors: `genpref_c` (GenPref,
centred), `freq_prob_c` (P(alpha in corpus) − 0.5), `log_total_c` (log total
exposures, centred).

---

## Analysis 1: Signed Gap — Final Checkpoint

Formula: `signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom)`

```{r sg-analysis1-fit, cache=TRUE, cache.lazy=FALSE}
sg_final_data <- setNames(lapply(MODELS_LIST, function(m) {
  model_data %>%
    filter(model == m) %>%
    group_by(binom) %>%
    filter(tokens == max(tokens)) %>%
    ungroup()
}), MODELS_LIST)

sg_fits_final <- setNames(lapply(MODELS_LIST, function(m) {
  slug <- tolower(gsub("[^a-z0-9]", "_", m))
  brm(
    signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
    data    = sg_final_data[[m]],
    family  = gaussian(),
    prior   = BRMS_PRIORS,
    iter    = BRMS_ITER,
    warmup  = BRMS_WARMUP,
    chains  = BRMS_CHAINS,
    cores   = BRMS_CORES,
    file      = file.path(MODELS_DIR, paste0("sg_final_", slug)),
    seed      = BRMS_SEED,
    backend   = BRMS_BACKEND,
    normalize = BRMS_NORMALIZE
  )
}), MODELS_LIST)

for (m in MODELS_LIST) {
  cat(sprintf("\n=== Signed gap — final checkpoint: %s ===\n", m))
  print(summary(sg_fits_final[[m]]))
}
```

### Plot 1: Coefficients — Signed Gap, Final Checkpoint

```{r sg-plot1, fig.width=13, fig.height=7}
TERM_LEVELS <- c("freq_prob_c", "genpref_c", "log_total_c",
                 "freq_prob_c:log_total_c", "genpref_c:log_total_c")
TERM_LABELS <- c("RelFreq\n(P(alpha)−0.5)", "GenPref",
                 "Overall freq\n(ln total)",
                 "RelFreq ×\nOverall freq", "GenPref ×\nOverall freq")

sg_coefs_final <- purrr::map_dfr(MODELS_LIST, function(m) {
  extract_coefs(sg_fits_final[[m]], m, TERM_LEVELS, TERM_LABELS)
})

coef_plot(
  sg_coefs_final,
  title    = "Predictors of signed gap — final checkpoint",
  subtitle = "signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  brms  |  * = CI excludes zero",
  ylab     = "Coefficient  (95% credible interval)"
)
```

---

## Analysis 2: Signed Gap — Checkpoints Pooled

Same formula as Analysis 1 but using **all checkpoints**.

```{r sg-analysis2-fit, cache=TRUE, cache.lazy=FALSE}
sg_fits_pooled <- setNames(lapply(MODELS_LIST, function(m) {
  slug <- tolower(gsub("[^a-z0-9]", "_", m))
  brm(
    signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
    data    = filter(model_data, model == m),
    family  = gaussian(),
    prior   = BRMS_PRIORS,
    iter    = BRMS_ITER,
    warmup  = BRMS_WARMUP,
    chains  = BRMS_CHAINS,
    cores   = BRMS_CORES,
    file      = file.path(MODELS_DIR, paste0("sg_pooled_", slug)),
    seed      = BRMS_SEED,
    backend   = BRMS_BACKEND,
    normalize = BRMS_NORMALIZE
  )
}), MODELS_LIST)

for (m in MODELS_LIST) {
  cat(sprintf("\n=== Signed gap — checkpoints pooled: %s ===\n", m))
  print(summary(sg_fits_pooled[[m]]))
}
```

### Plot 2: Coefficients — Signed Gap, Checkpoints Pooled

```{r sg-plot2, fig.width=13, fig.height=7}
sg_coefs_pooled <- purrr::map_dfr(MODELS_LIST, function(m) {
  extract_coefs(sg_fits_pooled[[m]], m, TERM_LEVELS, TERM_LABELS)
})

coef_plot(
  sg_coefs_pooled,
  title    = "Predictors of signed gap — all checkpoints pooled",
  subtitle = "signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  brms  |  * = CI excludes zero",
  ylab     = "Coefficient  (95% credible interval)"
)
```

---

## Analysis 3: Signed Gap — Checkpoint as Continuous Predictor

`tokens` (z-scored within each model) is interacted with every other predictor,
letting us ask how each effect evolves over training.

Formula: `signed_gap ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1 | binom)`

```{r sg-analysis3-fit, cache=TRUE, cache.lazy=FALSE}
TERM_LEVELS_CK <- c(
  "genpref_c", "freq_prob_c", "log_total_c", "tokens_c",
  "genpref_c:tokens_c", "freq_prob_c:tokens_c", "log_total_c:tokens_c"
)
TERM_LABELS_CK <- c(
  "GenPref", "RelFreq\n(P(alpha)−0.5)", "Overall freq\n(ln total)",
  "Checkpoint\n(tokens, z)",
  "GenPref ×\nCheckpoint", "RelFreq ×\nCheckpoint", "Overall freq ×\nCheckpoint"
)

sg_fits_ckpt <- setNames(lapply(MODELS_LIST, function(m) {
  slug <- tolower(gsub("[^a-z0-9]", "_", m))
  brm(
    signed_gap ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1 | binom),
    data    = filter(model_data_ck, model == m),
    family  = gaussian(),
    prior   = BRMS_PRIORS,
    iter    = BRMS_ITER,
    warmup  = BRMS_WARMUP,
    chains  = BRMS_CHAINS,
    cores   = BRMS_CORES,
    file      = file.path(MODELS_DIR, paste0("sg_ckpt_", slug)),
    seed      = BRMS_SEED,
    backend   = BRMS_BACKEND,
    normalize = BRMS_NORMALIZE
  )
}), MODELS_LIST)

for (m in MODELS_LIST) {
  cat(sprintf("\n=== Signed gap — checkpoint-interacted: %s ===\n", m))
  print(summary(sg_fits_ckpt[[m]]))
}
```

### Plot 3a: Coefficients — Signed Gap, Checkpoint-Interacted

```{r sg-plot3a, fig.width=14, fig.height=7}
sg_coefs_ckpt <- purrr::map_dfr(MODELS_LIST, function(m) {
  extract_coefs(sg_fits_ckpt[[m]], m, TERM_LEVELS_CK, TERM_LABELS_CK)
})

coef_plot(
  sg_coefs_ckpt,
  title    = "Predictors of signed gap — checkpoint as continuous predictor",
  subtitle = "signed_gap ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1|binom)  |  brms  |  * = CI excludes zero",
  ylab     = "Coefficient  (95% credible interval)"
)
```

### Plot 3b: Marginal Effects Over Training — Signed Gap

```{r sg-plot3b, fig.width=14, fig.height=10}
marginal_plots_for(sg_fits_ckpt, model_data_ck, ylab = "Predicted signed gap")
```

---

# Part II: Preference Analyses

`preference = alpha_logprob − nonalpha_logprob`

Same structure as Part I, modelling raw preference rather than the gap relative
to corpus frequency.

---

## Analysis 4: Preference — Final Checkpoint

Formula: `preference ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom)`

```{r pref-analysis4-fit, cache=TRUE, cache.lazy=FALSE}
pref_final_data <- setNames(lapply(MODELS_LIST, function(m) {
  model_data %>%
    filter(model == m) %>%
    group_by(binom) %>%
    filter(tokens == max(tokens)) %>%
    ungroup()
}), MODELS_LIST)

pref_fits_final <- setNames(lapply(MODELS_LIST, function(m) {
  slug <- tolower(gsub("[^a-z0-9]", "_", m))
  brm(
    preference ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
    data    = pref_final_data[[m]],
    family  = gaussian(),
    prior   = BRMS_PRIORS,
    iter    = BRMS_ITER,
    warmup  = BRMS_WARMUP,
    chains  = BRMS_CHAINS,
    cores   = BRMS_CORES,
    file      = file.path(MODELS_DIR, paste0("pref_final_", slug)),
    seed      = BRMS_SEED,
    backend   = BRMS_BACKEND,
    normalize = BRMS_NORMALIZE
  )
}), MODELS_LIST)

for (m in MODELS_LIST) {
  cat(sprintf("\n=== Preference — final checkpoint: %s ===\n", m))
  print(summary(pref_fits_final[[m]]))
}
```

### Plot 4: Coefficients — Preference, Final Checkpoint

```{r pref-plot4, fig.width=13, fig.height=7}
pref_coefs_final <- purrr::map_dfr(MODELS_LIST, function(m) {
  extract_coefs(pref_fits_final[[m]], m, TERM_LEVELS, TERM_LABELS)
})

coef_plot(
  pref_coefs_final,
  title    = "Predictors of preference — final checkpoint",
  subtitle = "preference ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  brms  |  * = CI excludes zero",
  ylab     = "Coefficient  (95% credible interval)"
)
```

---

## Analysis 5: Preference — Checkpoints Pooled

```{r pref-analysis5-fit, cache=TRUE, cache.lazy=FALSE}
pref_fits_pooled <- setNames(lapply(MODELS_LIST, function(m) {
  slug <- tolower(gsub("[^a-z0-9]", "_", m))
  brm(
    preference ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1 | binom),
    data    = filter(model_data, model == m),
    family  = gaussian(),
    prior   = BRMS_PRIORS,
    iter    = BRMS_ITER,
    warmup  = BRMS_WARMUP,
    chains  = BRMS_CHAINS,
    cores   = BRMS_CORES,
    file      = file.path(MODELS_DIR, paste0("pref_pooled_", slug)),
    seed      = BRMS_SEED,
    backend   = BRMS_BACKEND,
    normalize = BRMS_NORMALIZE
  )
}), MODELS_LIST)

for (m in MODELS_LIST) {
  cat(sprintf("\n=== Preference — checkpoints pooled: %s ===\n", m))
  print(summary(pref_fits_pooled[[m]]))
}
```

### Plot 5: Coefficients — Preference, Checkpoints Pooled

```{r pref-plot5, fig.width=13, fig.height=7}
pref_coefs_pooled <- purrr::map_dfr(MODELS_LIST, function(m) {
  extract_coefs(pref_fits_pooled[[m]], m, TERM_LEVELS, TERM_LABELS)
})

coef_plot(
  pref_coefs_pooled,
  title    = "Predictors of preference — all checkpoints pooled",
  subtitle = "preference ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  brms  |  * = CI excludes zero",
  ylab     = "Coefficient  (95% credible interval)"
)
```

---

## Analysis 6: Preference — Checkpoint as Continuous Predictor

Formula: `preference ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1 | binom)`

```{r pref-analysis6-fit, cache=TRUE, cache.lazy=FALSE}
pref_fits_ckpt <- setNames(lapply(MODELS_LIST, function(m) {
  slug <- tolower(gsub("[^a-z0-9]", "_", m))
  brm(
    preference ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1 | binom),
    data    = filter(model_data_ck, model == m),
    family  = gaussian(),
    prior   = BRMS_PRIORS,
    iter    = BRMS_ITER,
    warmup  = BRMS_WARMUP,
    chains  = BRMS_CHAINS,
    cores   = BRMS_CORES,
    file      = file.path(MODELS_DIR, paste0("pref_ckpt_", slug)),
    seed      = BRMS_SEED,
    backend   = BRMS_BACKEND,
    normalize = BRMS_NORMALIZE
  )
}), MODELS_LIST)

for (m in MODELS_LIST) {
  cat(sprintf("\n=== Preference — checkpoint-interacted: %s ===\n", m))
  print(summary(pref_fits_ckpt[[m]]))
}
```

### Plot 6a: Coefficients — Preference, Checkpoint-Interacted

```{r pref-plot6a, fig.width=14, fig.height=7}
pref_coefs_ckpt <- purrr::map_dfr(MODELS_LIST, function(m) {
  extract_coefs(pref_fits_ckpt[[m]], m, TERM_LEVELS_CK, TERM_LABELS_CK)
})

coef_plot(
  pref_coefs_ckpt,
  title    = "Predictors of preference — checkpoint as continuous predictor",
  subtitle = "preference ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1|binom)  |  brms  |  * = CI excludes zero",
  ylab     = "Coefficient  (95% credible interval)"
)
```

### Plot 6b: Marginal Effects Over Training — Preference

```{r pref-plot6b, fig.width=14, fig.height=10}
marginal_plots_for(pref_fits_ckpt, model_data_ck, ylab = "Predicted preference")
```
