---
title: "GenPref × Signed Gap & Preference — Results Only (No Refitting)"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Overview

This document loads pre-fitted brms models from disk and renders all coefficient
plots and marginal effect plots. No models are refit and no data is loaded from
DuckDB. All six analyses (3 signed-gap × 3 preference) are covered.

Analyses 2 & 5 use the **per-checkpoint trajectory** approach: 20 evenly-spaced
checkpoints per model, each fit as a separate brms model with `log_total_c`
z-scored within that checkpoint's items. Files are named
`sg_traj_{slug}_ck{01-20}.rds` / `pref_traj_{slug}_ck{01-20}.rds`.

---

```{r libraries}
library(tidyverse)
library(brms)
library(patchwork)
```

```{r config}
# Adjust this if your models/ folder is not relative to this .Rmd file.
# Common alternatives:
#   MODELS_DIR <- "models"          # same directory as this .Rmd
#   MODELS_DIR <- "../models"       # one level up
#   MODELS_DIR <- "/absolute/path/to/models"
MODELS_DIR <- "models"

# ── Term levels/labels ────────────────────────────────────────────────────────
TERM_LEVELS <- c("freq_prob_c", "genpref_c", "log_total_c",
                 "freq_prob_c:log_total_c", "genpref_c:log_total_c")
TERM_LABELS <- c("RelFreq\n(P(alpha)−0.5)", "GenPref",
                 "Overall freq\n(ln total)",
                 "RelFreq ×\nOverall freq", "GenPref ×\nOverall freq")

TERM_LEVELS_CK <- c(
  "genpref_c", "freq_prob_c", "log_total_c", "tokens_c",
  "genpref_c:tokens_c", "freq_prob_c:tokens_c", "log_total_c:tokens_c"
)
TERM_LABELS_CK <- c(
  "GenPref", "RelFreq\n(P(alpha)−0.5)", "Overall freq\n(ln total)",
  "Checkpoint\n(tokens, z)",
  "GenPref ×\nCheckpoint", "RelFreq ×\nCheckpoint", "Overall freq ×\nCheckpoint"
)

SIZE_LEVELS  <- c("125M", "350M", "1.3B")
SIZE_COLOURS <- c("125M" = "#1b7837", "350M" = "#762a83", "1.3B" = "#b2182b")

N_TRAJ_CHECKPOINTS <- 20   # must match the value used when fitting

# Path to the checkpoint CSV — used only to recover per-checkpoint token counts
# for trajectory plot x-axes.  Set to NULL to disable (x-axis will be NA).
CSV_PATH <- tryCatch(
  normalizePath("../Data/processed/checkpoint_results_with_exposures.csv",
                mustWork = TRUE),
  error = function(e) NULL
)

PRED_SPECS <- list(
  list(
    var    = "genpref_c",
    vals   = c(-0.25, 0, 0.25),
    labels = c("Low GenPref (−0.25)", "Neutral (0)", "High GenPref (+0.25)"),
    cols   = c("Low GenPref (−0.25)"  = "#d73027",
               "Neutral (0)"          = "#999999",
               "High GenPref (+0.25)" = "#1a9850"),
    title  = "Effect of GenPref over training"
  ),
  list(
    var    = "freq_prob_c",
    vals   = c(-0.25, 0, 0.25),
    labels = c("Low RelFreq (−0.25)", "Neutral (0)", "High RelFreq (+0.25)"),
    cols   = c("Low RelFreq (−0.25)"  = "#d73027",
               "Neutral (0)"          = "#999999",
               "High RelFreq (+0.25)" = "#1a9850"),
    title  = "Effect of RelFreq over training"
  ),
  list(
    var    = "log_total_c",
    vals   = c(-1, 0, 1),
    labels = c("Low overall freq (−1 SD)", "Mean overall freq", "High overall freq (+1 SD)"),
    cols   = c("Low overall freq (−1 SD)"  = "#d73027",
               "Mean overall freq"          = "#999999",
               "High overall freq (+1 SD)"  = "#1a9850"),
    title  = "Effect of Overall Frequency over training"
  )
)
```

```{r build-traj-meta, message=TRUE}
# Build a lookup table: slug × ck_idx → actual token count.
# brms only stores formula variables in fit$data, so 'tokens' is absent from the
# per-checkpoint .rds files.  We recover token counts by replaying the same
# sample_checkpoints() logic used during fitting — a fast, read-only DuckDB query.
if (!is.null(CSV_PATH)) {
  library(duckdb)
  CSV_PATH_FWD <- gsub("\\\\", "/", CSV_PATH)
  con_meta <- dbConnect(duckdb::duckdb())
  dbExecute(con_meta, sprintf(
    "CREATE VIEW ckpt AS SELECT * FROM read_csv_auto('%s')", CSV_PATH_FWD
  ))
  models_meta <- sort(dbGetQuery(con_meta,
    "SELECT DISTINCT model FROM ckpt")$model)

  TRAJ_META <- bind_rows(lapply(models_meta, function(m) {
    slug      <- tolower(gsub("[^a-z0-9]", "_", m))
    all_ckpts <- dbGetQuery(con_meta, sprintf(
      "SELECT DISTINCT tokens FROM ckpt WHERE model = '%s' ORDER BY tokens",
      gsub("'", "''", m)
    ))$tokens
    n   <- min(N_TRAJ_CHECKPOINTS, length(all_ckpts))
    idx <- unique(round(seq(1, length(all_ckpts), length.out = n)))
    tibble(slug = slug, ck_idx = seq_along(idx), tokens = all_ckpts[idx])
  }))

  dbDisconnect(con_meta, shutdown = TRUE); rm(con_meta, CSV_PATH_FWD, models_meta)
  message(sprintf("TRAJ_META: %d models, %d checkpoint rows total",
                  n_distinct(TRAJ_META$slug), nrow(TRAJ_META)))
} else {
  TRAJ_META <- NULL
  warning("CSV not found — trajectory x-axis will be NA (set CSV_PATH in the config chunk)")
}
```

```{r helpers}
parse_corpus <- function(x) {
  case_when(
    grepl("babylm",  x, ignore.case = TRUE) ~ "BabyLM",
    grepl("opt.c4",  x, ignore.case = TRUE) ~ "C4",   # . matches - or _ (slug or original)
    TRUE ~ NA_character_
  )
}

parse_size <- function(x) {
  case_when(
    grepl("125m",    x, ignore.case = TRUE) ~ "125M",
    grepl("350m",    x, ignore.case = TRUE) ~ "350M",
    grepl("1[._]3b", x, ignore.case = TRUE) ~ "1.3B",  # matches 1.3b or 1_3b (slug)
    TRUE ~ NA_character_
  )
}

normalise_terms <- function(term) {
  dplyr::recode(term,
    "tokens_c:genpref_c"      = "genpref_c:tokens_c",
    "tokens_c:freq_prob_c"    = "freq_prob_c:tokens_c",
    "tokens_c:log_total_c"    = "log_total_c:tokens_c",
    "log_total_c:genpref_c"   = "genpref_c:log_total_c",
    "log_total_c:freq_prob_c" = "freq_prob_c:log_total_c"
  )
}

extract_coefs <- function(fit, model_name, term_levels, term_labels) {
  fe <- as.data.frame(fixef(fit))
  fe$term <- normalise_terms(rownames(fe))
  fe %>%
    as_tibble() %>%
    rename(estimate = Estimate, se = Est.Error, lo = Q2.5, hi = Q97.5) %>%
    filter(term != "Intercept") %>%
    mutate(
      term       = factor(term, levels = term_levels, labels = term_labels),
      sig        = if_else(lo > 0 | hi < 0, "*", ""),
      model_name = model_name,
      corpus     = parse_corpus(model_name),
      model_size = factor(parse_size(model_name), levels = SIZE_LEVELS)
    )
}

coef_plot <- function(coef_df, title, subtitle,
                      ylab = "Coefficient  (95% credible interval)") {
  if (nrow(coef_df) == 0) {
    message("No data available for: ", title, " — skipping plot.")
    return(invisible(NULL))
  }
  ggplot(coef_df, aes(x = term, y = estimate, colour = model_size)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_errorbar(aes(ymin = lo, ymax = hi),
                  width = 0.2, linewidth = 0.8,
                  position = position_dodge(0.5)) +
    geom_point(size = 3, position = position_dodge(0.5)) +
    geom_text(aes(label = sig, y = hi + 0.02),
              position = position_dodge(0.5), size = 4, vjust = 0) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_discrete(name = NULL) +
    scale_y_continuous(name = ylab) +
    facet_grid(corpus ~ ., scales = "free_y") +
    labs(title = title, subtitle = subtitle) +
    theme_bw(base_size = 11) +
    theme(legend.position  = "bottom",
          panel.grid.minor = element_blank(),
          axis.text.x      = element_text(size = 9))
}

# ── Marginal predictions from a single loaded fit ────────────────────────────
# Uses fitted(..., re_formula = NA) so no group-level draws are needed,
# keeping memory and compute low.
pred_grid_for <- function(predictor, levels_vec, label_vec, fit) {
  tk_seq <- seq(-2, 2, length.out = 30)
  grid <- expand.grid(
    tokens_c    = tk_seq,
    pred_val    = levels_vec,
    genpref_c   = 0,
    freq_prob_c = 0,
    log_total_c = 0
  )
  grid[[predictor]] <- grid$pred_val
  pred <- fitted(fit, newdata = grid, re_formula = NA, summary = TRUE)
  grid %>%
    mutate(
      estimate = pred[, "Estimate"],
      lo       = pred[, "Q2.5"],
      hi       = pred[, "Q97.5"],
      pred_f   = factor(pred_val, levels = levels_vec, labels = label_vec)
    ) %>%
    select(tokens_c, pred_f, estimate, lo, hi)
}

# ── Predictor summary from a loaded fit ───────────────────────────────────────
pred_summary <- function(fit, slug) {
  d    <- fit$data
  vars <- intersect(c("genpref_c", "freq_prob_c", "log_total_c", "tokens_c"), names(d))
  var_labels <- c(
    genpref_c   = "GenPref",
    freq_prob_c = "RelFreq (P(alpha)−0.5)",
    log_total_c = "Overall freq (ln total)",
    tokens_c    = "Checkpoint (tokens, z)"
  )
  purrr::map_dfr(vars, function(v) {
    x <- d[[v]]
    tibble(
      model    = slug,
      variable = var_labels[v],
      n        = sum(!is.na(x)),
      min      = round(min(x,    na.rm = TRUE), 3),
      median   = round(median(x, na.rm = TRUE), 3),
      mean     = round(mean(x,   na.rm = TRUE), 3),
      max      = round(max(x,    na.rm = TRUE), 3)
    )
  })
}

# ── Discover saved models from the models/ directory ─────────────────────────
# Each single-model fit: <prefix><slug>.rds
#   e.g. sg_final_babylm_125m.rds
# Each trajectory fit:   <traj_prefix><slug>_ck<nn>.rds
#   e.g. sg_traj_babylm_125m_ck01.rds … sg_traj_babylm_125m_ck20.rds
list_fits <- function(prefix) {
  pat   <- paste0("^", prefix, ".+\\.rds$")
  paths <- list.files(MODELS_DIR, pattern = pat, full.names = TRUE)
  if (length(paths) == 0)
    warning(sprintf("No .rds files found for prefix '%s' in '%s'", prefix, MODELS_DIR))
  paths
}

# ── Trajectory plot ───────────────────────────────────────────────────────────
trajectory_plot <- function(traj_df, title, subtitle,
                            ylab = "Coefficient  (95% credible interval)") {
  if (nrow(traj_df) == 0) {
    message("No trajectory data — skipping: ", title)
    return(invisible(NULL))
  }
  ggplot(traj_df, aes(x = tokens / 1e6, y = estimate,
                      colour = model_size, fill = model_size,
                      group = model_name)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.8) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous(name = "Training tokens (millions)") +
    scale_y_continuous(name = ylab) +
    facet_grid(corpus ~ term, scales = "free_y") +
    labs(title = title, subtitle = subtitle) +
    theme_bw(base_size = 11) +
    theme(legend.position  = "bottom",
          panel.grid.minor = element_blank(),
          strip.text       = element_text(size = 8),
          axis.text.x      = element_text(size = 9))
}

# ── Stats plot (predictor distributions over checkpoints) ────────────────────
stats_plot <- function(stats_df, title) {
  if (nrow(stats_df) == 0) {
    message("No stats data available — skipping: ", title)
    return(invisible(NULL))
  }
  if (all(is.na(stats_df$tokens))) {
    message("All token counts are NA (TRAJ_META not built) — skipping: ", title)
    return(invisible(NULL))
  }
  has_log_total <- !all(is.na(stats_df$log_total_mean))

  p_n <- ggplot(stats_df,
                aes(x = tokens / 1e6, y = n_items,
                    colour = model_size, group = model_name)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous("Training tokens (millions)") +
    scale_y_continuous("N item pairs with data") +
    facet_wrap(~ corpus, scales = "free_x") +
    labs(title = title, subtitle = "Item coverage per checkpoint") +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())

  if (has_log_total) {
    p_exp <- ggplot(stats_df,
                    aes(x = tokens / 1e6, colour = model_size, fill = model_size,
                        group = model_name)) +
      geom_ribbon(aes(ymin = log_total_mean - log_total_sd,
                      ymax = log_total_mean + log_total_sd),
                  alpha = 0.15, colour = NA) +
      geom_line(aes(y = log_total_mean), linewidth = 0.8) +
      geom_point(aes(y = log_total_mean), size = 2) +
      scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
      scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
      scale_x_continuous("Training tokens (millions)") +
      scale_y_continuous("Mean log(total_seen)  [raw, before z-scoring]") +
      facet_wrap(~ corpus, scales = "free_x") +
      labs(subtitle = "Mean ± 1 SD of raw log(total_seen) — justifies per-checkpoint z-scoring") +
      theme_bw(base_size = 11) +
      theme(legend.position = "bottom", panel.grid.minor = element_blank())
    print(p_exp / p_n)
  } else {
    print(p_n)
  }
}

# ── GenPref significance timeline ────────────────────────────────────────────
# Supplementary strip chart: one coloured square per model × sampled checkpoint.
#   Blue  = 95% CI entirely above 0 (reliably positive)
#   Red   = 95% CI entirely below 0 (reliably negative)
#   Grey  = CI spans 0 (n.s.)
# Square size also encodes |estimate| so you can see effect-size trends.
# Use directly after trajectory_plot() — same traj_df input.
genpref_sig_timeline <- function(traj_df, title = NULL,
                                 focus_term = "GenPref") {
  if (nrow(traj_df) == 0) return(invisible(NULL))

  df <- traj_df %>%
    filter(as.character(term) == focus_term)

  if (nrow(df) == 0) {
    message("Term '", focus_term, "' not found in trajectory data — skipping sig timeline.")
    return(invisible(NULL))
  }
  if (all(is.na(df$tokens))) {
    message("All token counts NA — skipping sig timeline.")
    return(invisible(NULL))
  }

  df <- df %>%
    mutate(
      # P(beta > 0): normal approximation from posterior mean and SD
      # (posterior is nearly Gaussian for brms fixef; exact draws not needed)
      p_gt0     = pnorm(0, mean = estimate, sd = se, lower.tail = FALSE),
      pct_label = sprintf("%.0f%%", p_gt0 * 100)
    )

  ggplot(df, aes(x = tokens / 1e6, y = estimate)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    # CI bar in model-size colour (consistent with other plots)
    geom_linerange(aes(ymin = lo, ymax = hi, colour = model_size),
                   linewidth = 0.7, alpha = 0.55) +
    # Point filled by P(β > 0): blue = posterior mostly positive,
    #                            red  = posterior mostly negative,
    #                            grey = uncertain (~50%)
    geom_point(aes(fill = p_gt0), shape = 21,
               size = 3.2, colour = "grey25", stroke = 0.35) +
    # Percentage label nudged above each point
    geom_text(aes(label = pct_label, y = hi),
              size = 2.1, vjust = -0.5, colour = "grey25") +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_fill_gradient2(
      low      = "#b2182b", mid = "grey93", high = "#2166ac",
      midpoint = 0.5, limits = c(0, 1),
      name     = "P(\u03b2 > 0)",
      labels   = scales::percent_format(accuracy = 1)
    ) +
    scale_x_continuous("Training tokens (millions)") +
    scale_y_continuous(paste(focus_term, "coefficient (95% CI)")) +
    # One panel per model: corpus rows × size columns
    facet_grid(corpus ~ model_size, scales = "free_y") +
    labs(
      title    = title %||% paste(focus_term, "\u2014 coefficient & P(\u03b2 > 0) over training"),
      subtitle = paste0(
        "Point fill = % posterior with \u03b2 > 0  |  ",
        "blue > 50% (positive), red < 50% (negative), grey \u2248 50% (uncertain)  |  ",
        "bar = 95% CI"
      )
    ) +
    theme_bw(base_size = 11) +
    theme(
      legend.position  = "bottom",
      panel.grid.minor = element_blank(),
      strip.text       = element_text(size = 9)
    )
}

# ── Raw signed_gap descriptive summaries from trajectory fits ─────────────────
# Extracts fit$data to compute mean signed_gap and mean |signed_gap| per
# checkpoint. Returns a single tibble with one row per model × checkpoint.
load_traj_gap_summary <- function(traj_prefix) {
  paths <- list_fits(traj_prefix)
  if (length(paths) == 0) return(tibble())

  all_rows <- vector("list", length(paths))

  for (i in seq_along(paths)) {
    path  <- paths[i]
    fname <- sub("\\.rds$", "", basename(path))
    fname_no_prefix <- sub(paste0("^", traj_prefix), "", fname)
    slug   <- sub("_ck[0-9]+$", "", fname_no_prefix)
    ck_idx <- as.integer(sub("^.*_ck([0-9]+)$", "\\1", fname_no_prefix))

    ck_tokens <- if (!is.null(TRAJ_META)) {
      row <- filter(TRAJ_META, .data$slug == !!slug, .data$ck_idx == !!ck_idx)
      if (nrow(row) == 1) row$tokens else NA_real_
    } else NA_real_

    message("Descriptive load: ", basename(path))
    fit <- readRDS(path)
    d   <- fit$data
    rm(fit); gc()

    n    <- nrow(d)
    m    <- mean(d$signed_gap,        na.rm = TRUE)
    m_ab <- mean(abs(d$signed_gap),   na.rm = TRUE)
    s    <- sd(d$signed_gap,          na.rm = TRUE)
    s_ab <- sd(abs(d$signed_gap),     na.rm = TRUE)

    all_rows[[i]] <- tibble(
      slug = slug, ck_idx = ck_idx, tokens = ck_tokens,
      corpus     = parse_corpus(slug),
      model_size = factor(parse_size(slug), levels = SIZE_LEVELS),
      n      = n,
      mean   = m,    se   = s    / sqrt(n),
      mean_abs = m_ab, se_abs = s_ab / sqrt(n)
    ) %>%
      mutate(lo95     = mean     - 1.96 * se,
             hi95     = mean     + 1.96 * se,
             lo95_abs = mean_abs - 1.96 * se_abs,
             hi95_abs = mean_abs + 1.96 * se_abs)
  }

  bind_rows(all_rows)
}

# ── Load all checkpoint fits for one trajectory prefix ───────────────────────
# brms only stores formula variables in fit$data, so 'tokens' is not present
# in the per-checkpoint .rds files.  Token counts are recovered from TRAJ_META
# (built once from DuckDB in the setup chunk below).
# Returns list(coefs = tibble, stats = tibble).
load_traj_results <- function(traj_prefix, term_levels, term_labels) {
  paths <- list_fits(traj_prefix)
  if (length(paths) == 0) return(list(coefs = tibble(), stats = tibble()))

  all_coefs <- vector("list", length(paths))
  all_stats <- vector("list", length(paths))

  for (i in seq_along(paths)) {
    path  <- paths[i]
    fname <- sub("\\.rds$", "", basename(path))
    # Strip prefix → e.g. "babylm_125m_ck01"; then strip "_ck##" → slug
    fname_no_prefix <- sub(paste0("^", traj_prefix), "", fname)
    slug            <- sub("_ck[0-9]+$", "", fname_no_prefix)
    ck_idx          <- as.integer(sub("^.*_ck([0-9]+)$", "\\1", fname_no_prefix))
    message("Loading: ", basename(path))

    # Look up the actual token count from the pre-built metadata table
    ck_tokens <- if (!is.null(TRAJ_META)) {
      row <- filter(TRAJ_META, .data$slug == !!slug, .data$ck_idx == !!ck_idx)
      if (nrow(row) == 1) row$tokens else NA_real_
    } else {
      NA_real_
    }

    fit <- readRDS(path)

    all_coefs[[i]] <- extract_coefs(fit, slug, term_levels, term_labels) %>%
      mutate(tokens = ck_tokens)

    all_stats[[i]] <- tibble(
      model_name = slug,
      tokens     = ck_tokens,
      corpus     = parse_corpus(slug),
      model_size = factor(parse_size(slug), levels = SIZE_LEVELS),
      n_items    = nrow(fit$data),
      # log_total (raw, pre-z-score) is not stored in fit$data; stats_plot()
      # will skip the exposure panel when these are NA.
      log_total_mean = NA_real_, log_total_sd  = NA_real_,
      log_total_min  = NA_real_, log_total_max = NA_real_
    )

    rm(fit); gc()
  }

  list(coefs = bind_rows(all_coefs), stats = bind_rows(all_stats))
}
```

---

## Discover fitted models

```{r path-diagnostic}
cat("Working directory:", getwd(), "\n")
cat("MODELS_DIR resolves to:", normalizePath(MODELS_DIR, mustWork = FALSE), "\n")
cat("Directory exists:", dir.exists(MODELS_DIR), "\n\n")

all_rds <- list.files(MODELS_DIR, pattern = "\\.rds$", full.names = FALSE)
cat(sprintf("%d .rds file(s) found in models dir:\n", length(all_rds)))
cat(paste(" ", all_rds, collapse = "\n"), "\n")
```

```{r discover-models}
# Single-model fits (one .rds per model):
PREFIXES <- c(
  sg_final   = "sg_final_",
  sg_ckpt    = "sg_ckpt_",
  pref_final = "pref_final_",
  pref_ckpt  = "pref_ckpt_"
)

# Trajectory fits (20 .rds per model, named <prefix><slug>_ck<nn>.rds):
TRAJ_PREFIXES <- c(
  sg_traj   = "sg_traj_",
  pref_traj = "pref_traj_"
)

available      <- map(PREFIXES,      list_fits)
available_traj <- map(TRAJ_PREFIXES, list_fits)

cat("=== Single-model fits ===\n")
walk2(names(available), available, ~ {
  cat(sprintf("%-12s : %d file(s)\n", .x, length(.y)))
  walk(.y, ~ cat("  ", basename(.x), "\n"))
})

cat("\n=== Trajectory fits ===\n")
walk2(names(available_traj), available_traj, ~ {
  cat(sprintf("%-12s : %d file(s)\n", .x, length(.y)))
})
```

---

# Descriptive: Signed Gap Over Training

Raw outcome variable (signed gap) aggregated across items at each sampled
checkpoint. No regression — pure descriptive summary. Uses the same
`sg_traj_` files as Analysis 2.

```{r desc-gap-load, message=TRUE}
sg_gap_summary <- load_traj_gap_summary("sg_traj_")
```

## Plot D1: Mean Signed Gap

Average signed gap across all items at each checkpoint. Shows whether
models develop a directional bias (toward alpha or beta) over training and
whether that bias is stable, grows, or reverses.

```{r desc-gap-mean, fig.width=12, fig.height=5}
if (nrow(sg_gap_summary) > 0 && !all(is.na(sg_gap_summary$tokens))) {
  ggplot(sg_gap_summary,
         aes(x = tokens / 1e6, y = mean,
             colour = model_size, fill = model_size, group = slug)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_ribbon(aes(ymin = lo95, ymax = hi95), alpha = 0.15, colour = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.8) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous("Training tokens (millions)") +
    scale_y_continuous("Mean signed gap  (95% CI of mean)") +
    facet_grid(corpus ~ ., scales = "free_x") +
    labs(
      title    = "Mean signed gap over training — all items pooled",
      subtitle = "Positive = model rates alpha higher on average; negative = rates beta higher  |  ribbon = 95% CI of mean"
    ) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
} else {
  message("No data for mean signed gap plot.")
}
```

## Plot D2: Mean Absolute Signed Gap

Average of |signed gap| at each checkpoint. Shows whether models become
more decisive (larger gaps) or more uncertain (gaps shrinking toward zero)
as training progresses, regardless of direction.

```{r desc-gap-abs, fig.width=12, fig.height=5}
if (nrow(sg_gap_summary) > 0 && !all(is.na(sg_gap_summary$tokens))) {
  ggplot(sg_gap_summary,
         aes(x = tokens / 1e6, y = mean_abs,
             colour = model_size, fill = model_size, group = slug)) +
    geom_ribbon(aes(ymin = lo95_abs, ymax = hi95_abs), alpha = 0.15, colour = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.8) +
    scale_colour_manual(values = SIZE_COLOURS, name = "Model size") +
    scale_fill_manual(values   = SIZE_COLOURS, name = "Model size") +
    scale_x_continuous("Training tokens (millions)") +
    scale_y_continuous("Mean |signed gap|  (95% CI of mean)") +
    facet_grid(corpus ~ ., scales = "free_x") +
    labs(
      title    = "Mean absolute signed gap over training — all items pooled",
      subtitle = "Higher = model makes larger rating differences between alpha and beta  |  ribbon = 95% CI of mean"
    ) +
    theme_bw(base_size = 11) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank())
} else {
  message("No data for absolute signed gap plot.")
}
```

---

# Part I — Signed Gap

---

## Analysis 1: Signed Gap — Final Checkpoint

```{r sg-final-coefs, fig.width=13, fig.height=7}
sg_results_final <- map(available$sg_final, function(path) {
  slug <- sub("\\.rds$", "", sub("^sg_final_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS, TERM_LABELS),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
sg_coefs_final <- map_dfr(sg_results_final, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(sg_results_final, "summ"), n = Inf)

coef_plot(
  sg_coefs_final,
  title    = "Predictors of signed gap — final checkpoint",
  subtitle = "signed_gap ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  * = CI excludes zero"
)
```

---

## Analysis 2: Signed Gap — Per-Checkpoint Trajectory

20 evenly-spaced checkpoints per model; `log_total_c` z-scored within each
checkpoint's items.  Files: `sg_traj_{slug}_ck{01-20}.rds`.

### Plot 2a: Predictor Statistics Over Training

```{r sg-traj-stats, fig.width=13, fig.height=8}
sg_traj <- load_traj_results("sg_traj_", TERM_LEVELS, TERM_LABELS)
stats_plot(sg_traj$stats, "Item predictor statistics over training — signed gap models")
```

### Plot 2b: Coefficients Over Training

```{r sg-traj-coefs, fig.width=16, fig.height=7}
trajectory_plot(
  sg_traj$coefs,
  title    = "Signed gap predictors over training — 20 sampled checkpoints",
  subtitle = "brms at each checkpoint  |  log_total_c z-scored within checkpoint  |  ribbon = 95% CI  |  * = CI excludes zero"
)
```

### Plot 2c: GenPref Significance Over Training

```{r sg-traj-sig, fig.width=12, fig.height=3.5}
genpref_sig_timeline(
  sg_traj$coefs,
  title = "GenPref significance over training — signed gap models"
)
```

---

## Analysis 3: Signed Gap — Checkpoint-Interacted

### Plot 3a: Coefficients

```{r sg-ckpt-coefs, fig.width=14, fig.height=7}
sg_results_ckpt <- map(available$sg_ckpt, function(path) {
  slug <- sub("\\.rds$", "", sub("^sg_ckpt_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS_CK, TERM_LABELS_CK),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
sg_coefs_ckpt <- map_dfr(sg_results_ckpt, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(sg_results_ckpt, "summ"), n = Inf)

coef_plot(
  sg_coefs_ckpt,
  title    = "Predictors of signed gap — checkpoint as continuous predictor",
  subtitle = "signed_gap ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1|binom)  |  * = CI excludes zero"
)
```

### Plot 3b: Marginal Effects Over Training

```{r sg-ckpt-marginal, fig.width=14, fig.height=10}
# Load each ckpt fit once and compute all three marginal grids, then plot.
sg_marginal_all <- tibble()

for (path in available$sg_ckpt) {
  slug   <- sub("\\.rds$", "", sub("^sg_ckpt_", "", basename(path)))
  message("Marginals: ", basename(path))
  fit    <- readRDS(path)

  for (spec in PRED_SPECS) {
    sg_marginal_all <- bind_rows(
      sg_marginal_all,
      pred_grid_for(spec$var, spec$vals, spec$labels, fit) %>%
        mutate(
          predictor   = spec$var,
          model_label = slug,
          corpus      = parse_corpus(slug),
          model_size  = factor(parse_size(slug), levels = SIZE_LEVELS)
        )
    )
  }
  rm(fit); gc()
}

if (nrow(sg_marginal_all) == 0) {
  message("No sg_ckpt marginal data available — skipping plots.")
} else {
  for (spec in PRED_SPECS) {
    df <- filter(sg_marginal_all, predictor == spec$var)
    if (nrow(df) == 0) { message("No marginal data for: ", spec$title); next }
    p  <- ggplot(df, aes(x = tokens_c, y = estimate,
                         colour = pred_f, fill = pred_f)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
      geom_line(linewidth = 0.9) +
      scale_colour_manual(values = spec$cols, name = NULL) +
      scale_fill_manual(values   = spec$cols, name = NULL) +
      scale_x_continuous(name = "Checkpoint  (tokens, z-scored within model)") +
      scale_y_continuous(name = "Predicted signed gap") +
      facet_wrap(~ model_label, scales = "free_y", ncol = 3) +
      labs(title    = spec$title,
           subtitle = "Other predictors held at 0  |  95% posterior credible ribbon  |  population-level") +
      theme_bw(base_size = 11) +
      theme(legend.position = "bottom", panel.grid.minor = element_blank())
    print(p)
  }
}
```

---

# Part II — Preference

---

## Analysis 4: Preference — Final Checkpoint

```{r pref-final-coefs, fig.width=13, fig.height=7}
pref_results_final <- map(available$pref_final, function(path) {
  slug <- sub("\\.rds$", "", sub("^pref_final_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS, TERM_LABELS),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
pref_coefs_final <- map_dfr(pref_results_final, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(pref_results_final, "summ"), n = Inf)

coef_plot(
  pref_coefs_final,
  title    = "Predictors of preference — final checkpoint",
  subtitle = "preference ~ genpref_c * log_total_c + freq_prob_c * log_total_c + (1|binom)  |  * = CI excludes zero"
)
```

---

## Analysis 5: Preference — Per-Checkpoint Trajectory

Same approach as Analysis 2, applied to raw preference.
Files: `pref_traj_{slug}_ck{01-20}.rds`.

### Plot 5a: Predictor Statistics Over Training

```{r pref-traj-stats, fig.width=13, fig.height=8}
pref_traj <- load_traj_results("pref_traj_", TERM_LEVELS, TERM_LABELS)
stats_plot(pref_traj$stats, "Item predictor statistics over training — preference models")
```

### Plot 5b: Coefficients Over Training

```{r pref-traj-coefs, fig.width=16, fig.height=7}
trajectory_plot(
  pref_traj$coefs,
  title    = "Preference predictors over training — 20 sampled checkpoints",
  subtitle = "brms at each checkpoint  |  log_total_c z-scored within checkpoint  |  ribbon = 95% CI  |  * = CI excludes zero"
)
```

### Plot 5c: GenPref Significance Over Training

```{r pref-traj-sig, fig.width=12, fig.height=3.5}
genpref_sig_timeline(
  pref_traj$coefs,
  title = "GenPref significance over training — preference models"
)
```

---

## Analysis 6: Preference — Checkpoint-Interacted

### Plot 6a: Coefficients

```{r pref-ckpt-coefs, fig.width=14, fig.height=7}
pref_results_ckpt <- map(available$pref_ckpt, function(path) {
  slug <- sub("\\.rds$", "", sub("^pref_ckpt_", "", basename(path)))
  message("Loading: ", basename(path))
  fit  <- readRDS(path)
  res  <- list(coefs = extract_coefs(fit, slug, TERM_LEVELS_CK, TERM_LABELS_CK),
               summ  = pred_summary(fit, slug))
  rm(fit); gc()
  res
})
pref_coefs_ckpt <- map_dfr(pref_results_ckpt, "coefs")

cat("=== Predictor ranges ===\n")
print(map_dfr(pref_results_ckpt, "summ"), n = Inf)

coef_plot(
  pref_coefs_ckpt,
  title    = "Predictors of preference — checkpoint as continuous predictor",
  subtitle = "preference ~ (genpref_c + freq_prob_c + log_total_c) * tokens_c + (1|binom)  |  * = CI excludes zero"
)
```

### Plot 6b: Marginal Effects Over Training

```{r pref-ckpt-marginal, fig.width=14, fig.height=10}
pref_marginal_all <- tibble()

for (path in available$pref_ckpt) {
  slug   <- sub("\\.rds$", "", sub("^pref_ckpt_", "", basename(path)))
  message("Marginals: ", basename(path))
  fit    <- readRDS(path)

  for (spec in PRED_SPECS) {
    pref_marginal_all <- bind_rows(
      pref_marginal_all,
      pred_grid_for(spec$var, spec$vals, spec$labels, fit) %>%
        mutate(
          predictor   = spec$var,
          model_label = slug,
          corpus      = parse_corpus(slug),
          model_size  = factor(parse_size(slug), levels = SIZE_LEVELS)
        )
    )
  }
  rm(fit); gc()
}

if (nrow(pref_marginal_all) == 0) {
  message("No pref_ckpt marginal data available — skipping plots.")
} else {
  for (spec in PRED_SPECS) {
    df <- filter(pref_marginal_all, predictor == spec$var)
    if (nrow(df) == 0) { message("No marginal data for: ", spec$title); next }
    p  <- ggplot(df, aes(x = tokens_c, y = estimate,
                         colour = pred_f, fill = pred_f)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
      geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
      geom_line(linewidth = 0.9) +
      scale_colour_manual(values = spec$cols, name = NULL) +
      scale_fill_manual(values   = spec$cols, name = NULL) +
      scale_x_continuous(name = "Checkpoint  (tokens, z-scored within model)") +
      scale_y_continuous(name = "Predicted preference") +
      facet_wrap(~ model_label, scales = "free_y", ncol = 3) +
      labs(title    = spec$title,
           subtitle = "Other predictors held at 0  |  95% posterior credible ribbon  |  population-level") +
      theme_bw(base_size = 11) +
      theme(legend.position = "bottom", panel.grid.minor = element_blank())
    print(p)
  }
}
```
