---
title: "Pythia & OLMo: Abstract Preference vs. Corpus Frequency"
subtitle: "Replication of abspref_casestudy analyses for Pythia (The Pile) and OLMo-3 (Dolma)"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    theme: flatly
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 12, fig.height = 6, dpi = 150,
                      out.width = "100%")
```

# Overview

This document replicates the core analyses from `abspref_casestudy.Rmd` for two
additional model families:

- **Pythia** (EleutherAI): 160M, 410M, 1.4B, 2.8B — trained on The Pile.
  Checkpoint data is in `training_attested.csv` (18 pre-selected steps per model).
- **OLMo-3** (AllenAI): 7B, 32B — trained on Dolma.
  Checkpoint data is in per-step CSVs under `Data/checkpoint_results/`.
  31 checkpoints available; 20 are sampled evenly on a log scale.

Analyses:
1. **ΔLL vs. perplexity** across training checkpoints (Oh & Schuler 2023 method)
2. **Model vs. human α-proportion** scatter (final checkpoint & over training)
3. **Disagreement items** — where RelFreq predicts the wrong direction
4. **Training trajectories** — case-study binomials over training
5. **Accuracy by frequency quintile** — mirrors Analysis 2c from `analysis2.Rmd`
6. **Correlation trajectories** — model pref ~ RelFreq, human pref, and each individual
   constraint over training; preference deviation ~ each constraint (mirrors Analysis
   2a/2b raw from `analysis2.Rmd`)
7. **AbsPref coefficient** — mixed-model trajectory of abstract preference effect

------------------------------------------------------------------------

```{r libraries}
library(tidyverse)
library(brms)
library(patchwork)
library(ggrepel)
library(scales)
library(kableExtra)
library(future)
library(furrr)
```

```{r config}
SCRIPT_DIR <- {
  knitr_input <- tryCatch(knitr::current_input(), error = function(e) NULL)
  if (!is.null(knitr_input) && nchar(knitr_input) > 0) {
    dirname(normalizePath(knitr_input, mustWork = FALSE))
  } else if (requireNamespace("rstudioapi", quietly = TRUE) && rstudioapi::isAvailable()) {
    dirname(normalizePath(rstudioapi::getActiveDocumentContext()$path, mustWork = FALSE))
  } else {
    normalizePath(".", mustWork = FALSE)
  }
}

DATA_DIR    <- normalizePath(file.path(SCRIPT_DIR, "../Data/processed"), mustWork = FALSE)
BASE_DIR    <- normalizePath(file.path(SCRIPT_DIR, ".."),               mustWork = FALSE)
RAW_DIR     <- file.path(BASE_DIR, "Data")

TRAIN_CSV        <- file.path(DATA_DIR, "training_attested.csv")
HUMAN_CSV        <- file.path(RAW_DIR,  "all_human_data.csv")
BINOMS_CSV       <- file.path(RAW_DIR,  "nonce_and_attested_binoms.csv")
CK_RESULTS_DIR   <- file.path(RAW_DIR,  "checkpoint_results")
PILE_FREQ_CSV    <- file.path(SCRIPT_DIR, "oh-schuler-replication", "pile_corpus_freq.csv")

# Sampling: number of checkpoints to keep per model for trajectory analyses
N_TRAJ_CHECKPOINTS <- 20

# ── Pythia model metadata ─────────────────────────────────────────────────────
PYTHIA_SIZES  <- c("160M", "410M", "1.4B", "2.8B")
PYTHIA_COLOURS <- c("160M" = "#c6dbef", "410M" = "#6baed6",
                    "1.4B" = "#2171b5", "2.8B" = "#08306b")

# ── OLMo model metadata ───────────────────────────────────────────────────────
OLMO_SIZES   <- c("7B", "32B")
OLMO_COLOURS <- c("7B" = "#fd8d3c", "32B" = "#bd0026")

MODELS_DIR <- file.path(SCRIPT_DIR, "models")
dir.create(MODELS_DIR, showWarnings = FALSE, recursive = TRUE)

# ── Parallelisation (mirrors analysis2.Rmd) ───────────────────────────────────
N_CORES           <- parallel::detectCores(logical = FALSE)
USABLE_CORES      <- 28L    # cap at 28 of 32 cores (7 models × 4 chains each)
N_PARALLEL_MODELS <- 7
CHAINS            <- 4L
CORES             <- CHAINS
ITER              <- 6000
WARMUP            <- 3000

PRIORS_GAUSSIAN <- c(
  prior(normal(0, 1), class = b),
  prior(normal(0, 1), class = Intercept),
  prior(exponential(1), class = sd),
  prior(exponential(1), class = sigma)
)

message(sprintf(
  "Cores: %d total | %d usable | %d parallel models | %d chains/cores per model",
  N_CORES, USABLE_CORES, N_PARALLEL_MODELS, CHAINS
))

stopifnot(file.exists(TRAIN_CSV), file.exists(HUMAN_CSV), dir.exists(CK_RESULTS_DIR))
```

```{r helpers}
make_slug <- function(x) tolower(gsub("[^a-z0-9]", "_", x))

# ── Model name parsers ────────────────────────────────────────────────────────
parse_pythia_size <- function(x) case_when(
  grepl("160m",  x, TRUE) ~ "160M",
  grepl("410m",  x, TRUE) ~ "410M",
  grepl("1[._]4b", x, TRUE) ~ "1.4B",
  grepl("2[._]8b", x, TRUE) ~ "2.8B",
  TRUE ~ NA_character_
)

parse_olmo_size <- function(x) case_when(
  grepl("7b",  x, TRUE) ~ "7B",
  grepl("32b", x, TRUE) ~ "32B",
  TRUE ~ NA_character_
)

# ── Log-scale step sampler ────────────────────────────────────────────────────
# Samples n steps from a sorted vector, evenly spaced on a log scale so that
# early training (rapid change) is better represented than late training.
sample_steps_log <- function(steps, n = N_TRAJ_CHECKPOINTS) {
  steps <- sort(unique(steps))
  if (length(steps) <= n) return(steps)
  # Add 1 before log to handle step 0
  log_steps <- log(steps + 1)
  breaks    <- seq(min(log_steps), max(log_steps), length.out = n)
  selected  <- vapply(breaks, function(b) steps[which.min(abs(log_steps - b))], numeric(1))
  unique(selected)
}

# ── freq_prob_c from Pile frequencies (if available) ─────────────────────────
use_pile_freq <- file.exists(PILE_FREQ_CSV)
if (use_pile_freq) {
  pile_freq <- read_csv(PILE_FREQ_CSV, show_col_types = FALSE) |>
    select(binom, pile_freq_prob_c = freq_prob_c, pile_log_total = log_total)
  message(sprintf("Pile corpus frequencies loaded: %d binomials", nrow(pile_freq)))
} else {
  message("Pile corpus frequencies not found — using Google Books RelFreq as fallback.")
}
```

```{r setup-parallel}
plan(multicore, workers = N_PARALLEL_MODELS)
message(sprintf("furrr plan: %s with %d workers", class(plan())[1], N_PARALLEL_MODELS))
```

------------------------------------------------------------------------

# Data Loading

```{r load-pythia}
message("Loading Pythia data from training_attested.csv ...")
train_all <- read_csv(TRAIN_CSV, show_col_types = FALSE)

pythia_raw <- train_all |>
  filter(grepl("pythia", model, TRUE)) |>
  mutate(
    model_family = "Pythia",
    model_size   = factor(parse_pythia_size(model), levels = PYTHIA_SIZES),
    corpus       = "The Pile"   # all Pythia models trained on The Pile
  ) |>
  filter(!is.na(model_size))

message(sprintf("  Pythia: %d rows, %d models, %d unique steps",
                nrow(pythia_raw),
                n_distinct(pythia_raw$model),
                n_distinct(pythia_raw$step)))
```

```{r load-olmo}
message("Loading OLMo-3 checkpoint CSVs ...")
olmo_files <- list.files(CK_RESULTS_DIR, pattern = "^Olmo.*\\.csv$",
                         full.names = TRUE)

# Sample steps log-uniformly per model
olmo_all_steps <- tibble(path = olmo_files) |>
  mutate(
    fname      = basename(path),
    model_name = sub("_stage.*", "", fname),          # e.g. "Olmo-3-1025-7B"
    step       = as.integer(sub(".*step(\\d+)\\.csv$", "\\1", fname))
  )

selected_olmo <- olmo_all_steps |>
  group_by(model_name) |>
  group_modify(function(d, k) {
    kept <- sample_steps_log(d$step, N_TRAJ_CHECKPOINTS)
    filter(d, step %in% kept)
  }) |>
  ungroup()

message(sprintf("  OLMo: %d files selected from %d total",
                nrow(selected_olmo), nrow(olmo_all_steps)))

olmo_raw <- selected_olmo |>
  group_by(model_name) |>
  group_modify(function(d, k) {
    map_dfr(d$path, read_csv, show_col_types = FALSE)
  }) |>
  ungroup() |>
  mutate(
    model_family = "OLMo-3",
    model_size   = factor(parse_olmo_size(model), levels = OLMO_SIZES),
    corpus       = "Dolma"
  ) |>
  filter(!is.na(model_size))

message(sprintf("  OLMo loaded: %d rows", nrow(olmo_raw)))
```

```{r load-human}
human_trials <- read_csv(HUMAN_CSV, show_col_types = FALSE) |>
  mutate(
    binom      = Alpha,
    resp_alpha = as.integer(resp == "alpha")
  ) |>
  select(binom, resp_alpha, RelFreq, GenPref, OverallFreq)

binoms_meta <- read_csv(BINOMS_CSV, show_col_types = FALSE) |>
  select(binom = Alpha, Attested)

attested_binoms <- binoms_meta |> filter(Attested == 1) |> pull(binom)
human_attested  <- human_trials |> filter(binom %in% attested_binoms)

message(sprintf("Human data: %d trials, %d attested binomials",
                nrow(human_attested), n_distinct(human_attested$binom)))
```

```{r attach-freq}
# Attach corpus frequency to model preference data.
#
# Dynamic per-checkpoint exposure frequencies (alpha_seen/beta_seen) are NOT
# available for Pythia or OLMo in the current data — only our custom OPT models
# have them via checkpoint_results_with_exposures.csv. Therefore we use:
#   - Pile freq_prob_c (from compute_pile_freq.py) if available — best proxy for
#     both Pythia (The Pile) and OLMo (Dolma, similar web-text distribution)
#   - Google Books RelFreq as fallback if Pile CSV not yet computed
# In both cases the frequency is static (not checkpoint-specific).

static_binom_freq <- human_trials |>
  select(binom, RelFreq) |>
  distinct()

if (use_pile_freq) {
  static_binom_freq <- static_binom_freq |>
    left_join(pile_freq, by = "binom")
}

attach_corpus_freq <- function(df) {
  # Only join columns from static_binom_freq that aren't already in df,
  # to avoid RelFreq.x / RelFreq.y collisions when training_attested.csv
  # already carries a RelFreq column (e.g. for Pythia data).
  new_cols <- setdiff(names(static_binom_freq), c("binom", names(df)))
  if (length(new_cols) > 0) {
    df <- df |> left_join(static_binom_freq |> select(binom, all_of(new_cols)), by = "binom")
  }
  if (use_pile_freq) {
    # Use Pile freq_prob_c; fall back to centred Google Books RelFreq if NA
    df <- df |> mutate(freq_use = coalesce(pile_freq_prob_c, RelFreq - 0.5))
  } else {
    df <- df |> mutate(freq_use = RelFreq - 0.5)
  }
  df
}

# Final-checkpoint data (per model: latest step available)
pythia_final <- pythia_raw |>
  group_by(model) |>
  filter(step == max(step)) |>
  ungroup() |>
  attach_corpus_freq()

olmo_final <- olmo_raw |>
  group_by(model) |>
  filter(step == max(step)) |>
  ungroup() |>
  attach_corpus_freq()
```

------------------------------------------------------------------------

# §1 — ΔLL vs. Perplexity

ΔLL = logLik(resp_alpha ~ freq + model_pref) − logLik(resp_alpha ~ freq), computed
per checkpoint from individual human binary choices. Perplexity = exp(training loss)
from `tokens` column (approx) or matched from available loss curves.

For Pythia/OLMo the training loss is not separately stored, so we approximate
perplexity from the model's checkpoint step and token count where available.

```{r delta-ll-pythia}
# ── Pythia ΔLL trajectory ─────────────────────────────────────────────────────
# Sample checkpoints
pythia_ck <- pythia_raw |>
  group_by(model, model_size) |>
  group_modify(function(d, k) {
    kept <- sample_steps_log(d$step, N_TRAJ_CHECKPOINTS)
    filter(d, step %in% kept)
  }) |>
  ungroup() |>
  attach_corpus_freq() |>
  filter(binom %in% attested_binoms)

# Join human trial outcomes
pythia_trials_ck <- human_attested |>
  select(binom, resp_alpha) |>
  inner_join(pythia_ck |> select(model, model_size, step, binom, preference, freq_use, tokens),
             by = "binom")

# ΔLL per checkpoint
pythia_delta_ll <- pythia_trials_ck |>
  group_by(model, model_size, step, tokens) |>
  group_modify(function(d, k) {
    d <- filter(d, !is.na(freq_use), !is.na(preference))
    if (nrow(d) < 5) return(tibble(delta_ll = NA_real_, n_trials = nrow(d), n_items = 0L))
    baseline <- glm(resp_alpha ~ freq_use,              family = binomial, data = d)
    full     <- glm(resp_alpha ~ freq_use + preference, family = binomial, data = d)
    tibble(delta_ll = as.numeric(logLik(full)) - as.numeric(logLik(baseline)),
           n_trials = nrow(d), n_items = n_distinct(d$binom))
  }) |>
  ungroup() |>
  mutate(
    # Approximate perplexity from token count: each step ≈ 2M tokens for Pythia
    # More precisely: Pythia reports training loss but we use tokens as a proxy
    # for training progress on the x-axis
    corpus = "The Pile"
  )

message(sprintf("Pythia ΔLL computed: %d checkpoint rows", nrow(pythia_delta_ll)))
```

```{r delta-ll-olmo}
# ── OLMo ΔLL trajectory ───────────────────────────────────────────────────────
olmo_ck <- olmo_raw |>
  filter(binom %in% attested_binoms) |>
  attach_corpus_freq()

olmo_trials_ck <- human_attested |>
  select(binom, resp_alpha) |>
  inner_join(olmo_ck |> select(model, model_size, step, binom, preference, freq_use, tokens),
             by = "binom")

olmo_delta_ll <- olmo_trials_ck |>
  group_by(model, model_size, step, tokens) |>
  group_modify(function(d, k) {
    d <- filter(d, !is.na(freq_use), !is.na(preference))
    if (nrow(d) < 5) return(tibble(delta_ll = NA_real_, n_trials = nrow(d), n_items = 0L))
    baseline <- glm(resp_alpha ~ freq_use,              family = binomial, data = d)
    full     <- glm(resp_alpha ~ freq_use + preference, family = binomial, data = d)
    tibble(delta_ll = as.numeric(logLik(full)) - as.numeric(logLik(baseline)),
           n_trials = nrow(d), n_items = n_distinct(d$binom))
  }) |>
  ungroup() |>
  mutate(corpus = "Dolma")

message(sprintf("OLMo ΔLL computed: %d checkpoint rows", nrow(olmo_delta_ll)))
```

```{r delta-ll-slope-tests}
# Slope tests: does ΔLL increase as training progresses (more tokens)?
# Matches abspref_casestudy slope test logic but uses log10(tokens) as x-axis
# instead of log2(perplexity) since per-checkpoint loss is unavailable.
delta_ll_all <- bind_rows(
  pythia_delta_ll |> mutate(model_family = "Pythia",
                              model_size   = factor(model_size, levels = PYTHIA_SIZES)),
  olmo_delta_ll   |> mutate(model_family = "OLMo-3",
                              model_size   = factor(model_size, levels = OLMO_SIZES))
) |> filter(!is.na(delta_ll), tokens > 0)

slope_tests <- delta_ll_all |>
  group_by(model_family, model_size) |>
  group_modify(function(d, k) {
    if (nrow(d) < 3) return(tibble(slope = NA_real_, se = NA_real_, t = NA_real_, p = NA_real_, n = nrow(d)))
    fit <- lm(delta_ll ~ log10(tokens), data = d)
    cf  <- summary(fit)$coefficients
    tibble(slope = cf["log10(tokens)", "Estimate"],
           se    = cf["log10(tokens)", "Std. Error"],
           t     = cf["log10(tokens)", "t value"],
           p     = cf["log10(tokens)", "Pr(>|t|)"],
           n     = nrow(d))
  }) |>
  ungroup() |>
  mutate(
    sig   = case_when(p < .001 ~ "***", p < .01 ~ "**", p < .05 ~ "*", p < .10 ~ ".", TRUE ~ "ns"),
    label = sprintf("β = %.3f, t(%d) = %.2f, p %s %s",
                    slope, n - 2L, t,
                    ifelse(p < .001, "<", "="),
                    ifelse(p < .001, ".001", sprintf("%.3f", p)), sig)
  )

cat("\nSlope tests: log10(tokens) → ΔLL\n")
print(slope_tests |> select(model_family, model_size, slope, se, t, p, sig, n), n = Inf)
```

```{r delta-ll-plot}
# Use tokens as x-axis (training progress) since per-checkpoint loss not available.
# Tokens are a monotone proxy for training progress and allow cross-model comparison.

delta_ll_all |>
  ggplot(aes(x = tokens, y = delta_ll, colour = model_size)) +
  facet_grid(model_family ~ corpus, scales = "free") +
  geom_point(size = 1.8, alpha = 0.5) +
  geom_smooth(method = "loess", formula = y ~ x, se = TRUE,
              linetype = "dashed", linewidth = 0.9, alpha = 0.12) +
  scale_x_continuous("Training tokens (log scale)", trans = "log10",
                     labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous("ΔLL") +
  scale_colour_manual(
    values = c(PYTHIA_COLOURS, OLMO_COLOURS),
    name   = "Model size"
  ) +
  labs(
    title    = "ΔLL vs. training tokens — Pythia & OLMo-3",
    subtitle = paste0(
      "ΔLL = logLik(resp_alpha ~ ",
      ifelse(use_pile_freq, "Pile freq", "Google Books RelFreq"),
      " + model pref) − logLik(baseline). Dashed = loess smoother with 95% CI."
    )
  ) +
  theme_classic(base_size = 13) +
  theme(
    panel.grid.major = element_line(colour = "grey92", linewidth = 0.4),
    strip.background = element_rect(fill = "grey95"),
    legend.position  = "right"
  )
```

------------------------------------------------------------------------

# §2 — Model vs. Human α-Ordering Proportions

For each binomial, the **model proportion** is the fraction of the 51 sentence-frame
prompts where the model prefers the α ordering (preference > 0). The **human proportion**
is the fraction of human participants who chose the α ordering.

```{r alpha-prop-compute}
# Human proportion per binomial
human_prop <- human_attested |>
  group_by(binom) |>
  summarise(human_prop_alpha = mean(resp_alpha), .groups = "drop")

# Model proportion at final checkpoint per (model, binom)
compute_model_prop <- function(df) {
  df |>
    filter(binom %in% attested_binoms) |>
    group_by(model, model_size, corpus, binom) |>
    summarise(model_prop_alpha = mean(preference > 0), .groups = "drop") |>
    left_join(human_prop, by = "binom")
}

pythia_prop <- compute_model_prop(
  pythia_final |> mutate(corpus = "The Pile")
)
olmo_prop   <- compute_model_prop(
  olmo_final |> mutate(corpus = "Dolma")
)

all_prop <- bind_rows(
  pythia_prop |> mutate(model_family = "Pythia",
                         model_size   = factor(model_size, levels = PYTHIA_SIZES)),
  olmo_prop   |> mutate(model_family = "OLMo-3",
                         model_size   = factor(model_size, levels = OLMO_SIZES))
)
```

## Plot 1 — Final checkpoint scatter

```{r alpha-prop-final-plot}
r_labels <- all_prop |>
  group_by(model_family, model_size) |>
  summarise(
    r   = tryCatch(cor(human_prop_alpha, model_prop_alpha, use = "complete.obs"),
                   error = function(e) NA_real_),
    .groups = "drop"
  ) |>
  mutate(lbl = sprintf("r = %.2f", r))

ggplot(all_prop, aes(x = human_prop_alpha, y = model_prop_alpha)) +
  facet_grid(model_family ~ model_size) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted", colour = "grey60") +
  geom_point(size = 1.8, alpha = 0.5,
             aes(colour = model_family)) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE,
              colour = "black", linewidth = 0.7, alpha = 0.15) +
  geom_text(data = r_labels,
            aes(label = lbl, x = 0.15, y = 0.9),
            hjust = 0, size = 3.5, fontface = "bold") +
  scale_x_continuous("Proportion of humans choosing α", limits = c(0, 1)) +
  scale_y_continuous("Proportion of prompts where model prefers α", limits = c(0, 1)) +
  scale_colour_manual(values = c("Pythia" = "#2171b5", "OLMo-3" = "#bd0026"),
                      guide = "none") +
  labs(
    title    = "Model vs. human α-proportion — final checkpoint",
    subtitle = "Each point = one attested binomial. Diagonal = perfect agreement."
  ) +
  theme_classic(base_size = 12) +
  theme(strip.background = element_rect(fill = "grey95"),
        panel.grid.major = element_line(colour = "grey92", linewidth = 0.3))
```

## Plot 2 — Pearson r over training (tokens)

```{r alpha-prop-traj}
# Compute per-checkpoint model α-proportion, then correlate with human
compute_r_traj <- function(ck_df, corpus_label, model_size_levels) {
  ck_df |>
    filter(binom %in% attested_binoms) |>
    group_by(model, model_size, step, tokens, binom) |>
    summarise(model_prop_alpha = mean(preference > 0), .groups = "drop") |>
    left_join(human_prop, by = "binom") |>
    group_by(model_size, step, tokens) |>
    summarise(
      r = tryCatch(
        cor(human_prop_alpha, model_prop_alpha, use = "complete.obs"),
        error = function(e) NA_real_
      ),
      .groups = "drop"
    ) |>
    mutate(
      model_size = factor(model_size, levels = model_size_levels),
      corpus     = corpus_label
    )
}

r_traj_pythia <- compute_r_traj(
  pythia_ck |> filter(binom %in% attested_binoms) |>
    group_by(model, model_size, step, tokens, binom) |>
    summarise(preference = mean(preference), .groups = "drop"),
  "The Pile", PYTHIA_SIZES
)

r_traj_olmo <- compute_r_traj(
  olmo_ck |> filter(binom %in% attested_binoms) |>
    group_by(model, model_size, step, tokens, binom) |>
    summarise(preference = mean(preference), .groups = "drop"),
  "Dolma", OLMO_SIZES
)

bind_rows(
  r_traj_pythia |> mutate(model_family = "Pythia"),
  r_traj_olmo   |> mutate(model_family = "OLMo-3")
) |>
  filter(tokens > 0) |>
  ggplot(aes(x = tokens, y = r, colour = model_size)) +
  facet_grid(model_family ~ ., scales = "free_x") +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous("Training tokens (log scale)", trans = "log10",
                     labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous("Pearson r (model vs. human α-proportion)") +
  scale_colour_manual(values = c(PYTHIA_COLOURS, OLMO_COLOURS), name = "Model size") +
  labs(
    title    = "Model–human agreement over training — Pythia & OLMo-3",
    subtitle = "Pearson r between per-binomial model α-proportion and human α-proportion."
  ) +
  theme_classic(base_size = 13) +
  theme(strip.background = element_rect(fill = "grey95"),
        panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),
        legend.position  = "right")
```

------------------------------------------------------------------------

# §3 — Disagreement Items

Binomials where corpus RelFreq predicts the *wrong* direction (sign(model_pref) ≠ sign(freq_prob_c))
at the final checkpoint of the smallest model in each family (160M for Pythia, 7B for OLMo).

```{r disagreement}
# Per-binom GenPref (constant across participants): join once from human data
binom_genpref <- human_trials |>
  select(binom, GenPref) |>
  distinct()

find_disagreement <- function(final_df, discovery_size, size_levels, family_label) {
  # Add GenPref only if not already present (training_attested.csv carries it)
  if (!"GenPref" %in% names(final_df)) {
    final_df <- final_df |> left_join(binom_genpref, by = "binom")
  }
  # Discovery set: smallest model, items where RelFreq predicts wrong direction
  disc <- final_df |>
    filter(model_size == discovery_size, !is.na(freq_use), !is.na(preference)) |>
    mutate(
      freq_dir  = sign(freq_use),
      pref_dir  = sign(preference),
      genpref_dir = sign(GenPref)
    ) |>
    # Consistent with abspref_casestudy: RelFreq wrong AND GenPref/model in same direction
    filter(freq_dir != 0, pref_dir != 0, freq_dir != pref_dir,
           !is.na(genpref_dir), genpref_dir != 0, genpref_dir == pref_dir) |>
    select(binom, freq_dir, pref_dir) |>
    distinct()

  final_df |>
    filter(binom %in% disc$binom, !is.na(freq_use), !is.na(preference)) |>
    left_join(disc, by = "binom") |>
    mutate(
      model_family = family_label,
      model_size   = factor(model_size, levels = size_levels),
      pref_prob    = plogis(preference) - 0.5
    )
}

pythia_disagree <- find_disagreement(
  pythia_final |> mutate(corpus = "The Pile"),
  "160M", PYTHIA_SIZES, "Pythia"
)
olmo_disagree   <- find_disagreement(
  olmo_final |> mutate(corpus = "Dolma"),
  "7B", OLMO_SIZES, "OLMo-3"
)

cat(sprintf("Pythia disagreement items (160M): %d binomials\n",
            n_distinct(pythia_disagree$binom)))
cat(sprintf("OLMo-3 disagreement items (7B):   %d binomials\n",
            n_distinct(olmo_disagree$binom)))
```

```{r disagreement-plot}
bind_rows(pythia_disagree, olmo_disagree) |>
  filter(!is.na(GenPref)) |>
  ggplot(aes(x = GenPref, y = pref_prob, colour = model_size)) +
  facet_grid(model_family ~ .) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_smooth(method = "lm", formula = y ~ x, se = TRUE, alpha = 0.12,
              linewidth = 0.8) +
  scale_x_continuous("GenPref (human absolute preference; centred at 0)") +
  scale_y_continuous("Model preference P(α) − 0.5") +
  scale_colour_manual(values = c(PYTHIA_COLOURS, OLMO_COLOURS), name = "Model size") +
  labs(
    title    = "Disagreement items: model pref vs. GenPref",
    subtitle = "Items where corpus RelFreq predicts the wrong direction in the smallest model."
  ) +
  theme_classic(base_size = 12) +
  theme(strip.background = element_rect(fill = "grey95"),
        panel.grid.major = element_line(colour = "grey92", linewidth = 0.3))
```

```{r disagreement-table}
bind_rows(
  pythia_disagree |> select(model_family, model_size, binom, GenPref, pref_prob) |>
    group_by(model_family, binom, GenPref) |>
    summarise(pref_prob = mean(pref_prob), .groups = "drop"),
  olmo_disagree |> select(model_family, model_size, binom, GenPref, pref_prob) |>
    group_by(model_family, binom, GenPref) |>
    summarise(pref_prob = mean(pref_prob), .groups = "drop")
) |>
  arrange(model_family, desc(abs(GenPref))) |>
  mutate(across(where(is.numeric), round, 3)) |>
  kable(caption = "Disagreement items (mean across model sizes)", booktabs = TRUE) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "condensed"))
```

------------------------------------------------------------------------

# §4 — Training Trajectories (Case Studies)

The 6 strongest disagreement binomials (by |GenPref|) from the 160M Pythia model,
traced across all Pythia sizes over training.

```{r case-binoms}
PYTHIA_CASE_BINOMS <- pythia_disagree |>
  filter(model_size == "160M") |>
  group_by(binom) |>
  summarise(absGP = abs(first(GenPref)), .groups = "drop") |>
  slice_max(absGP, n = 6) |>
  pull(binom)

OLMO_CASE_BINOMS <- olmo_disagree |>
  filter(model_size == "7B") |>
  group_by(binom) |>
  summarise(absGP = abs(first(GenPref)), .groups = "drop") |>
  slice_max(absGP, n = 6) |>
  pull(binom)

cat("Pythia case binomials:", paste(PYTHIA_CASE_BINOMS, collapse = "; "), "\n")
cat("OLMo-3 case binomials:", paste(OLMO_CASE_BINOMS,  collapse = "; "), "\n")
```

## Pythia trajectories

```{r pythia-traj-plot}
pythia_traj <- pythia_ck |>
  filter(binom %in% PYTHIA_CASE_BINOMS) |>
  mutate(pref_prob  = plogis(preference) - 0.5,
         model_size = factor(parse_pythia_size(model), levels = PYTHIA_SIZES))

ggplot(pythia_traj |> filter(!is.na(pref_prob)),
       aes(x = tokens, y = pref_prob, colour = model_size)) +
  facet_wrap(~ binom, scales = "free_y", ncol = 3) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_x_continuous("Training tokens (log)", trans = "log10",
                     labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous("P(α) − 0.5") +
  scale_colour_manual(values = PYTHIA_COLOURS, name = "Model size") +
  labs(title    = "Pythia — training trajectories for disagreement binomials",
       subtitle = "Items where 160M corpus RelFreq predicts wrong direction.") +
  theme_classic(base_size = 11) +
  theme(strip.background = element_rect(fill = "grey95"),
        panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),
        legend.position  = "bottom")
```

## OLMo-3 trajectories

```{r olmo-traj-plot}
olmo_traj <- olmo_ck |>
  filter(binom %in% OLMO_CASE_BINOMS) |>
  mutate(pref_prob  = plogis(preference) - 0.5,
         model_size = factor(parse_olmo_size(model), levels = OLMO_SIZES))

ggplot(olmo_traj |> filter(!is.na(pref_prob)),
       aes(x = tokens, y = pref_prob, colour = model_size)) +
  facet_wrap(~ binom, scales = "free_y", ncol = 3) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5, alpha = 0.7) +
  scale_x_continuous("Training tokens (log)", trans = "log10",
                     labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous("P(α) − 0.5") +
  scale_colour_manual(values = OLMO_COLOURS, name = "Model size") +
  labs(title    = "OLMo-3 — training trajectories for disagreement binomials",
       subtitle = "Items where 7B corpus RelFreq predicts wrong direction.") +
  theme_classic(base_size = 11) +
  theme(strip.background = element_rect(fill = "grey95"),
        panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),
        legend.position  = "bottom")
```

------------------------------------------------------------------------

# §5 — Prediction Accuracy by Frequency Quintile

Mirrors Analysis 2c in `analysis2.Rmd`. Builds a per-binomial aggregated dataset
from raw checkpoint data (equivalent to `attested_agg.csv`) and computes accuracy
by overall-frequency quintile across training.

```{r constraint-meta}
CONSTRAINT_COLS <- c("Form", "Percept", "Culture", "Power", "Intense",
                     "Icon", "Freq", "Len", "Lapse", "BStress")
CONSTRAINT_COLS_REDUCED <- setdiff(CONSTRAINT_COLS, c("Lapse", "Percept"))

# Load individual constraint scores per binomial from the metadata CSV.
# Falls back gracefully if the columns are absent.
binoms_constraints <- read_csv(BINOMS_CSV, show_col_types = FALSE) |>
  filter(Attested == 1) |>
  select(binom = Alpha, any_of(CONSTRAINT_COLS))

# Human aggregated stats — majority-pref, RelFreq, OverallFreq, per-binom constraints
human_agg_full <- human_attested |>
  group_by(binom) |>
  summarise(
    hum_pref    = mean(resp_alpha),
    RelFreq     = first(RelFreq),
    OverallFreq = first(OverallFreq),
    GenPref     = first(GenPref),
    .groups     = "drop"
  ) |>
  left_join(binoms_constraints, by = "binom")
```

```{r build-model-agg}
# Build per-binom, per-checkpoint aggregated data (mean preference across prompts)
build_model_agg <- function(ck_df, model_size_levels, corpus_label) {
  ck_df |>
    filter(binom %in% attested_binoms) |>
    group_by(model, model_size, step, tokens, binom) |>
    summarise(preference = mean(preference, na.rm = TRUE), .groups = "drop") |>
    left_join(human_agg_full, by = "binom") |>
    mutate(
      model_size    = factor(model_size, levels = model_size_levels),
      corpus        = corpus_label,
      freq_quintile = cut(log(pmax(OverallFreq, 1e-9)), breaks = 5,
                          labels = c("Q1 (lowest)", "Q2", "Q3", "Q4", "Q5 (highest)"))
    ) |>
    filter(!is.na(freq_quintile))
}

pythia_agg <- build_model_agg(pythia_ck, PYTHIA_SIZES, "The Pile")
olmo_agg   <- build_model_agg(olmo_ck,   OLMO_SIZES,   "Dolma")

# Subsample to 20 checkpoint steps per model (mirrors 2c sampling logic)
all_agg <- bind_rows(
  pythia_agg |> mutate(model_family = "Pythia"),
  olmo_agg   |> mutate(model_family = "OLMo-3")
) |>
  group_by(model) |>
  mutate(step_rank = dense_rank(tokens)) |>
  filter(step_rank %in% unique(round(seq(1, max(step_rank), length.out = N_TRAJ_CHECKPOINTS)))) |>
  ungroup()
```

## Accuracy by frequency quintile

```{r acc-by-freq-quintile, fig.height=7}
acc_by_freq <- all_agg |>
  mutate(correct = (preference > 0) == (hum_pref > 0.5)) |>
  group_by(model_family, model_size, corpus, tokens, freq_quintile) |>
  summarise(accuracy = mean(correct, na.rm = TRUE), .groups = "drop")

ggplot(acc_by_freq, aes(x = freq_quintile, y = accuracy,
                         colour = tokens / 1e6, group = tokens)) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  geom_point(size = 1.8) +
  scale_colour_gradient(low = "#c7e9b4", high = "#154733",
                        name = "Training tokens\n(millions)") +
  scale_y_continuous("Accuracy", limits = c(0.3, 1),
                     labels = scales::percent_format()) +
  scale_x_discrete("Overall frequency quintile") +
  facet_grid(model_family ~ model_size) +
  labs(title = "Prediction accuracy by overall frequency quintile — Pythia & OLMo-3") +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title   = element_text(face = "bold", size = 14),
        strip.text   = element_text(face = "bold", size = 13),
        axis.text.x  = element_text(size = 10, angle = 20, hjust = 1),
        axis.text.y  = element_text(colour = "black", size = 11),
        axis.title   = element_text(colour = "black", size = 13),
        legend.text  = element_text(size = 11),
        legend.title = element_text(size = 12))
```

## Correlation with human preference by frequency quintile

```{r cor-humpref-quintile, fig.height=7}
cor_by_freq <- all_agg |>
  group_by(model_family, model_size, corpus, tokens, freq_quintile) |>
  summarise(
    r = tryCatch(
      cor(preference, hum_pref, use = "complete.obs"),
      error = function(e) NA_real_
    ),
    .groups = "drop"
  )

ggplot(cor_by_freq, aes(x = freq_quintile, y = r,
                         colour = tokens / 1e6, group = tokens)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  geom_point(size = 1.8) +
  scale_colour_gradient(low = "#c7e9b4", high = "#154733",
                        name = "Training tokens\n(millions)") +
  scale_y_continuous("Pearson r (model pref ~ human pref)", limits = c(-0.5, 1)) +
  scale_x_discrete("Overall frequency quintile") +
  facet_grid(model_family ~ model_size) +
  labs(title = "Correlation with human preference by frequency quintile — Pythia & OLMo-3") +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title   = element_text(face = "bold", size = 14),
        strip.text   = element_text(face = "bold", size = 13),
        axis.text.x  = element_text(size = 10, angle = 20, hjust = 1),
        axis.text.y  = element_text(colour = "black", size = 11),
        axis.title   = element_text(colour = "black", size = 13),
        legend.text  = element_text(size = 11),
        legend.title = element_text(size = 12))
```

------------------------------------------------------------------------

# §6 — Correlation Trajectories

Mirrors the raw correlation trajectory plots in `analysis2.Rmd` (sections 2a raw,
2b raw, and the RelFreq–preference correlation trajectory).

## Model preference ~ RelFreq over training

```{r cor-relfreq-traj, fig.height=5}
cor_traj_relfreq <- all_agg |>
  group_by(model_family, model_size, corpus, tokens) |>
  summarise(r = tryCatch(cor(preference, RelFreq, use = "complete.obs"),
                         error = function(e) NA_real_),
            .groups = "drop")

ggplot(cor_traj_relfreq, aes(x = tokens, y = r, colour = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous("Training tokens (log scale)", trans = "log10",
                     labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous("Pearson r (model pref ~ relative freq)", limits = c(-1, 1)) +
  scale_colour_manual(values = c(PYTHIA_COLOURS, OLMO_COLOURS), name = "Model size") +
  facet_wrap(~ model_family, scales = "free_x") +
  labs(title = "Correlation of model preference with relative frequency over training") +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title      = element_text(face = "bold", size = 14),
        strip.text      = element_text(face = "bold", size = 13),
        axis.text       = element_text(colour = "black", size = 11),
        axis.title      = element_text(colour = "black", size = 13),
        legend.position = "bottom")
```

## Model preference ~ human preference over training

```{r cor-humpref-traj, fig.height=5}
cor_traj_hum <- all_agg |>
  group_by(model_family, model_size, corpus, tokens) |>
  summarise(r = tryCatch(cor(preference, hum_pref, use = "complete.obs"),
                         error = function(e) NA_real_),
            .groups = "drop")

ggplot(cor_traj_hum, aes(x = tokens, y = r, colour = model_size)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous("Training tokens (log scale)", trans = "log10",
                     labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous("Pearson r (model pref ~ human pref)", limits = c(-0.1, 1)) +
  scale_colour_manual(values = c(PYTHIA_COLOURS, OLMO_COLOURS), name = "Model size") +
  facet_wrap(~ model_family, scales = "free_x") +
  labs(title = "Raw trajectory: correlation of model preference with human preference") +
  theme_bw(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        plot.title      = element_text(face = "bold", size = 14),
        strip.text      = element_text(face = "bold", size = 13),
        axis.text       = element_text(colour = "black", size = 11),
        axis.title      = element_text(colour = "black", size = 13),
        legend.position = "bottom")
```

## Model preference ~ each individual constraint over training

```{r cor-constraints-traj, fig.height=10}
have_constraints <- all(CONSTRAINT_COLS_REDUCED %in% names(all_agg))

if (have_constraints) {
  cor_traj_constr <- all_agg |>
    group_by(model_family, model_size, corpus, tokens) |>
    summarise(across(all_of(c(CONSTRAINT_COLS_REDUCED, "RelFreq")),
                     ~ tryCatch(cor(preference, .x, use = "complete.obs"),
                                error = function(e) NA_real_)),
              .groups = "drop") |>
    pivot_longer(cols = all_of(c(CONSTRAINT_COLS_REDUCED, "RelFreq")),
                 names_to = "constraint", values_to = "r") |>
    mutate(constraint = factor(constraint, levels = c(CONSTRAINT_COLS_REDUCED, "RelFreq")))

  print(
    ggplot(cor_traj_constr, aes(x = tokens, y = r, colour = model_size)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
      geom_line(linewidth = 0.7) +
      geom_point(size = 1.5) +
      scale_x_continuous("Training tokens (log scale)", trans = "log10",
                         labels = label_number(scale_cut = cut_short_scale())) +
      scale_y_continuous("Pearson r (model pref ~ constraint)", limits = c(-1, 1)) +
      scale_colour_manual(values = c(PYTHIA_COLOURS, OLMO_COLOURS), name = "Model size") +
      facet_grid(constraint ~ model_family) +
      labs(title = "Raw trajectories: correlation of model preference with each constraint") +
      theme_bw(base_size = 11) +
      theme(panel.grid.minor = element_blank(),
            plot.title      = element_text(face = "bold", size = 12),
            strip.text.x    = element_text(face = "bold", size = 11),
            strip.text.y    = element_text(size = 9),
            axis.text       = element_text(colour = "black", size = 9),
            axis.title      = element_text(colour = "black", size = 11),
            legend.position = "bottom")
  )
} else {
  message("Individual constraint columns not available in aggregated data — skipping. ",
          "Check that BINOMS_CSV contains columns: ",
          paste(CONSTRAINT_COLS_REDUCED, collapse = ", "))
}
```

## Preference deviation ~ each individual constraint over training

Preference deviation = model preference − RelFreq (both on probability scale).

```{r cor-prefdev-constraints, fig.height=10}
if (have_constraints) {
  cor_traj_prefdev <- all_agg |>
    mutate(pref_dev = preference - RelFreq) |>
    group_by(model_family, model_size, corpus, tokens) |>
    summarise(across(all_of(CONSTRAINT_COLS_REDUCED),
                     ~ tryCatch(cor(pref_dev, .x, use = "complete.obs"),
                                error = function(e) NA_real_)),
              .groups = "drop") |>
    pivot_longer(cols = all_of(CONSTRAINT_COLS_REDUCED),
                 names_to = "constraint", values_to = "r") |>
    mutate(constraint = factor(constraint, levels = CONSTRAINT_COLS_REDUCED))

  print(
    ggplot(cor_traj_prefdev, aes(x = tokens, y = r, colour = model_size)) +
      geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
      geom_line(linewidth = 0.7) +
      geom_point(size = 1.5) +
      scale_x_continuous("Training tokens (log scale)", trans = "log10",
                         labels = label_number(scale_cut = cut_short_scale())) +
      scale_y_continuous("Pearson r (pref deviation ~ constraint)", limits = c(-1, 1)) +
      scale_colour_manual(values = c(PYTHIA_COLOURS, OLMO_COLOURS), name = "Model size") +
      facet_grid(constraint ~ model_family) +
      labs(title = "Raw trajectories: correlation of preference deviation with each constraint",
           subtitle = "Preference deviation = model preference \u2212 RelFreq") +
      theme_bw(base_size = 11) +
      theme(panel.grid.minor = element_blank(),
            plot.title      = element_text(face = "bold", size = 12),
            plot.subtitle   = element_text(size = 10, colour = "grey40"),
            strip.text.x    = element_text(face = "bold", size = 11),
            strip.text.y    = element_text(size = 9),
            axis.text       = element_text(colour = "black", size = 9),
            axis.title      = element_text(colour = "black", size = 11),
            legend.position = "bottom")
  )
} else {
  message("Individual constraint columns not available — skipping preference deviation plot.")
}
```

------------------------------------------------------------------------

# §7 — AbsPref Coefficient Trajectory (Bayesian Mixed Model)

Per checkpoint, fit:

$$\text{preference} \sim \text{AbsPref\_c} + \text{rf\_c} + \text{ovf\_c} + \text{AbsPref\_c} \times \text{ovf\_c} + \text{rf\_c} \times \text{ovf\_c} + (1 \mid \text{binom}) + (1 \mid \text{prompt})$$

using brms (Gaussian), matching the prior and sampling settings in `analysis2.Rmd`.
Fits are saved to `models/abspref_traj_<slug>_ck<nn>.rds` and skipped on re-run.

------------------------------------------------------------------------

## Fitting

```{r abspref-meta}
# Per-binomial predictor metadata (AbsPref = GenPref, rf = RelFreq, ovf = log(OverallFreq))
binom_meta_for_reg <- human_trials |>
  group_by(binom) |>
  summarise(
    AbsPref       = first(GenPref),
    RelFreq_g     = first(RelFreq),
    OverallFreq_g = log(pmax(first(OverallFreq), 1)),
    .groups       = "drop"
  )
if (use_pile_freq) {
  binom_meta_for_reg <- binom_meta_for_reg |>
    left_join(pile_freq |> rename(RelFreq_p     = pile_freq_prob_c,
                                   OverallFreq_p = pile_log_total), by = "binom")
}

# Build traj_meta for abspref: slug × ck_idx → step + tokens
# (needed to recover x-axis token counts when loading saved .rds files)
build_abspref_traj_meta <- function(ck_df, model_family_label) {
  ck_df |>
    filter(binom %in% attested_binoms) |>
    group_by(model) |>
    group_modify(function(d, k) {
      steps_sel  <- sample_steps_log(d$step, N_TRAJ_CHECKPOINTS)
      tokens_sel <- map_dbl(steps_sel, ~ d$tokens[d$step == .x][1])
      tibble(
        slug     = make_slug(k$model[1]),
        ck_idx   = seq_along(steps_sel),
        step     = steps_sel,
        tokens   = tokens_sel
      )
    }) |>
    ungroup() |>
    mutate(model_family = model_family_label)
}

abspref_traj_meta <- bind_rows(
  build_abspref_traj_meta(pythia_ck, "Pythia"),
  build_abspref_traj_meta(olmo_ck,   "OLMo-3")
)
message(sprintf("abspref_traj_meta: %d rows (%d models × up to %d checkpoints)",
                nrow(abspref_traj_meta), n_distinct(abspref_traj_meta$slug),
                N_TRAJ_CHECKPOINTS))

# Save to results dir so prepare_results.R can use token counts as x-axis values
RESULTS_DIR <- normalizePath(file.path(DATA_DIR, "results"), mustWork = FALSE)
dir.create(RESULTS_DIR, showWarnings = FALSE, recursive = TRUE)
readr::write_csv(abspref_traj_meta,
                 file.path(RESULTS_DIR, "abspref_traj_meta.csv"))
```

```{r fit-abspref-traj}
fit_abspref_traj <- function(ck_df, model_family_label) {
  models <- sort(unique(ck_df$model))
  message(sprintf("\n=== [abspref traj] %s — %d models ===", model_family_label, length(models)))

  future_map(models, function(model_name) {
    slug      <- make_slug(model_name)
    meta_rows <- filter(abspref_traj_meta,
                        .data$slug == slug, model_family == model_family_label)
    mdata <- filter(ck_df, model == model_name) |>
      filter(binom %in% attested_binoms) |>
      left_join(binom_meta_for_reg, by = "binom") |>
      mutate(
        rf  = if (use_pile_freq && "RelFreq_p"     %in% names(binom_meta_for_reg))
                coalesce(RelFreq_p,     RelFreq_g)
              else RelFreq_g,
        ovf = if (use_pile_freq && "OverallFreq_p" %in% names(binom_meta_for_reg))
                coalesce(OverallFreq_p, OverallFreq_g)
              else OverallFreq_g
      )

    for (i in seq_len(nrow(meta_rows))) {
      s      <- meta_rows$step[i]
      ck_idx <- meta_rows$ck_idx[i]
      ck     <- sprintf("ck%02d", ck_idx)
      fpath  <- file.path(MODELS_DIR,
                          sprintf("abspref_traj_%s_%s.rds", slug, ck))

      if (file.exists(fpath)) {
        message(sprintf("  %s %s — exists, skipping", slug, ck)); next
      }

      cdat <- filter(mdata, step == s) |>
        filter(!is.na(AbsPref), !is.na(rf), !is.na(ovf)) |>
        mutate(
          AbsPref_c = as.numeric(scale(AbsPref, scale = FALSE)),
          rf_c      = as.numeric(scale(rf,      scale = FALSE)),
          ovf_c     = as.numeric(scale(ovf,     scale = FALSE))
        )

      if (nrow(cdat) < 20) {
        warning(sprintf("Skipping %s %s — only %d rows", slug, ck, nrow(cdat))); next
      }

      has_prompt <- "prompt" %in% names(cdat)
      fml <- if (has_prompt)
        preference ~ AbsPref_c + rf_c + ovf_c +
          AbsPref_c:ovf_c + rf_c:ovf_c + (1 | binom) + (1 | prompt)
      else
        preference ~ AbsPref_c + rf_c + ovf_c +
          AbsPref_c:ovf_c + rf_c:ovf_c + (1 | binom)

      message(sprintf("  Fitting %s %s  step=%d  n=%d  has_prompt=%s",
                      slug, ck, s, nrow(cdat), has_prompt))
      fit <- brm(
        formula = fml,
        data    = cdat,
        family  = gaussian(),
        prior   = PRIORS_GAUSSIAN,
        iter    = ITER, warmup = WARMUP, chains = CHAINS, cores = CORES,
        backend = "cmdstanr", silent = 2, refresh = 0
      )
      saveRDS(fit, fpath); rm(fit); gc()
      message(sprintf("    Saved: %s", basename(fpath)))
    }
  }, .options = furrr_options(seed = TRUE))
  invisible(NULL)
}

fit_abspref_traj(pythia_ck, "Pythia")
fit_abspref_traj(olmo_ck,   "OLMo-3")
message("\nabspref_traj fitting done.")
```

------------------------------------------------------------------------

## Results

```{r plot-abspref-traj, fig.height=6}
load_abspref_traj_coefs <- function(model_family_label, size_levels) {
  meta_rows <- filter(abspref_traj_meta, model_family == model_family_label)

  map_dfr(unique(meta_rows$slug), function(slug) {
    slug_meta <- filter(meta_rows, .data$slug == slug)
    map_dfr(seq_len(nrow(slug_meta)), function(i) {
      ck_idx <- slug_meta$ck_idx[i]
      tokens <- slug_meta$tokens[i]
      ck     <- sprintf("ck%02d", ck_idx)
      fpath  <- file.path(MODELS_DIR,
                          sprintf("abspref_traj_%s_%s.rds", slug, ck))
      if (!file.exists(fpath)) return(tibble())
      fit <- readRDS(fpath)
      fe  <- as.data.frame(fixef(fit)); rm(fit); gc()
      if (!"AbsPref_c" %in% rownames(fe)) return(tibble())
      msize <- if (model_family_label == "Pythia")
                 parse_pythia_size(slug)
               else
                 parse_olmo_size(slug)
      tibble(
        slug         = slug,
        ck_idx       = ck_idx,
        tokens       = tokens,
        estimate     = fe["AbsPref_c", "Estimate"],
        lo           = fe["AbsPref_c", "Q2.5"],
        hi           = fe["AbsPref_c", "Q97.5"],
        model_size   = factor(msize, levels = size_levels),
        model_family = model_family_label
      )
    })
  })
}

pythia_abspref_coefs <- load_abspref_traj_coefs("Pythia",  PYTHIA_SIZES)
olmo_abspref_coefs   <- load_abspref_traj_coefs("OLMo-3",  OLMO_SIZES)

bind_rows(pythia_abspref_coefs, olmo_abspref_coefs) |>
  filter(tokens > 0) |>
  ggplot(aes(x = tokens, y = estimate, colour = model_size,
             fill = model_size, ymin = lo, ymax = hi)) +
  facet_grid(model_family ~ ., scales = "free_x") +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  geom_ribbon(alpha = 0.12, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  scale_x_continuous("Training tokens (log scale)", trans = "log10",
                     labels = label_number(scale_cut = cut_short_scale())) +
  scale_y_continuous("AbsPref coefficient  (95% credible interval)") +
  scale_colour_manual(values = c(PYTHIA_COLOURS, OLMO_COLOURS), name = "Model size") +
  scale_fill_manual(values   = c(PYTHIA_COLOURS, OLMO_COLOURS), guide = "none") +
  labs(
    title    = "AbsPref coefficient over training — Pythia & OLMo-3",
    subtitle = paste0(
      "brm(preference ~ AbsPref_c + rf_c + ovf_c + AbsPref_c:ovf_c + rf_c:ovf_c + ",
      "(1|binom) + (1|prompt)). Ribbon = 95% credible interval."
    )
  ) +
  theme_classic(base_size = 13) +
  theme(strip.background = element_rect(fill = "grey95"),
        panel.grid.major = element_line(colour = "grey92", linewidth = 0.3),
        legend.position  = "right")
```
