---
title: "writeup"
colm-mode: submission  # submission (anonymous + line numbers) | preprint (non-anonymous) | final (camera-ready)
format:
  colm-pdf: default
  html: default
editor: visual
bibliography: references.bib
author:
  - name:
      literal: Zachary Houghton
    email: znhoughton@ucdavis.edu
    affiliations:
      - name: Vail Systems, Inc.
      - name: University of Oregon
  - name:
      literal: 
    affiliations:
      - name: 
  - name:
      literal:
    affiliations:
      - name:
abstract: |
  (Abstract goes here.)
---

# Introduction

# Model Training

In order to examine the representation of binomials in language models, we trained six autoregressive language models from scratch using the OPT architecture [@zhangOPTOpenPretrained2022], crossing two training corpora with three model sizes (125M, 350M, and 1.3B parameters), for a total of six models. All models share a common tokenizer: a Byte-Pair Encoding (BPE) tokenizer with a vocabulary of 8,192 tokens and a context window of 1,024 tokens, trained once on the full BabyLM corpus and reused across all six models to ensure that differences between models reflect training data and scale rather than tokenization. Models were trained on 2× NVIDIA H100 GPUs using PyTorch Distributed Data Parallel (DDP) in bfloat16 precision, with AdamW optimization and a linear learning rate schedule with 4,000 warmup steps.

The first corpus is BabyLM[Version 3, @charpentier2025babylmturns3], a developmentally plausible corpus of approximately 150M tokens consisting of child-directed speech (CHILDES), children's books, and simple web text. Each model was trained for 64 epochs over this corpus, yielding approximately 9.1B total training tokens. The second corpus is a 10B-token subset of C4 [@raffelExploringLimitsTransfer2020], a filtered English web text corpus. C4 models were trained for a single pass over this corpus, also yielding approximately 10B training tokens. Comparing the two corpora allows us to examine whether the content and register of the training data shapes how models acquire binomial ordering preferences over training.

Checkpoints were saved approximately every 10M tokens throughout training and pushed to HuggingFace Hub, giving us dense coverage of each model's learning trajectory. All models and checkpoints will be made publicly available. @tbl-models summarizes the key training parameters for each model.

```{r}
#| label: tbl-models
#| tbl-cap: "Training configuration for all six models. All models use the OPT architecture, a shared BPE tokenizer (vocab 8,192; context 1,024 tokens), AdamW optimization, and a linear LR schedule with 4,000 warmup steps."
#| echo: false
#| message: false
#| warning: false

library(knitr)

data.frame(
  Corpus       = c("BabyLM", "BabyLM", "BabyLM", "C4", "C4", "C4"),
  Params       = c("125M", "350M", "1.3B", "125M", "350M", "1.3B"),
  `Total tokens` = c("~9.1B", "~9.1B", "~9.1B", "~10B", "~10B", "~10B"),
  Checkpoints  = c("~910", "~910", "~860", "~1,010", "~1,005", "~960"),
  LR           = c("$3\\times10^{-4}$", "$1\\times10^{-4}$", "$1\\times10^{-4}$",
                   "$3\\times10^{-4}$", "$1\\times10^{-4}$", "$1\\times10^{-4}$"),
  check.names  = FALSE
) |>
  kable(align = c("l", "r", "r", "r", "r"))
```

To verify training quality, we calculated the perplexity (exponentiated cross-entropy loss) at the final checkpoint for each model (@tbl-perplexity; @fig-loss-curves). All models began near random --- initial perplexity $\approx$ 8,000--13,000, consistent with the theoretical ceiling of $e^{\ln 8192} = 8{,}192$ for a uniform distribution over our vocabulary --- and converged substantially, with final perplexity ranging from 5.2 to 34.2. The 1.3B model achieves the lowest perplexity in both corpora; within each corpus, the 125M model outperforms the 350M, consistent with the larger model being relatively data-hungry at this training budget. This is reflected in convergence of the loss over training (@fig-loss-curves).

```{r}
#| label: tbl-perplexity
#| tbl-cap: "Training perplexity at initialization and final checkpoint for all six models. Perplexity = $e^{\\text{loss}}$; values are measured on the training set. Initial perplexity reflects random initialization; all models converge well below this baseline."
#| echo: false
#| message: false
#| warning: false

library(dplyr)
library(knitr)

ppl  <- read.csv("../Data/training_quality/perplexity_results.csv")
loss <- read.csv("../Data/training_quality/loss_curves.csv")

initial_ppl <- loss |>
  group_by(corpus, params) |>
  slice_min(order_by = step, n = 1) |>
  mutate(initial_ppl = round(exp(loss))) |>
  select(Corpus = corpus, Params = params, `Initial PPL` = initial_ppl)

ppl |>
  rename(`Final PPL` = Perplexity) |>
  left_join(initial_ppl, by = c("Corpus", "Params")) |>
  mutate(Params = factor(Params, levels = c("125M", "350M", "1.3B"))) |>
  arrange(Corpus, Params) |>
  select(Corpus, Params, `Initial PPL`, `Final PPL`) |>
  kable(align = c("l", "r", "r", "r"))
```

```{r}
#| label: fig-loss-curves
#| fig-cap: "Training loss over the course of training for all six models, separated by corpus. All models start near the theoretical random baseline ($\\ln 8192 \\approx 9.01$) and converge smoothly; the 1.3B model achieves the lowest final loss in both corpora."
#| echo: false
#| message: false
#| warning: false

library(ggplot2)

df <- read.csv("../Data/training_quality/loss_curves.csv")
df$params <- factor(df$params, levels = c("125M", "350M", "1.3B"))

ggplot(df, aes(x = tokens_seen / 1e9, y = loss, color = params)) +
  geom_line(linewidth = 0.7, alpha = 0.85) +
  facet_wrap(~corpus, scales = "free_x") +
  geom_hline(yintercept = log(8192), linetype = "dashed", color = "gray50", linewidth = 0.4) +
  scale_color_manual(values = c("125M" = "#2166ac", "350M" = "#74add1", "1.3B" = "#d6604d")) +
  labs(x = "Tokens seen (B)", y = "Cross-entropy loss", color = "Params") +
  theme_classic(base_size = 10) +
  theme(strip.background = element_blank(), strip.text = element_text(face = "bold"))
```
